### 2011-07-01 papers/paper1/paper.tex
Initial import of paper.

\documentclass[letterpaper]{article}

\usepackage{acronym}

\begin{document}

\title{The Input/Ouput Automata Library}
\author{Justin R. Wilson \and Christopher D. Gill}
\date{}

\maketitle

\begin{abstract}
The abstract abstract.
\end{abstract}

\section{Introduction}

Current abstractions for concurrency and asynchrony are suitable for a narrow set of application domains.
In order to build the systems that we desire to build, we need better abstractions.

Devices are asynchronous and concurrent.
Networks are asynchronous and concurrent.
Why not start with a model that is asynchronous and concurrent?

What inspires us:
\begin{itemize}
  \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
    ``Let's make synchronous function calls over a network.''
    Assumes a thread of control.
    Asynchronous calls were added later (Did these result in obfuscation?)

  \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
    Examples: X server, OS2 Presentation Manager, Java Swing

  \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.

  \item Michi Henning - ICE

  \item Steve Vinoski - Toolbox of programming languages

\end{itemize}

We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.

\section{System Model and Formalization}

\begin{itemize}
  \item State
  \item Communication
  \item Asynchrony
  \item Concurrency
\end{itemize}

I/O Automata
\begin{itemize}
  \item Compare with UNITY
  \item Compare with esterel
  \item Compare with pi calculus
  \item Compare with Ptolemy
\end{itemize}

\section{I/O Automata Representation and Programming Model\label{programming_model}}

Conclude with our recommendations for programming, highlight need for exploration.

\section{Design and Implementation}

\begin{itemize}
  \item Problem
  \item Design forces
  \item Solution
  \item Consequences
\end{itemize}

\section{Evaluation and Discussion}

\begin{itemize}
\item Translation of automaton in \emph{Distributed Algorithms} to C++.
\item Simulate a protocol then replace with network components
\item Compare a protocol using: single-threaded event-based (select loop), multi-threaded, I/O automata
\item Show a buggy program and then apply an invariant to find the bug.
\end{itemize}

Bring up style mentioned in section \ref{programming_model}.

\section{Related Work}

pthreads (?)
Edward Lee - The Problem with Threads
Herb Sutter - The Free Lunch is Over
Early work on concurrency

\section{Conclusion and Future Work}

\begin{itemize}
  \item We are going to use it to build the substrate.
  \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
  \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
  \item We can take advantage of multi-core in a very straight-forward way
  \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
  \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
\end{itemize}

\end{document}

### 2011-07-05 papers/paper1/paper.tex
Cleared send set upon failure.

@@
 
-\usepackage{acronym}
+\usepackage{outline}
 
@@
 \section{Introduction}
+
+\begin{outline}
+\item We desire to build complex distributed systems
+\item These systems are characterized by asynchrony and concurrency at both the macroscopic and microscopic levels.
+  \begin{outline}
+  \item Macroscopic - Operate in a network where computation is naturally concurrent and asynchronous.
+  \item Microscopic - Individual devices perform more than one task and the physical world is asynchronous.
+  \end{outline}
+\item In spite of the importance of asynchrony and concurrency, we build these systems using the thread abstraction which is awkward and difficult.
+
+\end{outline}
 

### 2011-07-05 lib/udp_sender_automaton.cpp

@@ -143,32 +143,36 @@ namespace ioa {
   }
 
   void udp_sender_automaton::write_ready_effect () {
-    assert (!m_send_queue.empty ());
-    std::pair<aid_t, send_arg*> item = m_send_queue.front ();
-    m_send_queue.pop_front ();
-    m_send_set.erase (item.first);
+    if (!m_send_queue.empty ()) {
+      std::pair<aid_t, send_arg*> item = m_send_queue.front ();
+      m_send_queue.pop_front ();
+      m_send_set.erase (item.first);
       
-    if (sendto (m_fd, item.second->buffer->data (), item.second->buffer->size (), 0, item.second->address.get_sockaddr (), item.second->address.get_socklen ()) != -1) {
-      // Success.
-      add_to_complete_map (item.first, 0);
-    }
-    else {
-      // Fail.
-      m_errno = errno;
-      close (m_fd);
-      m_fd = -1;
-      add_to_complete_map (item.first, m_errno);
-      
-      // Drain the send queue.
-      for (std::list<std::pair<aid_t, send_arg*> >::const_iterator pos = m_send_queue.begin ();
-	   pos != m_send_queue.end ();
-	   ++pos) {
-	add_to_complete_map (pos->first, m_errno);
-	delete pos->second;
+      if (sendto (m_fd, item.second->buffer->data (), item.second->buffer->size (), 0, item.second->address.get_sockaddr (), item.second->address.get_socklen ()) != -1) {
+	// Success.
+	add_to_complete_map (item.first, 0);
       }
-      m_send_queue.clear ();
-    }
+      else {
+	// Fail.
+	m_errno = errno;
+	close (m_fd);
+	m_fd = -1;
+	add_to_complete_map (item.first, m_errno);
+	
+	// Drain the send queue.
+	for (std::list<std::pair<aid_t, send_arg*> >::const_iterator pos = m_send_queue.begin ();
+	     pos != m_send_queue.end ();
+	     ++pos) {
+	  add_to_complete_map (pos->first, m_errno);
+	  delete pos->second;
+	}
+	m_send_queue.clear ();
+	m_send_set.clear ();
+      }
 
+      delete item.second;
+    }
+      
     if (m_send_queue.empty ()) {
       // No more messages to send.
       m_state = SEND_WAIT;
@@ -177,8 +181,6 @@ namespace ioa {
       // Go again.
       m_state = SCHEDULE_WRITE_READY;
     }
-      
[…]

### 2011-07-06 papers/paper1/paper.tex
Rough draft of introduction.

--- 2011-07-01 papers/paper1/paper.tex
@@
-\documentclass[letterpaper]{article}
+\documentclass[letterpaper, twocolumn]{article}
 
-\usepackage{acronym}
+\usepackage{outline}
 
@@
 
-\title{The Input/Ouput Automata Library}
+\title{Threads Considered Harmful}
 \author{Justin R. Wilson \and Christopher D. Gill}
@@
 \begin{abstract}
-The abstract abstract.
+Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
+Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
+In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
+We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
+Finally, we conclude with an evaluation of our I/O automata framework.
 \end{abstract}
@@
 
-Current abstractions for concurrency and asynchrony are suitable for a narrow set of application domains.
-In order to build the systems that we desire to build, we need better abstractions.
+Advances in embedded systems and networking are paving the way for distributed systems of unprecedented sophistication.
+Cyber-physical systems go beyond the traditional embedded system paradigm by explicitly modeling the dynamics of the physical system and communication network in the computational task.
+The inclusion of wireless network interfaces in modern electronic devices is transforming our homes, offices, and hospitals into pervasive computing environments.
+The interactions these systems have with the physical environment and its community of users require a degree of interoperability, configurability, adaptability, scalability, robustness, and security not found in existing systems.
 
-Devices are asynchronous and concurrent.
-Networks are asynchronous and concurrent.
-Why not start with a model that is asynchronous and concurrent?
+One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
+In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
+The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
+While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.
 
-What inspires us:
-\begin{itemize}
-  \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
-    ``Let's make synchronous function calls over a network.''
-    Assumes a thread of control.
-    Asynchronous calls were added later (Did these result in obfuscation?)
+Threads are the de facto implementation vehicle for most of computing including distributed systems.
+Threads dominate because modern processors, programming languages, and operating systems were all developed to support the thread model.
+Furthermore, the widespread availability of thread libraries makes them a natural choice for system development.
+While threads dominate in practice, using threads to develop systems has a number of significant drawbacks.
 
-  \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
-    Examples: X server, OS2 Presentation Manager, Java Swing
+First, ad hoc communication among threads prohibits software reuse.
+To illustrate this problem consider the challenge of developing an application with a graphical user interface that also uses a multi-threaded middleware library.
+Graphical user interfaces often use a single-threaded event-based design, e.g., X, Swing.
+The system integrator must write glue code that bridges the single-threaded event-based semantics to multi-threaded semantics.
+Bridging between two thread models is achievable, however, the systems we desire to build will rely on many libraries each with their own thread model.
+The importance of a unified approach to concurrency becomes apparent when one considers bridging between different threading models, e.g., event-based, reactor, proactor, active objects, and their attributes, e.g., priority, thread-creation, blocking vs.
+non-blocking I/O.
 
-  \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.
+Second, threads are difficult to treat formally and therefore \emph{very} difficult to treat informally.
+The correctness of programs is reasoned about either formally as part of formal software design and verification and/or informally during the debugging process.
+Since most programs will never be formally verified, the correctness of most programs hinges solely on the developers ability to reason about the program.
+Based on the historical example of structured programming~\cite{goto_considered_harmful}, there is a strong correlation between what is ``easy'' in the formal arena and what is ``easy'' in the informal arena, especially when the implementation mimics the formal model, e.g., structured programming languages.
+It is well known that a formal treatment of threads is difficult due to the arbitrary interleaving of critical sections and ability to define critical sections of arbitrary scope~\cite{lee_threads}.
+Thus, the correctness of most multi-threaded programs hinges on the developer's ability to explore arbitrary interleavings of critical sections, i.e., model checking, in their head.
 
-  \item Michi Henning - ICE
+Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
+The distributed systems we desire to build are asynchronous at both the device and network levels.
+Embedded devices must respond to asynchronous events in the environment.
+The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
+The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
+Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
+Signals are primitive, rigid, and difficult to use correctly.
+I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
 
-  \item Steve Vinoski - Toolbox of programming languages
+Difficulties in developing asynchronous and concurrent software using threads has prompted a number of concurrent programming languages, e.g., Esterel, Erlang, Ptolemy, Autopipe, and language extensions, e.g., OpenMP, Cilk, Split-C.
+Solutions that involve creating a new language or extending an existing language suffer from two main problems.
+First, such solutions are often domain-specific and do not generalize to different applications, a necessity for heterogeneous distributed systems.
+Furthermore, solutions not connected to a general-purpose formal model lack the important property that the complete system including the environment can be modeled as another component.
+Second, such solutions often fail to become mainstream due to practical considerations.
+For example, the combination of C and pthreads is much more popular than Erlang due to the popularity of C-like languages and C's connection with UNIX operating system.
+Most language extensions are targeted at niche markets and therefore never standardized and implemented in mainstream compilers.
 
-\end{itemize}
+Thus, in order to build sophisticated distributed systems, we require a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
+%% The model must be general to support the wide variety of devices and protocols necessary for cyber-physical and pervasive computing systems.
+%% The model must facilitate the development of concurrent modules that can be reused.
+%% The model must have a strong connection with a formal model to simplify the task of reasoning about individual components and complete systems.
+%% The model must be implemented in such a way that it is easily approachable.
+In section~\ref{system_model}, we expound upon the requirements of the model and connect the requirements with the I/O automata formal model.
+Section~\ref{representation} describes how I/O automata are represented in our I/O automata framework.
+In section~\ref{design}, we describe the design and implementation of our I/O automata framework.
+Section~\ref{evaluation} provides an evaluation and discussion of our framework.
+Section~\ref{related_work} contains an overview of related work and we offer our conclusions and thoughts for future work in section~\ref{conclusion}.
 
-We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
+%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
 
-\section{System Model and Formalization}
+%%formal models, e.g., UNITY, I/O Automata,
+
+%% Devices are asynchronous and concurrent.
+%% Networks are asynchronous and concurrent.
+%% Why not start with a model that is asynchronous and concurrent?
+
+%% What inspires us:
+%% \begin{itemize}
+%%   \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
+%%     ``Let's make synchronous function calls over a network.''
+%%     Assumes a thread of control.
+%%     Asynchronous calls were added later (Did these result in obfuscation?)
+
+%%   \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
+%%     Examples: X server, OS2 Presentation Manager, Java Swing
+
+%%   \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.
+
+%%   \item Michi Henning - ICE
+
+%%   \item Steve Vinoski - Toolbox of programming languages
+
+%% \end{itemize}
+
+%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
+
+\section{System Model and Formalization\label{system_model}}
 
@@
 
-\section{I/O Automata Representation and Programming Model\label{programming_model}}
+\section{I/O Automata Representation and Programming Model\label{representation}}
 
@@
 
-\section{Design and Implementation}
+\section{Design and Implementation\label{design}}
 
@@
 
-\section{Evaluation and Discussion}
+\section{Evaluation and Discussion\label{evaluation}}
 
@@
 
-Bring up style mentioned in section \ref{programming_model}.
+Bring up style mentioned in section \ref{representation}.
 
-\section{Related Work}
+\section{Related Work\label{related_work}}
 
@@
 
-\section{Conclusion and Future Work}
+\section{Conclusion and Future Work\label{conclusion}}
 

### 2011-07-06 papers/paper1/paper.tex
Temp update.

@@
 
-\begin{itemize}
+\begin{outline}
 \item State
+  \begin{outline}
+  \item Local state only
+  \item Shared state
+    \begin{outline}
+      \item Impossible in distributed systems
+      \item Dangerous in local systems
+    \end{outline}
+  \end{outline}
 \item Communication
+  \begin{outline}
+  \item Atomic asynchronous message passing
+  \item Network sets size of atom (UDP)
+  \item Can build reliable streams (TCP)
+  \item Local equivalent is passing a value
+  \item Model should lend itself to writing protocols
+  \end{outline}
 \item Asynchrony
+  \begin{outline}
+    \item Model must have natural support for asynchrony, i.e., event-based
+    \item Leads to a more efficient implementation because changed state and enabled actions become obvious
+  \end{outine}
 \item Concurrency
-\end{itemize}
+  \begin{outline}
+    \item Reason about systems using non-deterministic interleaving of atomic actions
+    \item Model should admit implementations that execute concurrently
+  \end{outline}
+\item Dynamics
+  \begin{outline}
+    \item Configuration - Edges in graph of communicating components can change at run-time.
+      \begin{outline}
+      \item Already required in distributed settings
+      \item Not addressed in formal models
+      \end{outline}
+    \item Extension - Nodes in graph of communicating components can change at run-time.
+  \end{outline}
+\end{outline}
 

### 2011-07-13 papers/paper1/paper.tex
Worked on system model.

--- 2011-07-06 papers/paper1/paper.tex
@@
-\documentclass[letterpaper, twocolumn]{article}
+\documentclass[letterpaper]{article}
 
@@
 
-\title{Threads Considered Harmful}
+\title{An I/O Automata Framework for Distributed System Programming}
 \author{Justin R. Wilson \and Christopher D. Gill}
@@
 \begin{abstract}
-Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
-Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
-We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
-Finally, we conclude with an evaluation of our I/O automata framework.
+In an effort to be portable, system developers often favor existing architectures, languages, operating systems, and standard libraries.
+The dominant abstraction in established computing technologies for concurrency and therefore distributed system development is threads.
+Consequently, distributed system developers are locked into using threads and experience the difficulties of thread-based development including obscure race conditions, poor support for the semantics of distributed systems, and difficulties integrating software developed using different thread models.
+Furthermore, the demand for portability and generality dismisses many of the techniques proposed in research that involve new programming languages, language extensions, or domain-specific solutions.
+This paper presents a development framework based on the I/O automata formal model implemented in C++.
+The I/O automata model is a proven technique for distributed system design because its semantics match those found in distributed systems.
+With the framework, developers can reason about correctness directly from the program text using the I/O automata formalism.
+The I/O  automata model also provides a common abstraction for concurrency which facilitates software integration and promotes reuse.
+TODO:  What are we doing for evaluation?
+%% Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
+%% Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
+%% In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
+%% We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
+%% Finally, we conclude with an evaluation of our I/O automata framework.
 \end{abstract}
@@
 In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
-The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
+The mismatch between the conceptual model and implementation method introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
 While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.
@@
 Signals are primitive, rigid, and difficult to use correctly.
-I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
+Synchronous I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
+Asynchronous I/O improves efficiency by avoiding the dispatching necessary to determine the I/O operation but are subject to the same limitations of shared-state thread-based programming.
 
@@
 
+\pagebreak
+
 \section{System Model and Formalization\label{system_model}}
 
-\begin{itemize}
-  \item State
-  \item Communication
-  \item Asynchrony
-  \item Concurrency
-\end{itemize}
+Our system model is influenced by the semantics of distributed systems, the semantics of devices, and the desire for conceptual integrity.
 
-I/O Automata
-\begin{itemize}
-  \item Compare with UNITY
-  \item Compare with esterel
-  \item Compare with pi calculus
-  \item Compare with Ptolemy
-\end{itemize}
+\paragraph{Local State.}
+First, we require that all modules manipulate local state only including modules that might be collocated on the same distributed node. 
+Said another way, there is no global or shared state.
+Computation designed around the manipulation of local state only is easily modularized and consequently, easier to distribute and execute concurrently.
+
+\paragraph{Message Passing.}
+Nodes in distributed systems interact by passing messages over communication links.
+Message passing can either be synchronous or asynchronous.
+In synchronous message passing, the events of message production and delivery are the same, i.e., atomic.
+In asynchronous message passing, the events of message production and delivery are logically distinct events subject to being interposed by other events.
+Synchronous message passing is most often used between modules on the same node while asynchronous message passing is most often used between nodes.
+Both synchronous and asynchronous message passing are required.
+Synchronous message passing is necessary for building composable systems because of it association with atomicity.
+Asynchronous message passing is necessary for building real distributed systems as this matches the semantics of the networks found in practice.
+
+\paragraph{Asynchronous Events.}
+Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
+This implies the model that contains a general treatment of events, i.e., modules can both produce and consume events.
+Programs, therefore, are reactive and structured as a set of event produces and consumers.
+
+%% Communicating with external resources such as the operating system or physical environment is foundational to writing useful programs.
+%% Consider the history of I/O in the C run-time library.
+%% The synchronous function call semantics of the C programming language made it possible to write system calls, e.g., read and write, that transfer control to the operating system which then completes the operation and return control to the process after the system call.
+%% This is called \emph{blocking I/O} because the process blocks while it waits for the operating system to complete the I/O.
+%% Blocking I/O has the severe restriction that only one I/O source can be used at a time.
+%% To remedy this, system calls such as select and poll were introduced that allow a process to monitor I/O sources for events and then respond to these events.
+%% Asynchronous I/O was a further advancement that allows for the system (instead of the process) to dispatch a function upon the completion of I/O.
+%% Compare this to UNIX signals that are neither general (fixed-number with no associated data) nor principled.
+%% Interrupts are difficult to use correctly.
+
+\paragraph{Dynamics.}
+The sets of nodes and communication links are dynamic in real distributed systems.
+By extension, the number of modules in an individual node and the configuration governing the interaction of modules should also be allowed to vary.
+Our system model must support this reality.
+
+\paragraph{Reflection.}
+The ability to dynamically configure the set of modules and their interactions implies that modules can know the status of their configuration.
+For example, a module might require that a produced event be consumed.
+
+%% \begin{outline}
+%% \item State
+%%   \begin{outline}
+%%   \item There is no shared state in a 
+%%   \item Local state only
+%%   \item Shared state
+%%     \begin{outline}
+%%       \item Impossible in distributed systems
+%%       \item Dangerous in local systems
+%%     \end{outline}
+%%   \end{outline}
+%% \item Communication
+%%   \begin{outline}
+%%   \item Atomic asynchronous message passing
+%%   \item Network sets size of atom (UDP)
+%%   \item Can build reliable streams (TCP)
+%%   \item Local equivalent is passing a value
+%%   \item Model should lend itself to writing protocols
+%%   \end{outline}
+%% \item Asynchrony
+%%   \begin{outline}
+%%     \item Model must have natural support for asynchrony, i.e., event-based
+%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
+%%   \end{outline}
+%% \item Concurrency
+%%   \begin{outline}
+%%     \item Reason about systems using non-deterministic interleaving of atomic actions
+%%     \item Model should admit implementations that execute concurrently
+%%   \end{outline}
+%% \item Dynamics
+%%   \begin{outline}
+%%     \item Configuration - Edges in graph of communicating components can change at run-time.
+%%       \begin{outline}
+%%       \item Already required in distributed settings
+%%       \item Not addressed in formal models
+%%       \end{outline}
+%%     \item Extension - Nodes in graph of communicating components can change at run-time.
+%%   \end{outline}
+%%   \item Reflection
+%% \end{outline}
+
+%% I/O Automata
+%% \begin{itemize}
+%%   \item Compare with UNITY
+%%   \item Compare with esterel
+%%   \item Compare with pi calculus
+%%   \item Compare with Ptolemy
+%% \end{itemize}
 

### 2011-07-13 papers/paper1/paper.tex
Replaced the myriad of system call results with enums.

@@
 
-Our system model is influenced by the semantics of distributed systems, the semantics of devices, and the desire for conceptual integrity.
+\subsection{Requirements}
+
+Our system model is motivated by the semantics of devices, the semantics of distributed systems, and the desire for conceptual integrity.
+We adopt a recursive view that nodes, i.e., devices, in a distributed system may themselves be represented as a distributed system consisting of different sensors, actuators, and processes.
+We use the term \emph{node} to refer to a participant in a typical distributed system, i.e., one that uses standard network links.
+We use term \emph{module} to refer to a computational element within a node.
+Modules that reside on the same node are said to be \emph{collocated}.
 
 \paragraph{Local State.}
-First, we require that all modules manipulate local state only including modules that might be collocated on the same distributed node. 
+We require that all modules and by extension all nodes only manipulate local state.
 Said another way, there is no global or shared state.
@@
 
-\paragraph{Message Passing.}
-Nodes in distributed systems interact by passing messages over communication links.
-Message passing can either be synchronous or asynchronous.
-In synchronous message passing, the events of message production and delivery are the same, i.e., atomic.
-In asynchronous message passing, the events of message production and delivery are logically distinct events subject to being interposed by other events.
-Synchronous message passing is most often used between modules on the same node while asynchronous message passing is most often used between nodes.
-Both synchronous and asynchronous message passing are required.
-Synchronous message passing is necessary for building composable systems because of it association with atomicity.
-Asynchronous message passing is necessary for building real distributed systems as this matches the semantics of the networks found in practice.
+%% \paragraph{Message Passing.}
+%% Nodes in distributed systems interact by passing messages over communication links.
+%% Message passing can either be synchronous or asynchronous.
+%% In synchronous message passing, the events of message production and delivery are the same, i.e., atomic.
+%% In asynchronous message passing, the events of message production and delivery are logically distinct events subject to being interposed by other events.
+%% Synchronous message passing is most often used between modules on the same node while asynchronous message passing is most often used between nodes.
+%% Both synchronous and asynchronous message passing are required.
+%% Synchronous message passing is necessary for building composable systems because of it association with atomicity.
+%% Asynchronous message passing is necessary for building real distributed systems as this matches the semantics of the networks found in practice.
 
-\paragraph{Asynchronous Events.}
+\paragraph{Asynchronous Atomic Events.}
 Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
-This implies the model that contains a general treatment of events, i.e., modules can both produce and consume events.
-Programs, therefore, are reactive and structured as a set of event produces and consumers.
+This implies that the model contains a general treatment of events, i.e., modules can both produce and consume events.
+Modules, therefore, are reactive and structured as a set of atomic event produces and consumers.
+The guarantee of atomicity obviates the need for concurrency control primitives and improves our ability to reason about the behavior of the module.
+Modules communicate using events according to a \emph{configuration} that determines which events are delivered to each module.
 
@@
 \paragraph{Dynamics.}
-The sets of nodes and communication links are dynamic in real distributed systems.
-By extension, the number of modules in an individual node and the configuration governing the interaction of modules should also be allowed to vary.
-Our system model must support this reality.
+The set of nodes and communication links is dynamic in real distributed systems.
+Similarly, the set of modules and their configuration in real devices can change as the device may be re-purposed by users over time.
+In general, our system model must support a dynamic number of modules and, consequently, dynamic configuration of those modules.
 
 \paragraph{Reflection.}
-The ability to dynamically configure the set of modules and their interactions implies that modules can know the status of their configuration.
-For example, a module might require that a produced event be consumed.
+The ability to dynamically configure a module implies that a module must be able to know its configuration status to preserve certain properties.
+For example, consider a module for a reliable FIFO queue.
+Since the queue can be dynamically configured, the queue must be able to determine if the event produced by removing and returning an item from the queue will be delivered.
+Otherwise, the queue will lose data items and therefore not be reliable.
+
+\subsection{I/O Automata}
+
+The Input/Output (I/O) automata formal model developed by Lynch~\cite{distributed_algorithms} meets our requirements for local state and atomic asynchronous events.
+An I/O automaton consists of state variables and atomic actions that manipulate the state.
+The three action types are inputs, outputs, and internal actions.
+Input actions receive a signal or a value and output actions produce a signal or a value.
+Internal actions just manipulate the state of the automaton.
+Output and internal actions are \emph{locally controlled actions}~\cite{distributed_algorithms} consisting of a precondition and an effect.
+Input and output actions are \emph{external actions}~\cite{distributed_algorithms} because they allow an automaton to communicate with other automata.
+Automata may \emph{composed} by concatenating the state variables of the constituent automata and incorporating the effects of input actions into output actions with the same name.
+A useful operation in composite automata is \emph{hiding} which converts an output action to an internal action.
+Composing automata results in a static configuration, i.e., the set of automata and their interactions are fixed.
+Execution in the I/O automata model consists of repeatedly selecting a local action and then applying the effect if the precondition is true.
+The scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not executed) infinitely often.
+
+\subsection{Extending I/O Automata for Dynamics}
+
+A standard technique for modeling dynamic modules in I/O automata is to assume they exist but are inactive.
+Dynamic automata can then be activated and deactivated using inputs.
+This approach relies on the static configuration produced by composition and is therefore not suitable for an implementation.
+
+We extend I/O automata for dynamic configuration by introducing a \emph{system automaton} that records the configuration of all automata in the system.
+We differentiate between automaton types and automaton instances.
+Every automaton instances has a unique \emph{automaton identifier (aid)}.
+The state of the system automaton consists of the set of aids $A$ indicating what automata exist and a set of binding records $B$ indicating the relationship between output and input actions.
+A binding record $(out, output, in, input)$ is tuple consisting of the aid of the output automaton $out$, the name of the output action $output$, the aid of the input automaton $in$, and the name of the input action $input$.
+
+The configuration contained in set of binding records must be a valid static configuration to ensure that the behavior of the complete system remains analyzable under the I/O automata model.
+TODO:  Output must exist.
+TODO:  Input must exist.
+TODO:  Binding must not exist.
+First, an input can only be bound to one output.
+Thus, a given $in$ and $input$ combination can appear at most once in $B$.
+Second, an output cannot be bound to an input in the same automaton.
+Thus, $out \neq in$ for every $(out, output, in, input)$ in $B$.
+Third, an output cannot be bound to two inputs in the same automaton.
+Thus, $in_1 \neq in_2$ for all $(out, output, in_1, input_1), (out, output, in_2, input_2)$ in $B$.
+
+The system automaton contains input actions for creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton and output actions that indicate the result of a create, bind, unbind, or destroy.
+All automata are composed with (not bound to) the system automaton and contain outputs for creating, binding, unbinding, and destroying and inputs for receiving results.
+Systems calls (create, bind, unbind, destroy) resemble a request-response protocol where an automaton sends a request, the system automaton receives the request, the system automaton then sends the result, and the automaton receives the result.
+This has the important consequence that \emph{configuration is asynchronous}.
+Destroying an automaton has the effect of removing all of its bindings.
+An automaton that destroys itself can therefore cause the system to deliver of unbound and destroyed events without a corresponding unbind or destroy.
+
+\subsection{Extending I/O Automata for Reflection}
+
+Automata are endowed with additional inputs driven by system outputs indicating when an action has been bound and unbound.
+Automata might use the bound input to enable processing, i.e., someone is listening so I'll start producing data.
+The unbound input is useful for cleaning data structures and inhibiting further processing, i.e., no one is listening so I'll stop producing data.
+The bound and unbound inputs are useful but they are also asynchronous, thus, an automaton cannot trust the status reported by the last bound or unbound for use in, for example, a precondition.
+
+The asynchronous nature of bound and unbound requires a primitive that can examine the contents of $B$ instantaneously.
+Thus, we make one exception to the rule of local state that says automata can examine the contents of $B$ to determine their current configuration.
+This is permissible so long as concurrent implementations provide multiple-reader/single-writer semantics to $B$.
+
+\subsection{Equivalence}
+
+Extending the I/O automata model for dynamics and reflection is useless unless we can relate the extensions back to the original model or establish equivalence.
+To do this, we will take advantage of the fair traces (?) property of I/O automata~\cite{distributed_algorithms}.
+%% We accept an execution trace in the dynamic setting if it is a fair trace of the static model.
+
+%% Modeling a system with I/O automata implies a fixed set of automata given by $A_0$ and set of binding records $B_0$.
+
+Start with a trace of dynamic $T_d$.
+Remove system actions from $T_d$ so $T_d'$.
+We accept $T_d'$ if it is a fair trace of the static model.
+
+Let $A_0$ be the set of automata and $B_0$ be the set of binding records in the static model.
+If you want to prove correspondence, you must prove that eventually $A = A_0$ and $B = B_0$.
+We also need to show that no output is executed before it is fully bound.
+
+%% supports the actions of creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton.
+%% Upon creation, All automata are bound
+%% The four operations necessary for dynamic configuration are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
 

### 2011-07-13 include/ioa/action_executor.hpp

@@ -508,13 +508,13 @@ namespace ioa {
 	m_binder (binder),
 	m_key (key)
       {
-	m_system_scheduler.bound (m_binder, m_key);
+	m_system_scheduler.bound (m_binder, automaton::BOUND_RESULT, m_key);
 	m_system_scheduler.output_bound (m_output);
 	m_system_scheduler.input_bound (*m_input.get ());
       }
       
       ~record () {
-	m_system_scheduler.unbound (m_binder, m_key);
+	m_system_scheduler.unbound (m_binder, automaton::UNBOUND_RESULT, m_key);
 	m_system_scheduler.output_unbound (m_output);
 	m_system_scheduler.input_unbound (*m_input.get ());
       }

### 2011-07-13 include/ioa/automaton.hpp

@@ -69,57 +69,94 @@ namespace ioa {
   private:
     void schedule () const;
 
+  private:
     bool sys_create_precondition () const;
     std::pair<const_shared_ptr<generator_interface>, void*> sys_create_effect ();
-
-    bool sys_bind_precondition () const;
-    std::pair<shared_ptr<bind_executor_interface>, void*> sys_bind_effect ();
-
-    bool sys_unbind_precondition () const;
-    void* sys_unbind_effect ();
-
-    bool sys_destroy_precondition () const;
-    void* sys_destroy_effect ();
-
-    bool sys_self_destruct_precondition () const;
-    void* sys_self_destruct_effect ();
-
-    void sys_create_key_exists_effect (void* const &);
-    void sys_instance_exists_effect (void* const &);
-    void sys_automaton_created_effect (std::pair<void* COMMA aid_t> const &);
-    void sys_bind_key_exists_effect (void* const &);
-    void sys_output_automaton_dne_effect (void* const &);
-    void sys_input_automaton_dne_effect (void* const &);
-    void sys_binding_exists_effect (void* const &);
-    void sys_output_action_unavailable_effect (void* const &);
-    void sys_input_action_unavailable_effect (void* const &);
-    void sys_bound_effect (void* const &);
-    void sys_bind_key_dne_effect (void* const &);
-    void sys_unbound_effect (void* const &);
-    void sys_create_key_dne_effect (void* const &);
-    void sys_automaton_destroyed_effect (void* const &);
-
   public:
     SYSTEM_OUTPUT (automaton, sys_create, std::pair<const_shared_ptr<generator_interface> COMMA void*>);
+
+  private:
+    bool sys_bind_precondition () const;
+    std::pair<shared_ptr<bind_executor_interface>, void*> sys_bind_effect ();
+  public:
     SYSTEM_OUTPUT (automaton, sys_bind, std::pair<shared_ptr<bind_executor_interface> COMMA void*>);
+
+  private:
+    bool sys_unbind_precondition () const;
+    void* sys_unbind_effect ();
+  public:
     SYSTEM_OUTPUT (automaton, sys_unbind, void*);
+
+  private:
[…]

### 2011-07-13 include/ioa/system_scheduler_interface.hpp

@@ -1,6 +1,8 @@
 #ifndef __system_scheduler_interface_hpp__
 #define __system_scheduler_interface_hpp__
 
+#include <ioa/automaton.hpp>
+
 namespace ioa {
 
   /*
@@ -31,57 +33,30 @@ namespace ioa {
 
     virtual void self_destruct (const aid_t automaton) = 0;
     
-    virtual void create_key_exists (const aid_t automaton,
-				    void* const key) = 0;
-    
-    virtual void instance_exists (const aid_t automaton,
-				  void* const key) = 0;
-    
-    virtual void automaton_created (const aid_t automaton,
-				    void* const key,
-				    const aid_t child) = 0;
-    
-    virtual void bind_key_exists (const aid_t automaton,
-				  void* const key) = 0;
-    
-    virtual void output_automaton_dne (const aid_t automaton,
-				       void* const key) = 0;
-    
-    virtual void input_automaton_dne (const aid_t automaton,
-				      void* const key) = 0;
-    
-    virtual void binding_exists (const aid_t automaton,
-				 void* const key) = 0;
-    
-    virtual void input_action_unavailable (const aid_t automaton,
-					   void* const key) = 0;
-    
-    virtual void output_action_unavailable (const aid_t automaton,
-					    void* const key) = 0;
+    virtual void created (const aid_t automaton,
+			  const automaton::created_t,
+			  void* const key,
+			  const aid_t child) = 0;
     
     virtual void bound (const aid_t automaton,
+			const automaton::bound_t,
 			void* const key) = 0;
     
     virtual void output_bound (const output_executor_interface&) = 0;
     
     virtual void input_bound (const input_executor_interface&) = 0;
     
-    virtual void bind_key_dne (const aid_t automaton,
-			       void* const key) = 0;
-    
     virtual void unbound (const aid_t automaton,
+			  const automaton::unbound_t,
 			  void* const key) = 0;
     
     virtual void output_unbound (const output_executor_interface&) = 0;
     
     virtual void input_unbound (const input_executor_interface&) = 0;
     
-    virtual void create_key_dne (const aid_t automaton,
-				 void* const key) = 0;
-    
-    virtual void automaton_destroyed (const aid_t automaton,
-				      void* const key) = 0;
-    
+    virtual void destroyed (const aid_t automaton,
+			    const automaton::destroyed_t,
+			    void* const key) = 0;
   };
 
 }

### 2011-07-13 lib/automaton.cpp

@@ -256,160 +256,179 @@ namespace ioa {
     return 0;
   }
 
-  void automaton::sys_create_key_exists_effect (void* const &) {
-    // We prevent this in create.
-    assert (false);
-  }
-
-  void automaton::sys_instance_exists_effect (void* const & t) {
-    // Find the helper (sanity check).
-    std::set<system_automaton_manager_interface*>::const_iterator pos = m_create_recv.find (static_cast<system_automaton_manager_interface*> (t));
-    assert (pos != m_create_recv.end ());
-    (*pos)->instance_exists ();
-    // The creation failed so erase.
-    m_destroy_send.erase (*pos);
-    m_create_recv.erase (pos);
-  }
-
-  void automaton::sys_automaton_created_effect (std::pair<void*, aid_t> const & t) {
-    // Find the helper (sanity check).
-    std::set<system_automaton_manager_interface*>::const_iterator pos = m_create_recv.find (static_cast<system_automaton_manager_interface*> (t.first));
-    assert (pos != m_create_recv.end ());
-    (*pos)->automaton_created (t.second);
-    // The create succeeded.  Move to done.
-    m_create_done.insert (*pos);
-    m_create_recv.erase (pos);
-  }
-
-  void automaton::sys_bind_key_exists_effect (void* const & t) {
-    // We prevent this in bind.
-    assert (false);
+  void automaton::sys_created_effect (const created_arg_t& arg) {
+    switch (arg.type) {
+    case CREATE_KEY_EXISTS_RESULT:
+      // We prevent this in create.
+      assert (false);
+      break;
+    case INSTANCE_EXISTS_RESULT:
+      {
+	// Find the helper (sanity check).
+	std::set<system_automaton_manager_interface*>::const_iterator pos = m_create_recv.find (static_cast<system_automaton_manager_interface*> (arg.key));
+	assert (pos != m_create_recv.end ());
+	(*pos)->instance_exists ();
+	// The creation failed so erase.
+	m_destroy_send.erase (*pos);
+	m_create_recv.erase (pos);
+      }
+      break;
+    case AUTOMATON_CREATED_RESULT:
+      // Find the helper (sanity check).
+      std::set<system_automaton_manager_interface*>::const_iterator pos = m_create_recv.find (static_cast<system_automaton_manager_interface*> (arg.key));
[…]

### 2011-07-13 lib/automaton_record.cpp

@@ -37,13 +37,13 @@ namespace ioa {
 				    automaton_record* child) {
     m_children.insert (std::make_pair (key, child));
     // Tell the parent that the child was created.
-    m_system_scheduler.automaton_created (m_aid, key, child->m_aid);
+    m_system_scheduler.created (m_aid, automaton::AUTOMATON_CREATED_RESULT, key, child->m_aid);
   }
 
   void automaton_record::remove_child (void* const key) {
     m_children.erase (key);
     // Tell the parent that the child was destroyed.
-    m_system_scheduler.automaton_destroyed (m_aid, key);
+    m_system_scheduler.destroyed (m_aid, automaton::AUTOMATON_DESTROYED_RESULT, key);
   }
 
   automaton_record* automaton_record::get_child (void* const key) const {

### 2011-07-13 lib/global_fifo_scheduler.cpp

@@ -447,55 +447,17 @@ namespace ioa {
       schedule_configq (new self_destruct_runnable (automaton));
     }
 
-    void create_key_exists (const aid_t aid,
-			    void* const key) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_create_key_exists, key, system_input_category ()));
-    }
-
-    void instance_exists (const aid_t aid,
-			  void* const key) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_instance_exists, key, system_input_category ()));
+    void created (const aid_t aid,
+		  const automaton::created_t t,
+		  void* const key,
+		  const aid_t child) {
+      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_created, automaton::created_arg_t (t, key, child), system_input_category ()));
     }
   
-    void automaton_created (const aid_t aid,
-			    void* const key,
-			    const aid_t child) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_automaton_created, std::make_pair (key, child), system_input_category ()));
-    }
-  
-    void bind_key_exists (const aid_t aid,
-			  void* const key) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_bind_key_exists, key, system_input_category ()));
-    }
-
-    void output_automaton_dne (const aid_t aid,
-			       void* const key) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_output_automaton_dne, key, system_input_category ()));
-    }
-
-    void input_automaton_dne (const aid_t aid,
-			      void* const key) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_input_automaton_dne, key, system_input_category ()));
-    }
-  
-    void binding_exists (const aid_t aid,
-			 void* const key) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_binding_exists, key, system_input_category ()));
-    }
-  
-    void input_action_unavailable (const aid_t aid,
-				   void* const key) {
-      schedule_configq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_input_action_unavailable, key, system_input_category ()));
-    }
-  
[…]

### 2011-07-13 lib/model.cpp

@@ -81,7 +81,7 @@ namespace ioa {
 
     if (m_records[creator_aid]->create_key_exists (key)) {
       // Create key already in use.
-      m_system_scheduler.create_key_exists (creator_aid, key);
+      m_system_scheduler.created (creator_aid, automaton::CREATE_KEY_EXISTS_RESULT, key, -1);
       return -1;
     }
     
@@ -101,7 +101,7 @@ namespace ioa {
     if (m_instances.count (instance) != 0) {
       // Return the aid and inform the automaton that the instance already exists.
       m_aids.replace (aid);
-      m_system_scheduler.instance_exists (creator_aid, key);
+      m_system_scheduler.created (creator_aid, automaton::INSTANCE_EXISTS_RESULT, key, -1);
       return -1;
     }
     
@@ -127,7 +127,7 @@ namespace ioa {
     
     if (m_records[binder]->bind_key_exists (key)) {
       // Bind key already in use.
-      m_system_scheduler.bind_key_exists (binder, key);
+      m_system_scheduler.bound (binder, automaton::BIND_KEY_EXISTS_RESULT, key);
       return -1;
     }
     
@@ -139,12 +139,12 @@ namespace ioa {
     input.set_parameter (output.get_aid ());
     
     if (!output.fetch_instance (*this)) {
-      m_system_scheduler.output_automaton_dne (binder, key);
+      m_system_scheduler.bound (binder, automaton::OUTPUT_AUTOMATON_DNE_RESULT, key);
       return -1;
     }
     
     if (!input.fetch_instance (*this)) {
-      m_system_scheduler.input_automaton_dne (binder, key);
+      m_system_scheduler.bound (binder, automaton::INPUT_AUTOMATON_DNE_RESULT, key);
       return -1;
     }
     
@@ -154,7 +154,7 @@ namespace ioa {
     
     if (pos != m_bindings.end ()) {
       // Bound.
-      m_system_scheduler.binding_exists (binder, key);
+      m_system_scheduler.bound (binder, automaton::BINDING_EXISTS_RESULT, key);
       return -1;
     }
     
@@ -164,7 +164,7 @@ namespace ioa {
     
     if (in_pos != m_bindings.end ()) {
       // Input unavailable.
-      m_system_scheduler.input_action_unavailable (binder, key);
+      m_system_scheduler.bound (binder, automaton::INPUT_ACTION_UNAVAILABLE_RESULT, key);
[…]

### 2011-07-13 lib/simple_scheduler.cpp

@@ -556,55 +556,17 @@ namespace ioa {
       schedule_sysq (new self_destruct_runnable (automaton));
     }
 
-    void create_key_exists (const aid_t aid,
-			    void* const key) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_create_key_exists, key, system_input_category ()));
-    }
-
-    void instance_exists (const aid_t aid,
-			  void* const key) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_instance_exists, key, system_input_category ()));
+    void created (const aid_t aid,
+		  const automaton::created_t t,
+		  void* const key,
+		  const aid_t child) {
+      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_created, automaton::created_arg_t (t, key, child), system_input_category ()));
     }
   
-    void automaton_created (const aid_t aid,
-			    void* const key,
-			    const aid_t child) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_automaton_created, std::make_pair (key, child), system_input_category ()));
-    }
-  
-    void bind_key_exists (const aid_t aid,
-			  void* const key) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_bind_key_exists, key, system_input_category ()));
-    }
-
-    void output_automaton_dne (const aid_t aid,
-			       void* const key) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_output_automaton_dne, key, system_input_category ()));
-    }
-
-    void input_automaton_dne (const aid_t aid,
-			      void* const key) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_input_automaton_dne, key, system_input_category ()));
-    }
-  
-    void binding_exists (const aid_t aid,
-			 void* const key) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_binding_exists, key, system_input_category ()));
-    }
-  
-    void input_action_unavailable (const aid_t aid,
-				   void* const key) {
-      schedule_sysq (make_action_runnable (automaton_handle<automaton> (aid), &automaton::sys_input_action_unavailable, key, system_input_category ()));
-    }
-  
[…]

### 2011-07-13 test/action_executor.cpp

@@ -66,56 +66,30 @@ struct test_system_scheduler :
 
   void self_destruct (const ioa::aid_t automaton) { }
 
-  void create_key_exists (const ioa::aid_t automaton,
-			  void* const key) { }
-    
-  void instance_exists (const ioa::aid_t automaton,
-			void* const key) { }
-    
-  void automaton_created (const ioa::aid_t automaton,
-			  void* const key,
-			  const ioa::aid_t child) { }
-    
-  void bind_key_exists (const ioa::aid_t automaton,
-			void* const key) { }
-    
-  void output_automaton_dne (const ioa::aid_t automaton,
-			     void* const key) { }
-    
-  void input_automaton_dne (const ioa::aid_t automaton,
-			    void* const key) { }
-    
-  void binding_exists (const ioa::aid_t automaton,
-		       void* const key) { }
-    
-  void input_action_unavailable (const ioa::aid_t automaton,
-				 void* const key) { }
-    
-  void output_action_unavailable (const ioa::aid_t automaton,
-				  void* const key) { }
+  void created (const ioa::aid_t automaton,
+		const ioa::automaton::created_t,
+		void* const key,
+		const ioa::aid_t) { }
     
   void bound (const ioa::aid_t automaton,
+	      const ioa::automaton::bound_t,
 	      void* const key) { }
     
   void output_bound (const ioa::output_executor_interface&) { }
     
   void input_bound (const ioa::input_executor_interface&) { }
     
-  void bind_key_dne (const ioa::aid_t automaton,
-		     void* const key) { }
-    
   void unbound (const ioa::aid_t automaton,
+		const ioa::automaton::unbound_t,
 		void* const key) { }
     
   void output_unbound (const ioa::output_executor_interface&) { }
     
   void input_unbound (const ioa::input_executor_interface&) { }
     
-  void create_key_dne (const ioa::aid_t automaton,
-		       void* const key) { }
-    
-  void automaton_destroyed (const ioa::aid_t automaton,
-			    void* const key) { }
+  void destroyed (const ioa::aid_t automaton,
+		  const ioa::automaton::destroyed_t,
+		  void* const key) { }
 };
 
 static const char*

### 2011-07-13 test/model.cpp

@@ -8,6 +8,30 @@
 #include "instance_holder.hpp"
 #include <iostream>
 
+struct unbound_t {
+  const ioa::aid_t aid;
+  const ioa::automaton::unbound_t type;
+  void* const key;
+
+  unbound_t (const ioa::aid_t a,
+	     const ioa::automaton::unbound_t t,
+	     void* const k) :
+    aid (a),
+    type (t),
+    key (k)
+  { }
+
+  bool operator< (const unbound_t other) const {
+    if (aid != other.aid) {
+      return aid < other.aid;
+    }
+    if (type != other.type) {
+      return type < other.type;
+    }
+    return key < other.key;
+  }
+};
+
 struct unbound_record
 {
   ioa::aid_t m_aid;
@@ -31,38 +55,40 @@ struct unbound_record
   }
 };
 
+struct destroyed_t {
+  const ioa::aid_t aid;
+  const ioa::automaton::destroyed_t type;
+  void* const key;
+
+  destroyed_t (const ioa::aid_t a,
+	       const ioa::automaton::destroyed_t t,
+	       void* const k) :
+    aid (a),
+    type (t),
+    key (k)
+  { }
+
+  bool operator< (const destroyed_t other) const {
+    if (aid != other.aid) {
+      return aid < other.aid;
+    }
+    if (type != other.type) {
+      return type < other.type;
+    }
+    return key < other.key;
+  }
+};
+
 struct test_system_scheduler :
   public ioa::system_scheduler_interface
 {
-  ioa::aid_t m_create_key_exists_automaton;
-  void* m_create_key_exists_key;
-
-  ioa::aid_t m_instance_exists_automaton;
-  void* m_instance_exists_key;
-
-  ioa::aid_t m_automaton_created_automaton;
-  void* m_automaton_created_key;
-  ioa::aid_t m_automaton_created_child;
-
-  ioa::aid_t m_bind_key_exists_automaton;
-  void* m_bind_key_exists_key;
-
-  ioa::aid_t m_output_automaton_dne_automaton;
-  void* m_output_automaton_dne_key;
-
-  ioa::aid_t m_input_automaton_dne_automaton;
-  void* m_input_automaton_dne_key;
-
-  ioa::aid_t m_binding_exists_automaton;
-  void* m_binding_exists_key;
-
-  ioa::aid_t m_input_action_unavailable_automaton;
-  void* m_input_action_unavailable_key;
-
[…]

### 2011-07-13 papers/paper1/abstract.tex
Broke into separate files.

--- 2011-07-06 papers/paper1/paper.tex
@@
-\documentclass[letterpaper, twocolumn]{article}
-
-\usepackage{outline}
-
-\begin{document}
-
-\title{Threads Considered Harmful}
-\author{Justin R. Wilson \and Christopher D. Gill}
-\date{}
-
-\maketitle
-
 \begin{abstract}
-Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
-Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
-We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
-Finally, we conclude with an evaluation of our I/O automata framework.
+In an effort to be portable, system developers often favor existing architectures, languages, operating systems, and standard libraries.
+The dominant abstraction in established computing technologies for concurrency and therefore distributed system development is threads.
+Consequently, distributed system developers are locked into using threads and experience the difficulties of thread-based development including obscure race conditions, poor support for the semantics of distributed systems, and difficulties integrating software developed using different thread models.
+Furthermore, the demand for portability and generality dismisses many of the techniques proposed in research that involve new programming languages, language extensions, or domain-specific solutions.
+This paper presents a development framework based on the I/O automata formal model implemented in C++.
+The I/O automata model is a proven technique for distributed system design because its semantics match those found in distributed systems.
+With the framework, developers can reason about correctness directly from the program text using the I/O automata formalism.
+The I/O  automata model also provides a common abstraction for concurrency which facilitates software integration and promotes reuse.
+TODO:  What are we doing for evaluation?
+%% Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
+%% Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
+%% In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
+%% We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
+%% Finally, we conclude with an evaluation of our I/O automata framework.
 \end{abstract}
-
-\section{Introduction}
-
-Advances in embedded systems and networking are paving the way for distributed systems of unprecedented sophistication.
-Cyber-physical systems go beyond the traditional embedded system paradigm by explicitly modeling the dynamics of the physical system and communication network in the computational task.
-The inclusion of wireless network interfaces in modern electronic devices is transforming our homes, offices, and hospitals into pervasive computing environments.
-The interactions these systems have with the physical environment and its community of users require a degree of interoperability, configurability, adaptability, scalability, robustness, and security not found in existing systems.
-
-One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
-In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
-The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
-While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.
-
-Threads are the de facto implementation vehicle for most of computing including distributed systems.
-Threads dominate because modern processors, programming languages, and operating systems were all developed to support the thread model.
-Furthermore, the widespread availability of thread libraries makes them a natural choice for system development.
-While threads dominate in practice, using threads to develop systems has a number of significant drawbacks.
-
-First, ad hoc communication among threads prohibits software reuse.
-To illustrate this problem consider the challenge of developing an application with a graphical user interface that also uses a multi-threaded middleware library.
-Graphical user interfaces often use a single-threaded event-based design, e.g., X, Swing.
-The system integrator must write glue code that bridges the single-threaded event-based semantics to multi-threaded semantics.
-Bridging between two thread models is achievable, however, the systems we desire to build will rely on many libraries each with their own thread model.
-The importance of a unified approach to concurrency becomes apparent when one considers bridging between different threading models, e.g., event-based, reactor, proactor, active objects, and their attributes, e.g., priority, thread-creation, blocking vs.
-non-blocking I/O.
-
-Second, threads are difficult to treat formally and therefore \emph{very} difficult to treat informally.
-The correctness of programs is reasoned about either formally as part of formal software design and verification and/or informally during the debugging process.
-Since most programs will never be formally verified, the correctness of most programs hinges solely on the developers ability to reason about the program.
-Based on the historical example of structured programming~\cite{goto_considered_harmful}, there is a strong correlation between what is ``easy'' in the formal arena and what is ``easy'' in the informal arena, especially when the implementation mimics the formal model, e.g., structured programming languages.
-It is well known that a formal treatment of threads is difficult due to the arbitrary interleaving of critical sections and ability to define critical sections of arbitrary scope~\cite{lee_threads}.
-Thus, the correctness of most multi-threaded programs hinges on the developer's ability to explore arbitrary interleavings of critical sections, i.e., model checking, in their head.
-
-Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-The distributed systems we desire to build are asynchronous at both the device and network levels.
-Embedded devices must respond to asynchronous events in the environment.
-The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
-The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
-Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
-Signals are primitive, rigid, and difficult to use correctly.
-I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
-
-Difficulties in developing asynchronous and concurrent software using threads has prompted a number of concurrent programming languages, e.g., Esterel, Erlang, Ptolemy, Autopipe, and language extensions, e.g., OpenMP, Cilk, Split-C.
-Solutions that involve creating a new language or extending an existing language suffer from two main problems.
-First, such solutions are often domain-specific and do not generalize to different applications, a necessity for heterogeneous distributed systems.
-Furthermore, solutions not connected to a general-purpose formal model lack the important property that the complete system including the environment can be modeled as another component.
-Second, such solutions often fail to become mainstream due to practical considerations.
-For example, the combination of C and pthreads is much more popular than Erlang due to the popularity of C-like languages and C's connection with UNIX operating system.
-Most language extensions are targeted at niche markets and therefore never standardized and implemented in mainstream compilers.
-
-Thus, in order to build sophisticated distributed systems, we require a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-%% The model must be general to support the wide variety of devices and protocols necessary for cyber-physical and pervasive computing systems.
-%% The model must facilitate the development of concurrent modules that can be reused.
-%% The model must have a strong connection with a formal model to simplify the task of reasoning about individual components and complete systems.
-%% The model must be implemented in such a way that it is easily approachable.
-In section~\ref{system_model}, we expound upon the requirements of the model and connect the requirements with the I/O automata formal model.
-Section~\ref{representation} describes how I/O automata are represented in our I/O automata framework.
-In section~\ref{design}, we describe the design and implementation of our I/O automata framework.
-Section~\ref{evaluation} provides an evaluation and discussion of our framework.
-Section~\ref{related_work} contains an overview of related work and we offer our conclusions and thoughts for future work in section~\ref{conclusion}.
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-%%formal models, e.g., UNITY, I/O Automata,
-
-%% Devices are asynchronous and concurrent.
-%% Networks are asynchronous and concurrent.
-%% Why not start with a model that is asynchronous and concurrent?
-
-%% What inspires us:
-%% \begin{itemize}
-%%   \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
-%%     ``Let's make synchronous function calls over a network.''
-%%     Assumes a thread of control.
-%%     Asynchronous calls were added later (Did these result in obfuscation?)
-
-%%   \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
-%%     Examples: X server, OS2 Presentation Manager, Java Swing
-
-%%   \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.
-
-%%   \item Michi Henning - ICE
-
-%%   \item Steve Vinoski - Toolbox of programming languages
-
-%% \end{itemize}
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-\section{System Model and Formalization\label{system_model}}
-
-\begin{outline}
-\item State
-  \begin{outline}
-  \item Local state only
-  \item Shared state
-    \begin{outline}
-      \item Impossible in distributed systems
-      \item Dangerous in local systems
-    \end{outline}
-  \end{outline}
-\item Communication
-  \begin{outline}
-  \item Atomic asynchronous message passing
-  \item Network sets size of atom (UDP)
-  \item Can build reliable streams (TCP)
-  \item Local equivalent is passing a value
-  \item Model should lend itself to writing protocols
-  \end{outline}
-\item Asynchrony
-  \begin{outline}
-    \item Model must have natural support for asynchrony, i.e., event-based
-    \item Leads to a more efficient implementation because changed state and enabled actions become obvious
-  \end{outine}
-\item Concurrency
-  \begin{outline}
-    \item Reason about systems using non-deterministic interleaving of atomic actions
-    \item Model should admit implementations that execute concurrently
-  \end{outline}
-\item Dynamics
-  \begin{outline}
-    \item Configuration - Edges in graph of communicating components can change at run-time.
-      \begin{outline}
-      \item Already required in distributed settings
-      \item Not addressed in formal models
-      \end{outline}
-    \item Extension - Nodes in graph of communicating components can change at run-time.
-  \end{outline}
-\end{outline}
-
-I/O Automata
-\begin{itemize}
-  \item Compare with UNITY
-  \item Compare with esterel
-  \item Compare with pi calculus
-  \item Compare with Ptolemy
-\end{itemize}
-
-\section{I/O Automata Representation and Programming Model\label{representation}}
-
-Conclude with our recommendations for programming, highlight need for exploration.
-
-\section{Design and Implementation\label{design}}
-
-\begin{itemize}
-  \item Problem
-  \item Design forces
-  \item Solution
-  \item Consequences
-\end{itemize}
-
-\section{Evaluation and Discussion\label{evaluation}}
-
-\begin{itemize}
-\item Translation of automaton in \emph{Distributed Algorithms} to C++.
-\item Simulate a protocol then replace with network components
-\item Compare a protocol using: single-threaded event-based (select loop), multi-threaded, I/O automata
-\item Show a buggy program and then apply an invariant to find the bug.
-\end{itemize}
-
-Bring up style mentioned in section \ref{representation}.
-
-\section{Related Work\label{related_work}}
-
-pthreads (?)
-Edward Lee - The Problem with Threads
-Herb Sutter - The Free Lunch is Over
-Early work on concurrency
-
-\section{Conclusion and Future Work\label{conclusion}}
-
-\begin{itemize}
-  \item We are going to use it to build the substrate.
-  \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
-  \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
-  \item We can take advantage of multi-core in a very straight-forward way
-  \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
-  \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
-\end{itemize}
-
-\end{document}

### 2011-07-13 papers/paper1/conclusion.tex
Broke into separate files.

--- 2011-07-06 papers/paper1/paper.tex
@@
-\documentclass[letterpaper, twocolumn]{article}
-
-\usepackage{outline}
-
-\begin{document}
-
-\title{Threads Considered Harmful}
-\author{Justin R. Wilson \and Christopher D. Gill}
-\date{}
-
-\maketitle
-
-\begin{abstract}
-Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
-Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
-We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
-Finally, we conclude with an evaluation of our I/O automata framework.
-\end{abstract}
-
-\section{Introduction}
-
-Advances in embedded systems and networking are paving the way for distributed systems of unprecedented sophistication.
-Cyber-physical systems go beyond the traditional embedded system paradigm by explicitly modeling the dynamics of the physical system and communication network in the computational task.
-The inclusion of wireless network interfaces in modern electronic devices is transforming our homes, offices, and hospitals into pervasive computing environments.
-The interactions these systems have with the physical environment and its community of users require a degree of interoperability, configurability, adaptability, scalability, robustness, and security not found in existing systems.
-
-One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
-In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
-The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
-While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.
-
-Threads are the de facto implementation vehicle for most of computing including distributed systems.
-Threads dominate because modern processors, programming languages, and operating systems were all developed to support the thread model.
-Furthermore, the widespread availability of thread libraries makes them a natural choice for system development.
-While threads dominate in practice, using threads to develop systems has a number of significant drawbacks.
-
-First, ad hoc communication among threads prohibits software reuse.
-To illustrate this problem consider the challenge of developing an application with a graphical user interface that also uses a multi-threaded middleware library.
-Graphical user interfaces often use a single-threaded event-based design, e.g., X, Swing.
-The system integrator must write glue code that bridges the single-threaded event-based semantics to multi-threaded semantics.
-Bridging between two thread models is achievable, however, the systems we desire to build will rely on many libraries each with their own thread model.
-The importance of a unified approach to concurrency becomes apparent when one considers bridging between different threading models, e.g., event-based, reactor, proactor, active objects, and their attributes, e.g., priority, thread-creation, blocking vs.
-non-blocking I/O.
-
-Second, threads are difficult to treat formally and therefore \emph{very} difficult to treat informally.
-The correctness of programs is reasoned about either formally as part of formal software design and verification and/or informally during the debugging process.
-Since most programs will never be formally verified, the correctness of most programs hinges solely on the developers ability to reason about the program.
-Based on the historical example of structured programming~\cite{goto_considered_harmful}, there is a strong correlation between what is ``easy'' in the formal arena and what is ``easy'' in the informal arena, especially when the implementation mimics the formal model, e.g., structured programming languages.
-It is well known that a formal treatment of threads is difficult due to the arbitrary interleaving of critical sections and ability to define critical sections of arbitrary scope~\cite{lee_threads}.
-Thus, the correctness of most multi-threaded programs hinges on the developer's ability to explore arbitrary interleavings of critical sections, i.e., model checking, in their head.
-
-Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-The distributed systems we desire to build are asynchronous at both the device and network levels.
-Embedded devices must respond to asynchronous events in the environment.
-The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
-The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
-Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
-Signals are primitive, rigid, and difficult to use correctly.
-I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
-
-Difficulties in developing asynchronous and concurrent software using threads has prompted a number of concurrent programming languages, e.g., Esterel, Erlang, Ptolemy, Autopipe, and language extensions, e.g., OpenMP, Cilk, Split-C.
-Solutions that involve creating a new language or extending an existing language suffer from two main problems.
-First, such solutions are often domain-specific and do not generalize to different applications, a necessity for heterogeneous distributed systems.
-Furthermore, solutions not connected to a general-purpose formal model lack the important property that the complete system including the environment can be modeled as another component.
-Second, such solutions often fail to become mainstream due to practical considerations.
-For example, the combination of C and pthreads is much more popular than Erlang due to the popularity of C-like languages and C's connection with UNIX operating system.
-Most language extensions are targeted at niche markets and therefore never standardized and implemented in mainstream compilers.
-
-Thus, in order to build sophisticated distributed systems, we require a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-%% The model must be general to support the wide variety of devices and protocols necessary for cyber-physical and pervasive computing systems.
-%% The model must facilitate the development of concurrent modules that can be reused.
-%% The model must have a strong connection with a formal model to simplify the task of reasoning about individual components and complete systems.
-%% The model must be implemented in such a way that it is easily approachable.
-In section~\ref{system_model}, we expound upon the requirements of the model and connect the requirements with the I/O automata formal model.
-Section~\ref{representation} describes how I/O automata are represented in our I/O automata framework.
-In section~\ref{design}, we describe the design and implementation of our I/O automata framework.
-Section~\ref{evaluation} provides an evaluation and discussion of our framework.
-Section~\ref{related_work} contains an overview of related work and we offer our conclusions and thoughts for future work in section~\ref{conclusion}.
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-%%formal models, e.g., UNITY, I/O Automata,
-
-%% Devices are asynchronous and concurrent.
-%% Networks are asynchronous and concurrent.
-%% Why not start with a model that is asynchronous and concurrent?
-
-%% What inspires us:
-%% \begin{itemize}
-%%   \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
-%%     ``Let's make synchronous function calls over a network.''
-%%     Assumes a thread of control.
-%%     Asynchronous calls were added later (Did these result in obfuscation?)
-
-%%   \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
-%%     Examples: X server, OS2 Presentation Manager, Java Swing
-
-%%   \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.
-
-%%   \item Michi Henning - ICE
-
-%%   \item Steve Vinoski - Toolbox of programming languages
-
-%% \end{itemize}
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-\section{System Model and Formalization\label{system_model}}
-
-\begin{outline}
-\item State
-  \begin{outline}
-  \item Local state only
-  \item Shared state
-    \begin{outline}
-      \item Impossible in distributed systems
-      \item Dangerous in local systems
-    \end{outline}
-  \end{outline}
-\item Communication
-  \begin{outline}
-  \item Atomic asynchronous message passing
-  \item Network sets size of atom (UDP)
-  \item Can build reliable streams (TCP)
-  \item Local equivalent is passing a value
-  \item Model should lend itself to writing protocols
-  \end{outline}
-\item Asynchrony
-  \begin{outline}
-    \item Model must have natural support for asynchrony, i.e., event-based
-    \item Leads to a more efficient implementation because changed state and enabled actions become obvious
-  \end{outine}
-\item Concurrency
-  \begin{outline}
-    \item Reason about systems using non-deterministic interleaving of atomic actions
-    \item Model should admit implementations that execute concurrently
-  \end{outline}
-\item Dynamics
-  \begin{outline}
-    \item Configuration - Edges in graph of communicating components can change at run-time.
-      \begin{outline}
-      \item Already required in distributed settings
-      \item Not addressed in formal models
-      \end{outline}
-    \item Extension - Nodes in graph of communicating components can change at run-time.
-  \end{outline}
-\end{outline}
-
-I/O Automata
-\begin{itemize}
-  \item Compare with UNITY
-  \item Compare with esterel
-  \item Compare with pi calculus
-  \item Compare with Ptolemy
-\end{itemize}
-
-\section{I/O Automata Representation and Programming Model\label{representation}}
-
-Conclude with our recommendations for programming, highlight need for exploration.
-
-\section{Design and Implementation\label{design}}
-
-\begin{itemize}
-  \item Problem
-  \item Design forces
-  \item Solution
-  \item Consequences
-\end{itemize}
-
-\section{Evaluation and Discussion\label{evaluation}}
-
-\begin{itemize}
-\item Translation of automaton in \emph{Distributed Algorithms} to C++.
-\item Simulate a protocol then replace with network components
-\item Compare a protocol using: single-threaded event-based (select loop), multi-threaded, I/O automata
-\item Show a buggy program and then apply an invariant to find the bug.
-\end{itemize}
-
-Bring up style mentioned in section \ref{representation}.
-
-\section{Related Work\label{related_work}}
-
-pthreads (?)
-Edward Lee - The Problem with Threads
-Herb Sutter - The Free Lunch is Over
-Early work on concurrency
-
 \section{Conclusion and Future Work\label{conclusion}}
@@
 \end{itemize}
-
-\end{document}

### 2011-07-13 papers/paper1/design.tex
Broke into separate files.

--- 2011-07-06 papers/paper1/paper.tex
@@
-\documentclass[letterpaper, twocolumn]{article}
-
-\usepackage{outline}
-
-\begin{document}
-
-\title{Threads Considered Harmful}
-\author{Justin R. Wilson \and Christopher D. Gill}
-\date{}
-
-\maketitle
-
-\begin{abstract}
-Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
-Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
-We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
-Finally, we conclude with an evaluation of our I/O automata framework.
-\end{abstract}
-
-\section{Introduction}
-
-Advances in embedded systems and networking are paving the way for distributed systems of unprecedented sophistication.
-Cyber-physical systems go beyond the traditional embedded system paradigm by explicitly modeling the dynamics of the physical system and communication network in the computational task.
-The inclusion of wireless network interfaces in modern electronic devices is transforming our homes, offices, and hospitals into pervasive computing environments.
-The interactions these systems have with the physical environment and its community of users require a degree of interoperability, configurability, adaptability, scalability, robustness, and security not found in existing systems.
-
-One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
-In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
-The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
-While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.
-
-Threads are the de facto implementation vehicle for most of computing including distributed systems.
-Threads dominate because modern processors, programming languages, and operating systems were all developed to support the thread model.
-Furthermore, the widespread availability of thread libraries makes them a natural choice for system development.
-While threads dominate in practice, using threads to develop systems has a number of significant drawbacks.
-
-First, ad hoc communication among threads prohibits software reuse.
-To illustrate this problem consider the challenge of developing an application with a graphical user interface that also uses a multi-threaded middleware library.
-Graphical user interfaces often use a single-threaded event-based design, e.g., X, Swing.
-The system integrator must write glue code that bridges the single-threaded event-based semantics to multi-threaded semantics.
-Bridging between two thread models is achievable, however, the systems we desire to build will rely on many libraries each with their own thread model.
-The importance of a unified approach to concurrency becomes apparent when one considers bridging between different threading models, e.g., event-based, reactor, proactor, active objects, and their attributes, e.g., priority, thread-creation, blocking vs.
-non-blocking I/O.
-
-Second, threads are difficult to treat formally and therefore \emph{very} difficult to treat informally.
-The correctness of programs is reasoned about either formally as part of formal software design and verification and/or informally during the debugging process.
-Since most programs will never be formally verified, the correctness of most programs hinges solely on the developers ability to reason about the program.
-Based on the historical example of structured programming~\cite{goto_considered_harmful}, there is a strong correlation between what is ``easy'' in the formal arena and what is ``easy'' in the informal arena, especially when the implementation mimics the formal model, e.g., structured programming languages.
-It is well known that a formal treatment of threads is difficult due to the arbitrary interleaving of critical sections and ability to define critical sections of arbitrary scope~\cite{lee_threads}.
-Thus, the correctness of most multi-threaded programs hinges on the developer's ability to explore arbitrary interleavings of critical sections, i.e., model checking, in their head.
-
-Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-The distributed systems we desire to build are asynchronous at both the device and network levels.
-Embedded devices must respond to asynchronous events in the environment.
-The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
-The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
-Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
-Signals are primitive, rigid, and difficult to use correctly.
-I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
-
-Difficulties in developing asynchronous and concurrent software using threads has prompted a number of concurrent programming languages, e.g., Esterel, Erlang, Ptolemy, Autopipe, and language extensions, e.g., OpenMP, Cilk, Split-C.
-Solutions that involve creating a new language or extending an existing language suffer from two main problems.
-First, such solutions are often domain-specific and do not generalize to different applications, a necessity for heterogeneous distributed systems.
-Furthermore, solutions not connected to a general-purpose formal model lack the important property that the complete system including the environment can be modeled as another component.
-Second, such solutions often fail to become mainstream due to practical considerations.
-For example, the combination of C and pthreads is much more popular than Erlang due to the popularity of C-like languages and C's connection with UNIX operating system.
-Most language extensions are targeted at niche markets and therefore never standardized and implemented in mainstream compilers.
-
-Thus, in order to build sophisticated distributed systems, we require a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-%% The model must be general to support the wide variety of devices and protocols necessary for cyber-physical and pervasive computing systems.
-%% The model must facilitate the development of concurrent modules that can be reused.
-%% The model must have a strong connection with a formal model to simplify the task of reasoning about individual components and complete systems.
-%% The model must be implemented in such a way that it is easily approachable.
-In section~\ref{system_model}, we expound upon the requirements of the model and connect the requirements with the I/O automata formal model.
-Section~\ref{representation} describes how I/O automata are represented in our I/O automata framework.
-In section~\ref{design}, we describe the design and implementation of our I/O automata framework.
-Section~\ref{evaluation} provides an evaluation and discussion of our framework.
-Section~\ref{related_work} contains an overview of related work and we offer our conclusions and thoughts for future work in section~\ref{conclusion}.
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-%%formal models, e.g., UNITY, I/O Automata,
-
-%% Devices are asynchronous and concurrent.
-%% Networks are asynchronous and concurrent.
-%% Why not start with a model that is asynchronous and concurrent?
-
-%% What inspires us:
-%% \begin{itemize}
-%%   \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
-%%     ``Let's make synchronous function calls over a network.''
-%%     Assumes a thread of control.
-%%     Asynchronous calls were added later (Did these result in obfuscation?)
-
-%%   \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
-%%     Examples: X server, OS2 Presentation Manager, Java Swing
-
-%%   \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.
-
-%%   \item Michi Henning - ICE
-
-%%   \item Steve Vinoski - Toolbox of programming languages
-
-%% \end{itemize}
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-\section{System Model and Formalization\label{system_model}}
-
-\begin{outline}
-\item State
-  \begin{outline}
-  \item Local state only
-  \item Shared state
-    \begin{outline}
-      \item Impossible in distributed systems
-      \item Dangerous in local systems
-    \end{outline}
-  \end{outline}
-\item Communication
-  \begin{outline}
-  \item Atomic asynchronous message passing
-  \item Network sets size of atom (UDP)
-  \item Can build reliable streams (TCP)
-  \item Local equivalent is passing a value
-  \item Model should lend itself to writing protocols
-  \end{outline}
-\item Asynchrony
-  \begin{outline}
-    \item Model must have natural support for asynchrony, i.e., event-based
-    \item Leads to a more efficient implementation because changed state and enabled actions become obvious
-  \end{outine}
-\item Concurrency
-  \begin{outline}
-    \item Reason about systems using non-deterministic interleaving of atomic actions
-    \item Model should admit implementations that execute concurrently
-  \end{outline}
-\item Dynamics
-  \begin{outline}
-    \item Configuration - Edges in graph of communicating components can change at run-time.
-      \begin{outline}
-      \item Already required in distributed settings
-      \item Not addressed in formal models
-      \end{outline}
-    \item Extension - Nodes in graph of communicating components can change at run-time.
-  \end{outline}
-\end{outline}
-
-I/O Automata
-\begin{itemize}
-  \item Compare with UNITY
-  \item Compare with esterel
-  \item Compare with pi calculus
-  \item Compare with Ptolemy
-\end{itemize}
-
-\section{I/O Automata Representation and Programming Model\label{representation}}
-
-Conclude with our recommendations for programming, highlight need for exploration.
-
 \section{Design and Implementation\label{design}}
@@
 \end{itemize}
-
-\section{Evaluation and Discussion\label{evaluation}}
-
-\begin{itemize}
-\item Translation of automaton in \emph{Distributed Algorithms} to C++.
-\item Simulate a protocol then replace with network components
-\item Compare a protocol using: single-threaded event-based (select loop), multi-threaded, I/O automata
-\item Show a buggy program and then apply an invariant to find the bug.
-\end{itemize}
-
-Bring up style mentioned in section \ref{representation}.
-
-\section{Related Work\label{related_work}}
-
-pthreads (?)
-Edward Lee - The Problem with Threads
-Herb Sutter - The Free Lunch is Over
-Early work on concurrency
-
-\section{Conclusion and Future Work\label{conclusion}}
-
-\begin{itemize}
-  \item We are going to use it to build the substrate.
-  \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
-  \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
-  \item We can take advantage of multi-core in a very straight-forward way
-  \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
-  \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
-\end{itemize}
-
-\end{document}

### 2011-07-13 papers/paper1/evaluation.tex
Broke into separate files.

--- 2011-07-06 papers/paper1/paper.tex
@@
-\documentclass[letterpaper, twocolumn]{article}
-
-\usepackage{outline}
-
-\begin{document}
-
-\title{Threads Considered Harmful}
-\author{Justin R. Wilson \and Christopher D. Gill}
-\date{}
-
-\maketitle
-
-\begin{abstract}
-Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
-Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
-We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
-Finally, we conclude with an evaluation of our I/O automata framework.
-\end{abstract}
-
-\section{Introduction}
-
-Advances in embedded systems and networking are paving the way for distributed systems of unprecedented sophistication.
-Cyber-physical systems go beyond the traditional embedded system paradigm by explicitly modeling the dynamics of the physical system and communication network in the computational task.
-The inclusion of wireless network interfaces in modern electronic devices is transforming our homes, offices, and hospitals into pervasive computing environments.
-The interactions these systems have with the physical environment and its community of users require a degree of interoperability, configurability, adaptability, scalability, robustness, and security not found in existing systems.
-
-One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
-In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
-The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
-While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.
-
-Threads are the de facto implementation vehicle for most of computing including distributed systems.
-Threads dominate because modern processors, programming languages, and operating systems were all developed to support the thread model.
-Furthermore, the widespread availability of thread libraries makes them a natural choice for system development.
-While threads dominate in practice, using threads to develop systems has a number of significant drawbacks.
-
-First, ad hoc communication among threads prohibits software reuse.
-To illustrate this problem consider the challenge of developing an application with a graphical user interface that also uses a multi-threaded middleware library.
-Graphical user interfaces often use a single-threaded event-based design, e.g., X, Swing.
-The system integrator must write glue code that bridges the single-threaded event-based semantics to multi-threaded semantics.
-Bridging between two thread models is achievable, however, the systems we desire to build will rely on many libraries each with their own thread model.
-The importance of a unified approach to concurrency becomes apparent when one considers bridging between different threading models, e.g., event-based, reactor, proactor, active objects, and their attributes, e.g., priority, thread-creation, blocking vs.
-non-blocking I/O.
-
-Second, threads are difficult to treat formally and therefore \emph{very} difficult to treat informally.
-The correctness of programs is reasoned about either formally as part of formal software design and verification and/or informally during the debugging process.
-Since most programs will never be formally verified, the correctness of most programs hinges solely on the developers ability to reason about the program.
-Based on the historical example of structured programming~\cite{goto_considered_harmful}, there is a strong correlation between what is ``easy'' in the formal arena and what is ``easy'' in the informal arena, especially when the implementation mimics the formal model, e.g., structured programming languages.
-It is well known that a formal treatment of threads is difficult due to the arbitrary interleaving of critical sections and ability to define critical sections of arbitrary scope~\cite{lee_threads}.
-Thus, the correctness of most multi-threaded programs hinges on the developer's ability to explore arbitrary interleavings of critical sections, i.e., model checking, in their head.
-
-Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-The distributed systems we desire to build are asynchronous at both the device and network levels.
-Embedded devices must respond to asynchronous events in the environment.
-The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
-The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
-Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
-Signals are primitive, rigid, and difficult to use correctly.
-I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
-
-Difficulties in developing asynchronous and concurrent software using threads has prompted a number of concurrent programming languages, e.g., Esterel, Erlang, Ptolemy, Autopipe, and language extensions, e.g., OpenMP, Cilk, Split-C.
-Solutions that involve creating a new language or extending an existing language suffer from two main problems.
-First, such solutions are often domain-specific and do not generalize to different applications, a necessity for heterogeneous distributed systems.
-Furthermore, solutions not connected to a general-purpose formal model lack the important property that the complete system including the environment can be modeled as another component.
-Second, such solutions often fail to become mainstream due to practical considerations.
-For example, the combination of C and pthreads is much more popular than Erlang due to the popularity of C-like languages and C's connection with UNIX operating system.
-Most language extensions are targeted at niche markets and therefore never standardized and implemented in mainstream compilers.
-
-Thus, in order to build sophisticated distributed systems, we require a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-%% The model must be general to support the wide variety of devices and protocols necessary for cyber-physical and pervasive computing systems.
-%% The model must facilitate the development of concurrent modules that can be reused.
-%% The model must have a strong connection with a formal model to simplify the task of reasoning about individual components and complete systems.
-%% The model must be implemented in such a way that it is easily approachable.
-In section~\ref{system_model}, we expound upon the requirements of the model and connect the requirements with the I/O automata formal model.
-Section~\ref{representation} describes how I/O automata are represented in our I/O automata framework.
-In section~\ref{design}, we describe the design and implementation of our I/O automata framework.
-Section~\ref{evaluation} provides an evaluation and discussion of our framework.
-Section~\ref{related_work} contains an overview of related work and we offer our conclusions and thoughts for future work in section~\ref{conclusion}.
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-%%formal models, e.g., UNITY, I/O Automata,
-
-%% Devices are asynchronous and concurrent.
-%% Networks are asynchronous and concurrent.
-%% Why not start with a model that is asynchronous and concurrent?
-
-%% What inspires us:
-%% \begin{itemize}
-%%   \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
-%%     ``Let's make synchronous function calls over a network.''
-%%     Assumes a thread of control.
-%%     Asynchronous calls were added later (Did these result in obfuscation?)
-
-%%   \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
-%%     Examples: X server, OS2 Presentation Manager, Java Swing
-
-%%   \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.
-
-%%   \item Michi Henning - ICE
-
-%%   \item Steve Vinoski - Toolbox of programming languages
-
-%% \end{itemize}
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-\section{System Model and Formalization\label{system_model}}
-
-\begin{outline}
-\item State
-  \begin{outline}
-  \item Local state only
-  \item Shared state
-    \begin{outline}
-      \item Impossible in distributed systems
-      \item Dangerous in local systems
-    \end{outline}
-  \end{outline}
-\item Communication
-  \begin{outline}
-  \item Atomic asynchronous message passing
-  \item Network sets size of atom (UDP)
-  \item Can build reliable streams (TCP)
-  \item Local equivalent is passing a value
-  \item Model should lend itself to writing protocols
-  \end{outline}
-\item Asynchrony
-  \begin{outline}
-    \item Model must have natural support for asynchrony, i.e., event-based
-    \item Leads to a more efficient implementation because changed state and enabled actions become obvious
-  \end{outine}
-\item Concurrency
-  \begin{outline}
-    \item Reason about systems using non-deterministic interleaving of atomic actions
-    \item Model should admit implementations that execute concurrently
-  \end{outline}
-\item Dynamics
-  \begin{outline}
-    \item Configuration - Edges in graph of communicating components can change at run-time.
-      \begin{outline}
-      \item Already required in distributed settings
-      \item Not addressed in formal models
-      \end{outline}
-    \item Extension - Nodes in graph of communicating components can change at run-time.
-  \end{outline}
-\end{outline}
-
-I/O Automata
-\begin{itemize}
-  \item Compare with UNITY
-  \item Compare with esterel
-  \item Compare with pi calculus
-  \item Compare with Ptolemy
-\end{itemize}
-
-\section{I/O Automata Representation and Programming Model\label{representation}}
-
-Conclude with our recommendations for programming, highlight need for exploration.
-
-\section{Design and Implementation\label{design}}
-
-\begin{itemize}
-  \item Problem
-  \item Design forces
-  \item Solution
-  \item Consequences
-\end{itemize}
-
 \section{Evaluation and Discussion\label{evaluation}}
@@
 Bring up style mentioned in section \ref{representation}.
-
-\section{Related Work\label{related_work}}
-
-pthreads (?)
-Edward Lee - The Problem with Threads
-Herb Sutter - The Free Lunch is Over
-Early work on concurrency
-
-\section{Conclusion and Future Work\label{conclusion}}
-
-\begin{itemize}
-  \item We are going to use it to build the substrate.
-  \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
-  \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
-  \item We can take advantage of multi-core in a very straight-forward way
-  \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
-  \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
-\end{itemize}
-
-\end{document}

### 2011-07-13 papers/paper1/paper.tex
Broke into separate files.

--- 2011-07-06 papers/paper1/paper.tex
@@
 
-\begin{abstract}
-Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
-Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
-We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
-Finally, we conclude with an evaluation of our I/O automata framework.
-\end{abstract}
+\input{abstract}
 
-\section{Introduction}
+\input{introduction}
 
-Advances in embedded systems and networking are paving the way for distributed systems of unprecedented sophistication.
-Cyber-physical systems go beyond the traditional embedded system paradigm by explicitly modeling the dynamics of the physical system and communication network in the computational task.
-The inclusion of wireless network interfaces in modern electronic devices is transforming our homes, offices, and hospitals into pervasive computing environments.
-The interactions these systems have with the physical environment and its community of users require a degree of interoperability, configurability, adaptability, scalability, robustness, and security not found in existing systems.
+\input{system_model}
 
-One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
-In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
-The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
-While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.
+\input{programming_model}
 
-Threads are the de facto implementation vehicle for most of computing including distributed systems.
-Threads dominate because modern processors, programming languages, and operating systems were all developed to support the thread model.
-Furthermore, the widespread availability of thread libraries makes them a natural choice for system development.
-While threads dominate in practice, using threads to develop systems has a number of significant drawbacks.
+\input{design}
 
-First, ad hoc communication among threads prohibits software reuse.
-To illustrate this problem consider the challenge of developing an application with a graphical user interface that also uses a multi-threaded middleware library.
-Graphical user interfaces often use a single-threaded event-based design, e.g., X, Swing.
-The system integrator must write glue code that bridges the single-threaded event-based semantics to multi-threaded semantics.
-Bridging between two thread models is achievable, however, the systems we desire to build will rely on many libraries each with their own thread model.
-The importance of a unified approach to concurrency becomes apparent when one considers bridging between different threading models, e.g., event-based, reactor, proactor, active objects, and their attributes, e.g., priority, thread-creation, blocking vs.
-non-blocking I/O.
+\input{evaluation}
 
-Second, threads are difficult to treat formally and therefore \emph{very} difficult to treat informally.
-The correctness of programs is reasoned about either formally as part of formal software design and verification and/or informally during the debugging process.
-Since most programs will never be formally verified, the correctness of most programs hinges solely on the developers ability to reason about the program.
-Based on the historical example of structured programming~\cite{goto_considered_harmful}, there is a strong correlation between what is ``easy'' in the formal arena and what is ``easy'' in the informal arena, especially when the implementation mimics the formal model, e.g., structured programming languages.
-It is well known that a formal treatment of threads is difficult due to the arbitrary interleaving of critical sections and ability to define critical sections of arbitrary scope~\cite{lee_threads}.
-Thus, the correctness of most multi-threaded programs hinges on the developer's ability to explore arbitrary interleavings of critical sections, i.e., model checking, in their head.
+\input{related_work}
 
-Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-The distributed systems we desire to build are asynchronous at both the device and network levels.
-Embedded devices must respond to asynchronous events in the environment.
-The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
-The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
-Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
-Signals are primitive, rigid, and difficult to use correctly.
-I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
-
-Difficulties in developing asynchronous and concurrent software using threads has prompted a number of concurrent programming languages, e.g., Esterel, Erlang, Ptolemy, Autopipe, and language extensions, e.g., OpenMP, Cilk, Split-C.
-Solutions that involve creating a new language or extending an existing language suffer from two main problems.
-First, such solutions are often domain-specific and do not generalize to different applications, a necessity for heterogeneous distributed systems.
-Furthermore, solutions not connected to a general-purpose formal model lack the important property that the complete system including the environment can be modeled as another component.
-Second, such solutions often fail to become mainstream due to practical considerations.
-For example, the combination of C and pthreads is much more popular than Erlang due to the popularity of C-like languages and C's connection with UNIX operating system.
-Most language extensions are targeted at niche markets and therefore never standardized and implemented in mainstream compilers.
-
-Thus, in order to build sophisticated distributed systems, we require a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-%% The model must be general to support the wide variety of devices and protocols necessary for cyber-physical and pervasive computing systems.
-%% The model must facilitate the development of concurrent modules that can be reused.
-%% The model must have a strong connection with a formal model to simplify the task of reasoning about individual components and complete systems.
-%% The model must be implemented in such a way that it is easily approachable.
-In section~\ref{system_model}, we expound upon the requirements of the model and connect the requirements with the I/O automata formal model.
-Section~\ref{representation} describes how I/O automata are represented in our I/O automata framework.
-In section~\ref{design}, we describe the design and implementation of our I/O automata framework.
-Section~\ref{evaluation} provides an evaluation and discussion of our framework.
-Section~\ref{related_work} contains an overview of related work and we offer our conclusions and thoughts for future work in section~\ref{conclusion}.
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-%%formal models, e.g., UNITY, I/O Automata,
-
-%% Devices are asynchronous and concurrent.
-%% Networks are asynchronous and concurrent.
-%% Why not start with a model that is asynchronous and concurrent?
-
-%% What inspires us:
-%% \begin{itemize}
-%%   \item CORBA - Remote Method Invocation (Remote Procedure Call with a ``this'' pointer).
-%%     ``Let's make synchronous function calls over a network.''
-%%     Assumes a thread of control.
-%%     Asynchronous calls were added later (Did these result in obfuscation?)
-
-%%   \item Patterns for asynchronous concurrency in a synchronous, thread-based world: Active objects, Reactor, Proactor, etc.
-%%     Examples: X server, OS2 Presentation Manager, Java Swing
-
-%%   \item AJAX - The modern Web is built on JavaScript + \emph{asynchronous} web page requests = AJAX.
-
-%%   \item Michi Henning - ICE
-
-%%   \item Steve Vinoski - Toolbox of programming languages
-
-%% \end{itemize}
-
-%% We provide a common semantics and model for building concurrent and asynchronous systems with straight-forward abstractions.
-
-\section{System Model and Formalization\label{system_model}}
-
-\begin{outline}
-\item State
-  \begin{outline}
-  \item Local state only
-  \item Shared state
-    \begin{outline}
-      \item Impossible in distributed systems
-      \item Dangerous in local systems
-    \end{outline}
-  \end{outline}
-\item Communication
-  \begin{outline}
-  \item Atomic asynchronous message passing
-  \item Network sets size of atom (UDP)
-  \item Can build reliable streams (TCP)
-  \item Local equivalent is passing a value
-  \item Model should lend itself to writing protocols
-  \end{outline}
-\item Asynchrony
-  \begin{outline}
-    \item Model must have natural support for asynchrony, i.e., event-based
-    \item Leads to a more efficient implementation because changed state and enabled actions become obvious
-  \end{outine}
-\item Concurrency
-  \begin{outline}
-    \item Reason about systems using non-deterministic interleaving of atomic actions
-    \item Model should admit implementations that execute concurrently
-  \end{outline}
-\item Dynamics
-  \begin{outline}
-    \item Configuration - Edges in graph of communicating components can change at run-time.
-      \begin{outline}
-      \item Already required in distributed settings
-      \item Not addressed in formal models
-      \end{outline}
-    \item Extension - Nodes in graph of communicating components can change at run-time.
-  \end{outline}
-\end{outline}
-
-I/O Automata
-\begin{itemize}
-  \item Compare with UNITY
-  \item Compare with esterel
-  \item Compare with pi calculus
-  \item Compare with Ptolemy
-\end{itemize}
-
-\section{I/O Automata Representation and Programming Model\label{representation}}
-
-Conclude with our recommendations for programming, highlight need for exploration.
-
-\section{Design and Implementation\label{design}}
-
-\begin{itemize}
-  \item Problem
-  \item Design forces
-  \item Solution
-  \item Consequences
-\end{itemize}
-
-\section{Evaluation and Discussion\label{evaluation}}
-
-\begin{itemize}
-\item Translation of automaton in \emph{Distributed Algorithms} to C++.
-\item Simulate a protocol then replace with network components
-\item Compare a protocol using: single-threaded event-based (select loop), multi-threaded, I/O automata
-\item Show a buggy program and then apply an invariant to find the bug.
-\end{itemize}
-
-Bring up style mentioned in section \ref{representation}.
-
-\section{Related Work\label{related_work}}
-
-pthreads (?)
-Edward Lee - The Problem with Threads
-Herb Sutter - The Free Lunch is Over
-Early work on concurrency
-
-\section{Conclusion and Future Work\label{conclusion}}
-
-\begin{itemize}
-  \item We are going to use it to build the substrate.
-  \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
-  \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
-  \item We can take advantage of multi-core in a very straight-forward way
-  \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
-  \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
-\end{itemize}
+\input{conclusion}
 

### 2011-07-13 papers/paper1/programming_model.tex
Broke into separate files.

\section{I/O Automata Representation and Programming Model\label{representation}}

Conclude with our recommendations for programming, highlight need for exploration.

### 2011-07-13 papers/paper1/related_work.tex
Broke into separate files.

\section{Related Work\label{related_work}}

pthreads (?)
Edward Lee - The Problem with Threads
Herb Sutter - The Free Lunch is Over
Early work on concurrency

### 2011-07-13 papers/paper1/system_model.tex
Broke into separate files.

\section{System Model and Formalization\label{system_model}}

\subsection{Requirements}

Our system model is motivated by the semantics of devices, the semantics of distributed systems, and the desire for conceptual integrity.
We adopt a recursive view that nodes, i.e., devices, in a distributed system may themselves be represented as a distributed system consisting of different sensors, actuators, and processes.
We use the term \emph{node} to refer to a participant in a typical distributed system, i.e., one that uses standard network links.
We use term \emph{module} to refer to a computational element within a node.
Modules that reside on the same node are said to be \emph{collocated}.

\paragraph{Local State.}
We require that all modules and by extension all nodes only manipulate local state.
Said another way, there is no global or shared state.
Computation designed around the manipulation of local state only is easily modularized and consequently, easier to distribute and execute concurrently.

%% \paragraph{Message Passing.}
%% Nodes in distributed systems interact by passing messages over communication links.
%% Message passing can either be synchronous or asynchronous.
%% In synchronous message passing, the events of message production and delivery are the same, i.e., atomic.
%% In asynchronous message passing, the events of message production and delivery are logically distinct events subject to being interposed by other events.
%% Synchronous message passing is most often used between modules on the same node while asynchronous message passing is most often used between nodes.
%% Both synchronous and asynchronous message passing are required.
%% Synchronous message passing is necessary for building composable systems because of it association with atomicity.
%% Asynchronous message passing is necessary for building real distributed systems as this matches the semantics of the networks found in practice.

\paragraph{Asynchronous Atomic Events.}
Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
This implies that the model contains a general treatment of events, i.e., modules can both produce and consume events.
Modules, therefore, are reactive and structured as a set of atomic event produces and consumers.
The guarantee of atomicity obviates the need for concurrency control primitives and improves our ability to reason about the behavior of the module.
Modules communicate using events according to a \emph{configuration} that determines which events are delivered to each module.

%% Communicating with external resources such as the operating system or physical environment is foundational to writing useful programs.
%% Consider the history of I/O in the C run-time library.
%% The synchronous function call semantics of the C programming language made it possible to write system calls, e.g., read and write, that transfer control to the operating system which then completes the operation and return control to the process after the system call.
%% This is called \emph{blocking I/O} because the process blocks while it waits for the operating system to complete the I/O.
%% Blocking I/O has the severe restriction that only one I/O source can be used at a time.
%% To remedy this, system calls such as select and poll were introduced that allow a process to monitor I/O sources for events and then respond to these events.
%% Asynchronous I/O was a further advancement that allows for the system (instead of the process) to dispatch a function upon the completion of I/O.
%% Compare this to UNIX signals that are neither general (fixed-number with no associated data) nor principled.
%% Interrupts are difficult to use correctly.

\paragraph{Dynamics.}
The set of nodes and communication links is dynamic in real distributed systems.
Similarly, the set of modules and their configuration in real devices can change as the device may be re-purposed by users over time.
In general, our system model must support a dynamic number of modules and, consequently, dynamic configuration of those modules.

\paragraph{Reflection.}
The ability to dynamically configure a module implies that a module must be able to know its configuration status to preserve certain properties.
For example, consider a module for a reliable FIFO queue.
Since the queue can be dynamically configured, the queue must be able to determine if the event produced by removing and returning an item from the queue will be delivered.
Otherwise, the queue will lose data items and therefore not be reliable.

\subsection{I/O Automata}

The Input/Output (I/O) automata formal model developed by Lynch~\cite{distributed_algorithms} meets our requirements for local state and atomic asynchronous events.
An I/O automaton consists of state variables and atomic actions that manipulate the state.
The three action types are inputs, outputs, and internal actions.
Input actions receive a signal or a value and output actions produce a signal or a value.
Internal actions just manipulate the state of the automaton.
Output and internal actions are \emph{locally controlled actions}~\cite{distributed_algorithms} consisting of a precondition and an effect.
Input and output actions are \emph{external actions}~\cite{distributed_algorithms} because they allow an automaton to communicate with other automata.
Automata may \emph{composed} by concatenating the state variables of the constituent automata and incorporating the effects of input actions into output actions with the same name.
A useful operation in composite automata is \emph{hiding} which converts an output action to an internal action.
Composing automata results in a static configuration, i.e., the set of automata and their interactions are fixed.
Execution in the I/O automata model consists of repeatedly selecting a local action and then applying the effect if the precondition is true.
The scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not executed) infinitely often.

\subsection{Extending I/O Automata for Dynamics}

A standard technique for modeling dynamic modules in I/O automata is to assume they exist but are inactive.
Dynamic automata can then be activated and deactivated using inputs.
This approach relies on the static configuration produced by composition and is therefore not suitable for an implementation.

We extend I/O automata for dynamic configuration by introducing a \emph{system automaton} that records the configuration of all automata in the system.
We differentiate between automaton types and automaton instances.
Every automaton instances has a unique \emph{automaton identifier (aid)}.
The state of the system automaton consists of the set of aids $A$ indicating what automata exist and a set of binding records $B$ indicating the relationship between output and input actions.
A binding record $(out, output, in, input)$ is tuple consisting of the aid of the output automaton $out$, the name of the output action $output$, the aid of the input automaton $in$, and the name of the input action $input$.

The configuration contained in set of binding records must be a valid static configuration to ensure that the behavior of the complete system remains analyzable under the I/O automata model.
First, the output automaton and input automaton must exist $\forall (out, output, in, input) \in B: out \in A \land in \in A$.
% TODO:  Binding must not exist.
Second, an input can only be bound to one output.
Thus, a given $in$ and $input$ combination can appear at most once in $B$.
Third, an output cannot be bound to an input in the same automaton.
Thus, $out \neq in$ for every $(out, output, in, input)$ in $B$.
Fourth, an output cannot be bound to two inputs in the same automaton.
Thus, $in_1 \neq in_2$ for all $(out, output, in_1, input_1), (out, output, in_2, input_2)$ in $B$.

The system automaton contains input actions for creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton and output actions that indicate the result of a create, bind, unbind, or destroy.
All automata are composed with (not bound to) the system automaton and contain outputs for creating, binding, unbinding, and destroying and inputs for receiving results.
Systems calls (create, bind, unbind, destroy) resemble a request-response protocol where an automaton sends a request, the system automaton receives the request, the system automaton then sends the result, and the automaton receives the result.
This has the important consequence that \emph{configuration is asynchronous}.
Destroying an automaton has the effect of removing all of its bindings.
An automaton that destroys itself can therefore cause the system to deliver of unbound and destroyed events without a corresponding unbind or destroy.

\subsection{Extending I/O Automata for Reflection}

Automata are endowed with additional inputs driven by system outputs indicating when an action has been bound and unbound.
Automata might use the bound input to enable processing, i.e., someone is listening so I'll start producing data.
The unbound input is useful for cleaning data structures and inhibiting further processing, i.e., no one is listening so I'll stop producing data.
The bound and unbound inputs are useful but they are also asynchronous, thus, an automaton cannot trust the status reported by the last bound or unbound for use in, for example, a precondition.

The asynchronous nature of bound and unbound requires a primitive that can examine the contents of $B$ instantaneously.
Thus, we make one exception to the rule of local state that says automata can examine the contents of $B$ to determine their current configuration.
This is permissible so long as concurrent implementations provide multiple-reader/single-writer semantics to $B$.

\subsection{Equivalence}

Extending the I/O automata model for dynamics and reflection is useless unless we can relate the extensions back to the original model or establish equivalence.
To do this, we will take advantage of the fair traces (?) property of I/O automata~\cite{distributed_algorithms}.
%% We accept an execution trace in the dynamic setting if it is a fair trace of the static model.

%% Modeling a system with I/O automata implies a fixed set of automata given by $A_0$ and set of binding records $B_0$.

Start with a trace of dynamic $T_d$.
Remove system actions from $T_d$ so $T_d'$.
We accept $T_d'$ if it is a fair trace of the static model.

Let $A_0$ be the set of automata and $B_0$ be the set of binding records in the static model.
If you want to prove correspondence, you must prove that eventually $A = A_0$ and $B = B_0$.
We also need to show that no output is executed before it is fully bound.

%% supports the actions of creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton.
%% Upon creation, All automata are bound
%% The four operations necessary for dynamic configuration are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.

%% \begin{outline}
%% \item State
%%   \begin{outline}
%%   \item There is no shared state in a 
%%   \item Local state only
%%   \item Shared state
%%     \begin{outline}
%%       \item Impossible in distributed systems
%%       \item Dangerous in local systems
%%     \end{outline}
%%   \end{outline}
%% \item Communication
%%   \begin{outline}
%%   \item Atomic asynchronous message passing
%%   \item Network sets size of atom (UDP)
%%   \item Can build reliable streams (TCP)
%%   \item Local equivalent is passing a value
%%   \item Model should lend itself to writing protocols
%%   \end{outline}
%% \item Asynchrony
%%   \begin{outline}
%%     \item Model must have natural support for asynchrony, i.e., event-based
%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
%%   \end{outline}
%% \item Concurrency
%%   \begin{outline}
%%     \item Reason about systems using non-deterministic interleaving of atomic actions
%%     \item Model should admit implementations that execute concurrently
%%   \end{outline}
%% \item Dynamics
%%   \begin{outline}
%%     \item Configuration - Edges in graph of communicating components can change at run-time.
%%       \begin{outline}
%%       \item Already required in distributed settings
%%       \item Not addressed in formal models
%%       \end{outline}
%%     \item Extension - Nodes in graph of communicating components can change at run-time.
%%   \end{outline}
%%   \item Reflection
%% \end{outline}

%% I/O Automata
%% \begin{itemize}
%%   \item Compare with UNITY
%%   \item Compare with esterel
%%   \item Compare with pi calculus
%%   \item Compare with Ptolemy
%% \end{itemize}

### 2011-07-19 papers/paper1/abstract.tex
Abstract edits.

@@
 \begin{abstract}
-In an effort to be portable, system developers often favor existing architectures, languages, operating systems, and standard libraries.
-The dominant abstraction in established computing technologies for concurrency and therefore distributed system development is threads.
-Consequently, distributed system developers are locked into using threads and experience the difficulties of thread-based development including obscure race conditions, poor support for the semantics of distributed systems, and difficulties integrating software developed using different thread models.
-Furthermore, the demand for portability and generality dismisses many of the techniques proposed in research that involve new programming languages, language extensions, or domain-specific solutions.
-This paper presents a development framework based on the I/O automata formal model implemented in C++.
-The I/O automata model is a proven technique for distributed system design because its semantics match those found in distributed systems.
-With the framework, developers can reason about correctness directly from the program text using the I/O automata formalism.
-The I/O  automata model also provides a common abstraction for concurrency which facilitates software integration and promotes reuse.
-TODO:  What are we doing for evaluation?
-%% Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
-%% Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
-%% In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
-%% We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
-%% Finally, we conclude with an evaluation of our I/O automata framework.
+The ability to realize sophisticated distributed systems such as those found in cyber-physical systems, pervasive computing, and enterprise computing is predicated on a suitable model and implementation of asynchrony and concurrency.
+The size and complexity of these systems implies that software re-use among their components is also essential.
+Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized in reusable software.
+% By different forms we meen single-threaded, multi-threaded, time triggered, event triggered, etc.
+
+However, existing concurrency abstractions are inadequate to the needs of many sophisticated distributed systems, and new approaches are needed.
+While threads match the semantics of widely available synchronous processors, thread-based development may introduce hazards for the distributed system developer including obscure race conditions, poor support for asynchrony, and difficulties integrating software developed using different thread models.
+An alternative concurrency abstraction that matches the semantics of distributed systems and facilitates software re-use is therefore needed.
+
+The reactive semantics I/O automata formal model match the asynchronous and concurrent semantics of distributed systems and the I/O automata formal model has been widely used in the design and verification of real distributed systems.
+However, there is to date insufficient experience \emph{building} and evaluating real distributed systems directly atop I/O automata.
+Previous efforts in this area have focused on formal software development, i.e., automata are described in a language suitable for theorem proving and simulation and aftward are translated into a production language for deployment.
+The translation from modeling language to production language complicates interfacing with system facilities and makes developing low-level protocols difficult, an area which could greatly benefit from the direct application of I/O automata.
+
+To this end, we contribute the \emph{ioa++ framework} for developing distributed systems.
+In this paper, we outline a system model for distributed systems and connect these requirements to the I/O automata formal model.
+We then describe how ioa++ is implemented in C++ and offer a number of examples showing how ioa++ can be used to develop distributed systems.
+
+Our results \comment{experiences?} show that ioa++ can help simplify developing distributed systems.
+Ioa++ allows developers to reason about asynchronous and concurrent programs directly from the source code using the I/O automata formal model.
+Where reasoning about the correctness of a module in a threaded program requires considering interactions among the various threads, reasoning about the correctness of a module in ioa++ is limited to the module itself.
+Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one thread model to another.
 \end{abstract}

### 2011-07-22 papers/paper1/abstract.tex
Updated abstract and introduction.

@@
 \begin{abstract}
+
+% What is asynchrony?
 The ability to realize sophisticated distributed systems such as those found in cyber-physical systems, pervasive computing, and enterprise computing is predicated on a suitable model and implementation of asynchrony and concurrency.
-The size and complexity of these systems implies that software re-use among their components is also essential.
+The size and complexity of these systems implies that software re-use in their development, maintenance, and evaluation is also essential.
 Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized in reusable software.
@@
 
-However, existing concurrency abstractions are inadequate to the needs of many sophisticated distributed systems, and new approaches are needed.
+% What is a thread model?
+However, existing concurrency abstractions are inadequate to the needs of many sophisticated distributed systems, and new approaches must be pursued.
 While threads match the semantics of widely available synchronous processors, thread-based development may introduce hazards for the distributed system developer including obscure race conditions, poor support for asynchrony, and difficulties integrating software developed using different thread models.
-An alternative concurrency abstraction that matches the semantics of distributed systems and facilitates software re-use is therefore needed.
+Events, while matching the semantics of distributed systems more closely, are often inflexible, ad hoc, and lack a standard calculus to facilitate the creation of re-usable event-based modules.
+Alternative concurrency abstractions that more closely match the semantics of distributed systems \emph{and} facilitate software re-use are therefore needed.
+% We aren't saying I/O automata are the best, just better than threads.
 
-The reactive semantics I/O automata formal model match the asynchronous and concurrent semantics of distributed systems and the I/O automata formal model has been widely used in the design and verification of real distributed systems.
-However, there is to date insufficient experience \emph{building} and evaluating real distributed systems directly atop I/O automata.
-Previous efforts in this area have focused on formal software development, i.e., automata are described in a language suitable for theorem proving and simulation and aftward are translated into a production language for deployment.
+The reactive semantics of the I/O automata model match the asynchronous and concurrent semantics of distributed systems well and has been proven effective in the formal design and verification of real world distributed systems.
+However, there is to date insufficient experience \emph{building} and \emph{evaluating} real distributed systems directly atop I/O automata.
+Previous efforts in this area have focused mainly on formal software development, i.e., automata are described in a language suitable for theorem proving and simulation and aftward are translated into a production language for deployment.
 The translation from modeling language to production language complicates interfacing with system facilities and makes developing low-level protocols difficult, an area which could greatly benefit from the direct application of I/O automata.
 
-To this end, we contribute the \emph{ioa++ framework} for developing distributed systems.
-In this paper, we outline a system model for distributed systems and connect these requirements to the I/O automata formal model.
+To this end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
+We outline a system model for distributed systems and connect these requirements to the I/O automata formal model.
 We then describe how ioa++ is implemented in C++ and offer a number of examples showing how ioa++ can be used to develop distributed systems.
 
-Our results \comment{experiences?} show that ioa++ can help simplify developing distributed systems.
+Our evalutations show that ioa++ can help simplify developing distributed systems.
 Ioa++ allows developers to reason about asynchronous and concurrent programs directly from the source code using the I/O automata formal model.
 Where reasoning about the correctness of a module in a threaded program requires considering interactions among the various threads, reasoning about the correctness of a module in ioa++ is limited to the module itself.
-Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one thread model to another.
+Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one thread model or event system to another.
 \end{abstract}

### 2011-07-25 papers/paper1/system_model.tex
New introduction and part of system model.

@@
 
-\subsection{Requirements}
+\subsection{Component Architecture}
 
-Our system model is motivated by the semantics of devices, the semantics of distributed systems, and the desire for conceptual integrity.
-We adopt a recursive view that nodes, i.e., devices, in a distributed system may themselves be represented as a distributed system consisting of different sensors, actuators, and processes.
-We use the term \emph{node} to refer to a participant in a typical distributed system, i.e., one that uses standard network links.
-We use term \emph{module} to refer to a computational element within a node.
-Modules that reside on the same node are said to be \emph{collocated}.
+Components are the basis of any re-usable concurrent software systems.
+We adopt Szyperski's definition of components~\cite{component_software}:
+\begin{quote}
+  A software component is a unit of composition with contractually specified interfaces and explicit context dependencies only.
+  A software component can be deployed independently and is subject to composition by third parties. 
+\end{quote}
+A component architecture defines components and how they execute and interact.
+For example, a component in a multi-threaded component architecture consists of private state that is accessed through synchronous function call.
+Components can be configured to use one another and can execute concurrently.
+This section outlines a component model and architecture for an event-based system.
 
-\paragraph{Local State.}
-We require that all modules and by extension all nodes only manipulate local state.
-Said another way, there is no global or shared state.
-Computation designed around the manipulation of local state only is easily modularized and consequently, easier to distribute and execute concurrently.
+\paragraph{Components.}
+A \emph{component} consists of state variables and a set of event handlers.
+Each event handler is only allowed to modify the state of the component to which it belongs.
+The absence of shared state ensures that each component can be deployed independently.
+Event handlers can either be an \emph{output} that produces a signal or value, an \emph{input} that consumes a signal or value, or an \emph{internal} that neither produces nor consumes a signal or value.
+Components can be created and destroyed at run-time.
+
+\paragraph{Binding.}
+A \emph{binding} is an association between an output and an input.
+An output can be \emph{bound} to zero or more inputs subject to type compatibility.
+An input can be bound at most once.
+Outputs and inputs can either be \emph{public}, available for binding by external components, or \emph{private}, only available for binding by the containing component.
+Internals are always private.
+Events can be bound and unbound at run-time.
+The component requesting that an output be bound to and input is said to be the \emph{owner} of the binding.
+Private outputs and inputs can only be bound by the containing component.
+
+\paragraph{Execution.}
+Execution proceeds by randomly selecting and atomically executing outputs and internals.
+When an output is executed, both the output all bound inputs are executed in one atomic step receiving the value produced by the output if applicable.
+Two events can be executed concurrently if the set of components implied by each event are disjoint.
+All components have outputs for creating, binding, unbinding, and destroying and inputs for receive results that are permanently bound to ``the system.''
+System calls follow a request-reply system where a component sends a request for a create, bind, unbind, or destroy, the system receives the request, the system processes the request, the system sends a reply, and the component receives the reply.
+
+\paragraph{Composition.}
+Binding, atomic execution, and the absence of shared state allows components to be composed, i.e., combined in such a way as to yield a valid component.
+Components can be composed by 1) concatenating their state variables, 2) renaming events so every event's name is unique, and 3) folding input events into the output events to which they are bound.
+Composition allows the programmer to reason about a constellation of components as a single component.
+Often, the properties of the resulting component can be reasoned about using the properties of the constituent components.
+Components can be composed at compile time using the process outlined above, however, it is often better to preserve component boundaries to take advantage of concurrent execution.
+
+\begin{figure}
+\center
+\includegraphics{system_model}
+\caption{Example composable event-based system.
+  Components are depicted as solid rectangles, e.g., Component3.
+  Events (outputs, inputs, and internals) are depicted as dotted rectangles, e.g., \emph{Ritz?}.
+  Output events are suffixed with \emph{!}.
+  Inputs are suffixed with \emph{?}.
+  Internals have no suffix.
+  Public events start with an uppercase letter.
+  Private events start with a lowercase letter.
+  Bindings are indicate by dotted lines with an arrow pointing to the input.}
+\label{sys_model}
+\end{figure}
+
+\paragraph{Example.}
+Figure~\ref{sys_model} depicts a fictitious system of three components.
+Component3 contains a public input \emph{Ritz?} and a public output \emph{Blitz!}.
+Component3 also contains a private output \emph{boom!(t1)} that produces a value of type \emph{t1} and a private input \emph{jot?}.
+Component3 contains two components: Component1 and Component2.
+Component1 contains a public input \emph{Bam?(t1)}, a public output \emph{Foo!}, and a private internal \emph{think}.
+Component2 contains a public input \emph{Bang?(t1)}, a public input \emph{Frob?}, and public output \emph{Jig!}.
+All bindings are owned by Component3.
+The output \emph{boom!(t1)} is bound to inputs \emph{Bam?(t1)} and \emph{Bang?(t1)} and shows how one output can be bound to multiple inputs.
+The output \emph{Foo!} is bound to input \emph{Frob?} and shows how an output in one component can be bound to input in another component by a third component.
+
+Opportunities for concurrent execution exist for the system depicted in figure~\ref{sys_model}.
+The set of components implied by \emph{Blitz!} is \{Component3\} while the set of components implied by \emph{Foo!} is \{Component1, Component2\}.
+Since the sets are disjoint, \emph{Blitz!} and \emph{Foo!} can be executed concurrently.
+The set of components implied by \emph{boom!(t1)} is \{Component1, Component2, Component3\}.
+Consequently, \emph{boom!(t1)} cannot be executed concurrently with any other event.
+
+%% \subsection{Requirements}
+
+%% Our system model is motivated by the semantics of devices, the semantics of distributed systems, and the desire for conceptual integrity.
+%% We adopt a recursive view that nodes, i.e., devices, in a distributed system may themselves be represented as a distributed system consisting of different sensors, actuators, and processes.
+%% We use the term \emph{node} to refer to a participant in a typical distributed system, i.e., one that uses standard network links.
+%% We use term \emph{module} to refer to a computational element within a node.
+%% Modules that reside on the same node are said to be \emph{collocated}.
+
+%% \paragraph{Local State.}
+%% We require that all modules and by extension all nodes only manipulate local state.
+%% Said another way, there is no global or shared state.
+%% Computation designed around the manipulation of local state only is easily modularized and consequently, easier to distribute and execute concurrently.
 
@@
 
-\paragraph{Asynchronous Atomic Events.}
-Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
-This implies that the model contains a general treatment of events, i.e., modules can both produce and consume events.
-Modules, therefore, are reactive and structured as a set of atomic event produces and consumers.
-The guarantee of atomicity obviates the need for concurrency control primitives and improves our ability to reason about the behavior of the module.
-Modules communicate using events according to a \emph{configuration} that determines which events are delivered to each module.
+%% \paragraph{Asynchronous Atomic Events.}
+%% Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
+%% This implies that the model contains a general treatment of events, i.e., modules can both produce and consume events.
+%% Modules, therefore, are reactive and structured as a set of atomic event produces and consumers.
+%% The guarantee of atomicity obviates the need for concurrency control primitives and improves our ability to reason about the behavior of the module.
+%% Modules communicate using events according to a \emph{configuration} that determines which events are delivered to each module.
 
@@
 
-\paragraph{Dynamics.}
-The set of nodes and communication links is dynamic in real distributed systems.
-Similarly, the set of modules and their configuration in real devices can change as the device may be re-purposed by users over time.
-In general, our system model must support a dynamic number of modules and, consequently, dynamic configuration of those modules.
+%% \paragraph{Dynamics.}
+%% The set of nodes and communication links is dynamic in real distributed systems.
+%% Similarly, the set of modules and their configuration in real devices can change as the device may be re-purposed by users over time.
+%% In general, our system model must support a dynamic number of modules and, consequently, dynamic configuration of those modules.
 
-\paragraph{Reflection.}
-The ability to dynamically configure a module implies that a module must be able to know its configuration status to preserve certain properties.
-For example, consider a module for a reliable FIFO queue.
-Since the queue can be dynamically configured, the queue must be able to determine if the event produced by removing and returning an item from the queue will be delivered.
-Otherwise, the queue will lose data items and therefore not be reliable.
+%% \paragraph{Reflection.}
+%% The ability to dynamically configure a module implies that a module must be able to know its configuration status to preserve certain properties.
+%% For example, consider a module for a reliable FIFO queue.
+%% Since the queue can be dynamically configured, the queue must be able to determine if the event produced by removing and returning an item from the queue will be delivered.
+%% Otherwise, the queue will lose data items and therefore not be reliable.
 
@@
 
-The Input/Output (I/O) automata formal model developed by Lynch~\cite{distributed_algorithms} meets our requirements for local state and atomic asynchronous events.
+The Input/Output (I/O) automata model was developed by Lynch~\cite{distributed_algorithms} to model reactive concurrent and asynchronous systems.
 An I/O automaton consists of state variables and atomic actions that manipulate the state.
@@
 Internal actions just manipulate the state of the automaton.
-Output and internal actions are \emph{locally controlled actions}~\cite{distributed_algorithms} consisting of a precondition and an effect.
+Output and internal actions are \emph{locally controlled actions}~\cite{distributed_algorithms} are often presented as precondition and an effect.
 Input and output actions are \emph{external actions}~\cite{distributed_algorithms} because they allow an automaton to communicate with other automata.
@@
 
-\subsection{Extending I/O Automata for Dynamics}
+The proposed system has a straightforward mapping to the I/O automata model: components are automata and events are actions.
+However, the I/O automata model assumes a static configuration while the proposed system requires the ability to deal with dynamic configurations, i.e., creating, binding, unbinding, and destroying.
+
+\subsection{I/O Automata and Dynamic Configuration}
 
@@
 Dynamic automata can then be activated and deactivated using inputs.
-This approach relies on the static configuration produced by composition and is therefore not suitable for an implementation.
+We will use a similar technique to model dynamic configurations.
 
-We extend I/O automata for dynamic configuration by introducing a \emph{system automaton} that records the configuration of all automata in the system.
+We introduce a \emph{system automaton} that records the configuration of all automata in the system.
 We differentiate between automaton types and automaton instances.
@@
 The system automaton contains input actions for creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton and output actions that indicate the result of a create, bind, unbind, or destroy.
-All automata are composed with (not bound to) the system automaton and contain outputs for creating, binding, unbinding, and destroying and inputs for receiving results.
-Systems calls (create, bind, unbind, destroy) resemble a request-response protocol where an automaton sends a request, the system automaton receives the request, the system automaton then sends the result, and the automaton receives the result.
+All automata are composed with the system automaton and contain outputs for creating, binding, unbinding, and destroying and inputs for receiving results.
+Systems calls (create, bind, unbind, destroy) resemble a request-response protocol where an automaton sends a request, the system automaton receives the request, the system processes the request, the system automaton then sends the result, and the automaton receives the result.
 This has the important consequence that \emph{configuration is asynchronous}.
 Destroying an automaton has the effect of removing all of its bindings.
-An automaton that destroys itself can therefore cause the system to deliver of unbound and destroyed events without a corresponding unbind or destroy.
+An automaton that destroys itself can therefore cause the system to deliver unbound and destroyed events without an unbind or destroy.
+
+Using the system automaton as a bookkeeping mechanism, we can relate a statically composed system to a dynamically configured one.
+The key idea is to show that actions belong to automaton $a$ are only executed when $a \in A$ and that an output action is only executed when $B$ contains all of bindings associated with the static configuration.
 

### 2011-08-15 papers/paper1/abstract.tex
Moving on to section 2.

@@
 
-% What is asynchrony?
 The ability to realize sophisticated distributed systems such as those found in cyber-physical systems, pervasive computing, and enterprise computing is predicated on a suitable model and implementation of asynchrony and concurrency.
@@
 Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized in reusable software.
-% By different forms we meen single-threaded, multi-threaded, time triggered, event triggered, etc.
-
-% What is a thread model?
-However, existing concurrency abstractions are inadequate to the needs of many sophisticated distributed systems, and new approaches must be pursued.
-While threads match the semantics of widely available synchronous processors, thread-based development may introduce hazards for the distributed system developer including obscure race conditions, poor support for asynchrony, and difficulties integrating software developed using different thread models.
-Events, while matching the semantics of distributed systems more closely, are often inflexible, ad hoc, and lack a standard calculus to facilitate the creation of re-usable event-based modules.
-Alternative concurrency abstractions that more closely match the semantics of distributed systems \emph{and} facilitate software re-use are therefore needed.
-% We aren't saying I/O automata are the best, just better than threads.
-
-The reactive semantics of the I/O automata model match the asynchronous and concurrent semantics of distributed systems well and has been proven effective in the formal design and verification of real world distributed systems.
-However, there is to date insufficient experience \emph{building} and \emph{evaluating} real distributed systems directly atop I/O automata.
-Previous efforts in this area have focused mainly on formal software development, i.e., automata are described in a language suitable for theorem proving and simulation and aftward are translated into a production language for deployment.
-The translation from modeling language to production language complicates interfacing with system facilities and makes developing low-level protocols difficult, an area which could greatly benefit from the direct application of I/O automata.
+% By ``different forms'' we mean single-threaded, multi-threaded, time triggered, event triggered, etc.
 
@@
 Ioa++ allows developers to reason about asynchronous and concurrent programs directly from the source code using the I/O automata formal model.
-Where reasoning about the correctness of a module in a threaded program requires considering interactions among the various threads, reasoning about the correctness of a module in ioa++ is limited to the module itself.
+Where reasoning about the correctness of modules in a threaded program requires considering interactions among the various threads, reasoning about the correctness of modules in ioa++ is limited to the modules themselves and a well-defined set of composition rules.
 Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one thread model or event system to another.

### 2011-08-15 papers/paper1/system_model.tex
Moving on to section 2.

@@
-\section{System Model and Formalization\label{system_model}}
+\section{Component Model and Formalization\label{system_model}}
 
-\subsection{Component Architecture}
+\subsection{Component Model}
 
-Components are the basis of any re-usable concurrent software systems.
-We adopt Szyperski's definition of components~\cite{component_software}:
-\begin{quote}
-  A software component is a unit of composition with contractually specified interfaces and explicit context dependencies only.
-  A software component can be deployed independently and is subject to composition by third parties. 
-\end{quote}
-A component architecture defines components and how they execute and interact.
-For example, a component in a multi-threaded component architecture consists of private state that is accessed through synchronous function call.
-Components can be configured to use one another and can execute concurrently.
-This section outlines a component model and architecture for an event-based system.
+Modularity, well-defined interfaces, and composition are properties of reusable software.
+Modularity implies that a unit of software can be deployed independently of other units.
+Well-defined interfaces and interactions allow modules to expose their functionality in a regular way that facilitates manual and automated reasoning about interactions with the module.
+Composition allows a group of interacting modules to be repackaged and understood as a single module.
+A unit of software that exhibits these properties is called a \emph{component}~\cite{szyperski2002component}.
+This section outlines the component model for our event-based system.
 
 \paragraph{Components.}
-A \emph{component} consists of state variables and a set of event handlers.
+A \emph{component} in our system consists of state variables and a set of event handlers.
 Each event handler is only allowed to modify the state of the component to which it belongs.
 The absence of shared state ensures that each component can be deployed independently.
-Event handlers can either be an \emph{output} that produces a signal or value, an \emph{input} that consumes a signal or value, or an \emph{internal} that neither produces nor consumes a signal or value.
-Components can be created and destroyed at run-time.
+An event handler can either be an \emph{output} that produces a signal or value, an \emph{input} that consumes a signal or value, or an \emph{internal} that neither produces nor consumes a signal or value.
+% Components can be created and destroyed at run-time.
 
@@
 An input can be bound at most once.
-Outputs and inputs can either be \emph{public}, available for binding by external components, or \emph{private}, only available for binding by the containing component.
+Outputs and inputs can either be \emph{public}, available for binding by external components, or \emph{private}, only available for binding by the component that contains them.
 Internals are always private.
-Events can be bound and unbound at run-time.
-The component requesting that an output be bound to and input is said to be the \emph{owner} of the binding.
-Private outputs and inputs can only be bound by the containing component.
+%Events can be bound and unbound at run-time.
+The component requesting that an output be bound to an input is said to be the \emph{owner} of the binding.
+Private outputs and inputs can only be bound by the component that contains them.
 
 \paragraph{Execution.}
-Execution proceeds by randomly selecting and atomically executing outputs and internals.
-When an output is executed, both the output all bound inputs are executed in one atomic step receiving the value produced by the output if applicable.
+Execution proceeds by non-deterministically selecting and atomically executing outputs and internals.
+When an output is executed, both the output and all bound inputs are executed in one atomic step receiving the value produced by the output if applicable.
 Two events can be executed concurrently if the set of components implied by each event are disjoint.
-All components have outputs for creating, binding, unbinding, and destroying and inputs for receive results that are permanently bound to ``the system.''
-System calls follow a request-reply system where a component sends a request for a create, bind, unbind, or destroy, the system receives the request, the system processes the request, the system sends a reply, and the component receives the reply.
+%% All components have outputs for creating, binding, unbinding, and destroying and inputs for receiving results that are permanently bound to ``the system.''
+%% System calls follow a request-reply system where a component sends a request for a create, bind, unbind, or destroy, the system receives the request, the system processes the request, the system sends a reply, and the component receives the reply.
 
 \paragraph{Composition.}
-Binding, atomic execution, and the absence of shared state allows components to be composed, i.e., combined in such a way as to yield a valid component.
-Components can be composed by 1) concatenating their state variables, 2) renaming events so every event's name is unique, and 3) folding input events into the output events to which they are bound.
+Binding, atomic execution, and the absence of shared state allows components to be composed.
+Components are composed by 1) concatenating their state variables, 2) renaming events so every event's name is unique, and 3) folding input events into the output events to which they are bound.
 Composition allows the programmer to reason about a constellation of components as a single component.
@@
 The set of components implied by \emph{Blitz!} is \{Component3\} while the set of components implied by \emph{Foo!} is \{Component1, Component2\}.
-Since the sets are disjoint, \emph{Blitz!} and \emph{Foo!} can be executed concurrently.
+Since the sets and therefore set of state variables are disjoint, \emph{Blitz!} and \emph{Foo!} can be executed concurrently.
 The set of components implied by \emph{boom!(t1)} is \{Component1, Component2, Component3\}.
@@
 
-\subsection{I/O Automata}
+\subsection{Formalization}
 
-The Input/Output (I/O) automata model was developed by Lynch~\cite{distributed_algorithms} to model reactive concurrent and asynchronous systems.
+The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
 An I/O automaton consists of state variables and atomic actions that manipulate the state.
@@
 Internal actions just manipulate the state of the automaton.
-Output and internal actions are \emph{locally controlled actions}~\cite{distributed_algorithms} are often presented as precondition and an effect.
-Input and output actions are \emph{external actions}~\cite{distributed_algorithms} because they allow an automaton to communicate with other automata.
+Output and internal actions are \emph{locally controlled actions}~\cite{lynch1996distributed} are often presented as precondition and an effect.
+The precondition is a predicate over the state variables that enables the effect when selected.
+The effect is a function that computes the next state of the automaton from the current state.
+Input and output actions are \emph{external actions}~\cite{lynch1996distributed} because they allow an automaton to communicate with other automata.
 Automata may \emph{composed} by concatenating the state variables of the constituent automata and incorporating the effects of input actions into output actions with the same name.
-A useful operation in composite automata is \emph{hiding} which converts an output action to an internal action.
+A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.
 Composing automata results in a static configuration, i.e., the set of automata and their interactions are fixed.
@@
 The proposed system has a straightforward mapping to the I/O automata model: components are automata and events are actions.
+
+\section{Dynamics}
+
 However, the I/O automata model assumes a static configuration while the proposed system requires the ability to deal with dynamic configurations, i.e., creating, binding, unbinding, and destroying.

### 2011-08-16 papers/paper1/dynamics.tex
Revision of Section 2.  New example.

--- 2011-07-13 papers/paper1/system_model.tex
@@
-\section{System Model and Formalization\label{system_model}}
+\section{Dynamics}
 
-\subsection{Requirements}
+However, the I/O automata model assumes a static configuration while the proposed system requires the ability to deal with dynamic configurations, i.e., creating, binding, unbinding, and destroying.
 
-Our system model is motivated by the semantics of devices, the semantics of distributed systems, and the desire for conceptual integrity.
-We adopt a recursive view that nodes, i.e., devices, in a distributed system may themselves be represented as a distributed system consisting of different sensors, actuators, and processes.
-We use the term \emph{node} to refer to a participant in a typical distributed system, i.e., one that uses standard network links.
-We use term \emph{module} to refer to a computational element within a node.
-Modules that reside on the same node are said to be \emph{collocated}.
-
-\paragraph{Local State.}
-We require that all modules and by extension all nodes only manipulate local state.
-Said another way, there is no global or shared state.
-Computation designed around the manipulation of local state only is easily modularized and consequently, easier to distribute and execute concurrently.
-
-%% \paragraph{Message Passing.}
-%% Nodes in distributed systems interact by passing messages over communication links.
-%% Message passing can either be synchronous or asynchronous.
-%% In synchronous message passing, the events of message production and delivery are the same, i.e., atomic.
-%% In asynchronous message passing, the events of message production and delivery are logically distinct events subject to being interposed by other events.
-%% Synchronous message passing is most often used between modules on the same node while asynchronous message passing is most often used between nodes.
-%% Both synchronous and asynchronous message passing are required.
-%% Synchronous message passing is necessary for building composable systems because of it association with atomicity.
-%% Asynchronous message passing is necessary for building real distributed systems as this matches the semantics of the networks found in practice.
-
-\paragraph{Asynchronous Atomic Events.}
-Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
-This implies that the model contains a general treatment of events, i.e., modules can both produce and consume events.
-Modules, therefore, are reactive and structured as a set of atomic event produces and consumers.
-The guarantee of atomicity obviates the need for concurrency control primitives and improves our ability to reason about the behavior of the module.
-Modules communicate using events according to a \emph{configuration} that determines which events are delivered to each module.
-
-%% Communicating with external resources such as the operating system or physical environment is foundational to writing useful programs.
-%% Consider the history of I/O in the C run-time library.
-%% The synchronous function call semantics of the C programming language made it possible to write system calls, e.g., read and write, that transfer control to the operating system which then completes the operation and return control to the process after the system call.
-%% This is called \emph{blocking I/O} because the process blocks while it waits for the operating system to complete the I/O.
-%% Blocking I/O has the severe restriction that only one I/O source can be used at a time.
-%% To remedy this, system calls such as select and poll were introduced that allow a process to monitor I/O sources for events and then respond to these events.
-%% Asynchronous I/O was a further advancement that allows for the system (instead of the process) to dispatch a function upon the completion of I/O.
-%% Compare this to UNIX signals that are neither general (fixed-number with no associated data) nor principled.
-%% Interrupts are difficult to use correctly.
-
-\paragraph{Dynamics.}
-The set of nodes and communication links is dynamic in real distributed systems.
-Similarly, the set of modules and their configuration in real devices can change as the device may be re-purposed by users over time.
-In general, our system model must support a dynamic number of modules and, consequently, dynamic configuration of those modules.
-
-\paragraph{Reflection.}
-The ability to dynamically configure a module implies that a module must be able to know its configuration status to preserve certain properties.
-For example, consider a module for a reliable FIFO queue.
-Since the queue can be dynamically configured, the queue must be able to determine if the event produced by removing and returning an item from the queue will be delivered.
-Otherwise, the queue will lose data items and therefore not be reliable.
-
-\subsection{I/O Automata}
-
-The Input/Output (I/O) automata formal model developed by Lynch~\cite{distributed_algorithms} meets our requirements for local state and atomic asynchronous events.
-An I/O automaton consists of state variables and atomic actions that manipulate the state.
-The three action types are inputs, outputs, and internal actions.
-Input actions receive a signal or a value and output actions produce a signal or a value.
-Internal actions just manipulate the state of the automaton.
-Output and internal actions are \emph{locally controlled actions}~\cite{distributed_algorithms} consisting of a precondition and an effect.
-Input and output actions are \emph{external actions}~\cite{distributed_algorithms} because they allow an automaton to communicate with other automata.
-Automata may \emph{composed} by concatenating the state variables of the constituent automata and incorporating the effects of input actions into output actions with the same name.
-A useful operation in composite automata is \emph{hiding} which converts an output action to an internal action.
-Composing automata results in a static configuration, i.e., the set of automata and their interactions are fixed.
-Execution in the I/O automata model consists of repeatedly selecting a local action and then applying the effect if the precondition is true.
-The scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not executed) infinitely often.
-
-\subsection{Extending I/O Automata for Dynamics}
+\subsection{I/O Automata and Dynamic Configuration}
 

### 2011-08-16 papers/paper1/system_model.tex
Revision of Section 2.  New example.

--- 2011-07-13 papers/paper1/system_model.tex
@@
-\section{System Model and Formalization\label{system_model}}
+\section{Component Model and Formalization\label{system_model}}
 
-\subsection{Requirements}
+\subsection{Component Model}
 
-Our system model is motivated by the semantics of devices, the semantics of distributed systems, and the desire for conceptual integrity.
-We adopt a recursive view that nodes, i.e., devices, in a distributed system may themselves be represented as a distributed system consisting of different sensors, actuators, and processes.
-We use the term \emph{node} to refer to a participant in a typical distributed system, i.e., one that uses standard network links.
-We use term \emph{module} to refer to a computational element within a node.
-Modules that reside on the same node are said to be \emph{collocated}.
+Modularity, well-defined interfaces, and composition are properties of reusable software.
+Modularity implies that a unit of software can be deployed independently of other units.
+Well-defined interfaces and interactions allow modules to expose their functionality in a regular way that facilitates manual and automated reasoning about interactions with the module.
+Composition allows a group of interacting modules to be repackaged and understood as a single module.
+A unit of software that exhibits these properties is called a \emph{component}~\cite{szyperski2002component}.
+This section outlines the component model for our event-based system.
 
-\paragraph{Local State.}
-We require that all modules and by extension all nodes only manipulate local state.
-Said another way, there is no global or shared state.
-Computation designed around the manipulation of local state only is easily modularized and consequently, easier to distribute and execute concurrently.
+\paragraph{Components.}
+A \emph{component} in our model consists of state variables and a set of event handlers.
+Each event handler is only allowed to modify the state of the component with which it is associated.
+The absence of shared state ensures that each component can be deployed independently.
+An event handler can either be an \emph{output handler} that produces a signal or value, an \emph{input handler} that consumes a signal or value, or an \emph{internal handler} that neither produces nor consumes a signal or value.
+Event handlers that produce or consume signals are said to be \emph{unvalued} while event handler that produce or consume values are said to be \emph{valued}.
+% Components can be created and destroyed at run-time.
 
-%% \paragraph{Message Passing.}
-%% Nodes in distributed systems interact by passing messages over communication links.
-%% Message passing can either be synchronous or asynchronous.
-%% In synchronous message passing, the events of message production and delivery are the same, i.e., atomic.
-%% In asynchronous message passing, the events of message production and delivery are logically distinct events subject to being interposed by other events.
-%% Synchronous message passing is most often used between modules on the same node while asynchronous message passing is most often used between nodes.
-%% Both synchronous and asynchronous message passing are required.
-%% Synchronous message passing is necessary for building composable systems because of it association with atomicity.
-%% Asynchronous message passing is necessary for building real distributed systems as this matches the semantics of the networks found in practice.
+\paragraph{Binding.}
+A \emph{binding} is an association between an output handler and an input handler.
+An output handler can be \emph{bound} to zero or more input handlers subject to type compatibility.
+For example, an input handler consuming an integer value can only be bound to an output handler that produces an integer value.
+An input handler can be bound at most once.\footnote{This restriction facilitates composition.}
+Output handlers and input handlers can either be \emph{public}, available for binding by external components, or \emph{private}, only available for binding by their associated component.
+Internal handlers cannot be bound and are always private.
+%Events can be bound and unbound at run-time.
+The component requiring that an output handler be bound to an input handler is said to be the \emph{owner} of the binding.
+A binding that involves a private handler can only be owned by the component associated with the private handler.
 
-\paragraph{Asynchronous Atomic Events.}
-Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
-This implies that the model contains a general treatment of events, i.e., modules can both produce and consume events.
-Modules, therefore, are reactive and structured as a set of atomic event produces and consumers.
-The guarantee of atomicity obviates the need for concurrency control primitives and improves our ability to reason about the behavior of the module.
-Modules communicate using events according to a \emph{configuration} that determines which events are delivered to each module.
+\paragraph{Execution.}
+Execution proceeds by non-deterministically selecting and atomically executing output handlers and internal handlers.
+When an output handler is executed, both the output handler and all bound input handlers are executed in one atomic step receiving the value produced by the output if applicable.\footnote{This restriction facilitates composition.}
+Multiple handlers can be executed concurrently if the sets of components implied by each handler are disjoint.
+%% All components have outputs for creating, binding, unbinding, and destroying and inputs for receiving results that are permanently bound to ``the system.''
+%% System calls follow a request-reply system where a component sends a request for a create, bind, unbind, or destroy, the system receives the request, the system processes the request, the system sends a reply, and the component receives the reply.
 
-%% Communicating with external resources such as the operating system or physical environment is foundational to writing useful programs.
-%% Consider the history of I/O in the C run-time library.
-%% The synchronous function call semantics of the C programming language made it possible to write system calls, e.g., read and write, that transfer control to the operating system which then completes the operation and return control to the process after the system call.
-%% This is called \emph{blocking I/O} because the process blocks while it waits for the operating system to complete the I/O.
-%% Blocking I/O has the severe restriction that only one I/O source can be used at a time.
-%% To remedy this, system calls such as select and poll were introduced that allow a process to monitor I/O sources for events and then respond to these events.
-%% Asynchronous I/O was a further advancement that allows for the system (instead of the process) to dispatch a function upon the completion of I/O.
-%% Compare this to UNIX signals that are neither general (fixed-number with no associated data) nor principled.
-%% Interrupts are difficult to use correctly.
+\paragraph{Composition.}
+Binding, atomic execution, and the absence of shared state allows components to be composed.
+Components are composed by (1) concatenating their state variables, (2) renaming event handlers so every handler's name is unique, and (3) folding input handlers into the output handlers to which they are bound.
+Composition allows the programmer to reason about a constellation of components as a single component.
+Often, the properties of the resulting component can be reasoned about using the properties of the constituent components.
+Components can be composed at compile time using the process outlined above, but it is often better to preserve component boundaries to take advantage of concurrent execution.
 
-\paragraph{Dynamics.}
-The set of nodes and communication links is dynamic in real distributed systems.
-Similarly, the set of modules and their configuration in real devices can change as the device may be re-purposed by users over time.
-In general, our system model must support a dynamic number of modules and, consequently, dynamic configuration of those modules.
+\paragraph{Parameters.}
+Event handlers are identified by their name and a unique \emph{parameter} that is set by binding.
+Event handlers that do not rely on a parameter, i.e., their parameter is always a null parameter, are said to be \emph{unparameterized} while event handlers that do rely on a parameter are said to be \emph{parameterized}.
+Parameters facilitate the definition of components with a dynamic number of event handlers.
+Since input handlers can only be bound to one output handler, fan-in can be accomplished by choosing a different parameter for each binding to an input handler.
+More generally, session semantics can be realized by associating the same parameter with all bindings involving the same component, i.e., each session corresponds to one parameter.
 
-\paragraph{Reflection.}
-The ability to dynamically configure a module implies that a module must be able to know its configuration status to preserve certain properties.
-For example, consider a module for a reliable FIFO queue.
-Since the queue can be dynamically configured, the queue must be able to determine if the event produced by removing and returning an item from the queue will be delivered.
-Otherwise, the queue will lose data items and therefore not be reliable.
+\paragraph{Event handler types.}
+The combination of input, output, and internal handlers, unvalued and valued, unparameterized and parameterized results in 10 possible event handler types:
+unvalued unparameterized output (uv-up output),
+unvalued parameterized output (uv-p output),
+valued unparameterized output (v-up output),
+valued parameterized output (v-p output),
+unvalued unparameterized input (uv-up input),
+unvalued parameterized input (uv-p input),
+valued unparameterized input (v-up input),
+valued parameterized input (v-p input),
+unparameterized internal (up internal), and
+parameterized internal (p internal).
 
-\subsection{I/O Automata}
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{system_model}
+\caption{Example remote sensor node expressed as a constellation of components, event handlers, and bindings.
+  Components are depicted as solid rectangles, e.g., Remote Sensor.
+  Event handlers are depicted as dotted rectangles, e.g., \emph{trigger!}.
+  Output handlers are suffixed with \emph{!}, e.g., \emph{trigger!}.
+  Input handlers are suffixed with \emph{?}, e.g., \emph{Trigger?}.
+  Internal handlers have no suffix, e.g., \emph{compute}.
+  Public handlers start with an uppercase letter, e.g., \emph{Trigger?}.
+  Private events start with a lowercase letter, e.g., \emph{trigger!}.
+  Bindings are indicate by dotted lines with an arrow pointing to the input.}
+\label{sys_model}
+\end{figure}
 
-The Input/Output (I/O) automata formal model developed by Lynch~\cite{distributed_algorithms} meets our requirements for local state and atomic asynchronous events.
+\paragraph{Example.}
+Figure~\ref{sys_model} depicts a remote sensor node expressed as a constellation of components, event handlers, and bindings.
+The Remote Sensor component consists of a Control component and three Channel components representing three network connections.
+The Control component consists of a Sensor component, a Filter component, and a Statistics component.
+The Control component starts the sampling process with the private uv-up output $Control.trigger!$.
+The public uv-up input $Sensor.Trigger?$ is executed atomically with $Control.trigger!$ due to the binding $Control.trigger! \to (Sensor.Trigger?)$.
+When the sampling process is complete, the public v-up output $Sensor.Sample!(f)$ distributes the sample $f$ to the Filter and Statistics components via the $Sensor.Sample!(f) \to (Filter.In?(t), Statistics.In?(u))$ binding.
+Note that $f$, $t$, and $u$ must be of the same type.
+The Filter component passes the filtered sample to the Control component via the $Filter.Out!(f) \to Control.sample?(f)$ binding which is owned by the Control component since $Control.sample?(f)$ is private.
+The Statistics component calculates the statistics incrementally using the up internal $Statistics.compute$.
+The statistics can be polled by the Control component using the $Control.request! \to (Statistics.Request?)$ and $Statistics.Response!(s) \to (Control.response?(s))$ bindings.
+The Control component contains a public vp-output $Control.Send![](m)$ and a public vp-input $Control.Recv?[](m)$ for sending and receiving network messages using Channel components.
+The parameters, i.e., 0, 1, and 2, are used to route messages to the appropriate network using the session idea outlined previously.
+Opportunities for concurrent execution exist for the remote sensor depicted in figure~\ref{sys_model}.
+The set of components implied by $Control.trigger!$ is $(Control, Sensor)$ while the set of components implied by $Statistics.compute$ is $(Statistics)$.
+Since the sets are disjoint, the two actions can be executed concurrently.
+
+\subsection{Formalization as I/O Automata}
+
+The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
 An I/O automaton consists of state variables and atomic actions that manipulate the state.
@@
 Internal actions just manipulate the state of the automaton.
-Output and internal actions are \emph{locally controlled actions}~\cite{distributed_algorithms} consisting of a precondition and an effect.
-Input and output actions are \emph{external actions}~\cite{distributed_algorithms} because they allow an automaton to communicate with other automata.
+Output and internal actions are \emph{locally controlled actions}~\cite{lynch1996distributed} are often presented as precondition and an effect.
+The precondition is a predicate over the state variables that enables the effect when selected.
+The effect is a function that computes the next state of the automaton from the current state.
+Input and output actions are \emph{external actions}~\cite{lynch1996distributed} because they allow an automaton to communicate with other automata.
 Automata may \emph{composed} by concatenating the state variables of the constituent automata and incorporating the effects of input actions into output actions with the same name.
-A useful operation in composite automata is \emph{hiding} which converts an output action to an internal action.
+A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.
 Composing automata results in a static configuration, i.e., the set of automata and their interactions are fixed.
@@
 
-\subsection{Extending I/O Automata for Dynamics}
-
-A standard technique for modeling dynamic modules in I/O automata is to assume they exist but are inactive.
-Dynamic automata can then be activated and deactivated using inputs.
-This approach relies on the static configuration produced by composition and is therefore not suitable for an implementation.
-
-We extend I/O automata for dynamic configuration by introducing a \emph{system automaton} that records the configuration of all automata in the system.
-We differentiate between automaton types and automaton instances.
-Every automaton instances has a unique \emph{automaton identifier (aid)}.
-The state of the system automaton consists of the set of aids $A$ indicating what automata exist and a set of binding records $B$ indicating the relationship between output and input actions.
-A binding record $(out, output, in, input)$ is tuple consisting of the aid of the output automaton $out$, the name of the output action $output$, the aid of the input automaton $in$, and the name of the input action $input$.
-
-The configuration contained in set of binding records must be a valid static configuration to ensure that the behavior of the complete system remains analyzable under the I/O automata model.
-First, the output automaton and input automaton must exist $\forall (out, output, in, input) \in B: out \in A \land in \in A$.
-% TODO:  Binding must not exist.
-Second, an input can only be bound to one output.
-Thus, a given $in$ and $input$ combination can appear at most once in $B$.
-Third, an output cannot be bound to an input in the same automaton.
-Thus, $out \neq in$ for every $(out, output, in, input)$ in $B$.
-Fourth, an output cannot be bound to two inputs in the same automaton.
-Thus, $in_1 \neq in_2$ for all $(out, output, in_1, input_1), (out, output, in_2, input_2)$ in $B$.
-
-The system automaton contains input actions for creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton and output actions that indicate the result of a create, bind, unbind, or destroy.
-All automata are composed with (not bound to) the system automaton and contain outputs for creating, binding, unbinding, and destroying and inputs for receiving results.
-Systems calls (create, bind, unbind, destroy) resemble a request-response protocol where an automaton sends a request, the system automaton receives the request, the system automaton then sends the result, and the automaton receives the result.
-This has the important consequence that \emph{configuration is asynchronous}.
-Destroying an automaton has the effect of removing all of its bindings.
-An automaton that destroys itself can therefore cause the system to deliver of unbound and destroyed events without a corresponding unbind or destroy.
-
-\subsection{Extending I/O Automata for Reflection}
-
-Automata are endowed with additional inputs driven by system outputs indicating when an action has been bound and unbound.
-Automata might use the bound input to enable processing, i.e., someone is listening so I'll start producing data.
-The unbound input is useful for cleaning data structures and inhibiting further processing, i.e., no one is listening so I'll stop producing data.
-The bound and unbound inputs are useful but they are also asynchronous, thus, an automaton cannot trust the status reported by the last bound or unbound for use in, for example, a precondition.
-
-The asynchronous nature of bound and unbound requires a primitive that can examine the contents of $B$ instantaneously.
-Thus, we make one exception to the rule of local state that says automata can examine the contents of $B$ to determine their current configuration.
-This is permissible so long as concurrent implementations provide multiple-reader/single-writer semantics to $B$.
-
-\subsection{Equivalence}
-
-Extending the I/O automata model for dynamics and reflection is useless unless we can relate the extensions back to the original model or establish equivalence.
-To do this, we will take advantage of the fair traces (?) property of I/O automata~\cite{distributed_algorithms}.
-%% We accept an execution trace in the dynamic setting if it is a fair trace of the static model.
-
-%% Modeling a system with I/O automata implies a fixed set of automata given by $A_0$ and set of binding records $B_0$.
-
-Start with a trace of dynamic $T_d$.
-Remove system actions from $T_d$ so $T_d'$.
-We accept $T_d'$ if it is a fair trace of the static model.
-
-Let $A_0$ be the set of automata and $B_0$ be the set of binding records in the static model.
-If you want to prove correspondence, you must prove that eventually $A = A_0$ and $B = B_0$.
-We also need to show that no output is executed before it is fully bound.
-
-%% supports the actions of creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton.
-%% Upon creation, All automata are bound
-%% The four operations necessary for dynamic configuration are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
-
-%% \begin{outline}
-%% \item State
-%%   \begin{outline}
-%%   \item There is no shared state in a 
-%%   \item Local state only
-%%   \item Shared state
-%%     \begin{outline}
-%%       \item Impossible in distributed systems
-%%       \item Dangerous in local systems
-%%     \end{outline}
-%%   \end{outline}
-%% \item Communication
-%%   \begin{outline}
-%%   \item Atomic asynchronous message passing
-%%   \item Network sets size of atom (UDP)
-%%   \item Can build reliable streams (TCP)
-%%   \item Local equivalent is passing a value
-%%   \item Model should lend itself to writing protocols
-%%   \end{outline}
-%% \item Asynchrony
-%%   \begin{outline}
-%%     \item Model must have natural support for asynchrony, i.e., event-based
-%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
-%%   \end{outline}
-%% \item Concurrency
-%%   \begin{outline}
-%%     \item Reason about systems using non-deterministic interleaving of atomic actions
-%%     \item Model should admit implementations that execute concurrently
-%%   \end{outline}
-%% \item Dynamics
-%%   \begin{outline}
-%%     \item Configuration - Edges in graph of communicating components can change at run-time.
-%%       \begin{outline}
-%%       \item Already required in distributed settings
-%%       \item Not addressed in formal models
-%%       \end{outline}
-%%     \item Extension - Nodes in graph of communicating components can change at run-time.
-%%   \end{outline}
-%%   \item Reflection
-%% \end{outline}
-
-%% I/O Automata
-%% \begin{itemize}
-%%   \item Compare with UNITY
-%%   \item Compare with esterel
-%%   \item Compare with pi calculus
-%%   \item Compare with Ptolemy
-%% \end{itemize}
+The proposed system has a straightforward mapping to the I/O automata model: components are automata and event handlers are actions.

### 2011-08-19 papers/paper1/system_model.tex
Second revision of Section 2.

@@
-\section{Component Model and Formalization\label{system_model}}
+\section{I/O Automata Component Model\label{system_model}}
 
-\subsection{Component Model}
+\subsection{I/O Automata as Components}
 
-Modularity, well-defined interfaces, and composition are properties of reusable software.
-Modularity implies that a unit of software can be deployed independently of other units.
-Well-defined interfaces and interactions allow modules to expose their functionality in a regular way that facilitates manual and automated reasoning about interactions with the module.
-Composition allows a group of interacting modules to be repackaged and understood as a single module.
+Modularity, well-defined interfaces, and composition are essential properties of reusable software.
 A unit of software that exhibits these properties is called a \emph{component}~\cite{szyperski2002component}.
-This section outlines the component model for our event-based system.
+Modularity implies that a component can be deployed independently of other components.
+Well-defined interfaces and interactions allow components to expose their functionality in a regular way that facilitates reasoning about different interactions with the component.
+Composition allows a group of interacting components to be understood as a cohesive unit.
 
-\paragraph{Components.}
-A \emph{component} in our model consists of state variables and a set of event handlers.
-Each event handler is only allowed to modify the state of the component with which it is associated.
-The absence of shared state ensures that each component can be deployed independently.
-An event handler can either be an \emph{output handler} that produces a signal or value, an \emph{input handler} that consumes a signal or value, or an \emph{internal handler} that neither produces nor consumes a signal or value.
-Event handlers that produce or consume signals are said to be \emph{unvalued} while event handler that produce or consume values are said to be \emph{valued}.
-% Components can be created and destroyed at run-time.
+The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
+An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
+An action is only permitted to manipulate the state variables of the automaton to which it belongs.
+An action is either an internal action, an output action, or an input action.
+An internal action just changes the state of the automaton.
+An output action changes the state of the automaton and produces a signal or value that is consumed by zero or more input actions.
+An input action changes the state of the automaton when it receives a signal or value produced by an output action.
+Output actions and input actions that produce or consume signals are said to be \emph{unvalued} while those that produce or consume values are said to be \emph{valued}.
+The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.
 
-\paragraph{Binding.}
-A \emph{binding} is an association between an output handler and an input handler.
-An output handler can be \emph{bound} to zero or more input handlers subject to type compatibility.
-For example, an input handler consuming an integer value can only be bound to an output handler that produces an integer value.
-An input handler can be bound at most once.\footnote{This restriction facilitates composition.}
-Output handlers and input handlers can either be \emph{public}, available for binding by external components, or \emph{private}, only available for binding by their associated component.
-Internal handlers cannot be bound and are always private.
-%Events can be bound and unbound at run-time.
-The component requiring that an output handler be bound to an input handler is said to be the \emph{owner} of the binding.
-A binding that involves a private handler can only be owned by the component associated with the private handler.
+The set of output and internal actions constitutes the \emph{local signature}~\cite{lynch1996distributed} of the automaton.
+Often, local actions are presented as a precondition and an effect.
+The precondition is a predicate over the state variables that enables the effect when selected.
+The effect is a function that computes the next state of the automaton from the current state.
+Input actions presented in this style consist solely of an effect.
 
-\paragraph{Execution.}
-Execution proceeds by non-deterministically selecting and atomically executing output handlers and internal handlers.
-When an output handler is executed, both the output handler and all bound input handlers are executed in one atomic step receiving the value produced by the output if applicable.\footnote{This restriction facilitates composition.}
-Multiple handlers can be executed concurrently if the sets of components implied by each handler are disjoint.
-%% All components have outputs for creating, binding, unbinding, and destroying and inputs for receiving results that are permanently bound to ``the system.''
-%% System calls follow a request-reply system where a component sends a request for a create, bind, unbind, or destroy, the system receives the request, the system processes the request, the system sends a reply, and the component receives the reply.
+The set of output and input actions constitutes the \emph{external signature}~\cite{lynch1996distributed} of the automaton.
+An automaton communicates with other automata through the actions in its external signature.
+I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effect of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
+Composing automata that have identically named output actions is ambiguous and not allowed.
+The result of composition is a new automaton thus the principles for reasoning about the behavior of a composed system are the same as the principles for reasoning about a single automaton.
+Often, the properties of the resulting aggregate automaton can be reasoned about using the properties of the constituent automata.
+%% A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.
 
-\paragraph{Composition.}
-Binding, atomic execution, and the absence of shared state allows components to be composed.
-Components are composed by (1) concatenating their state variables, (2) renaming event handlers so every handler's name is unique, and (3) folding input handlers into the output handlers to which they are bound.
-Composition allows the programmer to reason about a constellation of components as a single component.
-Often, the properties of the resulting component can be reasoned about using the properties of the constituent components.
-Components can be composed at compile time using the process outlined above, but it is often better to preserve component boundaries to take advantage of concurrent execution.
+Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
+To be general, the scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not executed) infinitely often.
+Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
+Executing a composed output action consists of atomically evaluating the precondition and simultaneously applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
+The atomic relationship between composed output actions and input actions means that multiple automata may atomically change state when executing an output action.
+The guarantee of atomicity allows a developer to relate the state of one automaton to another resulting in well-defined interactions that facilitate composition.
+
+I/O automata are components and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
+Each action is only allowed to modify the state of the automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
+The absence of shared state thus ensures that each automaton can be deployed independently.
+The external signature of an I/O automaton constitutes its interface and the result of interacting with an I/O automaton is well-defined since every action is atomic.
+Composition is guaranteed by the model itself.
+
+\subsection{Practical Considerations}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{system_model}
+\caption{Example event detection system.
+  Automata are depicted as solid rectangles, e.g., Processor.
+  Named automata are labeled by a box in their upper left corner, e.g., C.
+  Actions are depicted as dotted rectangles, e.g., \emph{detect!}.
+  Output actions are suffixed with \emph{!}, e.g., \emph{detect!}.
+  Input actions are suffixed with \emph{?}, e.g., \emph{detected?}.
+  Internal actions have no suffix, e.g., \emph{process1}.
+  The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.
+  Bindings are indicate by dotted lines with an arrow pointing to the input.}
+\label{sys_model}
+\end{figure}
+
+As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
+However, it lacks certain features needed to develop real components.
+This subsection introduces these features and lays the foundation for the dynamics discusses in section~\ref{dynamics}.
+
+\paragraph{Recursive Encapsulation.}
+Component frameworks that support recursive encapsulation allow components to contain other components.
+Recursive encapsulation encodes the idea of ``divide-and-conquer'' and is at the heart of both bottom-up and top-down design strategies.
+The contained component is the \emph{child} while the containing component is the \emph{parent}.
+The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
+Distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a component to contain multiple child of the same type.
+
+For I/O automata, recursive encapsulation means that an I/O automaton can contain and use other I/O automata.
+Recursive encapsulation implies that a parent automaton is the composition of itself and its children.
+The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively to it and its children.
+The sample event detector system of figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
+The Processor automaton is the root and C and D are its children.
+
+\paragraph{Explicit Composition.}
+Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
+Using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
+To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
+Such an association is called a \emph{binding} and is a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.
+
+Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
+The automaton the prescribes a bindings is said to \emph{own} that binding.
+The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
+A map for renaming inputs is generated by replacing action names with their fully-qualified name in all bindings by tracing from the root.
+For the example in figure~\ref{sys_model}, this map is $\{ (out!(s), C.in?(s)), (C.out!(t), D.in?(t)), (D.detect!, detected?) \}$.
+Composition as defined by the I/O automata model can be applied by using the map to rename all input actions to their corresponding output action.
+
+\paragraph{Binding rules.}
+To be compliant with the named-based composition of the I/O automata model, the set of bindings must adhere to certain rules.
+First, the output action and input action of a binding must agree on the type being produced as consumed.
+For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
+The system in figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value, i.e., a signal.
+Second, an input action can be bound to at most one output action.
+Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
+A binding map where all bindings that mention then same input action are equivalent satisfies this rule, however, we strengthen this rule to say that an input action can appear at most once in the global binding map.
+Consequently, every binding has exactly one owner.
+Third, an output action cannot be bound to more than one input action in the same automaton.
+This was rejected in the I/O automata model since the state of the automaton containing the input actions is not well-defined after the output is executed since the input actions affects will be applied in some non-deterministic order.
+Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
+The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B :: \pi (i) \neq \pi (j) \rangle \rangle$.
+Fourth, an output action cannot be bound to an input action in the same automaton.
+The I/O automata model prevents this requiring that all actions belonging to the same automaton have a unique name.
+Allowing it results in undefined behavior since the order of the output and input effects is undefined.
+While we usually think about the output happening before the input because a value must be generated, the I/O automata model admits an interpretation where the value is generated first and the effects are applied non-deterministically.
+To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
+
+\paragraph{Concurrent execution.}
+Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
+Since the state of each automaton is independent, opportunities exist for the concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
+Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
+For an internal action, this set consists of the automaton containing the internal action.
+For an output action, this set consists of the automaton containing the output action and the automata contain the inputs actions with which the output action is bound, i.e., composed.
+Two actions can be executed concurrently if their respective sets of implied automata and therefore state variables are disjoint.
+The implied automata sets for each local action in figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect!
+\to \{D, root\}$.
+Thus, $process1$ and $out!(s)$ can be executed concurrently while $process1$ and $detect!$ can not.
+
+These observations create some interesting opportunities for designing and analyzing concurrent software.
+Migrating intense computation to internal actions increases the level of parallelism in a system, i.e., the set of implied automata is small.
+This is equivalent to the known maxim that critical sections should be kept as small as possible.
+High degrees of fan-out, an output action being bound to many input actions, creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
+However, the input effects can be applied in parallel since the state of each automaton is independent.
+High degrees of fan-in, an automaton has many bound input actions, also creates a bottle-neck by increasing the probability that two set of implied actions will not be disjoint.
+An high-performance scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
+A slow performed constellation might be statically composed to take advantage of in-lining.
+A slow performing automaton might be factored into a number of child automata to increase parallelism.
 
 \paragraph{Parameters.}
-Event handlers are identified by their name and a unique \emph{parameter} that is set by binding.
-Event handlers that do not rely on a parameter, i.e., their parameter is always a null parameter, are said to be \emph{unparameterized} while event handlers that do rely on a parameter are said to be \emph{parameterized}.
-Parameters facilitate the definition of components with a dynamic number of event handlers.
-Since input handlers can only be bound to one output handler, fan-in can be accomplished by choosing a different parameter for each binding to an input handler.
-More generally, session semantics can be realized by associating the same parameter with all bindings involving the same component, i.e., each session corresponds to one parameter.
+Actions in the I/O automata model can defined using parameters.
+Parameters are useful for situations requiring fan-in or a session.
+For example, consider an automaton with an input actions $in?(s)$ that is to be bound to $N$ different output actions.
+The solution is to parameterize the input action with a parameter that indicates the associated output actions, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
+\emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
+Parameters, for actions that require them, are specified in the binding and serve to identify the action.
 
-\paragraph{Event handler types.}
-The combination of input, output, and internal handlers, unvalued and valued, unparameterized and parameterized results in 10 possible event handler types:
+\paragraph{Action types.}
+The combination of input, output, and internal actions, unvalued and valued, unparameterized and parameterized results in 10 possible action types:
 unvalued unparameterized output (uv-up output),
@@
 \center
-\includegraphics[width=\textwidth]{system_model}
-\caption{Example remote sensor node expressed as a constellation of components, event handlers, and bindings.
-  Components are depicted as solid rectangles, e.g., Remote Sensor.
-  Event handlers are depicted as dotted rectangles, e.g., \emph{trigger!}.
-  Output handlers are suffixed with \emph{!}, e.g., \emph{trigger!}.
-  Input handlers are suffixed with \emph{?}, e.g., \emph{Trigger?}.
-  Internal handlers have no suffix, e.g., \emph{compute}.
-  Public handlers start with an uppercase letter, e.g., \emph{Trigger?}.
-  Private events start with a lowercase letter, e.g., \emph{trigger!}.
-  Bindings are indicate by dotted lines with an arrow pointing to the input.}
-\label{sys_model}
+\includegraphics[width=\textwidth]{example1}
+\caption{Example remote sensor system.
+  Automata are depicted as solid rectangles, e.g., Remote Sensor.
+  Named automata are labeled by a box in their upper left corner, e.g., Z.
+  Actions are depicted as dotted rectangles, e.g., \emph{trigger!}.
+  Output actions are suffixed with \emph{!}, e.g., \emph{trigger!}.
+  Input actions are suffixed with \emph{?}, e.g., \emph{trigger?}.
+  Internal actions have no suffix, e.g., \emph{compute}.
+  The type associated with a valued action is in parentheses after the name, e.g., \emph{in?(s)}.
+  Parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
+  Bindings are indicate by dotted lines with an arrow pointing to the input.
+}
+\label{example1}
 \end{figure}
@@
 \paragraph{Example.}
-Figure~\ref{sys_model} depicts a remote sensor node expressed as a constellation of components, event handlers, and bindings.
-The Remote Sensor component consists of a Control component and three Channel components representing three network connections.
-The Control component consists of a Sensor component, a Filter component, and a Statistics component.
-The Control component starts the sampling process with the private uv-up output $Control.trigger!$.
-The public uv-up input $Sensor.Trigger?$ is executed atomically with $Control.trigger!$ due to the binding $Control.trigger! \to (Sensor.Trigger?)$.
-When the sampling process is complete, the public v-up output $Sensor.Sample!(f)$ distributes the sample $f$ to the Filter and Statistics components via the $Sensor.Sample!(f) \to (Filter.In?(t), Statistics.In?(u))$ binding.
-Note that $f$, $t$, and $u$ must be of the same type.
-The Filter component passes the filtered sample to the Control component via the $Filter.Out!(f) \to Control.sample?(f)$ binding which is owned by the Control component since $Control.sample?(f)$ is private.
-The Statistics component calculates the statistics incrementally using the up internal $Statistics.compute$.
-The statistics can be polled by the Control component using the $Control.request! \to (Statistics.Request?)$ and $Statistics.Response!(s) \to (Control.response?(s))$ bindings.
-The Control component contains a public vp-output $Control.Send![](m)$ and a public vp-input $Control.Recv?[](m)$ for sending and receiving network messages using Channel components.
-The parameters, i.e., 0, 1, and 2, are used to route messages to the appropriate network using the session idea outlined previously.
-Opportunities for concurrent execution exist for the remote sensor depicted in figure~\ref{sys_model}.
-The set of components implied by $Control.trigger!$ is $(Control, Sensor)$ while the set of components implied by $Statistics.compute$ is $(Statistics)$.
+Figure~\ref{example1} depicts a remote sensor systems as a composition of automata.
+The Remote Sensor automaton consists of the Control automaton $Z$ and three Channel automata $A$, $B$, and $C$ representing three network connections.
+The Control automaton $Z$ consists of the Sensor automaton $S$, the Filter automaton $F$, and the Statistics automaton $T$.
+The Control automaton $C$ starts the sampling process with the uv-up output $C.trigger!$.
+The uv-up input $S.trigger?$ is executed atomically with $C.trigger!$ due to the binding $(C.trigger!, S.trigger?)$.
+When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter $F$ and Statistics component $T$ via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
+Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
+The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $F.out!(t) \to C.sample?(t)$ binding.
+The Statistics automaton $T$ calculates the statistics incrementally using the up internal $T.compute$.
+The statistics can be polled by the Control automaton $Z$ using the bindings $(Z.request!, T.request?)$ and $(T.response!(u), Z.response?(u))$.
+The Control automaton $Z$ contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata $A$, $B$, and $C$.
+The parameters 0, 1, and 2, are used to route messages to Channel automata $A$, $B$, and $C$ respectively using the session idea mentioned previously.
+Opportunities for concurrent execution exist for the remote sensor depicted in figure~\ref{example1}.
+The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
 Since the sets are disjoint, the two actions can be executed concurrently.
-
-\subsection{Formalization as I/O Automata}
-
-The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
-An I/O automaton consists of state variables and atomic actions that manipulate the state.
-The three action types are inputs, outputs, and internal actions.
-Input actions receive a signal or a value and output actions produce a signal or a value.
-Internal actions just manipulate the state of the automaton.
-Output and internal actions are \emph{locally controlled actions}~\cite{lynch1996distributed} are often presented as precondition and an effect.
-The precondition is a predicate over the state variables that enables the effect when selected.
-The effect is a function that computes the next state of the automaton from the current state.
-Input and output actions are \emph{external actions}~\cite{lynch1996distributed} because they allow an automaton to communicate with other automata.
-Automata may \emph{composed} by concatenating the state variables of the constituent automata and incorporating the effects of input actions into output actions with the same name.
-A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.
-Composing automata results in a static configuration, i.e., the set of automata and their interactions are fixed.
-Execution in the I/O automata model consists of repeatedly selecting a local action and then applying the effect if the precondition is true.
-The scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not executed) infinitely often.
-
-The proposed system has a straightforward mapping to the I/O automata model: components are automata and event handlers are actions.

### 2011-08-19 tutorial/count_to_ten_automaton.cpp

@@ -12,18 +12,11 @@ private:
 
 public:
   count_to_ten_automaton () :
-    m_count (1)
-  {
-    schedule ();
+    m_count (1) {
+    increment_schedule ();
   }
   
 private:
-  void schedule () const {
-    if (increment_precondition ()) {
-      ioa::schedule (&count_to_ten_automaton::increment);
-    }
-  }
-
   bool increment_precondition () const {
     return m_count <= 10;
   }
@@ -34,7 +27,9 @@ private:
   }
 
   void increment_schedule () const {
-    schedule ();
+    if (increment_precondition ()) {
+      ioa::schedule (&count_to_ten_automaton::increment);
+    }
   }
 
   UP_INTERNAL (count_to_ten_automaton, increment);
@@ -42,8 +37,7 @@ private:
 
 int main () {
   ioa::global_fifo_scheduler sched;
-  std::auto_ptr<ioa::typed_generator_interface<count_to_ten_automaton> > x = ioa::make_generator<count_to_ten_automaton> ();
-  ioa::run (sched, x);
+  ioa::run (sched, ioa::make_generator<count_to_ten_automaton> ());
   return 0; 
 }
 

### 2011-08-22 papers/paper1/system_model.tex
Edits to Section 2.

@@
 An action is either an internal action, an output action, or an input action.
-An internal action just changes the state of the automaton.
+An internal action changes the state of the automaton.
 An output action changes the state of the automaton and produces a signal or value that is consumed by zero or more input actions.
 An input action changes the state of the automaton when it receives a signal or value produced by an output action.
-Output actions and input actions that produce or consume signals are said to be \emph{unvalued} while those that produce or consume values are said to be \emph{valued}.
+\emph{Unvalued} output and input actions produce and consume signals respectively.
+\emph{Valued} output and input actions produce and consume values respectively.
 The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.
 
-The set of output and internal actions constitutes the \emph{local signature}~\cite{lynch1996distributed} of the automaton.
+An automaton's output and internal actions constitutes its \emph{local signature}~\cite{lynch1996distributed}.
 Often, local actions are presented as a precondition and an effect.
@@
 
-The set of output and input actions constitutes the \emph{external signature}~\cite{lynch1996distributed} of the automaton.
+An automaton's output and input actions constitutes its \emph{external signature}~\cite{lynch1996distributed}.
 An automaton communicates with other automata through the actions in its external signature.
@@
 However, it lacks certain features needed to develop real components.
-This subsection introduces these features and lays the foundation for the dynamics discusses in section~\ref{dynamics}.
+This subsection introduces these features and lays the foundation for the dynamics discussed in section~\ref{dynamics}.
 
@@
 The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
-Distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a component to contain multiple child of the same type.
+Distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple child of the same type.
 
@@
 Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
-The automaton the prescribes a bindings is said to \emph{own} that binding.
+The automaton that prescribes a bindings is said to \emph{own} that binding.
 The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
@@
 For the example in figure~\ref{sys_model}, this map is $\{ (out!(s), C.in?(s)), (C.out!(t), D.in?(t)), (D.detect!, detected?) \}$.
-Composition as defined by the I/O automata model can be applied by using the map to rename all input actions to their corresponding output action.
+Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
 
@@
 To be compliant with the named-based composition of the I/O automata model, the set of bindings must adhere to certain rules.
-First, the output action and input action of a binding must agree on the type being produced as consumed.
+First, the output action and input action of a binding must agree on the type being produced and consumed.
 For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
@@
 Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
-A binding map where all bindings that mention then same input action are equivalent satisfies this rule, however, we strengthen this rule to say that an input action can appear at most once in the global binding map.
+A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
+For dynamics, however, we strengthen this rule to say that an input action can appear at most once in the global binding map.
 Consequently, every binding has exactly one owner.
 Third, an output action cannot be bound to more than one input action in the same automaton.
-This was rejected in the I/O automata model since the state of the automaton containing the input actions is not well-defined after the output is executed since the input actions affects will be applied in some non-deterministic order.
+This was rejected in the I/O automata model since the state of the automaton containing the input actions is not well-defined after the output is executed since the input action affects will be applied in undefined order.
 Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
-The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B :: \pi (i) \neq \pi (j) \rangle \rangle$.
+The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
 Fourth, an output action cannot be bound to an input action in the same automaton.
-The I/O automata model prevents this requiring that all actions belonging to the same automaton have a unique name.
-Allowing it results in undefined behavior since the order of the output and input effects is undefined.
-While we usually think about the output happening before the input because a value must be generated, the I/O automata model admits an interpretation where the value is generated first and the effects are applied non-deterministically.
+The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
+Allowing such bindings result in undefined behavior since the order of the output and input effects is undefined.
+To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
+However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
 To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
@@
 For an internal action, this set consists of the automaton containing the internal action.
-For an output action, this set consists of the automaton containing the output action and the automata contain the inputs actions with which the output action is bound, i.e., composed.
+For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound, i.e., composed.
 Two actions can be executed concurrently if their respective sets of implied automata and therefore state variables are disjoint.
@@
 Migrating intense computation to internal actions increases the level of parallelism in a system, i.e., the set of implied automata is small.
-This is equivalent to the known maxim that critical sections should be kept as small as possible.
 High degrees of fan-out, an output action being bound to many input actions, creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
 However, the input effects can be applied in parallel since the state of each automaton is independent.
-High degrees of fan-in, an automaton has many bound input actions, also creates a bottle-neck by increasing the probability that two set of implied actions will not be disjoint.
+High degrees of fan-in, an automaton has many bound input actions, also creates a bottle-neck by increasing the probability that two sets of implied automata will not be disjoint.
 An high-performance scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
-A slow performed constellation might be statically composed to take advantage of in-lining.
-A slow performing automaton might be factored into a number of child automata to increase parallelism.
+A slow composition of automata might be statically composed to take advantage of in-lining.
+A slow automaton might be factored into a number of child automata to increase parallelism.
 
@@
 The Control automaton $Z$ consists of the Sensor automaton $S$, the Filter automaton $F$, and the Statistics automaton $T$.
-The Control automaton $C$ starts the sampling process with the uv-up output $C.trigger!$.
-The uv-up input $S.trigger?$ is executed atomically with $C.trigger!$ due to the binding $(C.trigger!, S.trigger?)$.
+The Control automaton $C$ starts the sampling process with the uv-up output $Z.trigger!$.
+The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
 When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter $F$ and Statistics component $T$ via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
 Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
-The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $F.out!(t) \to C.sample?(t)$ binding.
+The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $F.out!(t) \to Z.sample?(t)$ binding.
 The Statistics automaton $T$ calculates the statistics incrementally using the up internal $T.compute$.

### 2011-08-22 tutorial/count_to_ten_automaton.cpp

@@ -40,4 +40,3 @@ int main () {
   ioa::run (sched, ioa::make_generator<count_to_ten_automaton> ());
   return 0; 
 }
-

### 2011-08-22 tutorial/producer_consumer.cpp

@@ -11,18 +11,11 @@ private:
 
 public:
   producer_automaton () :
-    m_count (1)
-  {
-    schedule ();
+    m_count (1) {
+    produce_schedule ();
   }
   
 private:
-  void schedule () const {
-    if (produce_precondition ()) {
-      ioa::schedule (&producer_automaton::produce);
-    }
-  }
-
   bool produce_precondition () const {
     return m_count <= 10;
   }
@@ -34,7 +27,9 @@ private:
   }
 
   void produce_schedule () const {
-    schedule ();
+    if (produce_precondition ()) {
+      ioa::schedule (&producer_automaton::produce);
+    }
   }
 
 public:
@@ -61,10 +56,12 @@ class producer_consumer_automaton :
 public:
   producer_consumer_automaton () {
     ioa::automaton_manager<producer_automaton>* producer =
-      ioa::make_automaton_manager (this, ioa::make_generator<producer_automaton> ());
+      ioa::make_automaton_manager (this,
+	  ioa::make_generator<producer_automaton> ());
 
     ioa::automaton_manager<consumer_automaton>* consumer =
-      ioa::make_automaton_manager (this, ioa::make_generator<consumer_automaton> ());
+      ioa::make_automaton_manager (this,
+          ioa::make_generator<consumer_automaton> ());
 
     ioa::make_binding_manager (this,
 			       producer, &producer_automaton::produce,

### 2011-08-22 tutorial/two_counters.cpp

@@ -13,16 +13,10 @@ public:
   count_to_ten_automaton () :
     m_count (1)
   {
-    schedule ();
+    increment_schedule ();
   }
   
 private:
-  void schedule () const {
-    if (increment_precondition ()) {
-      ioa::schedule (&count_to_ten_automaton::increment);
-    }
-  }
-
   bool increment_precondition () const {
     return m_count <= 10;
   }
@@ -35,7 +29,9 @@ private:
   }
 
   void increment_schedule () const {
-    schedule ();
+    if (increment_precondition ()) {
+      ioa::schedule (&count_to_ten_automaton::increment);
+    }
   }
 
   UP_INTERNAL (count_to_ten_automaton, increment);
@@ -46,8 +42,10 @@ class two_counter_automaton :
 {
 public:
   two_counter_automaton () {
-    ioa::make_automaton_manager (this, ioa::make_generator<count_to_ten_automaton> ());
-    ioa::make_automaton_manager (this, ioa::make_generator<count_to_ten_automaton> ());
+    ioa::make_automaton_manager (this,
+        ioa::make_generator<count_to_ten_automaton> ());
+    ioa::make_automaton_manager (this,
+	ioa::make_generator<count_to_ten_automaton> ());
   }
 
 };
@@ -57,4 +55,3 @@ int main () {
   ioa::run (sched, ioa::make_generator<two_counter_automaton> ());
   return 0; 
 }
-

### 2011-08-24 papers/paper1/dynamics.tex
Second revision of Section 2.
First draft of section 3.

@@
-\section{Dynamics}
+\section{Dynamics\label{dynamics}}
 
-However, the I/O automata model assumes a static configuration while the proposed system requires the ability to deal with dynamic configurations, i.e., creating, binding, unbinding, and destroying.
+The I/O automata model assumes that systems are composed of a finite or infinitely countable number of automata~\cite{lynch1996distributed}.
+Assuming an infinitely countable number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
+The set of interactions in the I/O automata model is also fixed since automata are statically composed.
+Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
+To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}.
 
-\subsection{I/O Automata and Dynamic Configuration}
+While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system results in over-provisioning or inflexibility.
+To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
+System resources are wasted if the actual number of participants is much smaller than $N$.
+If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
+Consequently, a system built using the component model described in section~\ref{system_model} requires the ability to deal with a dynamic set of automata and interactions.
 
-A standard technique for modeling dynamic modules in I/O automata is to assume they exist but are inactive.
-Dynamic automata can then be activated and deactivated using inputs.
-We will use a similar technique to model dynamic configurations.
+\paragraph{System actions.}
+The four indivisible operations for managing a dynamic set of automata are create, bind, unbind, and destroy.
+Create, bind, unbind, and destroy are collectively called \emph{system actions}.
+The \emph{create} operation allows an automaton to create a child automaton.
+The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
+The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.
 
-We introduce a \emph{system automaton} that records the configuration of all automata in the system.
-We differentiate between automaton types and automaton instances.
-Every automaton instances has a unique \emph{automaton identifier (aid)}.
-The state of the system automaton consists of the set of aids $A$ indicating what automata exist and a set of binding records $B$ indicating the relationship between output and input actions.
-A binding record $(out, output, in, input)$ is tuple consisting of the aid of the output automaton $out$, the name of the output action $output$, the aid of the input automaton $in$, and the name of the input action $input$.
+System actions can fail for a variety of reasons.
+A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
+A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in section~\ref{system_model} would be violated.
+An unbind fails if the binding does not exist.
+Note that the owner of binding is taken into account when determining if the binding exist so only the owner of a binding can dissolve it.
+A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy.
 
-The configuration contained in set of binding records must be a valid static configuration to ensure that the behavior of the complete system remains analyzable under the I/O automata model.
-First, the output automaton and input automaton must exist $\forall (out, output, in, input) \in B: out \in A \land in \in A$.
-% TODO:  Binding must not exist.
-Second, an input can only be bound to one output.
-Thus, a given $in$ and $input$ combination can appear at most once in $B$.
-Third, an output cannot be bound to an input in the same automaton.
-Thus, $out \neq in$ for every $(out, output, in, input)$ in $B$.
-Fourth, an output cannot be bound to two inputs in the same automaton.
-Thus, $in_1 \neq in_2$ for all $(out, output, in_1, input_1), (out, output, in_2, input_2)$ in $B$.
+%% Let $A$ be the set of automata that exist where each element in $A$ is a pair $(p, a)$ where $p$ is the parent of $a$.
+%% Let $B$ be the set of bindings that exist where each element in $B$ is a tuple $(w, o_a, o_n, o_p, i_a, i_n, i_p)$ where $w$ is the automaton that owns the binding, $o_a$ is the output automaton, $o_n$ is the name of the output action, $o_p$ is the parameter of the output action, $i_a$ is the input automaton, $i_n$ is the name of the input action, and $i_p$ is the parameter of the input action.
+%% Define the predicate $exists (a) = \langle \exists p, q : (p, q) \in A :: q = a\rangle$ that is true when automaton $a$ exists, i.e., has a parent\footnote{The root automaton's parent is the special symbol $\perp$.}.
+%% Define the predicate $existsb (a) = \langle \exists a : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: a = w \lor a = o_a \lor a = i_a \rangle$ that is true when automaton $a$ appears somewhere in $B$.
+%% We define the create operation as the Hoare triple $\{exists (p) \land \lnot exists (a)\} \quad create(a)_p \quad \{(p,a) \in A\}$ which says that the parent automaton $p$ can create automaton $a$ if $p$ exists and $a$ does not exist.
+%% The destroy operation can be defined $\{(p,a) \in A\} \quad destroy(a)_p \quad \{(p,a) \notin A \land \lnot existsb(a)\}$ which says that in order to destroy $a$ and remove all bindings associated with $a$, $p$ must be the parent of $a$.
+%% Define the predicate $free (a, n, p) = \langle \forall w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: \lnot (i_a = a \land i_n = n \land i_p = p) \rangle$ that is true when the input action argument is not bound.
+%% Define the predicate $involved (o, n, p, i) = \langle \exists w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: o_a = o \land o_n = n \land o_p = p \land i_a = i \rangle$ that is true when automaton $o$'s output action $(n, p, i)$ is bound to some input action in automaton $i$.
+%% The bind operation can be defined $\{exists (w) \land exists (o_a) \land exists (i_a) \land free (i_a, i_n, i_p) \land \lnot involved(o_a, o_n, o_p, i_a) \land o_a \neq i_a \} \quad bind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\}$ which says that in order for the binding to succeed the owner $w$ exists, automaton $o_a$ exists, automaton $i_a$ exists, the input action $(i_a, i_n, i_p)$ must not be bound, the output action $(o_n, o_p)$ of automaton $o_a$ cannot be bound to an input action in automaton $i_a$, and the automata $o_a$ and $i_a$ cannot be the same.
+%% The precondition for $bind$ is same as the binding rules derived in section~\ref{system_model} augmented with existence tests for the owner, output automaton, and input automaton.
+%% The unbind operation is similar to destroy: $\{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\} \quad unbind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \notin B\}$.
 
-The system automaton contains input actions for creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton and output actions that indicate the result of a create, bind, unbind, or destroy.
-All automata are composed with the system automaton and contain outputs for creating, binding, unbinding, and destroying and inputs for receiving results.
-Systems calls (create, bind, unbind, destroy) resemble a request-response protocol where an automaton sends a request, the system automaton receives the request, the system processes the request, the system automaton then sends the result, and the automaton receives the result.
-This has the important consequence that \emph{configuration is asynchronous}.
-Destroying an automaton has the effect of removing all of its bindings.
-An automaton that destroys itself can therefore cause the system to deliver unbound and destroyed events without an unbind or destroy.
+% I didn't do more complex transactions, i.e., create-and-bind, because its not clear how it fails.
+% Are partial successes allowed?
+% Furthermore, such  transactions can be arbitrarily complex which doesn't bode well for efficiency.
 
-Using the system automaton as a bookkeeping mechanism, we can relate a statically composed system to a dynamically configured one.
-The key idea is to show that actions belong to automaton $a$ are only executed when $a \in A$ and that an output action is only executed when $B$ contains all of bindings associated with the static configuration.
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{system_action}
+\caption{System actions.
+  An automaton contains output actions for creating, binding, unbinding, and destroying.
+  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
+  The system automaton returns a result using the appropriate parameterized output action.
+  The automaton contains input actions to receive the results.}
+\label{system_action}
+\end{figure}
 
-\subsection{Extending I/O Automata for Reflection}
+\paragraph{The system automaton.}
+Let the set of automata $A$ and set of bindings $B$ be managed by the \emph{system automaton} which represents the run-time system.
+Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
+The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
+Figure~\ref{system_action} shows the relationship between an automaton and the system automaton.
 
-Automata are endowed with additional inputs driven by system outputs indicating when an action has been bound and unbound.
-Automata might use the bound input to enable processing, i.e., someone is listening so I'll start producing data.
-The unbound input is useful for cleaning data structures and inhibiting further processing, i.e., no one is listening so I'll stop producing data.
-The bound and unbound inputs are useful but they are also asynchronous, thus, an automaton cannot trust the status reported by the last bound or unbound for use in, for example, a precondition.
+Conceptually, the state variables $A$ and $B$ belong to the system automaton.
+However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
+The scheduler uses $A$ to ensure that only actions belonging to automata exist are selected and uses $B$ to ensure that the appropriate set of inputs actions receive the value produced when an output action is executed.
+Thus, the set of actions implied by any action is the set defined in section~\ref{system_model} \emph{and} and the system automaton.
+However, the opportunities for concurrency explored in section~\ref{system_model} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
 
-The asynchronous nature of bound and unbound requires a primitive that can examine the contents of $B$ instantaneously.
-Thus, we make one exception to the rule of local state that says automata can examine the contents of $B$ to determine their current configuration.
-This is permissible so long as concurrent implementations provide multiple-reader/single-writer semantics to $B$.
+\paragraph{Binding predicates.}
+The static configuration of a system as prescribed by the I/O automata model guarantees that the set of implied automata for an action is always the same.
+This property is necessary and desirable for reasoning about the behavior of the system.
+Admitting dynamics removes this guarantee since the set of implied automata will change according to the various system actions being executed.
+To be specific, the set of implied automata for internal actions does not change but the set of implied automata for output actions can be changed with a successful bind, unbind, or destroy.
 
-\subsection{Equivalence}
+To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
+To do this, we conjunct the precondition of an output action with a predicate over the set of bindings $B$ to ensure that the output action is not executed until its set of implied automata is correct.
+Notice that this only requires reading $B$ and is permitted based on the required properties of the system automaton.
+Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
+A static system can be implemented by adding binding predicates and logic to create the constellation of automata.
 
-Extending the I/O automata model for dynamics and reflection is useless unless we can relate the extensions back to the original model or establish equivalence.
-To do this, we will take advantage of the fair traces (?) property of I/O automata~\cite{distributed_algorithms}.
-%% We accept an execution trace in the dynamic setting if it is a fair trace of the static model.
+%% All components have outputs for creating, binding, unbinding, and destroying and inputs for receiving results that are permanently bound to ``the system.''
+%% System calls follow a request-reply system where a component sends a request for a create, bind, unbind, or destroy, the system receives the request, the system processes the request, the system sends a reply, and the component receives the reply.
 
-%% Modeling a system with I/O automata implies a fixed set of automata given by $A_0$ and set of binding records $B_0$.
+%Events can be bound and unbound at run-time.
+%% The I/O automata model is static, i.e., the set of automata and their interactions are fixed.
+% Components can be created and destroyed at run-time.
 
-Start with a trace of dynamic $T_d$.
-Remove system actions from $T_d$ so $T_d'$.
-We accept $T_d'$ if it is a fair trace of the static model.
+%% \subsection{I/O Automata and Dynamic Configuration}
 
-Let $A_0$ be the set of automata and $B_0$ be the set of binding records in the static model.
-If you want to prove correspondence, you must prove that eventually $A = A_0$ and $B = B_0$.
-We also need to show that no output is executed before it is fully bound.
+%% A standard technique for modeling dynamic modules in I/O automata is to assume they exist but are inactive.
+%% Dynamic automata can then be activated and deactivated using inputs.
+%% We will use a similar technique to model dynamic configurations.
+
+%% We introduce a \emph{system automaton} that records the configuration of all automata in the system.
+%% We differentiate between automaton types and automaton instances.
+%% Every automaton instances has a unique \emph{automaton identifier (aid)}.
+%% The state of the system automaton consists of the set of aids $A$ indicating what automata exist and a set of binding records $B$ indicating the relationship between output and input actions.
+%% A binding record $(out, output, in, input)$ is tuple consisting of the aid of the output automaton $out$, the name of the output action $output$, the aid of the input automaton $in$, and the name of the input action $input$.
+
+%% The configuration contained in set of binding records must be a valid static configuration to ensure that the behavior of the complete system remains analyzable under the I/O automata model.
+%% First, the output automaton and input automaton must exist $\forall (out, output, in, input) \in B: out \in A \land in \in A$.
+%% % TODO:  Binding must not exist.
+%% Second, an input can only be bound to one output.
+%% Thus, a given $in$ and $input$ combination can appear at most once in $B$.
+%% Third, an output cannot be bound to an input in the same automaton.
+%% Thus, $out \neq in$ for every $(out, output, in, input)$ in $B$.
+%% Fourth, an output cannot be bound to two inputs in the same automaton.
+%% Thus, $in_1 \neq in_2$ for all $(out, output, in_1, input_1), (out, output, in_2, input_2)$ in $B$.
+
+%% The system automaton contains input actions for creating an automaton, binding an output to an input, unbinding an output and an input, and destroying an automaton and output actions that indicate the result of a create, bind, unbind, or destroy.
+%% All automata are composed with the system automaton and contain outputs for creating, binding, unbinding, and destroying and inputs for receiving results.
+%% Systems calls (create, bind, unbind, destroy) resemble a request-response protocol where an automaton sends a request, the system automaton receives the request, the system processes the request, the system automaton then sends the result, and the automaton receives the result.
+%% This has the important consequence that \emph{configuration is asynchronous}.
+%% Destroying an automaton has the effect of removing all of its bindings.
+%% An automaton that destroys itself can therefore cause the system to deliver unbound and destroyed events without an unbind or destroy.
+
+%% Using the system automaton as a bookkeeping mechanism, we can relate a statically composed system to a dynamically configured one.
+%% The key idea is to show that actions belong to automaton $a$ are only executed when $a \in A$ and that an output action is only executed when $B$ contains all of bindings associated with the static configuration.
+
+%% \subsection{Extending I/O Automata for Reflection}
+
+%% Automata are endowed with additional inputs driven by system outputs indicating when an action has been bound and unbound.
+%% Automata might use the bound input to enable processing, i.e., someone is listening so I'll start producing data.
+%% The unbound input is useful for cleaning data structures and inhibiting further processing, i.e., no one is listening so I'll stop producing data.
+%% The bound and unbound inputs are useful but they are also asynchronous, thus, an automaton cannot trust the status reported by the last bound or unbound for use in, for example, a precondition.
+
+%% The asynchronous nature of bound and unbound requires a primitive that can examine the contents of $B$ instantaneously.
+%% Thus, we make one exception to the rule of local state that says automata can examine the contents of $B$ to determine their current configuration.
+%% This is permissible so long as concurrent implementations provide multiple-reader/single-writer semantics to $B$.
+
+%% \subsection{Equivalence}
+
+%% Extending the I/O automata model for dynamics and reflection is useless unless we can relate the extensions back to the original model or establish equivalence.
+%% To do this, we will take advantage of the fair traces (?) property of I/O automata~\cite{distributed_algorithms}.
+%% %% We accept an execution trace in the dynamic setting if it is a fair trace of the static model.
+
+%% %% Modeling a system with I/O automata implies a fixed set of automata given by $A_0$ and set of binding records $B_0$.
+
+%% Start with a trace of dynamic $T_d$.
+%% Remove system actions from $T_d$ so $T_d'$.
+%% We accept $T_d'$ if it is a fair trace of the static model.
+
+%% Let $A_0$ be the set of automata and $B_0$ be the set of binding records in the static model.
+%% If you want to prove correspondence, you must prove that eventually $A = A_0$ and $B = B_0$.
+%% We also need to show that no output is executed before it is fully bound.
 

### 2011-08-24 examples/echo_client.cpp
Removed self-destruct.

@@ -72,7 +72,7 @@ private:
       strerror_r (val.err, buf, 256);
       std::cerr << "Couldn't connect: " << buf << std::endl;
 #endif
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
     else {
       m_connection = val.handle;
@@ -124,7 +124,7 @@ private:
       strerror_r (err, buf, 256);
       std::cerr << "Couldn't send: " << buf << std::endl;
 #endif
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
   }
 
@@ -143,13 +143,13 @@ private:
       strerror_r (val.err, buf, 256);
       std::cerr << "Couldn't receive: " << buf << std::endl;
 #endif
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
     else {
       assert (val.buffer.get () != 0);
       std::string s (static_cast<const char*> (val.buffer->data ()), val.buffer->size ());
       std::cout << "Received:\t" << s << std::endl;
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
   }
   

### 2011-08-24 examples/echo_server.cpp

@@ -80,7 +80,7 @@ private:
     }
     else {
       // Done.
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
   }
 
@@ -121,7 +121,7 @@ private:
       strerror_r (val.err, buf, 256);
       std::cerr << "Couldn't accept: " << buf << std::endl;
 #endif
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
   }
 

### 2011-08-24 examples/udp_receiver.cpp

@@ -48,7 +48,7 @@ private:
       strerror_r (v.err, buf, 256);
       std::cerr << "Couldn't receive udp_receiver_automaton: " << buf << std::endl;
 #endif
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
     else {
       assert (v.buffer.get () != 0);

### 2011-08-24 examples/udp_sender.cpp

@@ -78,7 +78,7 @@ private:
       strerror_r (result, buf, 256);
       std::cerr << "Couldn't send udp_sender_automaton: " << buf << std::endl;
 #endif
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
     else {
       m_state = SEND_COMPLETE;

### 2011-08-24 include/ioa/automaton.hpp

@@ -76,8 +76,6 @@ namespace ioa {
     std::set<system_binding_manager_interface*> m_unbind_send;
     std::set<system_binding_manager_interface*> m_unbind_recv;
 
-    bool m_self_destruct;
-
     // Automata can't be copied.
     automaton (const automaton&);
     automaton& operator= (const automaton&);
@@ -89,7 +87,6 @@ namespace ioa {
     void bind (system_binding_manager_interface* helper);
     void unbind (system_binding_manager_interface* helper);
     void destroy (system_automaton_manager_interface* helper);
-    void self_destruct ();
 
   private:
     void schedule () const;
@@ -121,13 +118,6 @@ namespace ioa {
     void sys_destroy_schedule () const { schedule (); }
   public:
     SYSTEM_OUTPUT (automaton, sys_destroy, void*);
-
-  private:
-    bool sys_self_destruct_precondition () const;
-    void* sys_self_destruct_effect ();
-    void sys_self_destruct_schedule () const { schedule (); }
-  public:
-    SYSTEM_OUTPUT (automaton, sys_self_destruct, void*);
 
   public:
     struct created_arg_t {

### 2011-08-24 include/ioa/global_fifo_scheduler.hpp

@@ -33,8 +33,6 @@ namespace ioa {
     
     void schedule (automaton::sys_destroy_type automaton::*ptr);
 
-    void schedule (automaton::sys_self_destruct_type automaton::*ptr);
-
     void schedule (action_runnable_interface*);
     
     void schedule_after (action_runnable_interface*,

### 2011-08-24 include/ioa/model_interface.hpp

@@ -38,7 +38,6 @@ namespace ioa {
     virtual int execute_sys_bind (const aid_t automaton) = 0;
     virtual int execute_sys_unbind (const aid_t automaton) = 0;
     virtual int execute_sys_destroy (const aid_t automaton) = 0;
-    virtual int execute_sys_self_destruct (const aid_t automaton) = 0;
 
     // Executing configuation actions.
     virtual aid_t create (const aid_t automaton,

### 2011-08-24 include/ioa/scheduler.hpp

@@ -19,7 +19,6 @@ namespace ioa {
   void schedule (automaton::sys_bind_type automaton::*ptr);
   void schedule (automaton::sys_unbind_type automaton::*ptr);
   void schedule (automaton::sys_destroy_type automaton::*ptr);
-  void schedule (automaton::sys_self_destruct_type automaton::*ptr);
 
   template <class I, class M>
   void schedule (M I::*member_ptr) {

### 2011-08-24 include/ioa/scheduler_interface.hpp

@@ -24,8 +24,6 @@ namespace ioa {
 
     virtual void schedule (automaton::sys_destroy_type automaton::*ptr) = 0;
 
-    virtual void schedule (automaton::sys_self_destruct_type automaton::*ptr) = 0;
-
     virtual void schedule (action_runnable_interface*) = 0;
 
     virtual void schedule_after (action_runnable_interface*,

### 2011-08-24 include/ioa/simple_scheduler.hpp

@@ -32,8 +32,6 @@ namespace ioa {
     
     void schedule (automaton::sys_destroy_type automaton::*ptr);
 
-    void schedule (automaton::sys_self_destruct_type automaton::*ptr);
-    
     void schedule (action_runnable_interface*);
     
     void schedule_after (action_runnable_interface*,

### 2011-08-24 include/ioa/system_scheduler_interface.hpp

@@ -32,8 +32,6 @@ namespace ioa {
     virtual void destroy (const aid_t automaton,
 			  void* const key) = 0;
 
-    virtual void self_destruct (const aid_t automaton) = 0;
-    
     virtual void created (const aid_t automaton,
 			  const created_t,
 			  void* const key,

### 2011-08-24 lib/automaton.cpp

@@ -4,9 +4,7 @@
 
 namespace ioa {
 
-  automaton::automaton () :
-    m_self_destruct (false)
-  { }
+  automaton::automaton () { }
 
   automaton::~automaton () {
     // Send the helpers a destroyed signal.
@@ -158,11 +156,6 @@ namespace ioa {
     schedule ();
   }
 
-  void automaton::self_destruct () {
-    m_self_destruct = true;
-    schedule ();
-  }
-
   bool automaton::sys_create_precondition () const {
     return !m_create_send.empty ();
   }
@@ -247,15 +240,6 @@ namespace ioa {
     m_destroy_send.erase (pos);
     m_destroy_recv.insert (helper);
     return helper;
-  }
-
-  bool automaton::sys_self_destruct_precondition () const {
-    return m_self_destruct;
-  }
-
-  void* automaton::sys_self_destruct_effect () {
-    m_self_destruct = false;
-    return 0;
   }
 
   void automaton::sys_created_effect (const created_arg_t& arg) {
@@ -445,9 +429,6 @@ namespace ioa {
     }
     if (sys_destroy_precondition ()) {
       ioa::schedule (&automaton::sys_destroy);
-    }
-    if (sys_self_destruct_precondition ()) {
-      ioa::schedule (&automaton::sys_self_destruct);
     }
   }
 

### 2011-08-24 lib/global_fifo_scheduler.cpp

@@ -14,13 +14,11 @@
 #include "sys_bind_runnable.hpp"
 #include "sys_unbind_runnable.hpp"
 #include "sys_destroy_runnable.hpp"
-#include "sys_self_destruct_runnable.hpp"
 
 #include "create_runnable.hpp"
 #include "bind_runnable.hpp"
 #include "unbind_runnable.hpp"
 #include "destroy_runnable.hpp"
-#include "self_destruct_runnable.hpp"
 
 #include "output_exec_runnable.hpp"
 #include "output_bound_runnable.hpp"
@@ -134,10 +132,6 @@ namespace ioa {
   
     void schedule (automaton::sys_destroy_type automaton::*member_ptr) {
       schedule_configq (new sys_destroy_runnable (get_current_aid ()));
-    }
-
-    void schedule (automaton::sys_self_destruct_type automaton::*member_ptr) {
-      schedule_configq (new sys_self_destruct_runnable (get_current_aid ()));
     }
 
     void schedule (action_runnable_interface* r) {
@@ -431,10 +425,6 @@ namespace ioa {
       schedule_configq (new destroy_runnable (automaton, key));
     }
 
-    void self_destruct (const aid_t automaton) {
-      schedule_configq (new self_destruct_runnable (automaton));
-    }
-
     void created (const aid_t aid,
 		  const created_t t,
 		  void* const key,
@@ -508,10 +498,6 @@ namespace ioa {
   }
   
   void global_fifo_scheduler::schedule (automaton::sys_destroy_type automaton::*ptr) {
-    m_impl->schedule (ptr);
-  }
-
-  void global_fifo_scheduler::schedule (automaton::sys_self_destruct_type automaton::*ptr) {
     m_impl->schedule (ptr);
   }
 

### 2011-08-24 lib/model.cpp

@@ -452,32 +452,6 @@ namespace ioa {
     return 0;
   }
 
-  int model::execute_sys_self_destruct (const aid_t aid) {
-    shared_lock lock (m_mutex);
-
-    if (!m_aids.contains (aid)) {
-      // Automaton does not exists.
-      return -1;
-    }
-
-    automaton* instance = get_instance (automaton_handle<automaton> (aid));
-
-    lock_automaton (aid);
-    m_system_scheduler.set_current_aid (aid);
-    if (instance->sys_self_destruct.precondition (*instance)) {
-      instance->sys_self_destruct (*instance);
-      m_system_scheduler.clear_current_aid ();
-      unlock_automaton (aid);
-      m_system_scheduler.self_destruct (aid);
-    }
-    else {
-      m_system_scheduler.clear_current_aid ();
-      unlock_automaton (aid);
-    }
-
-    return 0;
-  }
-
   int model::execute_output_bound (output_executor_interface& exec) {
     shared_lock lock (m_mutex);
     

### 2011-08-24 lib/model.hpp

@@ -129,7 +129,6 @@ namespace ioa {
     int execute_sys_bind (const aid_t automaton);
     int execute_sys_unbind (const aid_t automaton);
     int execute_sys_destroy (const aid_t automaton);
-    int execute_sys_self_destruct (const aid_t automaton);
     int execute_output_bound (output_executor_interface& exec);
     int execute_input_bound (input_executor_interface& exec);
     int execute_output_unbound (output_executor_interface& exec);

### 2011-08-24 lib/scheduler.cpp

@@ -29,11 +29,6 @@ namespace ioa {
     scheduler->schedule (ptr);
   }
 
-  void schedule (automaton::sys_self_destruct_type automaton::*ptr) {
-    assert (scheduler != 0);
-    scheduler->schedule (ptr);
-  }
-
   void close (int fd) {
     assert (scheduler != 0);
     scheduler->close (fd);

### 2011-08-24 lib/self_destruct_runnable.hpp

@@ -1,26 +1,0 @@
-#ifndef __self_destruct_runnable_hpp__
-#define __self_destruct_runnable_hpp__
-
-#include <ioa/runnable_interface.hpp>
-
-namespace ioa {
-  
-  class self_destruct_runnable :
-    public runnable_interface
-  {
-  private:
-    const aid_t m_automaton;
-    
-  public:
-    self_destruct_runnable (const aid_t automaton) :
-      m_automaton (automaton)
-    { }
-    
-    void operator() (model_interface& model) {
-      model.destroy (m_automaton);
-    }
-  };
-  
-}
-
-#endif

### 2011-08-24 lib/simple_scheduler.cpp

@@ -19,13 +19,11 @@
 #include "sys_bind_runnable.hpp"
 #include "sys_unbind_runnable.hpp"
 #include "sys_destroy_runnable.hpp"
-#include "sys_self_destruct_runnable.hpp"
 
 #include "create_runnable.hpp"
 #include "bind_runnable.hpp"
 #include "unbind_runnable.hpp"
 #include "destroy_runnable.hpp"
-#include "self_destruct_runnable.hpp"
 
 #include "output_exec_runnable.hpp"
 #include "output_bound_runnable.hpp"
@@ -419,10 +417,6 @@ namespace ioa {
       schedule_sysq (new sys_destroy_runnable (get_current_aid ()));
     }
 
-    void schedule (automaton::sys_self_destruct_type automaton::*member_ptr) {
-      schedule_sysq (new sys_self_destruct_runnable (get_current_aid ()));
-    }
-
     void schedule (action_runnable_interface* r) {
       schedule_execq (r);
     }
@@ -540,10 +534,6 @@ namespace ioa {
       schedule_sysq (new destroy_runnable (automaton, key));
     }
 
-    void self_destruct (const aid_t automaton) {
-      schedule_sysq (new self_destruct_runnable (automaton));
-    }
-
     void created (const aid_t aid,
 		  const created_t t,
 		  void* const key,
@@ -620,10 +610,6 @@ namespace ioa {
     m_impl->schedule (ptr);
   }
   
-  void simple_scheduler::schedule (automaton::sys_self_destruct_type automaton::*ptr) {
-    m_impl->schedule (ptr);
-  }
-
   void simple_scheduler::schedule (action_runnable_interface* r) {
     m_impl->schedule (r);
   }

### 2011-08-24 lib/sys_self_destruct_runnable.hpp

@@ -1,33 +1,0 @@
-#ifndef __sys_self_destruct_runnable_hpp__
-#define __sys_self_destruct_runnable_hpp__
-
-#include <ioa/action_runnable.hpp>
-#include <ioa/automaton.hpp>
-
-namespace ioa {
-  
-  class sys_self_destruct_runnable :
-    public action_runnable_interface
-  {
-  private:
-    const aid_t m_automaton;
-    const action_executor<automaton, automaton::sys_self_destruct_type> m_action;
-
-  public:
-    sys_self_destruct_runnable (const aid_t automaton) :
-      m_automaton (automaton),
-      m_action (automaton, &automaton::sys_self_destruct)
-    { }
-    
-    void operator() (model_interface& model) {
-      model.execute_sys_self_destruct (m_automaton);
-    }
-
-    const action_executor_interface& get_action () const {
-      return m_action;
-    }
-  };
-  
-}
-
-#endif

### 2011-08-24 lib/tcp_connection_automaton.cpp

@@ -43,7 +43,7 @@ namespace ioa {
     if ((o == &send_complete && send_complete.recent_op == UNBOUND) ||
 	(o == &receive && receive.recent_op == UNBOUND)) {
       // User is unbinding.  Time to die.
-      self_destruct ();
+      // TODO:  self_destruct ();
     }
     else if (o == &receive && receive.recent_op == BOUND && m_fd == -1 && m_receive_state == SCHEDULE_READ_READY) {
       // An automaton has bound to receive but we will never receive because m_fd is bad.

### 2011-08-24 test/action_executor.cpp

@@ -26,7 +26,6 @@ struct test_model :
   int execute_sys_bind (const ioa::aid_t automaton) { return -1; }
   int execute_sys_unbind (const ioa::aid_t automaton) { return -1; }
   int execute_sys_destroy (const ioa::aid_t automaton) { return -1; }
-  int execute_sys_self_destruct (const ioa::aid_t automaton) { return -1; }
 
   // Executing configuation actions.
   ioa::aid_t create (const ioa::aid_t automaton,
@@ -68,8 +67,6 @@ struct test_system_scheduler :
     
   void destroy (const ioa::aid_t automaton,
 		void* const key) { }
-
-  void self_destruct (const ioa::aid_t automaton) { }
 
   void created (const ioa::aid_t automaton,
 		const ioa::created_t,

### 2011-08-24 test/model.cpp

@@ -161,10 +161,6 @@ struct test_system_scheduler :
     assert (false);
   }
 
-  void self_destruct (const ioa::aid_t automaton) {
-    assert (false);
-  }
-
   void created (const ioa::aid_t automaton,
 		const ioa::created_t type,
 		void* const key,

### 2011-08-24 test/reuse_bind_key.cpp

@@ -33,7 +33,8 @@
 template <class OI, class OM, class II, class IM>
 class static_binding :
   private ioa::observer,
-  private ioa::system_binding_manager_interface
+  private ioa::system_binding_manager_interface,
+  public ioa::observable
 {
 private:
   bool m_bound;
@@ -91,6 +92,7 @@ private:
     case ioa::BOUND_RESULT:
       // Okay.
       m_bound = true;
+      notify_observers ();
       break;
     }
   }
@@ -104,6 +106,7 @@ private:
       // Okay.
       m_bound = false;
       ++m_unbound_count;
+      notify_observers ();
       if (m_unbound_count == 2) {
 	delete this;
       }
@@ -148,13 +151,8 @@ class automaton_B :
   public ioa::automaton
 {
 private:
-  void schedule () const { }
-  void input_effect () {
-    self_destruct ();
-  }
-  void input_schedule () const {
-    schedule ();
-  }
+  void input_effect () { }
+  void input_schedule () const { }
 
 public:
   UV_UP_INPUT (automaton_B, input);
@@ -170,11 +168,17 @@ private:
   bool m_rebind_flag;
 
   void observe (ioa::observable* o) {
-    if (o == m_b1) {
-      if (-1 == m_b1->get_handle ()) {
-	m_b1 = 0;
+    if (o == m_binding) {
+      if (!m_binding->is_bound ()) {
+	m_b1->destroy ();
       }
     }
+    else if (o == m_b1) {
+      if (-1 == m_b1->get_handle ()) {
+      	m_b1 = 0;
+      }
+    }
+
     schedule ();
   }
 
@@ -187,9 +191,7 @@ private:
     }
   }
   
-  bool output_precondition () const {
-    return m_b1 != 0 && ioa::binding_count (&automaton_A::output) != 0;
-  }
+  bool output_precondition () const { return false; }
   void output_effect () { }
   void output_schedule () const { schedule (); }
   UV_UP_OUTPUT (automaton_A, output);
@@ -215,6 +217,7 @@ public:
     m_self (ioa::get_aid ()),
     m_rebind_flag (false) {
     m_binding = new static_binding<automaton_A, output_type, automaton_B, automaton_B::input_type> (this);
+    add_observable (m_binding);
     m_b1 = new ioa::automaton_manager<automaton_B> (this, ioa::make_generator<automaton_B> ());
     add_observable (m_b1);
     m_binding->set_output (&m_self, &automaton_A::output);

### 2011-08-30 papers/paper1/system_model.tex
Revision of section 2.

@@
 \section{I/O Automata Component Model\label{system_model}}
-
-\subsection{I/O Automata as Components}
 
@@
 
+In this section we describe how I/O automata offer a natural realization of components, interfaces, and provide natural support for concurrency and composition.
 The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
@@
 
-An automaton's output and internal actions constitutes its \emph{local signature}~\cite{lynch1996distributed}.
-Often, local actions are presented as a precondition and an effect.
+An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
+Often, local actions consists of a precondition and an effect.
 The precondition is a predicate over the state variables that enables the effect when selected.
@@
 
-An automaton's output and input actions constitutes its \emph{external signature}~\cite{lynch1996distributed}.
+Together, an automaton's output and input actions constitute its \emph{external signature}~\cite{lynch1996distributed}.
 An automaton communicates with other automata through the actions in its external signature.
-I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effect of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
-Composing automata that have identically named output actions is ambiguous and not allowed.
-The result of composition is a new automaton thus the principles for reasoning about the behavior of a composed system are the same as the principles for reasoning about a single automaton.
-Often, the properties of the resulting aggregate automaton can be reasoned about using the properties of the constituent automata.
+I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effects of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
+Composing automata that have identically named output actions would be ambiguous and thus is not allowed.
+The result of composition is a new automaton, so that the principles for reasoning about the behavior of a composed system are the same as the principles for reasoning about a single automaton.
+Often, the properties of the resulting aggregate automaton can be reasoned about using only the properties of the constituent automata.
 %% A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.
@@
 Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
-To be general, the scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not executed) infinitely often.
+To be general, the scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
 Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
-Executing a composed output action consists of atomically evaluating the precondition and simultaneously applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
-The atomic relationship between composed output actions and input actions means that multiple automata may atomically change state when executing an output action.
-The guarantee of atomicity allows a developer to relate the state of one automaton to another resulting in well-defined interactions that facilitate composition.
+Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
+The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
+The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.
 
-I/O automata are components and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
+I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
 Each action is only allowed to modify the state of the automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
@@
 
-\subsection{Practical Considerations}
+\subsection{Practical Considerations\label{practical}}
 
@@
 \includegraphics[width=\textwidth]{system_model}
-\caption{Example event detection system.
-  Automata are depicted as solid rectangles, e.g., Processor.
-  Named automata are labeled by a box in their upper left corner, e.g., C.
-  Actions are depicted as dotted rectangles, e.g., \emph{detect!}.
-  Output actions are suffixed with \emph{!}, e.g., \emph{detect!}.
-  Input actions are suffixed with \emph{?}, e.g., \emph{detected?}.
-  Internal actions have no suffix, e.g., \emph{process1}.
-  The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.
-  Bindings are indicate by dotted lines with an arrow pointing to the input.}
+\caption{Example event detection system.}
 \label{sys_model}
@@
 The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
-Distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple child of the same type.
+A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.
 
-For I/O automata, recursive encapsulation means that an I/O automaton can contain and use other I/O automata.
-Recursive encapsulation implies that a parent automaton is the composition of itself and its children.
-The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively to it and its children.
-The sample event detector system of figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
+Recursive encapsulation means that an I/O automaton may contain and use other I/O automata.
+Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
+The kernel automaton allows a parent automaton to have its own state variables and actions.
+The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.
+
+Figure\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles, e.g., Processor.
+Named automata are labeled by a box in their upper left corner, e.g., C.
+The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
 The Processor automaton is the root and C and D are its children.
@@
 Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
-The automaton that prescribes a bindings is said to \emph{own} that binding.
+The automaton that prescribes a binding is said to \emph{own} that binding.
 The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
-A map for renaming inputs is generated by replacing action names with their fully-qualified name in all bindings by tracing from the root.
-For the example in figure~\ref{sys_model}, this map is $\{ (out!(s), C.in?(s)), (C.out!(t), D.in?(t)), (D.detect!, detected?) \}$.
+A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
 Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
 
+For the event detection system illustrated in Figure~\ref{sys_model}, actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
+Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have no suffix, e.g., \emph{process1}.
+The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.
+The map of bindings for the event detection system is $\{ (out!(s), C.in?(s)), (C.out!(t), D.in?(t)), (D.detect!, detected?) \}$.
+
 \paragraph{Binding rules.}
-To be compliant with the named-based composition of the I/O automata model, the set of bindings must adhere to certain rules.
+To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
 First, the output action and input action of a binding must agree on the type being produced and consumed.
 For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
-The system in figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value, i.e., a signal.
+The system shown in figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value (i.e., a signal).
 Second, an input action can be bound to at most one output action.
@@
 A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
-For dynamics, however, we strengthen this rule to say that an input action can appear at most once in the global binding map.
+For dynamics, which are discussed in section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
 Consequently, every binding has exactly one owner.
-Third, an output action cannot be bound to more than one input action in the same automaton.
-This was rejected in the I/O automata model since the state of the automaton containing the input actions is not well-defined after the output is executed since the input action affects will be applied in undefined order.
-Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
-The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
+For a graphical representation such as Figure~\ref{sys_model}, this rule means that an input is the sink of at most one arrow (binding).
+Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
+Graphically, arrows originating at the same output action terminate in different components.
+%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
+%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
 Fourth, an output action cannot be bound to an input action in the same automaton.
+Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
 The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
-Allowing such bindings result in undefined behavior since the order of the output and input effects is undefined.
 To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
 However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
-To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
+Graphically, an arrow originating in one component must terminate in another component.
+%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
 
@@
 Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
-Since the state of each automaton is independent, opportunities exist for the concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
+Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
 Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
@@
 Migrating intense computation to internal actions increases the level of parallelism in a system, i.e., the set of implied automata is small.
-High degrees of fan-out, an output action being bound to many input actions, creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
+A high degree of fan-out (an output action being bound to many input actions) creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
 However, the input effects can be applied in parallel since the state of each automaton is independent.
-High degrees of fan-in, an automaton has many bound input actions, also creates a bottle-neck by increasing the probability that two sets of implied automata will not be disjoint.
-An high-performance scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
-A slow composition of automata might be statically composed to take advantage of in-lining.
-A slow automaton might be factored into a number of child automata to increase parallelism.
+A high degree of fan-in (an automaton has many bound input actions) also creates a bottle-neck by increasing the probability that two sets of implied automata will not be disjoint.
+A scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
+Compositions of automata can be statically composed to take advantage of in-lining.
+An individual automaton can be factored into a number of child automata to increase parallelism.
 
@@
 \paragraph{Action types.}
-The combination of input, output, and internal actions, unvalued and valued, unparameterized and parameterized results in 10 possible action types:
+The allowed combinations of input, output, and internal actions, unvalued and valued actions, and unparameterized and parameterized actions result in 10 possible action types:
 unvalued unparameterized output (uv-up output),
@@
 \includegraphics[width=\textwidth]{example1}
-\caption{Example remote sensor system.
-  Automata are depicted as solid rectangles, e.g., Remote Sensor.
-  Named automata are labeled by a box in their upper left corner, e.g., Z.
-  Actions are depicted as dotted rectangles, e.g., \emph{trigger!}.
-  Output actions are suffixed with \emph{!}, e.g., \emph{trigger!}.
-  Input actions are suffixed with \emph{?}, e.g., \emph{trigger?}.
-  Internal actions have no suffix, e.g., \emph{compute}.
-  The type associated with a valued action is in parentheses after the name, e.g., \emph{in?(s)}.
-  Parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
-  Bindings are indicate by dotted lines with an arrow pointing to the input.
-}
+\caption{Example remote sensor system.}
 \label{example1}
@@
 \paragraph{Example.}
-Figure~\ref{example1} depicts a remote sensor systems as a composition of automata.
+Figure~\ref{example1} depicts a remote sensor systems as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
 The Remote Sensor automaton consists of the Control automaton $Z$ and three Channel automata $A$, $B$, and $C$ representing three network connections.
@@
 Since the sets are disjoint, the two actions can be executed concurrently.
+
+\subsection{Dynamics\label{dynamics}}
+
+The I/O automata model assumes that systems are composed of a finite or infinitely countable number of automata~\cite{lynch1996distributed}.
+Assuming an infinitely countable number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
+The set of interactions in the I/O automata model is also fixed since automata are statically composed.
+Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
+To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}.
+
+While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system results in over-provisioning or inflexibility.
+To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
+System resources are wasted if the actual number of participants is much smaller than $N$.
+If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
+Consequently, we require the ability to deal with a dynamic set of automata and interactions.
+
+\paragraph{System actions.}
+The four indivisible operations for managing a dynamic set of automata are create, bind, unbind, and destroy.
+Create, bind, unbind, and destroy are collectively called \emph{system actions}.
+The \emph{create} operation allows an automaton to create a child automaton.
+The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
+The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.
+
+System actions can fail for a variety of reasons.
+A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
+A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in section~\ref{practical} would be violated.
+An unbind fails if the binding does not exist.
+Note that the owner of binding is taken into account when determining if the binding exist so only the owner of a binding can dissolve it.
+A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy.
+
+%% Let $A$ be the set of automata that exist where each element in $A$ is a pair $(p, a)$ where $p$ is the parent of $a$.
+%% Let $B$ be the set of bindings that exist where each element in $B$ is a tuple $(w, o_a, o_n, o_p, i_a, i_n, i_p)$ where $w$ is the automaton that owns the binding, $o_a$ is the output automaton, $o_n$ is the name of the output action, $o_p$ is the parameter of the output action, $i_a$ is the input automaton, $i_n$ is the name of the input action, and $i_p$ is the parameter of the input action.
+%% Define the predicate $exists (a) = \langle \exists p, q : (p, q) \in A :: q = a\rangle$ that is true when automaton $a$ exists, i.e., has a parent\footnote{The root automaton's parent is the special symbol $\perp$.}.
+%% Define the predicate $existsb (a) = \langle \exists a : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: a = w \lor a = o_a \lor a = i_a \rangle$ that is true when automaton $a$ appears somewhere in $B$.
+%% We define the create operation as the Hoare triple $\{exists (p) \land \lnot exists (a)\} \quad create(a)_p \quad \{(p,a) \in A\}$ which says that the parent automaton $p$ can create automaton $a$ if $p$ exists and $a$ does not exist.
+%% The destroy operation can be defined $\{(p,a) \in A\} \quad destroy(a)_p \quad \{(p,a) \notin A \land \lnot existsb(a)\}$ which says that in order to destroy $a$ and remove all bindings associated with $a$, $p$ must be the parent of $a$.
+%% Define the predicate $free (a, n, p) = \langle \forall w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: \lnot (i_a = a \land i_n = n \land i_p = p) \rangle$ that is true when the input action argument is not bound.
+%% Define the predicate $involved (o, n, p, i) = \langle \exists w, o_a, o_n, o_p, i_a, i_n, i_p : (w, o_a, o_n, o_p, i_a, i_n, i_p) \in B :: o_a = o \land o_n = n \land o_p = p \land i_a = i \rangle$ that is true when automaton $o$'s output action $(n, p, i)$ is bound to some input action in automaton $i$.
+%% The bind operation can be defined $\{exists (w) \land exists (o_a) \land exists (i_a) \land free (i_a, i_n, i_p) \land \lnot involved(o_a, o_n, o_p, i_a) \land o_a \neq i_a \} \quad bind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\}$ which says that in order for the binding to succeed the owner $w$ exists, automaton $o_a$ exists, automaton $i_a$ exists, the input action $(i_a, i_n, i_p)$ must not be bound, the output action $(o_n, o_p)$ of automaton $o_a$ cannot be bound to an input action in automaton $i_a$, and the automata $o_a$ and $i_a$ cannot be the same.
+%% The precondition for $bind$ is same as the binding rules derived in section~\ref{system_model} augmented with existence tests for the owner, output automaton, and input automaton.
+%% The unbind operation is similar to destroy: $\{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\} \quad unbind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \notin B\}$.
+
+% I didn't do more complex transactions, i.e., create-and-bind, because its not clear how it fails.
+% Are partial successes allowed?
+% Furthermore, such  transactions can be arbitrarily complex which doesn't bode well for efficiency.
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{system_action}
+\caption{System actions.
+  An automaton contains output actions for creating, binding, unbinding, and destroying.
+  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
+  The system automaton returns a result using the appropriate parameterized output action.
+  The automaton contains input actions to receive the results.}
+\label{system_action}
+\end{figure}
+
+\paragraph{The system automaton.}
+Let the set of automata $A$ and set of bindings $B$ be managed by the \emph{system automaton} which represents the run-time system.
+Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
+The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
+Figure~\ref{system_action} shows the relationship between an automaton and the system automaton.
+
+Conceptually, the state variables $A$ and $B$ belong to the system automaton.
+However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
+The scheduler uses $A$ to ensure that only actions belonging to automata exist are selected and uses $B$ to ensure that the appropriate set of inputs actions receive the value produced when an output action is executed.
+Thus, the set of actions implied by any action is the set defined in section~\ref{system_model} \emph{and} and the system automaton.
+However, the opportunities for concurrency explored in section~\ref{system_model} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+
+\paragraph{Binding predicates.}
+The static configuration of a system as prescribed by the I/O automata model guarantees that the set of implied automata for an action is always the same.
+This property is necessary and desirable for reasoning about the behavior of the system.
+Admitting dynamics removes this guarantee since the set of implied automata will change according to the various system actions being executed.
+To be specific, the set of implied automata for internal actions does not change but the set of implied automata for output actions can be changed with a successful bind, unbind, or destroy.
+
+To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
+To do this, we conjunct the precondition of an output action with a predicate over the set of bindings $B$ to ensure that the output action is not executed until its set of implied automata is correct.
+Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as previously discussed.
+Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
+A static system can be implemented by adding binding predicates and logic to create the required constellation of automata.

### 2011-08-31 papers/paper1/component_model.tex
First draft of section 3.

--- 2011-08-30 papers/paper1/system_model.tex
@@
-\section{I/O Automata Component Model\label{system_model}}
+\section{I/O Automata Component Model\label{component_model}}
 
@@
 The scheduler uses $A$ to ensure that only actions belonging to automata exist are selected and uses $B$ to ensure that the appropriate set of inputs actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the set defined in section~\ref{system_model} \emph{and} and the system automaton.
-However, the opportunities for concurrency explored in section~\ref{system_model} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+Thus, the set of actions implied by any action is the set defined in section~\ref{practical} \emph{and} and the system automaton.
+However, the opportunities for concurrency explored in section~\ref{practical} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
 

### 2011-08-31 papers/paper1/design.tex
First draft of section 3.

@@
 
-\begin{itemize}
-  \item Problem
-  \item Design forces
-  \item Solution
-  \item Consequences
-\end{itemize}
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{architecture}
+\caption{Framework architecture.}
+\label{framework_architecture}
+\end{figure}
+
+An implementation of the model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
+The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
+The Scheduler has the resposibility of selecting and executing local actions subject to the I/O automata model and concurrency constraints of~\ref{component_model}.
+All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
+The services provided by the base class are indicated by the Automaton layer of Figure~\ref{framework_architecture}.
+The Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
+In most situations, the services provided by the Automaton base class are too low-level to be used directly by a User Automaton.
+An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Automaton base class.
+Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
+We assume a POSIX environment whose native concurrency mechanism is threads.
+
+\subsection{Scheduler}
+
+The scheduler has the responsibility of selecting and executing local actions.
+The two activities of the scheduler can be decomposed into a \emph{model} that executes actions and a \emph{controller} that selects the next action.
+
+\paragraph{The Model.}
+The model enforces the execution and concurrency constraints of section~\ref{component_model}.
+Input actions are executed with output actions according to the current set of bindings.
+The model allows executing actions concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
+Since concurrent access to the model is allowed, the module implementing the model must thread-safe.
+
+Our implementation of the model uses a two-level locking scheme to enforce the concurrent execution constraints.
+Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
+All other actions acquire a read lock to implement single-writer/multiple reader semantics.
+The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
+The locks are acquired in a fixed order to prevent deadlock.
+Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.
+
+\paragraph{The Controller.}
+The controller encodes a \emph{policy} that selects the next local action.
+For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler should be \emph{fair} as defined in section~\ref{component_model}.
+The run-time system can be initialized with different controllers to achieve different performance objectives.
+
+\paragraph{Explicit dynamic scheduling.}
+The I/O automata model assumes that the scheduler knows about every local action in the system.
+There are two approaches for achieving this in practice.
+In the first approach, automata declare \emph{statically} all of their actions to the scheduler.
+The controller then has the responsibility of cycling through the actions according to some fair policy.
+Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to trying each action with brute force.
+
+The approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
+The controller then has the responsibility of processing the set of requested actions according to some fair policy.
+Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
+Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
+Consequently, an automaton is given a chance to declare actions to the scheduler after one of its actions is executed, i.e., changes state.
+Since the state and therefore preconditions of all automata are independent, each automaton is responsible for scheduling its own actions.
+
+In this approach, scheduling disabled actions is unnecessary.
+Let $a$ be a scheduled disabled action.
+In order for $a$ to become enabled, the automaton associated with the action must change state by executing an enabled action $b$ sometime after $a$ is scheduled.
+Consequently, the scheduling of the disabled action $a$ can be deferred until the execution of action $b$.
+Notice that refraining from scheduling disabled actions does not affect the fairness of the scheduler since they cannot cause a state change when selected.
+This result leads to an programming idiom where the precondition of each action is checked before it is added to the schedule.
+
+In contrast, an automaton should ensure that the scheduler will consider all actions that are currently enabled.
+A simple technique for accomplishing this is to schedule all enabled actions.
+More efficient schemes are possible if one considers which actions enable each other and which actions have already been scheduled.
+Not declaring an enabled action violates the fairness of the controller since it will never select an action that may cause a state change.
+Not scheduling an (enabled) action is a common mistake.
+
+\paragraph{Actions of non-existent automata.}
+Every action in the set of actions considered by the controller was scheduled by its associated automaton.
+Since automata can be destroyed, an action is the set of actions owned by the controller might belong to an automaton that no longer exists.
+Our solution is to check if the automaton exists before executing an action.
+This check is performed by the model and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
+Our solution assumes that each automaton is assigned a unique identifier and that the set of actions turns over comparatively faster than the set of identifiers.
+
+\paragraph{Delayed actions.}
+Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities.
+The framework should permit the development of automata that act as timers and alarms.
+The corresponding feature that must be supported by the framework is scheduling an action at a future time.
+Conceptually, the action is held until the time associated with the action at which point it is added to the set of actions being considered by the scheduler.
+Actions scheduled in the past or present are released immediately to the scheduler.
+When the same action is scheduled at two different times, the earliest time is used.
+% Things I didn't say:  No canceling.
+
+\paragraph{File descriptor events.}
+File descriptors are used to communicate with the outside world in a POSIX environment.
+Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
+Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
+This implies the use of either the Proactor or Reactor pattern.
+For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.
+
+Our implementation of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
+A user automaton creates a file descriptors and places it in non-blocking mode.
+The automaton then tells the scheduler to release an action (typically an internal action) whenever the automaton becomes ready for reading (or writing).
+Subsequent requests made using the same file descriptor and event (ready for reading, ready for writing) are ignored.
+Once the file descriptor is ready, the action is released for consideration by the controller.
+The scheduler also provides a special function for closing file descriptors that purges them from the reactor.
+
+\paragraph{Global FIFO scheduler.}
+The scheduler included with the framework is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
+Scheduling is idempotent to prevent unbounded growth of the queue.
+The global FIFO scheduler uses the Reactor pattern to implement delayed actions and file descriptor events.
+
+\subsection{System Actions}
+
+\paragraph{System action protocol.}
+Our approach to system actions follows the model introduced in section~\ref{dynamics}.
+All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
+The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with either success or failure.
+Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
+Subjects allow an automaton to have multiple outstanding requests.
+
+Automata use a simple protocol for creating and destroying child automata.
+Each child automaton is represented by a unique subject which is in one or two of five states: CreateSend, CreateRecv, CreateDone, DestroySend, and DestroyRecv.
+Subjects start in the CreateSend state which indicates that the request for this subject is waiting to be sent to the system automaton.
+Once the request has been sent, the subject transitions to the CreateRecv state and waits for a response from the system automaton.
+Once the system automaton sends the response, the subject transitions to the CreateDone state.
+When the user automaton requests that the child automaton be destroyed, the subject is either (1) added to the DestroySend state if the subject is in CreateRecv or CreateDone or (2) removed from all states if the subject is in CreateSend.
+A subject appearing in both CreateDone and DestroySend transitions to the DestroyRecv state when the automaton makes the request to destroy.
+The system automaton can process system actions in any order.
+This protocol ensures that an automaton doesn't request that a child automaton be destroyed until it receives confirmation that it was created.
+The subject is forgotten when the system automaton responds with the result of the destroy.
+
+The protocol for binding is similar but addresses the possibility that a binding can be dissolved at any point due to the asynchronous destruction of automata.
+Define analogous states for the binding state machine: BindSend, BindRecv, BindDone, UnbindSend, and UnbindRecv.
+A binding subject can be in either BindRecv, BindDone, UnbindSend, or UnbindRecv when it receives a result from the system automaton indicating that the binding was dissolved.
+
+\paragraph{Automaton and Binding managers.}
+The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
+The Automaton Managers and Binding Managers provide an individualized facade to the system action state machines.
+Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
+Automaton Managers can be observed using the Observer pattern to determine when and how the creation/destruction process succeeds.
+Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.
+
+\paragraph{Generators.}
+A \emph{generator} is a single-use factory for creating an instance of an automaton.
+Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.

### 2011-09-03 papers/paper1/design.tex
First part of Representation and Programming Model.

@@
 Since the state and therefore preconditions of all automata are independent, each automaton is responsible for scheduling its own actions.
+For bootstrapping, each automaton is allowed to schedule actions when it is created.
 
@@
 Not scheduling an (enabled) action is a common mistake.
+
+\paragraph{Binding predicates and scheduling.}
+The preceding discussion on scheduling claims that an action in one automaton cannot affect preconditions in another automaton.
+This is not entirely correct when one considers the system automaton and binding predicates.
+Imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
+When the automaton is created, its only action is disabled.
+Consequently, it will not schedule its action and therefore not be executed.
+Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
+The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
+While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound; a situation that occurs quite often in practice.
+To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a bind or unbind.
 

### 2011-09-03 papers/paper1/programming_model.tex
Removed self-destruct.
First part of Representation and Programming Model.

@@
 
-Conclude with our recommendations for programming, highlight need for exploration.
+In this section, we show how I/O automata are represented using our framework and how they can be used to construct programs.
+We develop a simulation for a distributed algorithm to show the basic features of the framework and then show a real implementation of the same algorithm.
+The problem we consider is that of electing a leader in unidirectional ring.
+
+The algorithm we consider is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{} and Chang and Roberts~\cite{}.
+The LCR algorithm assumes that every node has a unique identifier.
+Each node begins by assuming that it is the leader and send its own unique id to its neighbor.
+If a node receives a unique id from its predecessor that is greater that the id that it believes to be the leader, then it changes its belief about the leader to the new id and sends the new id to its successor.
+When a node receives its own id, it elects itself the leader of the ring.
+
+\subsection{The AsynchLCR Automaton}
+
+Appendix~\ref{asynch_lcr} shows the code for the asynchronous LCR automaton.
+
+All automaton inherit from the \verb+ioa::automaton+ base class which implements the system call interface expected by the system automaton.
+The \verb+asynch_lcr_automaton+ inherits from the \verb+ioa::automaton+ base class (line 13-14).
+
+The state variables of the automaton are captured as members variables.
+The \verb+m_u+ variable stores the id of node that this id believes to be the leader (line 23).
+The \verb+m_send+ variable is a queue that stores ids that will be sent to the next node in the ring (line 24).
+The \verb+m+status+ variable is used by the leader to report its election (line 25).
+
+The constructor allows for initialization of the state variables and bootstrapping the scheduler.
+The unique id of the automaton is a random number with ties broken by the index of the automaton.
+Every node assumes that they are the leader initially \verb+i+ (line 29).
+All nodes start in the unknown state (line 30).
+Each node starts by sending their id to their neighbor (line 32).
+The automaton bootstraps the scheduler after its state variable are initialized (line 33).
+
+Actions consist of member variables that implement the Action concept.
+An easiest way to declare these variables is to use a macro for the appropriate action type.
+The \verb+asynch_lcr_automaton+ contains a \verb+send+ output action (line 53), a \verb+receive+ input action (line 73), and a \verb+leader+ output action (line 90).
+The macros assume that actions are encoded using two or three functions.
+Local actions require a precondition named \verb+action_name_precondition+ that returns true if the action is enabled (line 37, 76).
+Notice that the preconditions are declared \verb+const+ since they should not change the state of the automaton.
+All actions require an effect named \verb+action_name_effect+ that encodes the effect of the action (line 42, 56, 81).
+All actions also require a scheduling function named \verb+action_name_schedule+ that allows the automaton to schedule actions (line 48, 68, 85).
+Notice that the schedule functions are also declare \verb+const+ for the same reason as the preconditions.
+
+In the I/O automata model, the send precondition only requires that the send queue not be empty (line 38).
+However, if the send action executes before it is bound, the id at the front of the send queue will be lost.
+Consequently, we add a binding predicate that ensures that send output is bound to at least one input (line 39).
+The \verb+ioa::binding_count+ function returns the number of actions bound to the given action.
+This is non-negative for output actions, 0 or 1 for input actions, and 0 for internal actions.
+
+The send action is a valued unparameterized (v-up) output.
+The value produced by send has the type \verb+UID_t+ (line 42).
+Notice that this agrees with the type in the action declaration (line 53).
+Similarly, the receive action is a valued unparameterized (v-up) input action that accepts a value of type \verb+UID_t+ (line 56, 73).
+Notice that input action effects take a constant reference.
+
+The send action removes and returns the id at the front of the queue (line 42-46).
+The receive action updates the current leader sending it to nodes successor and confirms an election (line 56-66).
+Once a node is chosen as the leader, it reports it using the leader action (line 81-83).
+
+After every action is executed, the associated scheduling function is called.
+All scheduling calls in the \verb+asynch_lcr_automaton+ dispatch to the same member function (line 93-100).
+The \verb+asynch_lcr_automaton+ uses the strategy outlined in section~\ref{design} and checks the precondition of each action before scheduling it.
+
+\subsection{AsynchLCR Simulation}
+
+For a simulation of the AsynchLCR algorithm, we connect each node using a channel automaton.
+The channel automaton is introduced by Lynch in Chapter 8 of ~\cite{lynch1996distributed} and listed in appendix~\ref{channel}.
+The channel automaton takes a template parameter for the messages transported by the channel.
+
+Appendix~\ref{ring} contains an automaton that creates the ring.
+The template parameter \verb+T+ represents the automaton, i.e., \verb+asynch_lcr_automaton+, and the template parameter \verb+M+ represents the message type for the channel automata, i.e., \verb+UID_t+.
+The size of the ring to simulate is passed as the constructor argument \verb+N+ (line 9).
+
+The constructor for the ring starts by declaring vectors to hold automaton managers for the LCR automata (line 12) and automaton managers for the channels (line 13).
+The loop in (line 15-20) initializes the vectors.
+The \verb+ioa::make_automaton_manager+ function creates automaton managers.
+These require action to an \verb+ioa::automaton+ which is always provided by a \verb+this+ pointer and a generator as described in section~\ref{design}.
+The \verb+ioa::make_generator+ creates a generator.
+Generators take constructor arguments for the automaton they will produce and pass them to the constructor when the automaton is allocated.
+For example, the LCR automata are initialized with an index (line 17).
+
+The loop in (line 22-33) creates the various bindings necessary to create the ring and receive the leader election.
+The \verb+ioa::make_binding_manager+ function creates binding managers.
+This function requires access to an \verb+ioa::automaton+ to realize the system actions.
+The other arguments are the automaton managers and actions for the output action and input action.
+For example, the send action of the LCR automaton is bound to the send action of the successor channel in (line 23-25).
+Similarly, the receive action of the LCR automaton is bound to the receive action of the predecessor channel in (line 26-28).
+
+The ring automaton contains a single unvalued parameterized (uv-p) input action for receiving the leader (line 36-38).
+The parameter \verb+i+ is the index of the LCR automaton.
+The leader actions are bound in (line 29-31).
+Notice that the parameter for the leader input action is set in the call to \verb+ioa::make_binding_manager+.
+
+Binding to an automaton that already exists requires an object that behaves like an automaton manager but refers to an automaton that already exists.
+Objects of the \verb+ioa::handle_manager+ class satisfy this requirement.
+The ring automaton declares a handler manager \verb+m_self+ (line 6) and initializes it with its automaton identifier (line 10).
+Associated with each automaton is a unique identifier called an \emph{aid} which can be retrieved with \verb+ioa::get_aid+.
+
+Appendix~\ref{driver} contains the driver program for the LCR simulation.
+A global FIFO scheduler \verb+sched+ is declared on (line 11).
+The scheduler is started with the \verb+ioa::run+ function (line 12).
+The \verb+ioa::run+ function starts the scheduler using the root automaton returned by the specified generator which in this case is the ring automaton.
+
+\subsection{AsynchLCR Implementation}
+
+%% Conclude with our recommendations for programming, highlight need for exploration.
+
+%% Tips for Writing Programs with I/O Automata
+%% 1. Make sure that a parameter exists.
+%% 2. Move logic in observe to local actions.
+%% 3. Avoid observing an arbitrary number of managers.
+%% 4. Do not bind to managers allocated on the stack.
+%% 5. Stop when an error is encoutered and report it.
+%% 6. Make sure output actions are bound.
+%% 7. Make sure local actions are scheduled.
+%% 8. Do not pass pointers---use const_shared_ptr.
+%% 9. Check preconditions.
+%% 10. Check effects.
+%% 11. Order clauses of preconditions for efficiency.
+%% 12. Automata that should be destroyed at the same time should be created at the same time and vice versa.  (Group using a parent.)
+
+%% \begin{outline}
+%% \item State
+%%   \begin{outline}
+%%   \item There is no shared state in a 
+%%   \item Local state only
+%%   \item Shared state
+%%     \begin{outline}
+%%       \item Impossible in distributed systems
+%%       \item Dangerous in local systems
+%%     \end{outline}
+%%   \end{outline}
+%% \item Communication
+%%   \begin{outline}
+%%   \item Atomic asynchronous message passing
+%%   \item Network sets size of atom (UDP)
+%%   \item Can build reliable streams (TCP)
+%%   \item Local equivalent is passing a value
+%%   \item Model should lend itself to writing protocols
+%%   \end{outline}
+%% \item Asynchrony
+%%   \begin{outline}
+%%     \item Model must have natural support for asynchrony, i.e., event-based
+%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
+%%   \end{outline}
+%% \item Concurrency
+%%   \begin{outline}
+%%     \item Reason about systems using non-deterministic interleaving of atomic actions
+%%     \item Model should admit implementations that execute concurrently
+%%   \end{outline}
+%% \item Dynamics
+%%   \begin{outline}
+%%     \item Configuration - Edges in graph of communicating components can change at run-time.
+%%       \begin{outline}
+%%       \item Already required in distributed settings
+%%       \item Not addressed in formal models
+%%       \end{outline}
+%%     \item Extension - Nodes in graph of communicating components can change at run-time.
+%%   \end{outline}
+%%   \item Reflection
+%% \end{outline}
+
+%% I/O Automata
+%% \begin{itemize}
+%%   \item Compare with UNITY
+%%   \item Compare with esterel
+%%   \item Compare with pi calculus
+%%   \item Compare with Ptolemy
+%% \end{itemize}

### 2011-09-03 examples/unidirectional_ring_leader_election.hpp

@@ -9,6 +9,28 @@ class unidirectional_ring_leader_election :
   public ioa::automaton
 {
 private:
+  ioa::handle_manager<unidirectional_ring_leader_election> m_self;
+
+public:
+  unidirectional_ring_leader_election (const size_t N) :
+    m_self (ioa::get_aid ())
+  { 
+    std::vector<ioa::automaton_manager<T>*> T_managers;
+    std::vector<ioa::automaton_manager<channel_automaton<M> >*> channel_automaton_managers;
+    
+    for (size_t i = 0; i < N; ++i) {
+      T_managers.push_back (ioa::make_automaton_manager (this, ioa::make_generator<T> (i)));
+      channel_automaton_managers.push_back (ioa::make_automaton_manager (this, ioa::make_generator<channel_automaton<M> > ()));
+    }
+    
+    for (size_t i = 0; i < N; ++i) {
+      make_binding_manager (this, T_managers[i], &T::send, channel_automaton_managers[i], &channel_automaton<M>::send);
+      make_binding_manager (this, channel_automaton_managers[i], &channel_automaton<M>::receive, T_managers[(i + 1) % N], &T::receive);
+      make_binding_manager (this, T_managers[i], &T::leader, &m_self, &unidirectional_ring_leader_election::leader, i);
+    }
+  }
+
+private:
   void leader_effect (size_t i) {
     std::cout << i << " is the leader." << std::endl;
   }
@@ -16,29 +38,6 @@ private:
   void leader_schedule (size_t) const { }
 
   UV_P_INPUT (unidirectional_ring_leader_election, leader, size_t);
-
-  ioa::handle_manager<unidirectional_ring_leader_election> self;
-  std::vector<ioa::automaton_handle_interface<T>*> T_helpers;
-  std::vector<ioa::automaton_handle_interface<channel_automaton<M> >*> channel_automaton_managers;
-
-public:
-
-  unidirectional_ring_leader_election (const size_t N) :
-    self (ioa::get_aid ())
-  { 
-    for (size_t i = 0; i < N; ++i) {
-      T_helpers.push_back (new ioa::automaton_manager<T> (this, ioa::make_generator<T> (i)));
-      channel_automaton_managers.push_back (new ioa::automaton_manager<channel_automaton<M> > (this, ioa::make_generator<channel_automaton<M> > ()));
-    }
-    
-    for (size_t i = 0; i < N; ++i) {
[…]

### 2011-09-03 papers/paper1/asynch_lcr.cpp

@@ -1,0 +1,15 @@
+#include <ioa/global_fifo_scheduler.hpp>
+
+typedef unidirectional_ring_leader_election<asynch_lcr_automaton, UID_t> T;
+
+int main (int argc,
+	  char* argv[]) {
+  size_t node_count;
+  /* Initialize node_count. */
+  srand (time (0));
+
+  ioa::global_fifo_scheduler sched;
+  ioa::run (sched, ioa::make_generator<T> (node_count));
+
+  return 0; 
+}

### 2011-09-03 papers/paper1/asynch_lcr_automaton.hpp

@@ -1,0 +1,101 @@
+/*
+  AsyncLCR Automaton
+  Distributed Algorithms, p. 477.
+*/
+
+#include <ioa/ioa.hpp>
+#include <queue>
+#include <utility>
+#include <cstddef>
+
+typedef std::pair<int, size_t> UID_t;
+
+class asynch_lcr_automaton :
+  public ioa::automaton
+{
+private:
+  typedef enum {
+    UNKNOWN,
+    CHOSEN,
+    REPORTED
+  } status_t;
+
+  UID_t m_u;
+  std::queue<UID_t> m_send;
+  status_t m_status;
+
+public:
+  asynch_lcr_automaton (const size_t i) :
+    m_u (rand (), i),
+    m_status (UNKNOWN)
+  {
+    m_send.push (m_u);
+    schedule ();
+  }
+
+private:
+  bool send_precondition () const {
+    return !m_send.empty () &&
+      ioa::binding_count (&asynch_lcr_automaton::send) != 0;
+  }
+
+  UID_t send_effect () {
+    UID_t retval = m_send.front ();
+    m_send.pop ();
+    return retval;
+  }
+
+  void send_schedule () const {
+    schedule ();
+  }
+
+public:
+  V_UP_OUTPUT (asynch_lcr_automaton, send, UID_t);
+
+private:
+  void receive_effect (const UID_t& v) {
+    if (v > m_u) {
+      m_send.push (v);
+    }
+    else if (v == m_u) {
+      m_status = CHOSEN;
+    }
+    else {
+      // Do nothing.
+    }
+  }
+
+  void receive_schedule () const {
+    schedule ();
+  }
+
+public:
+  V_UP_INPUT (asynch_lcr_automaton, receive, UID_t);
+
+private:
+  bool leader_precondition () const {
+    return m_status == CHOSEN &&
+      ioa::binding_count (&asynch_lcr_automaton::leader) != 0;
+  }
+
+  void leader_effect () {
+    m_status = REPORTED;
+  }
+
+  void leader_schedule () const {
+    schedule ();
+  }
+
+public:
+  UV_UP_OUTPUT (asynch_lcr_automaton, leader);
+
+private:
+  void schedule () const {
+    if (send_precondition ()) {
+      ioa::schedule (&asynch_lcr_automaton::send);
+    }
+    if (leader_precondition ()) {
+      ioa::schedule (&asynch_lcr_automaton::leader);
+    }
+  }
+};

### 2011-09-03 papers/paper1/channel_automaton.hpp

@@ -1,6 +1,3 @@
-#ifndef __channel_automaton_hpp__
-#define __channel_automaton_hpp__
-
 /*
   Channel I/O Automaton
   Distributed Algorithms, p. 204.
@@ -29,7 +26,8 @@ public:
 
 private:
   bool receive_precondition () const {
-    return !m_queue.empty () && ioa::binding_count (&channel_automaton::receive) != 0;
+    return !m_queue.empty () &&
+      ioa::binding_count (&channel_automaton::receive) != 0;
   }
 
   T receive_effect () {
@@ -47,5 +45,3 @@ private:
 public:
   V_UP_OUTPUT (channel_automaton, receive, T);
 };
-
-#endif

### 2011-09-03 papers/paper1/unidirectional_ring_leader_election.hpp

@@ -1,0 +1,39 @@
+template <class T, class M>
+class unidirectional_ring_leader_election :
+  public ioa::automaton
+{
+private:
+  ioa::handle_manager<unidirectional_ring_leader_election> m_self;
+
+public:
+  unidirectional_ring_leader_election (const size_t N) :
+    m_self (ioa::get_aid ())
+  { 
+    std::vector<ioa::automaton_manager<T>*> T_managers;
+    std::vector<ioa::automaton_manager<channel_automaton<M> >*> channel_managers;
+    
+    for (size_t i = 0; i < N; ++i) {
+      T_managers.push_back (
+	  ioa::make_automaton_manager (this, ioa::make_generator<T> (i)));
+      channel_managers.push_back (
+	  ioa::make_automaton_manager (this, ioa::make_generator<channel_automaton<M> > ()));
+    }
+    
+    for (size_t i = 0; i < N; ++i) {
+      make_binding_manager (this,
+			    T_managers[i], &T::send,
+			    channel_managers[i], &channel_automaton<M>::send);
+      make_binding_manager (this,
+			    channel_managers[i], &channel_automaton<M>::receive,
+			    T_managers[(i + 1) % N], &T::receive);
+      make_binding_manager (this,
+			    T_managers[i], &T::leader,
+			    &m_self, &unidirectional_ring_leader_election::leader, i);
+    }
+  }
+
+private:
+  void leader_effect (size_t i) { /* i is the leader. */ }
+  void leader_schedule (size_t) const { }
+  UV_P_INPUT (unidirectional_ring_leader_election, leader, size_t);
+};

### 2011-09-07 lib/tcp_connection_automaton.cpp
Finished tcp_lcr.

@@ -20,6 +20,9 @@ namespace ioa {
     if (receive_precondition ()) {
       ioa::schedule (&tcp_connection_automaton::receive);
     }
+    if (connected_precondition ()) {
+      ioa::schedule (&tcp_connection_automaton::connected);
+    }
     if (error_precondition ()) {
       ioa::schedule (&tcp_connection_automaton::error);
     }

### 2011-09-13 papers/paper1/component_model.tex
Changed tcp_lcr to use asynch_lcr_automaton.

@@
 
-In this section we describe how I/O automata offer a natural realization of components, interfaces, and provide natural support for concurrency and composition.
+In this section we describe how I/O automata offer a natural realization of components and interfaces, and provide natural support for concurrency and composition.
 The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
@@
 An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
-Often, local actions consists of a precondition and an effect.
+Often, a local action consists of a precondition and an effect.
 The precondition is a predicate over the state variables that enables the effect when selected.
 The effect is a function that computes the next state of the automaton from the current state.
-Input actions presented in this style consist solely of an effect.
+An input action presented in this style consists solely of an effect.
 
@@
 Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
-To be general, the scheduler is assumed to be fair meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
+In general the scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
 Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
-Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
+Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions, using the value produced by the output action subject to the precondition.
 The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
@@
 However, it lacks certain features needed to develop real components.
-This subsection introduces these features and lays the foundation for the dynamics discussed in section~\ref{dynamics}.
+This subsection introduces these features and lays the foundation for adding the dynamics discussed in section~\ref{dynamics}.
 
@@
 Component frameworks that support recursive encapsulation allow components to contain other components.
-Recursive encapsulation encodes the idea of ``divide-and-conquer'' and is at the heart of both bottom-up and top-down design strategies.
+Recursive encapsulation is at the heart of both bottom-up and top-down design strategies.
 The contained component is the \emph{child} while the containing component is the \emph{parent}.
@@
 
-Recursive encapsulation means that an I/O automaton may contain and use other I/O automata.
+In ioa++, recursive encapsulation means that an I/O automaton may contain and use other I/O automata.
 Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
@@
 
-Figure\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles, e.g., Processor.
+Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles, e.g., Processor.
 Named automata are labeled by a box in their upper left corner, e.g., C.
@@
 For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
-The system shown in figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value (i.e., a signal).
+The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value (i.e., a signal).
 Second, an input action can be bound to at most one output action.

### 2011-09-13 papers/paper1/design.tex
Changed tcp_lcr to use asynch_lcr_automaton.

@@
 
-An implementation of the model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
+An implementation of the component model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
 The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
@@
 Every action in the set of actions considered by the controller was scheduled by its associated automaton.
-Since automata can be destroyed, an action is the set of actions owned by the controller might belong to an automaton that no longer exists.
+Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
 Our solution is to check if the automaton exists before executing an action.

### 2011-09-13 papers/paper1/programming_model.tex
Finished tcp_lcr.
Changed tcp_lcr to use asynch_lcr_automaton.

@@
-\section{I/O Automata Representation and Programming Model\label{representation}}
+\section{I/O Automata Programming Model\label{representation}}
 
-In this section, we show how I/O automata are represented using our framework and how they can be used to construct programs.
-We develop a simulation for a distributed algorithm to show the basic features of the framework and then show a real implementation of the same algorithm.
-The problem we consider is that of electing a leader in unidirectional ring.
+In this section, we show how I/O automata can be used to construct programs by implementing an algorithm for electing a leader in a unidirectional ring.
+The algorithm we consider is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
+The LCR algorithm assumes that every node has a unique identifier (UID).
+When the protocol begins, each node sends its UID to its successor.
+When a node receives a UID from its predecessor that is greater than its own UID, it forwards the UID to its successor.
+When a node receives its own UID, it elects itself the leader of the ring.
 
-The algorithm we consider is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{} and Chang and Roberts~\cite{}.
-The LCR algorithm assumes that every node has a unique identifier.
-Each node begins by assuming that it is the leader and send its own unique id to its neighbor.
-If a node receives a unique id from its predecessor that is greater that the id that it believes to be the leader, then it changes its belief about the leader to the new id and sends the new id to its successor.
-When a node receives its own id, it elects itself the leader of the ring.
+Automata are classes
+
+Actions (2/3 functions + MACRO)
+
+binding predicate (binding count)
+
+dynamics
 

### 2011-09-13 examples/asynch_lcr.cpp

@@ -21,6 +21,6 @@ int main (int argc,
   srand (time (0));
 
   ioa::global_fifo_scheduler sched;
-  ioa::run (sched, ioa::make_generator<unidirectional_ring_leader_election<asynch_lcr_automaton, UID_t> > (node_count));
+  ioa::run (sched, ioa::make_generator<unidirectional_ring_leader_election<asynch_lcr_automaton<UID_t> > > (node_count));
   return 0; 
 }

### 2011-09-13 examples/asynch_lcr_automaton.hpp

@@ -6,31 +6,25 @@
   Distributed Algorithms, p. 477.
 */
 
-#include "UID.hpp"
 #include <ioa/ioa.hpp>
 #include <queue>
-#include <iostream>
 
+template <typename UID>
 class asynch_lcr_automaton :
   public ioa::automaton
 {
 private:
-  typedef enum {
-    UNKNOWN,
-    CHOSEN,
-    REPORTED
-  } status_t;
-
-  UID_t m_u;
-  std::queue<UID_t> m_send;
-  status_t m_status;
+  UID m_u;
+  std::queue<UID> m_send;
+  bool m_report;
+  bool m_leader;
 
 public:
-  asynch_lcr_automaton (const size_t i) :
-    m_u (rand (), i),
-    m_status (UNKNOWN)
+  asynch_lcr_automaton (const UID& u) :
+    m_u (u),
+    m_report (false),
+    m_leader (false)
   {
-    std::cout << m_u.first << "\t" << m_u.second << std::endl;
     m_send.push (m_u);
     schedule ();
   }
@@ -50,8 +44,8 @@ private:
     return !m_send.empty () && ioa::binding_count (&asynch_lcr_automaton::send) != 0;
   }
 
-  UID_t send_effect () {
-    UID_t retval = m_send.front ();
+  UID send_effect () {
+    UID retval = m_send.front ();
     m_send.pop ();
     return retval;
   }
@@ -61,15 +55,22 @@ private:
   }
 
 public:
-  V_UP_OUTPUT (asynch_lcr_automaton, send, UID_t);
+  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
 
 private:
-  void receive_effect (const UID_t& v) {
+  void receive_effect (const UID& v) {
     if (v > m_u) {
       m_send.push (v);
+      if (m_leader) {
+	m_leader = false;
+	m_report = true;
+      }
     }
     else if (v == m_u) {
-      m_status = CHOSEN;
+      if (!m_leader) {
+	m_leader = true;
+	m_report = true;
+      }
     }
     else {
       // Do nothing.
@@ -81,15 +82,16 @@ private:
   }
 
 public:
-  V_UP_INPUT (asynch_lcr_automaton, receive, UID_t);
+  V_UP_INPUT (asynch_lcr_automaton, receive, UID);
 
 private:
   bool leader_precondition () const {
[…]

### 2011-09-13 examples/peterson_leader.cpp

@@ -21,6 +21,6 @@ int main (int argc,
   srand (time (0));
 
   ioa::global_fifo_scheduler sched;
-  ioa::run (sched, ioa::make_generator<unidirectional_ring_leader_election<peterson_leader_automaton, UID_t> > (node_count));
+  ioa::run (sched, ioa::make_generator<unidirectional_ring_leader_election<peterson_leader_automaton<UID_t> > > (node_count));
   return 0; 
 }

### 2011-09-13 examples/peterson_leader_automaton.hpp

@@ -6,11 +6,10 @@
   Distributed Algorithms, p. 483-484.
 */
 
-#include "UID.hpp"
 #include <ioa/ioa.hpp>
 #include <queue>
-#include <iostream>
 
+template <typename UID>
 class peterson_leader_automaton :
   public ioa::automaton
 {
@@ -28,22 +27,21 @@ private:
 
   mode_t m_mode;
   status_t m_status;
-  UID_t m_u;
-  UID_t m_uid[3];
+  UID m_u;
+  UID m_uid[3];
 
-  std::queue<UID_t> m_send;
-  std::queue<UID_t> m_receive;
+  std::queue<UID> m_send;
+  std::queue<UID> m_receive;
 
 public:
-  peterson_leader_automaton (const size_t i) :
+  peterson_leader_automaton (const UID& u) :
     m_mode (ACTIVE),
     m_status (UNKNOWN),
-    m_u (rand (), i)
+    m_u (u)
   {
-    std::cout << m_u.first << "\t" << m_u.second << std::endl;
     m_uid[0] = m_u;
-    m_uid[1] = std::make_pair (-1, -1);
-    m_uid[2] = std::make_pair (-1, -1);
+    m_uid[1] = UID ();
+    m_uid[2] = UID ();
     m_send.push (m_u);
     schedule ();
   }
@@ -78,8 +76,8 @@ private:
     return !m_send.empty () && ioa::binding_count (&peterson_leader_automaton::send) != 0;
   }
 
-  UID_t send_effect () {
-    UID_t retval = m_send.front ();
+  UID send_effect () {
+    UID retval = m_send.front ();
     m_send.pop ();
     return retval;
   }
@@ -89,10 +87,10 @@ private:
   }
 
 public:
-  V_UP_OUTPUT (peterson_leader_automaton, send, UID_t);
+  V_UP_OUTPUT (peterson_leader_automaton, send, UID);
 
 private:
-  void receive_effect (const UID_t& v) {
+  void receive_effect (const UID& v) {
     m_receive.push (v);
   }
 
@@ -101,15 +99,16 @@ private:
   }
 
 public:
-  V_UP_INPUT (peterson_leader_automaton, receive, UID_t);
+  V_UP_INPUT (peterson_leader_automaton, receive, UID);
 
 private:
   bool leader_precondition () const {
     return m_status == CHOSEN && ioa::binding_count (&peterson_leader_automaton::leader) != 0;
   }
 
[…]

### 2011-09-13 examples/tcp_lcr.cpp

@@ -2,11 +2,73 @@
 #include <ioa/tcp_acceptor_automaton.hpp>
 #include <ioa/tcp_connector_automaton.hpp>
 #include <ioa/global_fifo_scheduler.hpp>
+#include "asynch_lcr_automaton.hpp"
 
 #include <uuid/uuid.h>
 #include <iostream>
 
-#define TIMEOUT 10
+// In seconds.
+#define CONNECTION_TIMEOUT 10
+#define ELECT_TIMEOUT 10
+#define TOKEN_TIMEOUT 1
+
+struct uuid
+{
+  uuid_t id;
+
+  uuid () {
+    uuid_clear (id);
+  }
+
+  uuid (const uuid& other) {
+    uuid_copy (id, other.id);
+  }
+
+  void generate () {
+    uuid_generate (id);
+  }
+
+  bool operator> (const uuid& other) const {
+    return uuid_compare (id, other.id) > 0;
+  }
+
+  bool operator== (const uuid& other) const {
+    return uuid_compare (id, other.id) == 0;
+  }
+
+  bool operator!= (const uuid& other) const {
+    return uuid_compare (id, other.id) != 0;
+  }
+};
+
+enum message_type_t {
+  UNKNOWN,
+  ELECT,
+  TOKEN,
+};
+
+struct message_t
+{
+  message_type_t type;
+  uuid id;
+
+  message_t () :
+    type (UNKNOWN)
+  { }
+
+  message_t (const message_type_t t,
+	     const uuid& i) :
+    type (t),
+    id (i)
+  { }
+};
+
+std::ostream& operator<< (std::ostream& o,
+			  const uuid& id) {
+  char c[37];
+  uuid_unparse (id.id, c);
+  return (o << c);
+}
 
 class tcp_lcr_automaton :
   public ioa::automaton,
@@ -14,6 +76,11 @@ class tcp_lcr_automaton :
 {
 private:
   ioa::handle_manager<tcp_lcr_automaton> m_self;
+  uuid m_u;
+  std::queue<uuid> m_to_lcr;
+  bool m_init_lcr;
+  ioa::time m_last_token;
+  bool m_leader;
   const ioa::inet_address m_recv_address;
   const ioa::inet_address m_send_address;
   ioa::automaton_manager<ioa::tcp_acceptor_automaton>* m_acceptor;
@@ -23,34 +90,8 @@ private:
   ioa::automaton_manager<ioa::tcp_connector_automaton>* m_connector;
[…]

### 2011-09-13 examples/unidirectional_ring_leader_election.hpp

@@ -1,10 +1,11 @@
 #ifndef __unidirectional_ring_leader_election_hpp__
 #define __unidirectional_ring_leader_election_hpp__
 
+#include "UID.hpp"
 #include "channel_automaton.hpp"
 #include <iostream>
 
-template <class T, class M>
+template <class T>
 class unidirectional_ring_leader_election :
   public ioa::automaton
 {
@@ -16,28 +17,32 @@ public:
     m_self (ioa::get_aid ())
   { 
     std::vector<ioa::automaton_manager<T>*> T_managers;
-    std::vector<ioa::automaton_manager<channel_automaton<M> >*> channel_automaton_managers;
+    std::vector<ioa::automaton_manager<channel_automaton<UID_t> >*> channel_automaton_managers;
     
     for (size_t i = 0; i < N; ++i) {
-      T_managers.push_back (ioa::make_automaton_manager (this, ioa::make_generator<T> (i)));
-      channel_automaton_managers.push_back (ioa::make_automaton_manager (this, ioa::make_generator<channel_automaton<M> > ()));
+      UID_t u (rand (), i);
+      std::cout << u.first << "\t" << u.second << std::endl;
+      T_managers.push_back (ioa::make_automaton_manager (this, ioa::make_generator<T> (u)));
+      channel_automaton_managers.push_back (ioa::make_automaton_manager (this, ioa::make_generator<channel_automaton<UID_t> > ()));
     }
     
     for (size_t i = 0; i < N; ++i) {
-      make_binding_manager (this, T_managers[i], &T::send, channel_automaton_managers[i], &channel_automaton<M>::send);
-      make_binding_manager (this, channel_automaton_managers[i], &channel_automaton<M>::receive, T_managers[(i + 1) % N], &T::receive);
+      make_binding_manager (this, T_managers[i], &T::send, channel_automaton_managers[i], &channel_automaton<UID_t>::send);
+      make_binding_manager (this, channel_automaton_managers[i], &channel_automaton<UID_t>::receive, T_managers[(i + 1) % N], &T::receive);
       make_binding_manager (this, T_managers[i], &T::leader, &m_self, &unidirectional_ring_leader_election::leader, i);
     }
   }
 
 private:
-  void leader_effect (size_t i) {
-    std::cout << i << " is the leader." << std::endl;
+  void leader_effect (const bool& v, size_t i) {
+    if (v) {
[…]

### 2011-09-13 papers/paper1/programming_model.tex
ioa::automaton is virtual base.

@@
 
-In this section, we show how I/O automata can be used to construct programs by implementing an algorithm for electing a leader in a unidirectional ring.
+In this section, we show how I/O automata can be used to construct programs.
+
+We motivated our presentation using the classic problem of electing a leader in a unidirectional ring.
+
+\paragraph{Leader election in a unidirectional ring.}
+
+ by implementing an algorithm for electing a leader in a unidirectional ring.
 The algorithm we consider is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.

### 2011-09-13 examples/asynch_bcast_ack_automaton.hpp

@@ -21,7 +21,7 @@ enum message_type_t {
 typedef std::pair<message_type_t, size_t> message_t;
 
 class asynch_bcast_ack_automaton:
-  public ioa::automaton {
+  public virtual ioa::automaton {
 
 private:
 

### 2011-09-13 examples/asynch_bellman_ford_automaton.hpp

@@ -12,7 +12,7 @@
 #include <map>
 
 class asynch_bellman_ford_automaton :
-  public ioa::automaton {
+  public virtual ioa::automaton {
 private:
   typedef std::queue<size_t> distq;
 

### 2011-09-13 examples/asynch_bfs_automaton.hpp

@@ -14,7 +14,7 @@
 #include <queue>
 
 class asynch_bfs_automaton:
-  public ioa::automaton {
+  public virtual ioa::automaton {
 
 private:
 

### 2011-09-13 examples/asynch_lcr_automaton.hpp

@@ -11,7 +11,7 @@
 
 template <typename UID>
 class asynch_lcr_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   UID m_u;

### 2011-09-13 examples/asynch_spanning_tree_automaton.hpp

@@ -16,7 +16,7 @@ typedef enum {
 } search_t;
 
 class asynch_spanning_tree_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 
 private:

### 2011-09-13 examples/bidirectional_network.hpp

@@ -11,7 +11,7 @@
 
 template <class T, typename M>
 class bidirectional_network :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   std::vector<ioa::automaton_handle_interface<T>*> T_helpers;

### 2011-09-13 examples/bidirectional_spanning_tree.hpp

@@ -11,7 +11,7 @@
 
 template <class T, class M>
 class bidirectional_spanning_tree :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   void parent_effect(const size_t& j, size_t i) {

### 2011-09-13 examples/channel_automaton.hpp

@@ -11,7 +11,7 @@
 
 template <class T>
 class channel_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   std::queue<T> m_queue;

### 2011-09-13 examples/clock.cpp

@@ -4,7 +4,7 @@
 #include <ioa/global_fifo_scheduler.hpp>
 
 class trigger :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
 
@@ -34,7 +34,7 @@ public:
 */
 
 class clock_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   int m_counter;
@@ -106,7 +106,7 @@ public:
 
 
 class display :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   void clock_effect (int const & t) {
@@ -122,7 +122,7 @@ public:
 
 
 class composer :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 public:
   composer ()

### 2011-09-13 examples/echo_client.cpp

@@ -8,7 +8,7 @@
 #include <sys/ioctl.h>
 
 class stdin_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   enum state_t {
@@ -146,7 +146,7 @@ public:
 };
 
 class echo_client_automaton :
-  public ioa::automaton,
+  public virtual ioa::automaton,
   private ioa::observer
 {
 private:

### 2011-09-13 examples/echo_server.cpp

@@ -6,7 +6,7 @@
 #include <iostream>
 
 class client_handler_automaton :
-  public ioa::automaton,
+  public virtual ioa::automaton,
   private ioa::observer
 {
 private:
@@ -196,7 +196,7 @@ private:
 };
 
 class echo_server_automaton :
-  public ioa::automaton,
+  public virtual ioa::automaton,
   private ioa::observer
 {
 private:

### 2011-09-13 examples/peterson_leader_automaton.hpp

@@ -11,7 +11,7 @@
 
 template <typename UID>
 class peterson_leader_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   typedef enum {

### 2011-09-13 examples/slow_count.cpp

@@ -4,7 +4,7 @@
 #include <iostream>
 
 class count_to_ten :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   int m_count;

### 2011-09-13 examples/tcp_lcr.cpp

@@ -71,7 +71,7 @@ std::ostream& operator<< (std::ostream& o,
 }
 
 class tcp_lcr_automaton :
-  public ioa::automaton,
+  public virtual ioa::automaton,
   private ioa::observer
 {
 private:

### 2011-09-13 examples/udp_receiver.cpp

@@ -7,7 +7,7 @@
 #include <iostream>
 
 class udp_receiver :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   ioa::handle_manager<udp_receiver> m_self;

### 2011-09-13 examples/udp_sender.cpp

@@ -6,7 +6,7 @@
 #include <iostream>
 
 class udp_sender :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   enum state_t {

### 2011-09-13 examples/unidirectional_ring_leader_election.hpp

@@ -7,7 +7,7 @@
 
 template <class T>
 class unidirectional_ring_leader_election :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   ioa::handle_manager<unidirectional_ring_leader_election> m_self;

### 2011-09-13 examples/weighted_bidirectional_network.hpp

@@ -13,7 +13,7 @@
 
 template <class T, typename M>
 class weighted_bidirectional_network :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
 

### 2011-09-13 include/ioa/automaton.hpp

@@ -56,7 +56,8 @@ namespace ioa {
     virtual void unbound (const unbound_t result) = 0;
   };
 
-  class automaton {
+  class automaton
+  {
   private:
     // Invariant:  m_create_send INTERSECT m_create_recv INTERSECT m_destroy_send = empty set
     // Basically, a helper is in one place or another.

### 2011-09-13 include/ioa/tcp_acceptor_automaton.hpp

@@ -7,7 +7,7 @@
 namespace ioa {
 
   class tcp_acceptor_automaton :
-    public automaton
+    public virtual automaton
   {
   private:    
     enum state_t {

### 2011-09-13 include/ioa/tcp_connection_automaton.hpp

@@ -8,7 +8,7 @@
 namespace ioa {
 
   class tcp_connection_automaton :
-    public automaton
+    public virtual automaton
   {
   private:
     enum send_state_t {
@@ -107,7 +107,7 @@ namespace ioa {
   };
 
   class connection_init_automaton :
-    public automaton,
+    public virtual automaton,
     private observer
   {
   private:

### 2011-09-13 include/ioa/tcp_connector_automaton.hpp

@@ -6,7 +6,7 @@
 namespace ioa {
   
   class tcp_connector_automaton :
-    public automaton
+    public virtual automaton
   {
   private:
     handle_manager<tcp_connector_automaton> m_self;

### 2011-09-13 include/ioa/udp_receiver_automaton.hpp

@@ -9,7 +9,7 @@
 namespace ioa {
 
   class udp_receiver_automaton :
-    public automaton
+    public virtual automaton
   {
   public:
     struct receive_val {

### 2011-09-13 include/ioa/udp_sender_automaton.hpp

@@ -9,7 +9,7 @@
 namespace ioa {
 
   class udp_sender_automaton :
-    public automaton,
+    public virtual automaton,
     private observer
   {
   public:

### 2011-09-13 papers/paper1/asynch_lcr.cpp

@@ -1,15 +1,0 @@
-#include <ioa/global_fifo_scheduler.hpp>
-
-typedef unidirectional_ring_leader_election<asynch_lcr_automaton, UID_t> T;
-
-int main (int argc,
-	  char* argv[]) {
-  size_t node_count;
-  /* Initialize node_count. */
-  srand (time (0));
-
-  ioa::global_fifo_scheduler sched;
-  ioa::run (sched, ioa::make_generator<T> (node_count));
-
-  return 0; 
-}

### 2011-09-13 papers/paper1/asynch_lcr_automaton.hpp

@@ -1,101 +1,0 @@
-/*
-  AsyncLCR Automaton
-  Distributed Algorithms, p. 477.
-*/
-
-#include <ioa/ioa.hpp>
-#include <queue>
-#include <utility>
-#include <cstddef>
-
-typedef std::pair<int, size_t> UID_t;
-
-class asynch_lcr_automaton :
-  public ioa::automaton
-{
-private:
-  typedef enum {
-    UNKNOWN,
-    CHOSEN,
-    REPORTED
-  } status_t;
-
-  UID_t m_u;
-  std::queue<UID_t> m_send;
-  status_t m_status;
-
-public:
-  asynch_lcr_automaton (const size_t i) :
-    m_u (rand (), i),
-    m_status (UNKNOWN)
-  {
-    m_send.push (m_u);
-    schedule ();
-  }
-
-private:
-  bool send_precondition () const {
-    return !m_send.empty () &&
-      ioa::binding_count (&asynch_lcr_automaton::send) != 0;
-  }
-
-  UID_t send_effect () {
-    UID_t retval = m_send.front ();
-    m_send.pop ();
-    return retval;
-  }
-
-  void send_schedule () const {
-    schedule ();
-  }
-
-public:
-  V_UP_OUTPUT (asynch_lcr_automaton, send, UID_t);
-
-private:
-  void receive_effect (const UID_t& v) {
-    if (v > m_u) {
-      m_send.push (v);
-    }
-    else if (v == m_u) {
-      m_status = CHOSEN;
-    }
-    else {
-      // Do nothing.
-    }
-  }
-
-  void receive_schedule () const {
-    schedule ();
-  }
-
-public:
-  V_UP_INPUT (asynch_lcr_automaton, receive, UID_t);
-
-private:
-  bool leader_precondition () const {
-    return m_status == CHOSEN &&
-      ioa::binding_count (&asynch_lcr_automaton::leader) != 0;
-  }
-
-  void leader_effect () {
-    m_status = REPORTED;
-  }
-
-  void leader_schedule () const {
-    schedule ();
-  }
-
-public:
-  UV_UP_OUTPUT (asynch_lcr_automaton, leader);
-
-private:
-  void schedule () const {
-    if (send_precondition ()) {
-      ioa::schedule (&asynch_lcr_automaton::send);
-    }
-    if (leader_precondition ()) {
-      ioa::schedule (&asynch_lcr_automaton::leader);
-    }
-  }
-};

### 2011-09-13 papers/paper1/channel_automaton.hpp

@@ -1,47 +1,0 @@
-/*
-  Channel I/O Automaton
-  Distributed Algorithms, p. 204.
-*/
-
-#include <queue>
-#include <ioa/ioa.hpp>
-
-template <class T>
-class channel_automaton :
-  public ioa::automaton
-{
-private:
-  std::queue<T> m_queue;
-
-  void send_effect (const T& t) {
-    m_queue.push (t);
-  }
-  
-  void send_schedule () const {
-    receive_schedule ();
-  }
-
-public:
-  V_UP_INPUT (channel_automaton, send, T);
-
-private:
-  bool receive_precondition () const {
-    return !m_queue.empty () &&
-      ioa::binding_count (&channel_automaton::receive) != 0;
-  }
-
-  T receive_effect () {
-    T retval =  m_queue.front ();
-    m_queue.pop ();
-    return retval;
-  }
-
-  void receive_schedule () const {
-    if (receive_precondition ()) {
-      ioa::schedule (&channel_automaton::receive);
-    }
-  }
-
-public:
-  V_UP_OUTPUT (channel_automaton, receive, T);
-};

### 2011-09-13 papers/paper1/unidirectional_ring_leader_election.hpp

@@ -1,39 +1,0 @@
-template <class T, class M>
-class unidirectional_ring_leader_election :
-  public ioa::automaton
-{
-private:
-  ioa::handle_manager<unidirectional_ring_leader_election> m_self;
-
-public:
-  unidirectional_ring_leader_election (const size_t N) :
-    m_self (ioa::get_aid ())
-  { 
-    std::vector<ioa::automaton_manager<T>*> T_managers;
-    std::vector<ioa::automaton_manager<channel_automaton<M> >*> channel_managers;
-    
-    for (size_t i = 0; i < N; ++i) {
-      T_managers.push_back (
-	  ioa::make_automaton_manager (this, ioa::make_generator<T> (i)));
-      channel_managers.push_back (
-	  ioa::make_automaton_manager (this, ioa::make_generator<channel_automaton<M> > ()));
-    }
-    
-    for (size_t i = 0; i < N; ++i) {
-      make_binding_manager (this,
-			    T_managers[i], &T::send,
-			    channel_managers[i], &channel_automaton<M>::send);
-      make_binding_manager (this,
-			    channel_managers[i], &channel_automaton<M>::receive,
-			    T_managers[(i + 1) % N], &T::receive);
-      make_binding_manager (this,
-			    T_managers[i], &T::leader,
-			    &m_self, &unidirectional_ring_leader_election::leader, i);
-    }
-  }
-
-private:
-  void leader_effect (size_t i) { /* i is the leader. */ }
-  void leader_schedule (size_t) const { }
-  UV_P_INPUT (unidirectional_ring_leader_election, leader, size_t);
-};

### 2011-09-13 tutorial/count_to_ten_automaton.cpp

@@ -5,7 +5,7 @@
 #include <iostream>
 
 class count_to_ten_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   int m_count;

### 2011-09-13 tutorial/null_automaton.cpp

@@ -2,7 +2,7 @@
 #include <ioa/global_fifo_scheduler.hpp>
 
 class null_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 { };
 
 int main () {

### 2011-09-13 tutorial/producer_consumer.cpp

@@ -4,7 +4,7 @@
 #include <iostream>
 
 class producer_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   int m_count;
@@ -37,7 +37,7 @@ public:
 };
 
 class consumer_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   void consume_effect (const int& val) {
@@ -51,7 +51,7 @@ public:
 };
 
 class producer_consumer_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 public:
   producer_consumer_automaton () {

### 2011-09-13 tutorial/producer_consumer2.cpp

@@ -4,7 +4,7 @@
 #include <iostream>
 
 class producer_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   int m_count;
@@ -36,7 +36,7 @@ public:
 };
 
 class consumer_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   void consume_effect (const int& val) {
@@ -50,7 +50,7 @@ public:
 };
 
 class producer_consumer_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 public:
   producer_consumer_automaton () {

### 2011-09-13 tutorial/two_counters.cpp

@@ -4,7 +4,7 @@
 #include <iostream>
 
 class count_to_ten_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 private:
   int m_count;
@@ -38,7 +38,7 @@ private:
 };
 
 class two_counter_automaton :
-  public ioa::automaton
+  public virtual ioa::automaton
 {
 public:
   two_counter_automaton () {

### 2011-09-14 papers/paper1/programming_model.tex
Second revision of section 4.

--- 2011-09-13 papers/paper1/programming_model.tex
@@
 
-In this section, we show how I/O automata can be used to construct programs by implementing an algorithm for electing a leader in a unidirectional ring.
-The algorithm we consider is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
+In this section, we show how I/O automata can be used to construct programs.
+To motivate the presentation, we develop an automaton for electing a leader in a unidirectional ring using the LCR algorithm.
+
+\subsection{I/O Automata Representation}
+
+\paragraph{Leader election in a unidirectional ring.}
+Nodes in a distributed system are arranged in a ring so that two consecutive nodes are connected by a channel.
+The goal is to designate one of the nodes as the leader of the ring.
+The solution we adopt is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
 The LCR algorithm assumes that every node has a unique identifier (UID).
@@
 When a node receives its own UID, it elects itself the leader of the ring.
+Our implementation of the asynchronous LCR automaton is located in \verb+examples/asynch_lcr_automaton.hpp+ of the ioa++ package.
 
-Automata are classes
+\paragraph{Declaring an automaton.}
+All automata inherit from the common base class \verb+ioa::automaton+ which implements the system action interface.
+The following code declares the \verb+asynch_lcr_automaton+:
+\begin{lstlisting}
+template <typename UID>
+class asynch_lcr_automaton :
+  public virtual ioa::automaton
+{
+  ...
+};
+\end{lstlisting}
+Static composition can be accomplished by inheriting from, i.e., extending, other automata.
+A problem arises when an automaton extends more than one automaton since the common \verb+ioa::automaton+ base class will be duplicated for each inheritance path.
+The solution is to make the \verb+ioa::automaton+ base class a virtual base class which eliminates the duplicates and resolves all references appropriately.
+The \verb+asynch_lcr_automaton+ is declared as a template where the type \verb+UID+ is the type used for unique identifiers.
+This makes the \verb+asynch_lcr_automaton+ general since it can use any type that models the unique identifier concept.
 
-Actions (2/3 functions + MACRO)
+\paragraph{Declaring and initializing state variables.}
+The state variables of an automaton are expressed as member variables.
+The \verb+asynch_lcr_automaton+ has four state variables containing the unique id of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
+\begin{lstlisting}
+private:
+  const UID m_u;
+  std::queue<UID> m_send;
+  bool m_report;
+  bool m_leader;
+\end{lstlisting}
+Notice that the member variables are declared \verb+private+ which is in keeping with the notion that the state of each automaton is independent.
+State variables are initialized using a constructor.
+For the \verb+asynch_lcr_automaton+, the id is initialized, the send queue contains the id of the automaton, the automaton does not need to report its leader status, and the automaton is not the leader:
+\begin{lstlisting}
+public:
+  asynch_lcr_automaton (const UID& u) :
+    m_u (u),
+    m_report (false),
+    m_leader (false)
+  {
+    m_send.push (m_u);
+    schedule ();
+  }
+\end{lstlisting}
+The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function \verb+schedule+.
+We will return to scheduling later in the discussion.
 
-binding predicate (binding count)
+\paragraph{Declaring/defining actions.}
+Actions are member variables that model the action concept.
+An action has traits that indicate the type of action (input, output, internal), if the action is valued and the value type, and if the action is parameterized and the parameter type.
+Actions also contain methods for preconditions, effects, and scheduling.
+The traits are used to check that input actions are bound to output actions with the same value status and type and also control how the methods are executed.
 
-dynamics
+To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
+Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
+The macros assume the functions are named \verb+action_name_precondition+, \verb+action_name_effect+, and \verb+action_name_schedule+.
+For example, the send action of the \verb+asynch_lcr_automaton+ is declared:
+\begin{lstlisting}
+private:
+  bool send_precondition () const { ... }
+  UID send_effect () { ... }
+  void send_schedule () const { ... }
+public:
+  V_UP_OUTPUT (asynch_lcr_automaton, send, UID);
+\end{lstlisting}
+Input actions are declared similarly save the absence of a precondition.
+The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared \verb+const+.
+The macros for actions that are subject to binding by other automata belong in a \verb+public+ section.
 
-\subsection{The AsynchLCR Automaton}
+\paragraph{Scheduling.}
+Automata subject actions to the scheduler using the \verb+ioa::schedule+ function.
+A common pattern is to schedule all local actions using one function where each action is guarded by its precondition.
+For the \verb+asynch_lcr_automaton+, the scheduling function is:
+\begin{lstlisting}
+  void schedule () const {
+    if (send_precondition ()) {
+      ioa::schedule (&asynch_lcr_automaton::send);
+    }
+    if (leader_precondition ()) {
+      ioa::schedule (&asynch_lcr_automaton::leader);
+    }
+  }
+\end{lstlisting}
+The various scheduling functions in the \verb+asynch_lcr_automaton+ and the constructor all dispatch to this function.
 
-Appendix~\ref{asynch_lcr} shows the code for the asynchronous LCR automaton.
+\subsection{Dynamics}
 
-All automaton inherit from the \verb+ioa::automaton+ base class which implements the system call interface expected by the system automaton.
-The \verb+asynch_lcr_automaton+ inherits from the \verb+ioa::automaton+ base class (line 13-14).
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{tcp_lcr_automaton}
+\caption{TCP LCR automaton.}
+\label{tcp_lcr_automaton}
+\end{figure}
 
-The state variables of the automaton are captured as members variables.
-The \verb+m_u+ variable stores the id of node that this id believes to be the leader (line 23).
-The \verb+m_send+ variable is a queue that stores ids that will be sent to the next node in the ring (line 24).
-The \verb+m+status+ variable is used by the leader to report its election (line 25).
+\paragraph{Leader election in a ring using TCP.}
+The \verb+asynch_lcr_automaton+ is a reusable component designed to be used in the context of a larger system.
+The \verb+tcp_lcr_automaton+ listed in \verb+examples/tcp_lcr.cpp+ elects a leader in a dynamic ring of processes.
+The structure of the \verb+tcp_lcr_automaton+ is given in Figure~\ref{tcp_lcr_automaton}.
+The \verb+tcp_lcr_automaton+ is composed with a \verb+tcp_acceptor_automaton+ to accept a connection from its predecessor, a \verb+tcp_connector_automaton+ to create a connection to its successor, and two \verb+tcp_connection_automaton+ representing the predecessor and successor, and a \verb+asynch_lcr_automaton+ to realize the leader election algorithm.
+The leader of the ring periodically sends a token which is forwarded by all nodes in the ring.
+If a node has not received a token for a specified duration of time, it calls for an election.
+The \verb+asynch_lcr_automaton+ of Figure~\ref{tcp_lcr_automaton} can be substituted with a different automaton to realize a different leader election protocol or a different protocol altogether.
 
-The constructor allows for initialization of the state variables and bootstrapping the scheduler.
-The unique id of the automaton is a random number with ties broken by the index of the automaton.
-Every node assumes that they are the leader initially \verb+i+ (line 29).
-All nodes start in the unknown state (line 30).
-Each node starts by sending their id to their neighbor (line 32).
-The automaton bootstraps the scheduler after its state variable are initialized (line 33).
+\paragraph{Binding predicates.}
+The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified actions.
+The value returned by \verb+ioa::binding_count+ is zero for internal actions, one or zero for input actions, and non-negative for output actions.
+Most often, \verb+ioa::binding_count+ is used in the precondition of outputs to ensure that the value generated by the output will be received.
+This type of usage can be seen in the send action of the \verb+tcp_lcr_automaton+:
+\begin{lstlisting}
+  bool send_precondition (
+    ioa::automaton_manager<ioa::tcp_connection_automaton>* s) const
+ {
+    return s == m_successor &&
+           m_successor_connected &&
+           m_clear_to_send && !m_send.empty () &&
+           ioa::binding_count (&tcp_lcr_automaton::send, s) != 0;
+  }
+\end{lstlisting}
 
-Actions consist of member variables that implement the Action concept.
-An easiest way to declare these variables is to use a macro for the appropriate action type.
-The \verb+asynch_lcr_automaton+ contains a \verb+send+ output action (line 53), a \verb+receive+ input action (line 73), and a \verb+leader+ output action (line 90).
-The macros assume that actions are encoded using two or three functions.
-Local actions require a precondition named \verb+action_name_precondition+ that returns true if the action is enabled (line 37, 76).
-Notice that the preconditions are declared \verb+const+ since they should not change the state of the automaton.
-All actions require an effect named \verb+action_name_effect+ that encodes the effect of the action (line 42, 56, 81).
-All actions also require a scheduling function named \verb+action_name_schedule+ that allows the automaton to schedule actions (line 48, 68, 85).
-Notice that the schedule functions are also declare \verb+const+ for the same reason as the preconditions.
+\paragraph{Dynamic composition.}
+Dynamic composition is accomplished by creating child automata and binding to them.
+The \verb+tcp_lcr_automaton+ creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
+\begin{lstlisting}
+    ioa::automaton_manager<asynch_lcr_automaton<uuid> >* lcr =
+       ioa::make_automaton_manager (this,
+         ioa::make_generator<asynch_lcr_automaton<uuid> > (m_u));
+    ioa::make_binding_manager (this,
+			       lcr, &asynch_lcr_automaton<uuid>::send,
+			       &m_self, &tcp_lcr_automaton::send_lcr);
+    ioa::make_binding_manager (this,
+			       &m_self, &tcp_lcr_automaton::receive_lcr,
+			       lcr, &asynch_lcr_automaton<uuid>::receive);
+    ioa::make_binding_manager (this,
+			       lcr, &asynch_lcr_automaton<uuid>::leader,
+			       &m_self, &tcp_lcr_automaton::leader_lcr);
+    ioa::make_binding_manager (this,
+			       &m_self, &tcp_lcr_automaton::init_lcr,
+			       lcr, &asynch_lcr_automaton<uuid>::init);
+\end{lstlisting}
+Children automata are managed by \verb+ioa::automaton_manager+ objects which can be created via the \verb+ioa::make_automaton_manager+ function.
+An automaton manager object requires an \verb+ioa::automaton+ object for its system action implementation and a generator for allocating the child automaton.
+Bindings are managed by \verb+ioa::binding_manager+ objects which can created via the \verb+ioa::make_binding_manager+ function.
+A binding manager objects requires an \verb+ioa::automaton+ object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
+The \verb+m_self+ object of the example is a \verb+ioa::handle_manager+ which can be used to bind to automata that already exist.
+From the example, the \verb+tcp_lcr_automaton+ binds the \verb+asynch_lcr_automaton+ to itself.
 
-In the I/O automata model, the send precondition only requires that the send queue not be empty (line 38).
-However, if the send action executes before it is bound, the id at the front of the send queue will be lost.
-Consequently, we add a binding predicate that ensures that send output is bound to at least one input (line 39).
-The \verb+ioa::binding_count+ function returns the number of actions bound to the given action.
-This is non-negative for output actions, 0 or 1 for input actions, and 0 for internal actions.
-
-The send action is a valued unparameterized (v-up) output.
-The value produced by send has the type \verb+UID_t+ (line 42).
-Notice that this agrees with the type in the action declaration (line 53).
-Similarly, the receive action is a valued unparameterized (v-up) input action that accepts a value of type \verb+UID_t+ (line 56, 73).
-Notice that input action effects take a constant reference.
-
-The send action removes and returns the id at the front of the queue (line 42-46).
-The receive action updates the current leader sending it to nodes successor and confirms an election (line 56-66).
-Once a node is chosen as the leader, it reports it using the leader action (line 81-83).
-
-After every action is executed, the associated scheduling function is called.
-All scheduling calls in the \verb+asynch_lcr_automaton+ dispatch to the same member function (line 93-100).
-The \verb+asynch_lcr_automaton+ uses the strategy outlined in section~\ref{design} and checks the precondition of each action before scheduling it.
-
-\subsection{AsynchLCR Simulation}
-
-For a simulation of the AsynchLCR algorithm, we connect each node using a channel automaton.
-The channel automaton is introduced by Lynch in Chapter 8 of ~\cite{lynch1996distributed} and listed in appendix~\ref{channel}.
-The channel automaton takes a template parameter for the messages transported by the channel.
-
-Appendix~\ref{ring} contains an automaton that creates the ring.
-The template parameter \verb+T+ represents the automaton, i.e., \verb+asynch_lcr_automaton+, and the template parameter \verb+M+ represents the message type for the channel automata, i.e., \verb+UID_t+.
-The size of the ring to simulate is passed as the constructor argument \verb+N+ (line 9).
-
-The constructor for the ring starts by declaring vectors to hold automaton managers for the LCR automata (line 12) and automaton managers for the channels (line 13).
-The loop in (line 15-20) initializes the vectors.
-The \verb+ioa::make_automaton_manager+ function creates automaton managers.
-These require action to an \verb+ioa::automaton+ which is always provided by a \verb+this+ pointer and a generator as described in section~\ref{design}.
-The \verb+ioa::make_generator+ creates a generator.
-Generators take constructor arguments for the automaton they will produce and pass them to the constructor when the automaton is allocated.
-For example, the LCR automata are initialized with an index (line 17).
-
-The loop in (line 22-33) creates the various bindings necessary to create the ring and receive the leader election.
-The \verb+ioa::make_binding_manager+ function creates binding managers.
-This function requires access to an \verb+ioa::automaton+ to realize the system actions.
-The other arguments are the automaton managers and actions for the output action and input action.
-For example, the send action of the LCR automaton is bound to the send action of the successor channel in (line 23-25).
-Similarly, the receive action of the LCR automaton is bound to the receive action of the predecessor channel in (line 26-28).
-
-The ring automaton contains a single unvalued parameterized (uv-p) input action for receiving the leader (line 36-38).
-The parameter \verb+i+ is the index of the LCR automaton.
-The leader actions are bound in (line 29-31).
-Notice that the parameter for the leader input action is set in the call to \verb+ioa::make_binding_manager+.
-
-Binding to an automaton that already exists requires an object that behaves like an automaton manager but refers to an automaton that already exists.
-Objects of the \verb+ioa::handle_manager+ class satisfy this requirement.
-The ring automaton declares a handler manager \verb+m_self+ (line 6) and initializes it with its automaton identifier (line 10).
-Associated with each automaton is a unique identifier called an \emph{aid} which can be retrieved with \verb+ioa::get_aid+.
-
-Appendix~\ref{driver} contains the driver program for the LCR simulation.
-A global FIFO scheduler \verb+sched+ is declared on (line 11).
-The scheduler is started with the \verb+ioa::run+ function (line 12).
-The \verb+ioa::run+ function starts the scheduler using the root automaton returned by the specified generator which in this case is the ring automaton.
-
-\subsection{AsynchLCR Implementation}
+\subsection{Building Real Systems}
 

### 2011-09-14 examples/asynch_lcr_automaton.hpp

@@ -14,7 +14,7 @@ class asynch_lcr_automaton :
   public virtual ioa::automaton
 {
 private:
-  UID m_u;
+  const UID m_u;
   std::queue<UID> m_send;
   bool m_report;
   bool m_leader;
@@ -41,7 +41,8 @@ private:
   }
 
   bool send_precondition () const {
-    return !m_send.empty () && ioa::binding_count (&asynch_lcr_automaton::send) != 0;
+    return !m_send.empty () &&
+      ioa::binding_count (&asynch_lcr_automaton::send) != 0;
   }
 
   UID send_effect () {

### 2011-09-15 papers/paper1/programming_model.tex
Third revision of section 4.

@@
 
-In this section, we show how I/O automata can be used to construct programs.
-To motivate the presentation, we develop an automaton for electing a leader in a unidirectional ring using the LCR algorithm.
+In this section, we show how I/O automata can be used to construct real programs using ioa++.
+To motivate the presentation, we use the classic problem of electing a leader in a unidirectional ring.
+We first show how I/O automata are represented in the C++ language.
+We then discuss the mechanisms available for dynamic composition.
+We conclude with techniques for managing dynamic constellations of automata.
 
@@
 \paragraph{Leader election in a unidirectional ring.}
-Nodes in a distributed system are arranged in a ring so that two consecutive nodes are connected by a channel.
+Nodes in a distributed system are arranged in a ring so that two consecutive nodes are connected by a directed channel.
 The goal is to designate one of the nodes as the leader of the ring.
@@
 When the protocol begins, each node sends its UID to its successor.
-When a node receives a UID from its predecessor that is greater than its own UID, it forwards the UID to its successor.
+When a node receives a UID from its predecessor that is greater than its own, it forwards the UID to its successor.
 When a node receives its own UID, it elects itself the leader of the ring.
@@
 \end{lstlisting}
-Static composition can be accomplished by inheriting from, i.e., extending, other automata.
-A problem arises when an automaton extends more than one automaton since the common \verb+ioa::automaton+ base class will be duplicated for each inheritance path.
-The solution is to make the \verb+ioa::automaton+ base class a virtual base class which eliminates the duplicates and resolves all references appropriately.
-The \verb+asynch_lcr_automaton+ is declared as a template where the type \verb+UID+ is the type used for unique identifiers.
-This makes the \verb+asynch_lcr_automaton+ general since it can use any type that models the unique identifier concept.
+The \verb+asynch_lcr_automaton+ is declared as a template where the \verb+UID+ template parameter must model the unique identifier concept.
+The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
+The choice to declare \verb+ioa::automaton+ as a virtual base class is discussed in the paragraph on static composition.
 
 \paragraph{Declaring and initializing state variables.}
-The state variables of an automaton are expressed as member variables.
-The \verb+asynch_lcr_automaton+ has four state variables containing the unique id of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
+The state variables of an automaton are expressed as member variables of the corresponding class.
+The four state variables of the \verb+asynch_lcr_automaton+ include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
 \begin{lstlisting}
@@
 State variables are initialized using a constructor.
-For the \verb+asynch_lcr_automaton+, the id is initialized, the send queue contains the id of the automaton, the automaton does not need to report its leader status, and the automaton is not the leader:
+For the \verb+asynch_lcr_automaton+, the flags are both false and the UID is initialized and added to the send queue:
 \begin{lstlisting}
@@
 The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function \verb+schedule+.
-We will return to scheduling later in the discussion.
+The \verb+schedule+ member function is discussed in the paragraph on scheduling.
 
@@
 Actions are member variables that model the action concept.
-An action has traits that indicate the type of action (input, output, internal), if the action is valued and the value type, and if the action is parameterized and the parameter type.
-Actions also contain methods for preconditions, effects, and scheduling.
-The traits are used to check that input actions are bound to output actions with the same value status and type and also control how the methods are executed.
+Actions contain methods for preconditions, effects, and scheduling and traits that indicate the type of action (input, output, internal), if the action is valued and the value type, and if the action is parameterized and the parameter type.
+The traits are used to guide the execution of the methods supplying values and parameters as necessary.
+The traits are also used to check that input actions are bound to output actions with the same value status and type.
+Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.
 
@@
 \paragraph{Scheduling.}
-Automata subject actions to the scheduler using the \verb+ioa::schedule+ function.
-A common pattern is to schedule all local actions using one function where each action is guarded by its precondition.
+Automata submit actions to the scheduler using the \verb+ioa::schedule+ function.
+The \verb+ioa::schedule+ function can be called at any time.
+However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
+A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition.
 For the \verb+asynch_lcr_automaton+, the scheduling function is:
@@
 
+\paragraph{Static composition.}
+Static composition can be accomplished by inheriting from, i.e., extending, other automata.
+For example, an automaton wishing to statically compose an \verb+asynch_lcr_automaton+ with other automata could declare it as a base class and provide the necessary code to dispatch to the \verb+asynch_lcr_automaton+ actions.
+A problem arises when an automaton extends more than one automaton since the common \verb+ioa::automaton+ base class will be duplicated for each inheritance path.
+The solution is to make the \verb+ioa::automaton+ base class a virtual base class which eliminates the duplicates and resolves all references appropriately.
+
 \subsection{Dynamics}
@@
 The \verb+asynch_lcr_automaton+ is a reusable component designed to be used in the context of a larger system.
-The \verb+tcp_lcr_automaton+ listed in \verb+examples/tcp_lcr.cpp+ elects a leader in a dynamic ring of processes.
+The \verb+tcp_lcr_automaton+ listed in \verb+examples/tcp_lcr.cpp+ elects a leader in a dynamic ring of processes connected by TCP sockets.
 The structure of the \verb+tcp_lcr_automaton+ is given in Figure~\ref{tcp_lcr_automaton}.
@@
 If a node has not received a token for a specified duration of time, it calls for an election.
-The \verb+asynch_lcr_automaton+ of Figure~\ref{tcp_lcr_automaton} can be substituted with a different automaton to realize a different leader election protocol or a different protocol altogether.
 
 \paragraph{Binding predicates.}
-The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified actions.
-The value returned by \verb+ioa::binding_count+ is zero for internal actions, one or zero for input actions, and non-negative for output actions.
+Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
+The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified action.
+The value returned by \verb+ioa::binding_count+ is zero for internal actions, zero or one for input actions, and non-negative for output actions.
 Most often, \verb+ioa::binding_count+ is used in the precondition of outputs to ensure that the value generated by the output will be received.
@@
   {
-    return s == m_successor &&
-           m_successor_connected &&
-           m_clear_to_send && !m_send.empty () &&
-           ioa::binding_count (&tcp_lcr_automaton::send, s) != 0;
+    return ... && ioa::binding_count (&tcp_lcr_automaton::send, s) != 0;
   }
@@
 \paragraph{Dynamic composition.}
-Dynamic composition is accomplished by creating child automata and binding to them.
+Dynamic composition is accomplished by creating child automata and bindings.
 The \verb+tcp_lcr_automaton+ creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
@@
 A binding manager objects requires an \verb+ioa::automaton+ object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
+The \verb+ioa::automaton_manager+ class and \verb+ioa::binding_manager+ class contain methods for destroying and unbinding respectively.
 The \verb+m_self+ object of the example is a \verb+ioa::handle_manager+ which can be used to bind to automata that already exist.
 From the example, the \verb+tcp_lcr_automaton+ binds the \verb+asynch_lcr_automaton+ to itself.
+The wary reader might notice that the \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+s in the example are dynamically allocated but never recorded and thus never freed.
+Both classes are managed by the \verb+ioa::automaton+ and deallocated at the appropriate time.
 
-\subsection{Building Real Systems}
+\subsection{Managing Dynamic Constellations}
+
+In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the ``nuts and bolts'' presented in the previous section.
+This list is neither exhaustive nor authoritative but rather a distillation of our experiences working with I/O automata.
+
+\paragraph{Dynamics in the TCP LCR automaton.}
+A \verb+tcp_lcr_automaton+ continuously attempts to complete the ring by accepting a connection from its predecessor or connecting to its successor.
+Accepting can fail if the socket used for listening is currently unavailable.
+Connecting can fail if the successor automaton does not exist and is therefore not listening for incoming connections.
+The connections themselves will fail when nodes leave the ring.
+The \verb+tcp_lcr_automaton+ also contains constraints on the order in which automata are created with respect to each other and with respect to which actions are executed.
+For example, the predecessor connection must be created before it is passed to the acceptor via the accept action.
+Similarly, the successor connection must be created before it is passed via a constructor to the connector.
+
+\paragraph{System action events.}
+Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
+Specifically, an automaton must be able to receive events from its \verb+ioa::automaton+ base class.
+This is a straightforward application of the Observer Pattern~\cite{} where the user automaton observes events forwarded by the \verb+ioa::automaton+ base class.
+Conceptually, the system input actions of the \verb+ioa::automaton+ base class are mapped into the user automaton.
+
+\paragraph{Error reporting.}
+The event semantics of I/O automata provide a natural mechanism for reporting errors.
+An automaton that encounters an error can use an output action to signal the error.
+The automaton can continue processing if the error is transient or stop if the error is permanent.
+In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
+Using this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
+The two strategies for handling an error are to either reset or replace the portion of the constellation that encountered the error.
+
+\paragraph{Handling errors by resetting.}
+An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
+Errors need not be viewed as exceptional circumstances but a necessary part of the protocol implemented by the automaton.
+Automaton designers should strive to write automata that are capable of being reset as this is the simplest error handling strategy.
+
+\paragraph{Handling errors by replacing.}
+The non-deterministic and the asynchronous nature of system actions complicates replacing an automaton.
+To replace a child automaton, the parent automaton must destroy the child, create a new child, and bind to the new child.
+For reasons already discussed, the bind must occur after the create.
+This constraint is enforced by the \verb+ioa::binding_manager+ class.
+Additionally, the destroy must occur before the first bind to ensure that all bindings associated with the first child have been dissolved.
+Attempts to bind before the destroy will fail because either the output action or the input action will be unavailable, i.e., still bound to the first child.
+A general solution is to observe the automaton and wait for it to be destroyed.
+
+If all of the bindings of a child automaton are with its parent, then the child can be replaced using parameters.
+The parent automaton associates a parameter, typically the child automaton, with all bindings to the child automaton.
+When the child needs to be replaced, a new child is created and the current parameter is updated to reflect this change.
+The old bindings can be disabled by checking their parameter against the new parameter.
+When using replacement with parameters, there is no need to wait for the child automaton to be destroyed since the parameter causes all bindings to be different.
 

### 2011-09-15 lib/model.cpp

@@ -333,7 +333,7 @@ namespace ioa {
       // Automaton does not exist.
       return -1;
     }
-    
+
     exec (*this, m_system_scheduler);
     return 0;
   }

### 2011-09-15 papers/paper1/component_model.tex
Revision.

@@
 Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
-Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions, using the value produced by the output action subject to the precondition.
+Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions,  subject to the precondition using the value produced by the output action.
 The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
@@
 
-In ioa++, recursive encapsulation means that an I/O automaton may contain and use other I/O automata.
+Applying recursive encapsulation to I/O automata means that an I/O automaton may contain and use other I/O automata.
 Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
@@
 
-Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles, e.g., Processor.
-Named automata are labeled by a box in their upper left corner, e.g., C.
+Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
+The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
+The name of a named automata appears after its type, e.g., Converter C.
 The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
@@
 \paragraph{Parameters.}
-Actions in the I/O automata model can defined using parameters.
-Parameters are useful for situations requiring fan-in or a session.
-For example, consider an automaton with an input actions $in?(s)$ that is to be bound to $N$ different output actions.
-The solution is to parameterize the input action with a parameter that indicates the associated output actions, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
+Actions in the I/O automata model can be defined using parameters.
+Parameters are useful for situations requiring fan-in or session semantics.
+For example, consider an automaton with an input action $in?(s)$ that is to be bound to $N$ different output actions.
+To get around the limitation that an input can only be bound to one output, we augment the input action with a parameter that indicates the associated output action, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
 \emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
@@
 Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
-The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $F.out!(t) \to Z.sample?(t)$ binding.
+The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $(F.out!(t), Z.sample?(t))$ binding.
 The Statistics automaton $T$ calculates the statistics incrementally using the up internal $T.compute$.
@@
 \caption{System actions.
-  An automaton contains output actions for creating, binding, unbinding, and destroying.
+  A base automaton contains output actions for creating, binding, unbinding, and destroying.
   The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
   The system automaton returns a result using the appropriate parameterized output action.
-  The automaton contains input actions to receive the results.}
+  The base automaton contains input actions to receive the results.}
 \label{system_action}
@@
 The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
-Figure~\ref{system_action} shows the relationship between an automaton and the system automaton.
+Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.
 

### 2011-09-15 papers/paper1/design.tex
Revision.

@@
 All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
-The services provided by the base class are indicated by the Automaton layer of Figure~\ref{framework_architecture}.
-The Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
-In most situations, the services provided by the Automaton base class are too low-level to be used directly by a User Automaton.
-An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Automaton base class.
+The services provided by the base class are indicated by the Base Automaton layer of Figure~\ref{framework_architecture}.
+The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
+In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
+An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
 Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
@@
 
-\subsection{Scheduler}
+\subsection{Scheduler\label{scheduling}}
 
@@
 The model allows executing actions concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
-Since concurrent access to the model is allowed, the module implementing the model must thread-safe.
+Since concurrent access to the model is allowed, the module implementing the model must be thread-safe.
 
@@
 The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
-The locks are acquired in a fixed order to prevent deadlock.
+The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
 Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.
@@
 Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
-Consequently, an automaton is given a chance to declare actions to the scheduler after one of its actions is executed, i.e., changes state.
+Consequently, an automaton is given a opportunity to declare actions to the scheduler after one of its actions is executed.
 Since the state and therefore preconditions of all automata are independent, each automaton is responsible for scheduling its own actions.
@@
 A simple technique for accomplishing this is to schedule all enabled actions.
-More efficient schemes are possible if one considers which actions enable each other and which actions have already been scheduled.
+More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
 Not declaring an enabled action violates the fairness of the controller since it will never select an action that may cause a state change.
@@
 While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound; a situation that occurs quite often in practice.
-To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a bind or unbind.
+To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a successful bind or unbind.
 
@@
 This check is performed by the model and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
-Our solution assumes that each automaton is assigned a unique identifier and that the set of actions turns over comparatively faster than the set of identifiers.
+Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)} and that the set of actions in the scheduler turns over comparatively faster than the set of aids.
 
@@
 Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
-This implies the use of either the Proactor or Reactor pattern.
+This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
 For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.
@@
 
-\subsection{System Actions}
+\subsection{System Actions\label{system_action_section}}
 
@@
 Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
-Automaton Managers can be observed using the Observer pattern to determine when and how the creation/destruction process succeeds.
+Automaton Managers can be observed using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
 Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.
@@
 Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.
+
+\subsection{Implementation}
+
+Ioa++ is implemented in C++ and available at \url{http://jrwilson.github.com/ioa/}.

### 2011-09-15 papers/paper1/programming_model.tex
Revision.

@@
 class asynch_lcr_automaton :
-  public virtual ioa::automaton
+  public ioa::automaton
 {
@@
 The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
-The choice to declare \verb+ioa::automaton+ as a virtual base class is discussed in the paragraph on static composition.
 
@@
 Actions are member variables that model the action concept.
-Actions contain methods for preconditions, effects, and scheduling and traits that indicate the type of action (input, output, internal), if the action is valued and the value type, and if the action is parameterized and the parameter type.
-The traits are used to guide the execution of the methods supplying values and parameters as necessary.
+Actions contain methods for preconditions, effects, and scheduling and traits that indicate the actions's type (input, output, internal), value status, value type, parameter status, and parameter type.
+The traits are used to guide the execution of the methods by the model and supply values and parameters as necessary.
 The traits are also used to check that input actions are bound to output actions with the same value status and type.
@@
 However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
-A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition.
+A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
 For the \verb+asynch_lcr_automaton+, the scheduling function is:
@@
 
-\paragraph{Static composition.}
-Static composition can be accomplished by inheriting from, i.e., extending, other automata.
-For example, an automaton wishing to statically compose an \verb+asynch_lcr_automaton+ with other automata could declare it as a base class and provide the necessary code to dispatch to the \verb+asynch_lcr_automaton+ actions.
-A problem arises when an automaton extends more than one automaton since the common \verb+ioa::automaton+ base class will be duplicated for each inheritance path.
-The solution is to make the \verb+ioa::automaton+ base class a virtual base class which eliminates the duplicates and resolves all references appropriately.
-
-\subsection{Dynamics}
+\subsection{Dynamic Composition}
 
@@
 The structure of the \verb+tcp_lcr_automaton+ is given in Figure~\ref{tcp_lcr_automaton}.
-The \verb+tcp_lcr_automaton+ is composed with a \verb+tcp_acceptor_automaton+ to accept a connection from its predecessor, a \verb+tcp_connector_automaton+ to create a connection to its successor, and two \verb+tcp_connection_automaton+ representing the predecessor and successor, and a \verb+asynch_lcr_automaton+ to realize the leader election algorithm.
+The \verb+tcp_lcr_automaton+ is composed with a \verb+tcp_acceptor_automaton+ to accept a connection from its predecessor, a \verb+tcp_connector_automaton+ to create a connection to its successor, two \verb+tcp_connection_automaton+ representing the predecessor and successor, and a \verb+asynch_lcr_automaton+ to realize the leader election algorithm.
 The leader of the ring periodically sends a token which is forwarded by all nodes in the ring.
@@
 \paragraph{Dynamic composition.}
-Dynamic composition is accomplished by creating child automata and bindings.
+Dynamic composition is accomplished by creating child automata and bindings via \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+ objects.
 The \verb+tcp_lcr_automaton+ creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
@@
 The wary reader might notice that the \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+s in the example are dynamically allocated but never recorded and thus never freed.
-Both classes are managed by the \verb+ioa::automaton+ and deallocated at the appropriate time.
+Both classes are managed by the \verb+ioa::automaton+ and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.
 
@@
 Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
-Specifically, an automaton must be able to receive events from its \verb+ioa::automaton+ base class.
-This is a straightforward application of the Observer Pattern~\cite{} where the user automaton observes events forwarded by the \verb+ioa::automaton+ base class.
-Conceptually, the system input actions of the \verb+ioa::automaton+ base class are mapped into the user automaton.
+For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata supply the output action and input action are created before attempting to bind.
+User automata can observe \verb+ioa::automaton_manager+ objects and \verb+ioa::binding_manager+ objects to receive events generated upon the completion of a system action.
 
@@
 \paragraph{Handling errors by replacing.}
-The non-deterministic and the asynchronous nature of system actions complicates replacing an automaton.
+The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
 To replace a child automaton, the parent automaton must destroy the child, create a new child, and bind to the new child.

### 2011-09-16 papers/paper1/case_study.tex
Case study.

\section{Case Study:  The TCP LCR Automaton\label{case_study}}

In this section, we explore (1) if ioa++ offers a convenient representation for I/O automata and (2) and if ioa++ contains the features necessary for building real distributed systems.
Our approach is to (1) express an automaton in ioa++ using an existing listing, (2) simulate a distributed system built using the automaton as debugging step, and (3) build a real distributed system by composing the automaton with automata for network services.
For continuity, we use the asynchronous LCR automaton developed in section~\ref{representation}.

\paragraph{Translating the asynchronous LCR automaton into ioa++.}
The first version of the code for the asynchronous LCR automaton was translated directly from a listing in~\cite{lynch1996distributed}.
The final version contains an additional action to reinitialize the automaton and modifications that allow an automaton that elected itself leader to nullify its election.
The motivation for these modifications will be explained when we discuss its use in a real system.
The listing in~\cite{lynch1996distributed} consists of approximately 24 lines in a formal language that must be translated to a production language.
The raw listing for the final version consists of 121 lines (5x).
Scheduling accounts for 22 of the 121 lines (20\%) and is the main source of code that cannot be traced back to the formal listing.
On average, every local action contributed 6 lines of scheduling code and every input action contributed 3 lines of scheduling code.

We observe that having a representation of I/O automata in a production language that closely matches a formal representation is of great benefit because it allows one to reason about an automaton directly from the source code.
For example, an invariant in the asynchronous LCR automaton is that all of the UIDs in the send queue of automaton $i$ must be greater than or equal to the UID of automaton $i$.
This invariant can be quickly checked against the constructor and four actions of the \verb+asynch_lcr_automaton+.
Upon construction, the send queue contains the UID of the automaton so the invariant holds initially.
The leader and init actions do not modify the send queue so the invariant holds trivially.
The receive action only adds a UID to the send queue if the UID is greater than the automaton's UID.
The send action removes a UID from the send queue so if the invariant is true before, it will be true after.
Reasoning about an automaton directly from the source code allows one to use formal methods to develop and debug systems.

\paragraph{Unit testing with simulation.}
After the asynchronous LCR automaton was implemented, we developed a simulation to test it.
The \verb+examples/asynch_lcr.cpp+ file contains the driver for the simulation.
The simulation creates a unidirectional ring of a specified size consisting of alternating \verb+asynch_lcr_automaton+ and \verb+channel_automaton+.
(The \verb+channel_automaton+ was also a direct translation from~\cite{lynch1996distributed}.)

The simulation showed that the \verb+asynch_lcr_automaton+ was not correct.
The first bug we encountered was a silent failure where the ring always failed to elect a leader.
The problem was that the leader action was never being scheduled so the leader action was added to the scheduling function.
The second bug manifested itself by electing the wrong leader, i.e., one whose UID was not the maximum.
A closer inspection showed that certain UIDs were being sent but never received.
The key insight into solving the problem was to understand that the automata were executing as the ring was being constructed, i.e., we failed to consider dynamic composition.
The solution was to add binding predicates to the outputs of the \verb+asynch_lcr_automaton+ and \verb+channel_automaton+.

\paragraph{From abstraction to real system.}
The next phase was to use the asynchronous LCR automaton to perform leader election in a real unidirectional ring of processes connected by TCP sockets.
The leader of the ring periodically sends a token which is forwarded by all nodes in the ring.
If a node has not received a token for a specified duration of time, it calls for an election.
This is the motivation behind the action that resets the \verb+asynch_lcr_automaton+.
Nodes listen for connections from their predecessor and attempt to connect to their successor.
Nodes can join and leave the ring at any time.
The protocol should eventually succeed for complete rings and restart when a ring becomes incomplete.
These dynamics admit situations where a ring might lose its leader or contain multiple leaders.
A ring containing multiple leaders should stabilize to having a single leader.
This is the motivation behind the \verb+asynch_lcr_automaton+ being able to withdraw its election.
A rare but possible and valid situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.

The \verb+tcp_lcr_automaton+ which is listed in \verb+examples/tcp_lcr.cpp+ (484 lines) of the ioa++ package implements the protocol just described.
The \verb+tcp_lcr_automaton+ is a composition of a number of automata including an \verb+asynch_lcr_automaton+, an \verb+ioa::tcp_acceptor_automaton+, an \verb+ioa::tcp_connector_automaton+, and two \verb+ioa::tcp_connection_automaton+ representing the channel from the predecessor and channel to the successor.
The TCP automata provided by the ioa++ package are ordinary automata implemented using file descriptors and the techniques described in section~\ref{design}.
The \verb+tcp_lcr_automaton+ uses the techniques for managing dynamic constellations described in section~\ref{representation}.

The \verb+tcp_lcr_automaton+ is validation that I/O automata are a good foundation for building reusable concurrent modules.
The key idea is that all of these modules have well-defined interactions when composed while being inherently concurrent.
For example, one could substitute the \verb+asynch_lcr_automaton+ with a different automaton to realize a different leader election protocol or a different protocol altogether.
Alternatively, the \verb+tcp_lcr_automaton+ could be refactored to use a different communication mechanism, e.g., UDP.
One need only focus on the protocol implemented by an automaton when composing it with an existing system as opposed to worrying about interacting threads and race conditions.

%% \begin{figure}
%% \center
%% \includegraphics[width=\textwidth]{tcp_lcr_automaton}
%% \caption{TCP LCR automaton.}
%% \label{tcp_lcr_automaton}
%% \end{figure}

%% \paragraph{Dynamics in the TCP LCR automaton.}
%% A \verb+tcp_lcr_automaton+ continuously attempts to complete the ring by accepting a connection from its predecessor or connecting to its successor.
%% Accepting can fail if the socket used for listening is currently unavailable.
%% Connecting can fail if the successor automaton does not exist and is therefore not listening for incoming connections.
%% The connections themselves will fail when nodes leave the ring.
%% The \verb+tcp_lcr_automaton+ also contains constraints on the order in which automata are created with respect to each other and with respect to which actions are executed.
%% For example, the predecessor connection must be created before it is passed to the acceptor via the accept action.
%% Similarly, the successor connection must be created before it is passed via a constructor to the connector.

%% Conclude with our recommendations for programming, highlight need for exploration.

%% Tips for Writing Programs with I/O Automata
%% 1. Make sure that a parameter exists.
%% 2. Move logic in observe to local actions.
%% 3. Avoid observing an arbitrary number of managers.
%% 4. Do not bind to managers allocated on the stack.
%% 5. Stop when an error is encoutered and report it.
%% 6. Make sure output actions are bound.
%% 7. Make sure local actions are scheduled.
%% 8. Do not pass pointers---use const_shared_ptr.
%% 9. Check preconditions.
%% 10. Check effects.
%% 11. Order clauses of preconditions for efficiency.
%% 12. Automata that should be destroyed at the same time should be created at the same time and vice versa.  (Group using a parent.)

%% \begin{outline}
%% \item State
%%   \begin{outline}
%%   \item There is no shared state in a 
%%   \item Local state only
%%   \item Shared state
%%     \begin{outline}
%%       \item Impossible in distributed systems
%%       \item Dangerous in local systems
%%     \end{outline}
%%   \end{outline}
%% \item Communication
%%   \begin{outline}
%%   \item Atomic asynchronous message passing
%%   \item Network sets size of atom (UDP)
%%   \item Can build reliable streams (TCP)
%%   \item Local equivalent is passing a value
%%   \item Model should lend itself to writing protocols
%%   \end{outline}
%% \item Asynchrony
%%   \begin{outline}
%%     \item Model must have natural support for asynchrony, i.e., event-based
%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
%%   \end{outline}
%% \item Concurrency
%%   \begin{outline}
%%     \item Reason about systems using non-deterministic interleaving of atomic actions
%%     \item Model should admit implementations that execute concurrently
%%   \end{outline}
%% \item Dynamics
%%   \begin{outline}
%%     \item Configuration - Edges in graph of communicating components can change at run-time.
%%       \begin{outline}
%%       \item Already required in distributed settings
%%       \item Not addressed in formal models
%%       \end{outline}
%%     \item Extension - Nodes in graph of communicating components can change at run-time.
%%   \end{outline}
%%   \item Reflection
%% \end{outline}

%% I/O Automata
%% \begin{itemize}
%%   \item Compare with UNITY
%%   \item Compare with esterel
%%   \item Compare with pi calculus
%%   \item Compare with Ptolemy
%% \end{itemize}

### 2011-09-16 papers/paper1/programming_model.tex
Case study.

@@
 We conclude with techniques for managing dynamic constellations of automata.
-
-\subsection{I/O Automata Representation}
 
@@
 Our implementation of the asynchronous LCR automaton is located in \verb+examples/asynch_lcr_automaton.hpp+ of the ioa++ package.
+
+\subsection{I/O Automata Representation}
 
@@
 
-\subsection{Dynamic Composition}
-
-\begin{figure}
-\center
-\includegraphics[width=\textwidth]{tcp_lcr_automaton}
-\caption{TCP LCR automaton.}
-\label{tcp_lcr_automaton}
-\end{figure}
-
-\paragraph{Leader election in a ring using TCP.}
-The \verb+asynch_lcr_automaton+ is a reusable component designed to be used in the context of a larger system.
-The \verb+tcp_lcr_automaton+ listed in \verb+examples/tcp_lcr.cpp+ elects a leader in a dynamic ring of processes connected by TCP sockets.
-The structure of the \verb+tcp_lcr_automaton+ is given in Figure~\ref{tcp_lcr_automaton}.
-The \verb+tcp_lcr_automaton+ is composed with a \verb+tcp_acceptor_automaton+ to accept a connection from its predecessor, a \verb+tcp_connector_automaton+ to create a connection to its successor, two \verb+tcp_connection_automaton+ representing the predecessor and successor, and a \verb+asynch_lcr_automaton+ to realize the leader election algorithm.
-The leader of the ring periodically sends a token which is forwarded by all nodes in the ring.
-If a node has not received a token for a specified duration of time, it calls for an election.
-
-\paragraph{Binding predicates.}
-Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
-The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified action.
-The value returned by \verb+ioa::binding_count+ is zero for internal actions, zero or one for input actions, and non-negative for output actions.
-Most often, \verb+ioa::binding_count+ is used in the precondition of outputs to ensure that the value generated by the output will be received.
-This type of usage can be seen in the send action of the \verb+tcp_lcr_automaton+:
-\begin{lstlisting}
-  bool send_precondition (
-    ioa::automaton_manager<ioa::tcp_connection_automaton>* s) const
-  {
-    return ... && ioa::binding_count (&tcp_lcr_automaton::send, s) != 0;
-  }
-\end{lstlisting}
+\subsection{Dynamics}
 
 \paragraph{Dynamic composition.}
+The \verb+asynch_lcr_automaton+ is a reusable component designed to be composed in the context of a larger system.
 Dynamic composition is accomplished by creating child automata and bindings via \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+ objects.
-The \verb+tcp_lcr_automaton+ creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
+The \verb+tcp_lcr_automaton+ discussed in section~\ref{case_study} creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
 \begin{lstlisting}
@@
 
+\paragraph{Binding predicates.}
+Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
+The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified action.
+The value returned by \verb+ioa::binding_count+ is zero for internal actions, zero or one for input actions, and non-negative for output actions.
+Most often, \verb+ioa::binding_count+ is used in the precondition of outputs to ensure that the value generated by the output will be received.
+This type of usage can be seen in the send action of the \verb+asynch_lcr_automaton+:
+\begin{lstlisting}
+  bool send_precondition () const {
+    return !m_send.empty () &&
+      ioa::binding_count (&asynch_lcr_automaton::send) != 0;
+  }
+\end{lstlisting}
+
 \subsection{Managing Dynamic Constellations}
@@
 
-\paragraph{Dynamics in the TCP LCR automaton.}
-A \verb+tcp_lcr_automaton+ continuously attempts to complete the ring by accepting a connection from its predecessor or connecting to its successor.
-Accepting can fail if the socket used for listening is currently unavailable.
-Connecting can fail if the successor automaton does not exist and is therefore not listening for incoming connections.
-The connections themselves will fail when nodes leave the ring.
-The \verb+tcp_lcr_automaton+ also contains constraints on the order in which automata are created with respect to each other and with respect to which actions are executed.
-For example, the predecessor connection must be created before it is passed to the acceptor via the accept action.
-Similarly, the successor connection must be created before it is passed via a constructor to the connector.
-
 \paragraph{System action events.}
@@
 For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata supply the output action and input action are created before attempting to bind.
-User automata can observe \verb+ioa::automaton_manager+ objects and \verb+ioa::binding_manager+ objects to receive events generated upon the completion of a system action.
+Similarly, an automaton might compose with an automaton that it does not manage and therefore expects the automaton to exist before attempting to bind.
+To enable such behavior, user automata can observe \verb+ioa::automaton_manager+ objects and \verb+ioa::binding_manager+ objects to receive events generated upon the completion of a system action.
 
@@
 When using replacement with parameters, there is no need to wait for the child automaton to be destroyed since the parameter causes all bindings to be different.
-
-%% Conclude with our recommendations for programming, highlight need for exploration.
-
-%% Tips for Writing Programs with I/O Automata
-%% 1. Make sure that a parameter exists.
-%% 2. Move logic in observe to local actions.
-%% 3. Avoid observing an arbitrary number of managers.
-%% 4. Do not bind to managers allocated on the stack.
-%% 5. Stop when an error is encoutered and report it.
-%% 6. Make sure output actions are bound.
-%% 7. Make sure local actions are scheduled.
-%% 8. Do not pass pointers---use const_shared_ptr.
-%% 9. Check preconditions.
-%% 10. Check effects.
-%% 11. Order clauses of preconditions for efficiency.
-%% 12. Automata that should be destroyed at the same time should be created at the same time and vice versa.  (Group using a parent.)
-
-%% \begin{outline}
-%% \item State
-%%   \begin{outline}
-%%   \item There is no shared state in a 
-%%   \item Local state only
-%%   \item Shared state
-%%     \begin{outline}
-%%       \item Impossible in distributed systems
-%%       \item Dangerous in local systems
-%%     \end{outline}
-%%   \end{outline}
-%% \item Communication
-%%   \begin{outline}
-%%   \item Atomic asynchronous message passing
-%%   \item Network sets size of atom (UDP)
-%%   \item Can build reliable streams (TCP)
-%%   \item Local equivalent is passing a value
-%%   \item Model should lend itself to writing protocols
-%%   \end{outline}
-%% \item Asynchrony
-%%   \begin{outline}
-%%     \item Model must have natural support for asynchrony, i.e., event-based
-%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
-%%   \end{outline}
-%% \item Concurrency
-%%   \begin{outline}
-%%     \item Reason about systems using non-deterministic interleaving of atomic actions
-%%     \item Model should admit implementations that execute concurrently
-%%   \end{outline}
-%% \item Dynamics
-%%   \begin{outline}
-%%     \item Configuration - Edges in graph of communicating components can change at run-time.
-%%       \begin{outline}
-%%       \item Already required in distributed settings
-%%       \item Not addressed in formal models
-%%       \end{outline}
-%%     \item Extension - Nodes in graph of communicating components can change at run-time.
-%%   \end{outline}
-%%   \item Reflection
-%% \end{outline}
-
-%% I/O Automata
-%% \begin{itemize}
-%%   \item Compare with UNITY
-%%   \item Compare with esterel
-%%   \item Compare with pi calculus
-%%   \item Compare with Ptolemy
-%% \end{itemize}

### 2011-09-20 papers/paper1/abstract.tex
Revision.
Comments from Dr. Gill.

@@
 \begin{abstract}
-
-The ability to realize sophisticated distributed systems such as those found in cyber-physical systems, pervasive computing, and enterprise computing is predicated on a suitable model and implementation of asynchrony and concurrency.
-The size and complexity of these systems implies that software re-use in their development, maintenance, and evaluation is also essential.
-Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized in reusable software.
+The ability to realize sophisticated distributed systems such as those found in cyber-physical, pervasive, or enterprise computing environments is predicated on a suitable model and implementation of asynchrony and concurrency.
+The size and complexity of these systems implies that software re-use in their development, maintenance, and evolution is also essential.
+Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly and easily in reusable software.
 % By ``different forms'' we mean single-threaded, multi-threaded, time triggered, event triggered, etc.
 
-To this end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
-We outline a system model for distributed systems and connect these requirements to the I/O automata formal model.
-We then describe how ioa++ is implemented in C++ and offer a number of examples showing how ioa++ can be used to develop distributed systems.
+To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
+We outline a system model for asynchronous and concurrent systems, which is grounded in the design and implementation of the I/O automata formal model.
+We describe the design and implementation of ioa++ and show how ioa++ can be used to develop a real network protocol.
 
-Our evalutations show that ioa++ can help simplify developing distributed systems.
-Ioa++ allows developers to reason about asynchronous and concurrent programs directly from the source code using the I/O automata formal model.
-Where reasoning about the correctness of modules in a threaded program requires considering interactions among the various threads, reasoning about the correctness of modules in ioa++ is limited to the modules themselves and a well-defined set of composition rules.
+Our experiences demonstrate that ioa++ can help simplify developing distributed systems, by supporting reasoning about asynchronous and concurrent programs directly from the source code using the I/O automata formal model.
 Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one thread model or event system to another.
+Whereas reasoning about the correctness of modules in a threaded program requires considering interactions among the various threads, reasoning about the correctness of modules in ioa++ is limited to the modules themselves and a well-defined set of composition rules.
+
+We evaluate the ability to concurrently execute actions in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited by the interactions of the automata that comprise the system and the overhead of synchronization.
 \end{abstract}

### 2011-09-20 papers/paper1/case_study.tex
Comments from Dr. Gill.

@@
 Upon construction, the send queue contains the UID of the automaton so the invariant holds initially.
-The leader and init actions do not modify the send queue so the invariant holds trivially.
+The leader action does not modify the send queue so the invariant holds trivially.
+The init action adds the automaton's UID to the send queue.
 The receive action only adds a UID to the send queue if the UID is greater than the automaton's UID.
-The send action removes a UID from the send queue so if the invariant is true before, it will be true after.
+The send action removes a UID from the send queue.
 Reasoning about an automaton directly from the source code allows one to use formal methods to develop and debug systems.

### 2011-09-20 papers/paper1/evaluation.tex
Instrumentation.

@@
 
-\begin{itemize}
-\item Translation of automaton in \emph{Distributed Algorithms} to C++.
-\item Simulate a protocol then replace with network components
-\item Compare a protocol using: single-threaded event-based (select loop), multi-threaded, I/O automata
-\item Show a buggy program and then apply an invariant to find the bug.
-\end{itemize}
+The goal of this section is to provide a preliminary exploration of concurrent execution for I/O automata and establish a baseline for system overhead.
 
-Bring up style mentioned in section \ref{representation}.
+Time spent executing a system of automata falls into four categories.
+\emph{Ioa++} time indicates time spent executing framework code.
+\emph{Thread} time indicates time spent executing (blocking) calls to the pthreads library.
+\emph{Scheduling} time indicates time spent for calls into the framework to schedule an action.
+\emph{User} time indicates the time spent executing automaton code.
+The \emph{total} time is the sum of the four categories.
+
+\paragraph{Experimental design.}
+An automaton $R$ contains an input, output, and internal action.
+The output and internal action execute a routine whose complexity is proportional to the parameter $N$.
+The automaton schedules the internal action with probability $\rho$ and schedules the output action otherwise.
+The automaton executes a total of the 1,000 local actions.
+The system to be executed consists of two $R$ automata composed so the output action of one is composed with the input action of another.
+
+The number of threads, the parameter $N$, and the parameter $\rho$ constitute the configuration parameters for the experiment.
+The internal actions are independent and can be executed concurrently by independent threads.
+Increasing $N$ increases user time.
+When $\rho = 0$, the internal actions are never scheduled and the output actions are always scheduled.
+Consequently, no actions are executed concurrently.
+Conversely, when $\rho = 1$, the internal actions are always scheduled and every action can be executed concurrently.
+
+\paragraph{Results.}
+We executed the described system for $N = 10^{\{1, 2, 3, 4, 5, 6\}}$ and $\rho$ from $0.0$ to $1.0$ in $0.1$ increments.
+The results are shown in Figures~\ref{plot1}, \ref{plot10}, \ref{plot100}, \ref{plot1000}, \ref{plot10000}, \ref{plot100000}, and \ref{plot1000000}.
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{plot1}
+\label{plot1}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{plot10}
+\label{plot10}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{plot100}
+\label{plot100}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{plot1000}
+\label{plot1000}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{plot10000}
+\label{plot10000}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{plot100000}
+\label{plot100000}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{plot1000000}
+\label{plot1000000}
+\end{figure}
+
+0.000357291 0.000767172
+0.000359442 0.000769291

### 2011-09-20 include/ioa/global_fifo_scheduler.hpp

@@ -47,6 +47,10 @@ namespace ioa {
     void close (int fd);
     
     void run (std::auto_ptr<generator_interface> generator);
+
+    void begin_sys_call () { }
+
+    void end_sys_call () { }
   };
 
 }

### 2011-09-20 include/ioa/scheduler.hpp

@@ -99,6 +99,7 @@ namespace ioa {
 	    std::auto_ptr<typed_generator_interface<T> > generator) {
     scheduler = &s;
     scheduler->run (std::auto_ptr<generator_interface> (generator));
+    scheduler = 0;
   }
   
 }

### 2011-09-20 include/ioa/scheduler_interface.hpp

@@ -38,6 +38,10 @@ namespace ioa {
     virtual void close (int fd) = 0;
 
     virtual void run (std::auto_ptr<generator_interface> generator) = 0;
+
+    virtual void begin_sys_call () = 0;
+
+    virtual void end_sys_call () = 0;
   };
   
 }

### 2011-09-20 include/ioa/simple_scheduler.hpp

@@ -17,7 +17,7 @@ namespace ioa {
     void operator= (const simple_scheduler&) { }
 
   public:
-    simple_scheduler ();
+    simple_scheduler (int const threads = 1);
     ~simple_scheduler ();
     
     aid_t get_current_aid ();
@@ -46,6 +46,10 @@ namespace ioa {
     void close (int fd);
 
     void run (std::auto_ptr<generator_interface> generator);
+
+    void begin_sys_call ();
+
+    void end_sys_call ();
   };
 
 }

### 2011-09-20 lib/condition_variable.cpp

@@ -2,25 +2,35 @@
 #include "lock.hpp"
 #include <cassert>
 
+#include "profile.hpp"
+
 namespace ioa {
 
   condition_variable::condition_variable () {
+    BEGIN_SYS_CALL;
     int r = pthread_cond_init (&m_cond, 0);
+    END_SYS_CALL;
     assert (r == 0);
   }
 
   condition_variable::~condition_variable () {
+    BEGIN_SYS_CALL;
     int r = pthread_cond_destroy (&m_cond);
+    END_SYS_CALL;
     assert (r == 0);
   }
 
   void condition_variable::wait (lock& lock) {
+    BEGIN_SYS_CALL;
     int r = pthread_cond_wait (&m_cond, &lock.m_mutex.m_mutex);
+    END_SYS_CALL;
     assert (r == 0);
   }
 
   void condition_variable::notify_one () {
+    BEGIN_SYS_CALL;
     int r = pthread_cond_signal (&m_cond);
+    END_SYS_CALL;
     assert (r == 0);
   }
 

### 2011-09-20 lib/mutex.cpp

@@ -1,25 +1,35 @@
 #include <ioa/mutex.hpp>
 #include <cassert>
 
+#include "profile.hpp"
+
 namespace ioa {
 
   mutex::mutex () {
+    BEGIN_SYS_CALL;
     int r = pthread_mutex_init (&m_mutex, 0);
+    END_SYS_CALL;
     assert (r == 0);
   }
     
   mutex::~mutex () {
+    BEGIN_SYS_CALL;
     int r = pthread_mutex_destroy (&m_mutex);
+    END_SYS_CALL;
     assert (r == 0);
   }
     
   void mutex::lock () {
+    BEGIN_SYS_CALL;
     int r = pthread_mutex_lock (&m_mutex);
+    END_SYS_CALL;
     assert (r == 0);
   }
   
   void mutex::unlock () {
+    BEGIN_SYS_CALL;
     int r = pthread_mutex_unlock (&m_mutex);
+    END_SYS_CALL;
     assert (r == 0);
   }
 

### 2011-09-20 lib/shared_mutex.cpp

@@ -1,30 +1,42 @@
 #include <ioa/shared_mutex.hpp>
 #include <cassert>
 
+#include "profile.hpp"
+
 namespace ioa {
   
   shared_mutex::shared_mutex () {
+    BEGIN_SYS_CALL;
     int r = pthread_rwlock_init (&m_lock, 0);
+    END_SYS_CALL;
     assert (r == 0);
   }
 
   shared_mutex::~shared_mutex () {
+    BEGIN_SYS_CALL;
     int r = pthread_rwlock_destroy (&m_lock);
+    END_SYS_CALL;
     assert (r == 0);
   }
   
   void shared_mutex::unique_lock () {
+    BEGIN_SYS_CALL;
     int r = pthread_rwlock_wrlock (&m_lock);
+    END_SYS_CALL;
     assert (r == 0);
   }
 
   void shared_mutex::shared_lock () {
+    BEGIN_SYS_CALL;
     int r = pthread_rwlock_rdlock (&m_lock);
+    END_SYS_CALL;
     assert (r == 0);
   }
   
   void shared_mutex::unlock () {
+    BEGIN_SYS_CALL;
     int r = pthread_rwlock_unlock (&m_lock);
+    END_SYS_CALL;
     assert (r == 0);
   }
   

### 2011-09-20 lib/simple_scheduler.cpp

@@ -31,7 +31,16 @@
 #include "output_unbound_runnable.hpp"
 #include "input_unbound_runnable.hpp"
 
+#include <iostream>
+
 namespace ioa {
+
+  std::ostream& operator<< (std::ostream& os, const ioa::time& t) {
+    double d = t.usec ();
+    d /= 1000000.0;
+    d += t.sec ();
+    return (os << d);
+  }
 
   typedef std::pair<time, action_runnable_interface*> time_action;
   typedef std::pair<int, action_runnable_interface*> fd_action;
@@ -40,10 +49,80 @@ namespace ioa {
     public system_scheduler_interface
   {
   private:
+    class thread_context
+    {
+    private:
+      enum state_t {
+	NONE,
+	IOA,
+	THREAD,
+	USER,
+	SCHEDULE
+      };
+      state_t m_state;
+      ioa::time m_start;
+      ioa::time m_none;
+
+      void switch_to (state_t state) {
+	const ioa::time now = ioa::time::now ();
+	switch (m_state) {
+	case NONE:
+	  m_none += (now - m_start);
+	  break;
+	case IOA:
+	  m_ioa += (now - m_start);
+	  break;
+	case THREAD:
+	  m_thread += (now - m_start);
+	  break;
+	case USER:
+	  m_user += (now - m_start);
+	  break;
+	case SCHEDULE:
+	  m_schedule += (now - m_start);
+	  break;
+	}
+	m_state = state;
+	m_start = now;
+      }
+
+    public:
+      pthread_t m_id;
+      blocking_list<std::pair<bool, action_runnable_interface*> > m_execq;
+      ioa::time m_ioa;
+      ioa::time m_thread;
+      ioa::time m_user;
+      ioa::time m_schedule;
+
+      thread_context () :
+	m_state (NONE) { }
+
+      void switch_to_none () {
+	switch_to (NONE);
+      }
+
+      void switch_to_ioa () {
+	switch_to (IOA);
+      }
+
+      void switch_to_thread () {
+	switch_to (THREAD);
+      }
+
+      void switch_to_user () {
+	switch_to (USER);
+      }
+
+      void switch_to_schedule () {
+	switch_to (SCHEDULE);
+      }
+    };
 
     model m_model;
+    const int THREAD_COUNT;
     blocking_list<std::pair<bool, runnable_interface*> > m_sysq;
-    blocking_list<std::pair<bool, action_runnable_interface*> > m_execq;
[…]

### 2011-09-20 lib/thread.cpp

@@ -21,4 +21,7 @@ namespace ioa {
     assert (r == 0);
   }
 
+  pthread_t thread::get_id () const {
+    return m_thread;
+  }
 }

### 2011-09-20 lib/thread.hpp

@@ -62,6 +62,7 @@ namespace ioa {
 
     ~thread ();
     void join ();
+    pthread_t get_id () const;
   };
   
 }

### 2011-09-20 lib/thread_key.hpp

@@ -4,6 +4,8 @@
 #include <pthread.h>
 #include <cassert>
 
+#include "profile.hpp"
+
 namespace ioa {
 
   template <typename T>
@@ -14,23 +16,30 @@ namespace ioa {
 
   public:
     thread_key () {
+      BEGIN_SYS_CALL;
       int r = pthread_key_create (&m_key, 0);
+      END_SYS_CALL;
       assert (r == 0);
     }
 
     ~thread_key () {
+      BEGIN_SYS_CALL;
       int r = pthread_key_delete (m_key);
+      END_SYS_CALL;
       assert (r == 0);
     }
 
     T get () const {
+      // Using BEGIN_SYS_CALL/END_SYS_CALL results in infinite recursion.
       void* ptr = pthread_getspecific (m_key);
       return reinterpret_cast<T> (ptr);
     }
 
     void set (T t) {
       const void* ptr = reinterpret_cast<const void*> (t);
+      BEGIN_SYS_CALL;
       int r = pthread_setspecific (m_key, ptr);
+      END_SYS_CALL;
       assert (r == 0);
     }
   };

### 2011-09-20 papers/paper1/evaluation.tex
First draft of evaluation section.

--- 2011-07-13 papers/paper1/evaluation.tex
@@
-\section{Evaluation and Discussion\label{evaluation}}
+\section{Evaluation\label{evaluation}}
 
-\begin{itemize}
-\item Translation of automaton in \emph{Distributed Algorithms} to C++.
-\item Simulate a protocol then replace with network components
-\item Compare a protocol using: single-threaded event-based (select loop), multi-threaded, I/O automata
-\item Show a buggy program and then apply an invariant to find the bug.
-\end{itemize}
+Ioa++ permits independent actions to be executed concurrently.
+To test this capability, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
+We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
+We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
+We also vary the complexity of each action to gain insight into how synchronization affects concurrent execution.
 
-Bring up style mentioned in section \ref{representation}.
+An automaton $R$ contains an input, output, and internal action.
+The output and internal action execute a routine whose complexity is proportional to the parameter $N$.
+The automaton executes a fixed number of local actions which was set to 1,000 for all experiments.
+The automaton schedules the internal action with probability $\sqrt{\rho}$ and schedules the output action otherwise.
+The system $S$ to be executed consists of two $R$ automata composed so the output action of one is composed with the input action of another.
+If we divide the execution into rounds where each automaton executes a single action, the probability that both $R$ automata execute an internal action is $\sqrt{\rho} \times \sqrt{\rho} = \rho$.
+Thus, we can use the parameter $\rho$ to vary the independence of the automata.
+
+For this experiment, we developed a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action.
+When a single thread is used, all actions are executed by the same thread.
+When two threads are used, one thread executes the actions of one $R$ automaton while the other thread executes the actions of the other $R$ automaton.
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{speed_up}
+\caption{Speed-up for system $S$.\label{speed_up}}
+\end{figure}
+
+Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
+The parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$.
+The parameter $N$ was varied from $1$ to $1000000$ in factors of $10$.
+When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
+Consequently, every action must be serialized yielding a maximum speed-up of 1.
+When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.
+
+The speed-up shows a strong dependence on the duration of each local action which is proportional to $N$.
+A small value for $N$ means that very little time is spent executing automaton code.
+Consequently, a greater fraction of time is spent executing framework code which includes various synchronization calls to the pthreads library.
+For a small enough $N$, this overhead dominates the execution time.
+This situation is exacerbated by multiple threads since they will actively interfere with one another which can be seen in the slow-down for small values of $N$.
+Conversely, when $N$ is large, relatively little time is spent executing framework and synchronization code.
+Consequently, contention is reduced allowing for greater speed-ups as indicated by Figure~\ref{speed_up}.
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{overhead1}
+\caption{Overhead for system $S$ when using one thread.\label{overhead1}}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=\textwidth]{overhead2}
+\caption{Overhead for system $S$ when using two threads.\label{overhead2}}
+\end{figure}
+
+Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
+Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
+The fraction decreases as $\rho$ goes to 1 as the system transitions from acquiring two locks to one lock.
+The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
+For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.
+
+This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
+The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the \verb+ioa::schedule+ call.
+From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4.2115us with a standard deviation of .765us for the single threaded experiments and 4.858us with a standard deviation of 1.099us for the multi-threaded experiments.
+Similary, we calculated the scheduling overhead and found it to be .2275us with a standard deviation of .062us for the single threaded experiments and .0279us with a standard deviation of .079us for the multi-threaded experiments.
+Both of these can be improved with better scheduler design.
+The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.
+We plan to address scheduling and synchronization in future work.

### 2011-09-20 lib/simple_scheduler.cpp

@@ -661,10 +661,13 @@ namespace ioa {
       assert (aid != -1);
       thread_context* context = m_con.get ();
       if (context != 0) {
-	context->switch_to_user ();
+	context->switch_to_ioa ();
       }
       // This is to be used during generation so that any allocated memory can be associated with the automaton.
       m_current_aid.set (aid);
+      if (context != 0) {
+	context->switch_to_user ();
+      }
     }
   
     void clear_current_aid () {

### 2011-09-20 papers/paper1/component_model.tex
Dr. Gill's comments on the component model.

@@
 Modularity, well-defined interfaces, and composition are essential properties of reusable software.
-A unit of software that exhibits these properties is called a \emph{component}~\cite{szyperski2002component}.
+A unit of software that exhibits all of these properties is called a \emph{component}~\cite{szyperski2002component}.
 Modularity implies that a component can be deployed independently of other components.
-Well-defined interfaces and interactions allow components to expose their functionality in a regular way that facilitates reasoning about different interactions with the component.
+Well-defined interfaces and interactions allow components to expose their functionality in a regular way that facilitates reasoning about different interactions among them.
 Composition allows a group of interacting components to be understood as a cohesive unit.
 
-In this section we describe how I/O automata offer a natural realization of components and interfaces, and provide natural support for concurrency and composition.
+In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
 The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
@@
 An internal action changes the state of the automaton.
-An output action changes the state of the automaton and produces a signal or value that is consumed by zero or more input actions.
+An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
 An input action changes the state of the automaton when it receives a signal or value produced by an output action.
@@
 
-Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
-In general the scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
+Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
+The scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
 Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
@@
 However, it lacks certain features needed to develop real components.
-This subsection introduces these features and lays the foundation for adding the dynamics discussed in section~\ref{dynamics}.
+This subsection introduces these features and lays the foundation for adding the dynamics discussed in Section~\ref{dynamics}.
 
@@
 Component frameworks that support recursive encapsulation allow components to contain other components.
-Recursive encapsulation is at the heart of both bottom-up and top-down design strategies.
+Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
 The contained component is the \emph{child} while the containing component is the \emph{parent}.
@@
 Applying recursive encapsulation to I/O automata means that an I/O automaton may contain and use other I/O automata.
-Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
-The kernel automaton allows a parent automaton to have its own state variables and actions.
+Recursive encapsulation implies that a parent automaton is the result of composing its children.
+This definition excludes a parent from having any state or actions of its own.
+Since it is often convenient to associate state and actions with a parent, we allow a parent to be the composition of its children and an anonymous child that contains the state and actions attributed to the parent.
 The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.
@@
 The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
-The name of a named automata appears after its type, e.g., Converter C.
+The name of a named automaton appears after its type, e.g., Converter C.
+Actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
+Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have neither suffix, e.g., \emph{process1}.
+The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.
+
 The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
@@
 To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
-Such an association is called a \emph{binding} and is a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.
+Such an association is called a \emph{binding} and consists of a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.
 
@@
 A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
+For example, the fully-qualified name of the $in?(s)$ action of the Converter C in Figure~\ref{sys_model} is $.C.in?(s)$ due to the anonymous root automaton.
 Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
-
-For the event detection system illustrated in Figure~\ref{sys_model}, actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
-Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have no suffix, e.g., \emph{process1}.
-The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.
-The map of bindings for the event detection system is $\{ (out!(s), C.in?(s)), (C.out!(t), D.in?(t)), (D.detect!, detected?) \}$.
+The map of bindings for the event detection system is $\{ (.out!(s), .C.in?(s)), (.C.out!(t), .D.in?(t)), (.D.detect!, .detected?) \}$.
 
@@
 For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
-The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value (i.e., a signal).
+The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., a signal).
 Second, an input action can be bound to at most one output action.
@@
 A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
-For dynamics, which are discussed in section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
-Consequently, every binding has exactly one owner.
-For a graphical representation such as Figure~\ref{sys_model}, this rule means that an input is the sink of at most one arrow (binding).
+For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
+This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
 Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
-Graphically, arrows originating at the same output action terminate in different components.
 %% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
@@
 However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
-Graphically, an arrow originating in one component must terminate in another component.
+Graphically, an arrow originating in one component must terminate in another component, even if the latter is a child of the former as Figure~\ref{sys_model} illustrates.
 %% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
@@
 For an internal action, this set consists of the automaton containing the internal action.
-For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound, i.e., composed.
-Two actions can be executed concurrently if their respective sets of implied automata and therefore state variables are disjoint.
-The implied automata sets for each local action in figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect!
+For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
+Two actions can be executed concurrently if their respective sets of state variables are disjoint.
+The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect!
 \to \{D, root\}$.
@@
 Compositions of automata can be statically composed to take advantage of in-lining.
-An individual automaton can be factored into a number of child automata to increase parallelism.
+An individual automaton can be decomposed into a number of child automata to increase parallelism.
 
-\paragraph{Parameters.}
+\paragraph{Parameters and action types.}
 Actions in the I/O automata model can be defined using parameters.
@@
 
-\paragraph{Action types.}
 The allowed combinations of input, output, and internal actions, unvalued and valued actions, and unparameterized and parameterized actions result in 10 possible action types:
@@
 \paragraph{Example.}
-Figure~\ref{example1} depicts a remote sensor systems as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
-The Remote Sensor automaton consists of the Control automaton $Z$ and three Channel automata $A$, $B$, and $C$ representing three network connections.
-The Control automaton $Z$ consists of the Sensor automaton $S$, the Filter automaton $F$, and the Statistics automaton $T$.
-The Control automaton $C$ starts the sampling process with the uv-up output $Z.trigger!$.
+Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
+The Remote Sensor automaton consists of the Control automaton Z and three Channel automata A, B, and C representing three network connections.
+The Control automaton Z consists of the Sensor automaton S, the Filter automaton F, and the Statistics automaton T.
+The Control automaton C starts the sampling process with the uv-up output $Z.trigger!$.
 The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
-When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter $F$ and Statistics component $T$ via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
+When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter F and Statistics component T via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
 Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
-The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $(F.out!(t), Z.sample?(t))$ binding.
-The Statistics automaton $T$ calculates the statistics incrementally using the up internal $T.compute$.
-The statistics can be polled by the Control automaton $Z$ using the bindings $(Z.request!, T.request?)$ and $(T.response!(u), Z.response?(u))$.
-The Control automaton $Z$ contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata $A$, $B$, and $C$.
-The parameters 0, 1, and 2, are used to route messages to Channel automata $A$, $B$, and $C$ respectively using the session idea mentioned previously.
-Opportunities for concurrent execution exist for the remote sensor depicted in figure~\ref{example1}.
+The Filter automaton F passes the filtered sample to the Control automaton Z via the $(F.out!(t), Z.sample?(t))$ binding.
+The Statistics automaton T calculates the statistics incrementally using the up internal $T.compute$.
+The statistics can be polled by the Control automaton Z by issuing a request via the $(Z.request!, T.request?)$ binding and then receiving a response via the $(T.response!(u), Z.response?(u))$ binding.
+The Control automaton Z contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata A, B, and C.
+The parameters 0, 1, and 2, are used to route messages to Channel automata A, B, and C respectively using the session idea mentioned previously.
+Opportunities for concurrent execution exist for the remote sensor depicted in Figure~\ref{example1}.
 The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
@@
 
-The I/O automata model assumes that systems are composed of a finite or infinitely countable number of automata~\cite{lynch1996distributed}.
-Assuming an infinitely countable number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
+The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
+Assuming an countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
 The set of interactions in the I/O automata model is also fixed since automata are statically composed.
@@
 If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
-Consequently, we require the ability to deal with a dynamic set of automata and interactions.
+Consequently, we add the ability to deal with a dynamic set of automata and interactions.
 
 \paragraph{System actions.}
-The four indivisible operations for managing a dynamic set of automata are create, bind, unbind, and destroy.
+The four indivisible operations for managing a dynamic set of automata are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
 Create, bind, unbind, and destroy are collectively called \emph{system actions}.
@@
 A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
-A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in section~\ref{practical} would be violated.
+A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{practical} would be violated.
 An unbind fails if the binding does not exist.
-Note that the owner of binding is taken into account when determining if the binding exist so only the owner of a binding can dissolve it.
-A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy.
+Note that the owner of binding is taken into account when determining if the binding exists so only the owner of a binding can dissolve it.
+A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.
 
@@
 \paragraph{The system automaton.}
-Let the set of automata $A$ and set of bindings $B$ be managed by the \emph{system automaton} which represents the run-time system.
+The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
 Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
@@
 However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
-The scheduler uses $A$ to ensure that only actions belonging to automata exist are selected and uses $B$ to ensure that the appropriate set of inputs actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the set defined in section~\ref{practical} \emph{and} and the system automaton.
-However, the opportunities for concurrency explored in section~\ref{practical} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
+Thus, the set of actions implied by any action is the set defined in Section~\ref{practical} \emph{and} and the system automaton.
+However, the opportunities for concurrency explored in Section~\ref{practical} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
 
@@
 To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
-To do this, we conjunct the precondition of an output action with a predicate over the set of bindings $B$ to ensure that the output action is not executed until its set of implied automata is correct.
+To do this, we take the conjunct of the precondition of an output action with a predicate over the set of bindings $B$ called a \emph{binding predicate} to ensure that the output action is not executed until its set of implied automata is correct.
 Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as previously discussed.

### 2011-09-20 papers/paper1/design.tex
Dr. Gill's edits to design.

@@
 
-An implementation of the component model described in section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
+An implementation of the component model described in Section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
 The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
-The Scheduler has the resposibility of selecting and executing local actions subject to the I/O automata model and concurrency constraints of~\ref{component_model}.
+The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
 All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
-The services provided by the base class are indicated by the Base Automaton layer of Figure~\ref{framework_architecture}.
+The services provided by the base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
 The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
@@
 Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
-We assume a POSIX environment whose native concurrency mechanism is threads.
 
@@
 
-The scheduler has the responsibility of selecting and executing local actions.
-The two activities of the scheduler can be decomposed into a \emph{model} that executes actions and a \emph{controller} that selects the next action.
+The scheduler is responsible both for selecting and for executing local actions.
+These two activities can be decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
 
-\paragraph{The Model.}
-The model enforces the execution and concurrency constraints of section~\ref{component_model}.
+The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
 Input actions are executed with output actions according to the current set of bindings.
-The model allows executing actions concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
-Since concurrent access to the model is allowed, the module implementing the model must be thread-safe.
+The dispatcher allows executing actions concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
+Since concurrent access to the dispatcher is allowed, the module implementing the dispatcher must be thread-safe.
 
-Our implementation of the model uses a two-level locking scheme to enforce the concurrent execution constraints.
+Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
 Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
@@
 
-\paragraph{The Controller.}
 The controller encodes a \emph{policy} that selects the next local action.
-For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler should be \emph{fair} as defined in section~\ref{component_model}.
+For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler should be \emph{fair} as defined in Section~\ref{component_model}.
 The run-time system can be initialized with different controllers to achieve different performance objectives.
@@
 \paragraph{Binding predicates and scheduling.}
-The preceding discussion on scheduling claims that an action in one automaton cannot affect preconditions in another automaton.
+The preceding discussion on scheduling assumes that an action in one automaton cannot affect preconditions in another automaton.
 This is not entirely correct when one considers the system automaton and binding predicates.
-Imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
+For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
 When the automaton is created, its only action is disabled.
-Consequently, it will not schedule its action and therefore not be executed.
+Consequently, it will not schedule its action and therefore will not be executed.
 Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
 The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
-While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound; a situation that occurs quite often in practice.
+While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
 To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a successful bind or unbind.
 
-\paragraph{Actions of non-existent automata.}
-Every action in the set of actions considered by the controller was scheduled by its associated automaton.
+\paragraph{Actions of no-longer-existing automata.}
+Every action in the set of actions considered by the controller is scheduled by its associated automaton.
 Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
 Our solution is to check if the automaton exists before executing an action.
-This check is performed by the model and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
+This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
 Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)} and that the set of actions in the scheduler turns over comparatively faster than the set of aids.
@@
 Our implementation of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
-A user automaton creates a file descriptors and places it in non-blocking mode.
+A user automaton creates a file descriptor and configures it for non-blocking I/O.
 The automaton then tells the scheduler to release an action (typically an internal action) whenever the automaton becomes ready for reading (or writing).
@@
 \paragraph{System action protocol.}
-Our approach to system actions follows the model introduced in section~\ref{dynamics}.
+Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
 All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.

### 2011-09-22 papers/paper1/abstract.tex
Performed analysis again.

@@
 To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
-We outline a system model for asynchronous and concurrent systems, which is grounded in the design and implementation of the I/O automata formal model.
-We describe the design and implementation of ioa++ and show how ioa++ can be used to develop a real network protocol.
+We develop a component model for asynchronous and concurrent systems based on the I/O automata formal model and dynamic composition.
+We describe the design and implementation of ioa++ based on that system model and show how ioa++ can be used to develop protocols and other forms of distributed systems.
 
 Our experiences demonstrate that ioa++ can help simplify developing distributed systems, by supporting reasoning about asynchronous and concurrent programs directly from the source code using the I/O automata formal model.
-Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one thread model or event system to another.
+Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one concurrency model to another.
 Whereas reasoning about the correctness of modules in a threaded program requires considering interactions among the various threads, reasoning about the correctness of modules in ioa++ is limited to the modules themselves and a well-defined set of composition rules.
 
-We evaluate the ability to concurrently execute actions in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited by the interactions of the automata that comprise the system and the overhead of synchronization.
+We evaluate the ability to execute actions concurrently in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
 \end{abstract}
+
+\begin{IEEEkeywords}
+TODO; component; formatting; style; styling;
+\end{IEEEkeywords}

### 2011-09-22 papers/paper1/case_study.tex
Performed analysis again.

@@
 
-In this section, we explore (1) if ioa++ offers a convenient representation for I/O automata and (2) and if ioa++ contains the features necessary for building real distributed systems.
+In this section, we explore whether ioa++ offers a suitably convenient representation for reifying I/O automata and contains the features necessary for building real distributed systems.
 Our approach is to (1) express an automaton in ioa++ using an existing listing, (2) simulate a distributed system built using the automaton as debugging step, and (3) build a real distributed system by composing the automaton with automata for network services.
@@
 
-\paragraph{Translating the asynchronous LCR automaton into ioa++.}
+\paragraph*{Translating the asynchronous LCR automaton into ioa++}
 The first version of the code for the asynchronous LCR automaton was translated directly from a listing in~\cite{lynch1996distributed}.
@@
 For example, an invariant in the asynchronous LCR automaton is that all of the UIDs in the send queue of automaton $i$ must be greater than or equal to the UID of automaton $i$.
-This invariant can be quickly checked against the constructor and four actions of the \verb+asynch_lcr_automaton+.
+This invariant can be quickly checked against the constructor and four actions of the asynch\_lcr\_automaton.
 Upon construction, the send queue contains the UID of the automaton so the invariant holds initially.
@@
 
-\paragraph{Unit testing with simulation.}
+\paragraph*{Unit testing with simulation}
 After the asynchronous LCR automaton was implemented, we developed a simulation to test it.
-The \verb+examples/asynch_lcr.cpp+ file contains the driver for the simulation.
-The simulation creates a unidirectional ring of a specified size consisting of alternating \verb+asynch_lcr_automaton+ and \verb+channel_automaton+.
-(The \verb+channel_automaton+ was also a direct translation from~\cite{lynch1996distributed}.)
+The examples/asynch\_lcr.cpp file contains the driver for the simulation.
+The simulation creates a unidirectional ring of a specified size consisting of alternating asynch\_lcr\_automaton and channel\_automaton.
+(The channel\_automaton was also a direct translation from~\cite{lynch1996distributed}.)
 
-The simulation showed that the \verb+asynch_lcr_automaton+ was not correct.
+The simulation showed that the asynch\_lcr\_automaton was not correct.
 The first bug we encountered was a silent failure where the ring always failed to elect a leader.
@@
 The key insight into solving the problem was to understand that the automata were executing as the ring was being constructed, i.e., we failed to consider dynamic composition.
-The solution was to add binding predicates to the outputs of the \verb+asynch_lcr_automaton+ and \verb+channel_automaton+.
+The solution was to add binding predicates to the outputs of the asynch\_lcr\_automaton and channel\_automaton.
 
-\paragraph{From abstraction to real system.}
+\paragraph*{From abstraction to real system}
 The next phase was to use the asynchronous LCR automaton to perform leader election in a real unidirectional ring of processes connected by TCP sockets.
@@
 If a node has not received a token for a specified duration of time, it calls for an election.
-This is the motivation behind the action that resets the \verb+asynch_lcr_automaton+.
+This is the motivation behind the action that resets the asynch\_lcr\_automaton.
 Nodes listen for connections from their predecessor and attempt to connect to their successor.
@@
 A ring containing multiple leaders should stabilize to having a single leader.
-This is the motivation behind the \verb+asynch_lcr_automaton+ being able to withdraw its election.
+This is the motivation behind the asynch\_lcr\_automaton being able to withdraw its election.
 A rare but possible and valid situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
 
-The \verb+tcp_lcr_automaton+ which is listed in \verb+examples/tcp_lcr.cpp+ (484 lines) of the ioa++ package implements the protocol just described.
-The \verb+tcp_lcr_automaton+ is a composition of a number of automata including an \verb+asynch_lcr_automaton+, an \verb+ioa::tcp_acceptor_automaton+, an \verb+ioa::tcp_connector_automaton+, and two \verb+ioa::tcp_connection_automaton+ representing the channel from the predecessor and channel to the successor.
+The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp (484 lines) of the ioa++ package implements the protocol just described.
+The tcp\_lcr\_automaton is a composition of a number of automata including an asynch\_lcr\_automaton, an ioa::tcp\_acceptor\_automaton, an ioa::tcp\_connector\_automaton, and two ioa::tcp\_connection\_automaton representing the channel from the predecessor and channel to the successor.
 The TCP automata provided by the ioa++ package are ordinary automata implemented using file descriptors and the techniques described in section~\ref{design}.
-The \verb+tcp_lcr_automaton+ uses the techniques for managing dynamic constellations described in section~\ref{representation}.
+The tcp\_lcr\_automaton uses the techniques for managing dynamic constellations described in section~\ref{representation}.
 
-The \verb+tcp_lcr_automaton+ is validation that I/O automata are a good foundation for building reusable concurrent modules.
+The tcp\_lcr\_automaton is validation that I/O automata are a good foundation for building reusable concurrent modules.
 The key idea is that all of these modules have well-defined interactions when composed while being inherently concurrent.
-For example, one could substitute the \verb+asynch_lcr_automaton+ with a different automaton to realize a different leader election protocol or a different protocol altogether.
-Alternatively, the \verb+tcp_lcr_automaton+ could be refactored to use a different communication mechanism, e.g., UDP.
+For example, one could substitute the asynch\_lcr\_automaton with a different automaton to realize a different leader election protocol or a different protocol altogether.
+Alternatively, the tcp\_lcr\_automaton could be refactored to use a different communication mechanism, e.g., UDP.
 One need only focus on the protocol implemented by an automaton when composing it with an existing system as opposed to worrying about interacting threads and race conditions.

### 2011-09-22 papers/paper1/component_model.tex
Performed analysis again.

@@
 In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
-The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
+The Input/Output (I/O) automata model was developed by Lynch to model reactive concurrent and asynchronous systems~\cite{lynch1996distributed}.
 An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
@@
 An input action changes the state of the automaton when it receives a signal or value produced by an output action.
-\emph{Unvalued} output and input actions produce and consume signals respectively.
-\emph{Valued} output and input actions produce and consume values respectively.
 The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.
@@
 \center
-\includegraphics[width=\textwidth]{system_model}
+\includegraphics[width=\columnwidth]{system_model}
 \caption{Example event detection system.}
@@
 
-\paragraph{Recursive Encapsulation.}
+\paragraph*{Recursive Encapsulation}
 Component frameworks that support recursive encapsulation allow components to contain other components.
@@
 Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have neither suffix, e.g., \emph{process1}.
-The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.
+The type associated with the value produced or consumed by an action is in parentheses after the name, e.g., \emph{out!(s)}.
 
@@
 
-\paragraph{Explicit Composition.}
+\paragraph*{Explicit Composition}
 Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
@@
 A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
-For example, the fully-qualified name of the $in?(s)$ action of the Converter C in Figure~\ref{sys_model} is $.C.in?(s)$ due to the anonymous root automaton.
+For example, the fully-qualified name of the $in?(s)$ action of the Converter C in Figure~\ref{sys_model} is $[root].C.in?(s)$.
 Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
-The map of bindings for the event detection system is $\{ (.out!(s), .C.in?(s)), (.C.out!(t), .D.in?(t)), (.D.detect!, .detected?) \}$.
+The map of bindings for the event detection system is 
+\begin{displaymath}
+\begin{split}
+\{ ([root].out!(s), [root].C.in?(s)),\\
+   ([root].C.out!(t), [root].D.in?(t)),\\
+   ([root].D.detect!, [root].detected?) \}
+\end{split}
+\end{displaymath}
 
-\paragraph{Binding rules.}
+\paragraph*{Binding rules}
 To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
@@
 
-\paragraph{Concurrent execution.}
+\paragraph*{Concurrent execution}
 Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
@@
 
-\paragraph{Parameters and action types.}
+\paragraph*{Parameters and action types}
 Actions in the I/O automata model can be defined using parameters.
@@
 
+\emph{Unvalued} output and input actions produce and consume signals respectively.
+\emph{Valued} output and input actions produce and consume values respectively.
 The allowed combinations of input, output, and internal actions, unvalued and valued actions, and unparameterized and parameterized actions result in 10 possible action types:
@@
 \center
-\includegraphics[width=\textwidth]{example1}
+\includegraphics[width=\columnwidth]{example1}
 \caption{Example remote sensor system.}
@@
 
-\paragraph{Example.}
+\paragraph*{Example}
 Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
@@
 Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
-To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}.
+To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}\footnote{Use the actual paper.
+I think it has something to do with databases.}.
 
@@
 
-\paragraph{System actions.}
+\paragraph*{System actions}
 The four indivisible operations for managing a dynamic set of automata are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
@@
 \center
-\includegraphics[width=\textwidth]{system_action}
+\includegraphics[width=\columnwidth]{system_action}
 \caption{System actions.
@@
 
-\paragraph{The system automaton.}
+\paragraph*{The system automaton}
 The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
 Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
-The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
+The system automaton, in turn, has input actions for receiving requests for system actions, internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
 Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.
@@
 
-\paragraph{Binding predicates.}
+\paragraph*{Binding predicates}
 The static configuration of a system as prescribed by the I/O automata model guarantees that the set of implied automata for an action is always the same.

### 2011-09-22 papers/paper1/design.tex
Performed analysis again.

@@
 \center
-\includegraphics[width=\textwidth]{architecture}
+\includegraphics[width=\columnwidth]{architecture}
 \caption{Framework architecture.}
@@
 
-\paragraph{Explicit dynamic scheduling.}
+\paragraph*{Explicit dynamic scheduling}
 The I/O automata model assumes that the scheduler knows about every local action in the system.
@@
 
-\paragraph{Binding predicates and scheduling.}
+\paragraph*{Binding predicates and scheduling}
 The preceding discussion on scheduling assumes that an action in one automaton cannot affect preconditions in another automaton.
@@
 
-\paragraph{Actions of no-longer-existing automata.}
+\paragraph*{Actions of no-longer-existing automata}
 Every action in the set of actions considered by the controller is scheduled by its associated automaton.
@@
 
-\paragraph{Delayed actions.}
+\paragraph*{Delayed actions}
 Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities.
@@
 
-\paragraph{File descriptor events.}
+\paragraph*{File descriptor events}
 File descriptors are used to communicate with the outside world in a POSIX environment.
@@
 
-\paragraph{Global FIFO scheduler.}
+\paragraph*{Global FIFO scheduler}
 The scheduler included with the framework is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
@@
 
-\paragraph{System action protocol.}
+\paragraph*{System action protocol}
 Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
@@
 
-\paragraph{Automaton and Binding managers.}
+\paragraph*{Automaton and Binding managers}
 The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
@@
 
-\paragraph{Generators.}
+\paragraph*{Generators}
 A \emph{generator} is a single-use factory for creating an instance of an automaton.
@@
 
-Ioa++ is implemented in C++ and available at \url{http://jrwilson.github.com/ioa/}.
+Ioa++ is implemented in C++ for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.

### 2011-09-22 papers/paper1/evaluation.tex
Performed analysis again.

@@
 \center
-\includegraphics[width=\textwidth]{speed_up}
+\includegraphics[width=\columnwidth]{speed_up}
 \caption{Speed-up for system $S$.\label{speed_up}}
@@
 \center
-\includegraphics[width=\textwidth]{overhead1}
+\includegraphics[width=\columnwidth]{overhead1}
 \caption{Overhead for system $S$ when using one thread.\label{overhead1}}
@@
 \center
-\includegraphics[width=\textwidth]{overhead2}
+\includegraphics[width=\columnwidth]{overhead2}
 \caption{Overhead for system $S$ when using two threads.\label{overhead2}}
@@
 The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the \verb+ioa::schedule+ call.
-From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4.2115us with a standard deviation of .765us for the single threaded experiments and 4.858us with a standard deviation of 1.099us for the multi-threaded experiments.
-Similary, we calculated the scheduling overhead and found it to be .2275us with a standard deviation of .062us for the single threaded experiments and .0279us with a standard deviation of .079us for the multi-threaded experiments.
+From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4211.5ns with a standard deviation of 765ns for the single threaded experiments and 4858ns with a standard deviation of 1099ns for the multi-threaded experiments.
+Similary, we calculated the scheduling overhead and found it to be 227.5ns with a standard deviation of 62ns for the single threaded experiments and 27.9ns with a standard deviation of 79ns for the multi-threaded experiments.
 Both of these can be improved with better scheduler design.

### 2011-09-22 papers/paper1/paper.tex
Abstract edits.
Updated abstract and introduction.
New introduction and part of system model.
Moving on to section 2.
Revision of Section 2.  New example.
Revision of section 2.
First draft of section 3.
First part of Representation and Programming Model.
ioa::automaton is virtual base.
Second revision of section 4.
Revision.
Case study.
Performed analysis again.

@@
-\documentclass[letterpaper]{article}
+\documentclass[10pt, conference, compsocconf]{IEEEtran}
 
-\usepackage{outline}
+\usepackage{cite}
+\usepackage{graphicx}
+\usepackage{listings}
+\usepackage{url}
+\usepackage{amsmath}
+
+\lstset{language=C++, basicstyle=\footnotesize}
+
+\newif\ifjournal
+\journalfalse
 
@@
 
-\title{An I/O Automata Framework for Distributed System Programming}
-\author{Justin R. Wilson \and Christopher D. Gill}
-\date{}
+\title{Implementing Distributed Systems Directly with I/O Automata in Ioa++}
+\author{
+  \IEEEauthorblockN{Justin R. Wilson and Christopher D. Gill}
+  \IEEEauthorblockA{Department of Computer Science and Engineering\\
+    Campus Box 1045, Washington University\\
+    One Brookings Drive\\
+    St. Louis, MO 63130\\
+    {wilsonj,cdgill}@cse.wustl.edu}
+  \and
+  \IEEEauthorblockN{Guia-Catalin Roman}
+  \IEEEauthorblockA{TODO:  Dr. Roman's address}
+}
 
@@
 
-\input{system_model}
+\input{component_model}
+
+\input{design}
 
@@
 
-\input{design}
+\input{case_study}
 
@@
 
+\bibliographystyle{IEEEtran}
+\bibliography{bibliography}{}
+
 \end{document}

### 2011-09-22 papers/paper1/programming_model.tex
Performed analysis again.

@@
 
-\paragraph{Leader election in a unidirectional ring.}
+\paragraph*{Leader election in a unidirectional ring}
 Nodes in a distributed system are arranged in a ring so that two consecutive nodes are connected by a directed channel.
@@
 When a node receives its own UID, it elects itself the leader of the ring.
-Our implementation of the asynchronous LCR automaton is located in \verb+examples/asynch_lcr_automaton.hpp+ of the ioa++ package.
+Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp of the ioa++ package.
 
@@
 
-\paragraph{Declaring an automaton.}
-All automata inherit from the common base class \verb+ioa::automaton+ which implements the system action interface.
-The following code declares the \verb+asynch_lcr_automaton+:
+\paragraph*{Declaring an automaton}
+All automata inherit from the common base class ioa::automaton which implements the system action interface.
+The following code declares the asynch\_lcr\_automaton:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The \verb+asynch_lcr_automaton+ is declared as a template where the \verb+UID+ template parameter must model the unique identifier concept.
+The asynch\_lcr\_automaton is declared as a template where the UID template parameter must model the unique identifier concept.
 The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
 
-\paragraph{Declaring and initializing state variables.}
+\paragraph*{Declaring and initializing state variables}
 The state variables of an automaton are expressed as member variables of the corresponding class.
-The four state variables of the \verb+asynch_lcr_automaton+ include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
+The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Notice that the member variables are declared \verb+private+ which is in keeping with the notion that the state of each automaton is independent.
+Notice that the member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
 State variables are initialized using a constructor.
-For the \verb+asynch_lcr_automaton+, the flags are both false and the UID is initialized and added to the send queue:
+For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function \verb+schedule+.
-The \verb+schedule+ member function is discussed in the paragraph on scheduling.
+The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function schedule.
+The schedule member function is discussed in the paragraph on scheduling.
 
-\paragraph{Declaring/defining actions.}
+\paragraph*{Declaring/defining actions}
 Actions are member variables that model the action concept.
@@
 Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
-The macros assume the functions are named \verb+action_name_precondition+, \verb+action_name_effect+, and \verb+action_name_schedule+.
-For example, the send action of the \verb+asynch_lcr_automaton+ is declared:
+The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
+For example, the send action of the asynch\_lcr\_automaton is declared:
 \begin{lstlisting}
@@
 Input actions are declared similarly save the absence of a precondition.
-The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared \verb+const+.
-The macros for actions that are subject to binding by other automata belong in a \verb+public+ section.
+The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
+The macros for actions that are subject to binding by other automata belong in a public section.
 
-\paragraph{Scheduling.}
-Automata submit actions to the scheduler using the \verb+ioa::schedule+ function.
-The \verb+ioa::schedule+ function can be called at any time.
+\paragraph*{Scheduling}
+Automata submit actions to the scheduler using the ioa::schedule function.
+The ioa::schedule function can be called at any time.
 However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
 A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
-For the \verb+asynch_lcr_automaton+, the scheduling function is:
+For the asynch\_lcr\_automaton, the scheduling function is:
 \begin{lstlisting}
@@
 \end{lstlisting}
-The various scheduling functions in the \verb+asynch_lcr_automaton+ and the constructor all dispatch to this function.
+The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this function.
 
@@
 
-\paragraph{Dynamic composition.}
-The \verb+asynch_lcr_automaton+ is a reusable component designed to be composed in the context of a larger system.
-Dynamic composition is accomplished by creating child automata and bindings via \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+ objects.
-The \verb+tcp_lcr_automaton+ discussed in section~\ref{case_study} creates a \verb+asynch_lcr_automaton+ and binds to it with the following code:
+\paragraph*{Dynamic composition}
+The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
+Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
+The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
 \begin{lstlisting}
-    ioa::automaton_manager<asynch_lcr_automaton<uuid> >* lcr =
-       ioa::make_automaton_manager (this,
-         ioa::make_generator<asynch_lcr_automaton<uuid> > (m_u));
+ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
+lcr = ioa::make_automaton_manager (this,
+  ioa::make_generator<asynch_lcr_automaton<uuid> > (
+    m_u));
 ioa::make_binding_manager (this,
@@
 \end{lstlisting}
-Children automata are managed by \verb+ioa::automaton_manager+ objects which can be created via the \verb+ioa::make_automaton_manager+ function.
-An automaton manager object requires an \verb+ioa::automaton+ object for its system action implementation and a generator for allocating the child automaton.
-Bindings are managed by \verb+ioa::binding_manager+ objects which can created via the \verb+ioa::make_binding_manager+ function.
-A binding manager objects requires an \verb+ioa::automaton+ object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
-The \verb+ioa::automaton_manager+ class and \verb+ioa::binding_manager+ class contain methods for destroying and unbinding respectively.
-The \verb+m_self+ object of the example is a \verb+ioa::handle_manager+ which can be used to bind to automata that already exist.
-From the example, the \verb+tcp_lcr_automaton+ binds the \verb+asynch_lcr_automaton+ to itself.
-The wary reader might notice that the \verb+ioa::automaton_manager+ and \verb+ioa::binding_manager+s in the example are dynamically allocated but never recorded and thus never freed.
-Both classes are managed by the \verb+ioa::automaton+ and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.
+Children automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
+An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating the child automaton.
+Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
+A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
+The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
+The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
+From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
+The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
+Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.
 
-\paragraph{Binding predicates.}
+\paragraph*{Binding predicates}
 Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
-The binding predicate supplied by ioa++ is the \verb+ioa::binding_count+ function which returns the number of actions bound to the specified action.
-The value returned by \verb+ioa::binding_count+ is zero for internal actions, zero or one for input actions, and non-negative for output actions.
-Most often, \verb+ioa::binding_count+ is used in the precondition of outputs to ensure that the value generated by the output will be received.
-This type of usage can be seen in the send action of the \verb+asynch_lcr_automaton+:
+The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of actions bound to the specified action.
+The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
+Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received.
+This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
 \begin{lstlisting}
@@
   return !m_send.empty () &&
-      ioa::binding_count (&asynch_lcr_automaton::send) != 0;
+    ioa::binding_count (&asynch_lcr_automaton::send)
+      != 0;
 }
@@
 
-\paragraph{System action events.}
+\paragraph*{System action events}
 Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
@@
 Similarly, an automaton might compose with an automaton that it does not manage and therefore expects the automaton to exist before attempting to bind.
-To enable such behavior, user automata can observe \verb+ioa::automaton_manager+ objects and \verb+ioa::binding_manager+ objects to receive events generated upon the completion of a system action.
+To enable such behavior, user automata can observe ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.
 
-\paragraph{Error reporting.}
+\paragraph*{Error reporting}
 The event semantics of I/O automata provide a natural mechanism for reporting errors.
@@
 
-\paragraph{Handling errors by resetting.}
+\paragraph*{Handling errors by resetting}
 An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
@@
 
-\paragraph{Handling errors by replacing.}
+\paragraph*{Handling errors by replacing}
 The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
@@
 For reasons already discussed, the bind must occur after the create.
-This constraint is enforced by the \verb+ioa::binding_manager+ class.
+This constraint is enforced by the ioa::binding\_manager class.
 Additionally, the destroy must occur before the first bind to ensure that all bindings associated with the first child have been dissolved.

### 2011-09-22 papers/paper1/evaluation.tex
Revision to evaluation.

@@
 The output and internal action execute a routine whose complexity is proportional to the parameter $N$.
-The automaton executes a fixed number of local actions which was set to 1,000 for all experiments.
+The automaton executes a fixed number of local actions.
 The automaton schedules the internal action with probability $\sqrt{\rho}$ and schedules the output action otherwise.
-The system $S$ to be executed consists of two $R$ automata composed so the output action of one is composed with the input action of another.
+The system $S$ to be executed consists of two $R$ automata composed so the output action of one is composed with the input action of the other.
 If we divide the execution into rounds where each automaton executes a single action, the probability that both $R$ automata execute an internal action is $\sqrt{\rho} \times \sqrt{\rho} = \rho$.
@@
 When two threads are used, one thread executes the actions of one $R$ automaton while the other thread executes the actions of the other $R$ automaton.
+
+A trial consists of a choice for $\rho$, $N$, and the number of actions to be executed.
+For the experiments, the parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$ and the parameter $N$ was varied from $1$ to $1000000$ by factors of $10$.
+The number of actions executed by each automaton was fixed at 1000.
+Each trial was repeated 30 times.
+All calculations were performed assuming a normal distribution at a confidence level of 95\%.
+
+The driver for a single trial can be found in examples/random.cpp of the ioa++ package.
+The trials were performed on a MacBook Pro running OS X 10.5.8 with an 2.4 GHz Intel Core 2 Duo processor with 4GB of memory.
+The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.
 
@@
 Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
-The parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$.
-The parameter $N$ was varied from $1$ to $1000000$ in factors of $10$.
 When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
@@
 Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
+(Note that the confidence intervals are present but small.)
 Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
@@
 This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
-The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the \verb+ioa::schedule+ call.
-From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4211.5ns with a standard deviation of 765ns for the single threaded experiments and 4858ns with a standard deviation of 1099ns for the multi-threaded experiments.
-Similary, we calculated the scheduling overhead and found it to be 227.5ns with a standard deviation of 62ns for the single threaded experiments and 27.9ns with a standard deviation of 79ns for the multi-threaded experiments.
+The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
+From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4211ns $\pm$ 31ns for the single threaded experiments and 4858ns $\pm$ 32ns for the multi-threaded experiments.
+Similary, we calculated the scheduling overhead and found it to be 227ns $\pm$ 2.5ns for the single threaded experiments and 279ns $\pm$ 2.3ns for the multi-threaded experiments.
 Both of these can be improved with better scheduler design.

### 2011-09-23 papers/paper1/component_model.tex
Related work.

@@
 In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
-The Input/Output (I/O) automata model was developed by Lynch to model reactive concurrent and asynchronous systems~\cite{lynch1996distributed}.
+The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
 An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.

### 2011-09-23 papers/paper1/related_work.tex
Updated abstract and introduction.
Related work.

@@
 
-pthreads (?)
-Edward Lee - The Problem with Threads
-Herb Sutter - The Free Lunch is Over
-Early work on concurrency
+This work grew out of the frustration encountered when trying to develop distributed systems with threads.
+Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
+These frustrations will increase as concurrent programming is the only foreseeable way to increase performance given the leveling trend of clock frequencies and the advent of multi-core processors\cite{sutter2005software}.
+
+The idea to implement a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
+The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features we believe to be fundamental to developing distributed systems.
+\cite{lynch1996distributed} shows how I/O automata can be applied to a number of problems in distributed computing.
+Event-based proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} can be used to reason about I/O automata.
+
+Comparing I/O automata to every model for asynchronous, concurrent, and distributed computing is beyond the scope of this paper so we focus on exemplars.
+The I/O automata model is based on state transitions which matches the imperative style that dominates modern concurrent programming.
+This is similar to the motivation for UNITY~\cite{chandy1988parallel} and is in contrast to functional models such as the $\pi$-calculus~\cite{milner1992calculus} and actors~\cite{agha1986actors}.
+The I/O automata model assumes independent state with explicit communication like communicating sequential processes (CSP)~\cite{hoare1978communicating} but unlike threads~\cite{lee2006problem} and the UNITY model~\cite{chandy1988parallel}.
+All models for asynchronous, concurrent, and distributed computing admit non-determinism.
+Lee points out that the difficulties of thread-based programming come from a need to ``prune'' non-determinism when paired with shared state and adopts the view that software components should be deterministic with respect to concurrency save a few that introduce non-determinism in a controlled way~\cite{lee2006problem}.
+Programs in UNITY are composed by concatenating program texts and resolving shared variables~\cite{chandy1988parallel}.
+For this reason, they share many of the same problems as threads under composition.
+I/O automata can be viewed as taking the alternate approach by admitting non-determinism while prohibiting shared state.
+In I/O automata, the sequential flow control of threads~\cite{lee2006problem} and CSP~\cite{hoare1978communicating} is replaced by the non-deterministic execution of conditional atomic actions which resembles execution in the UNITY model~\cite{chandy1988parallel}.
+The ability to create new automata can be compared to actor creation in the actor model~\cite{agha1986actors}.
+Actors communicate asynchronously by name using a buffered mail system while I/O automata communicate synchronously and anonymously according to composition.
+
+Ioa++ can be compared directly to the the IOA language~\cite{garland2003ioa} developed by Lynch et al.
+IOA is programming language for I/O automata that can be used with formal verification tools.
+Recently, a compiler for IOA has been developed~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} and used it to implement a number of protocols~\cite{georgiou2009automated}.
+\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
+Programming written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
+IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
+IOA lacks support for procedure calls which complicates modeling and interfacing with external services; especially when a procedure call blocks.
+The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
+The next action to be executed is determined by a schedule function that resolves the non-determinism in the schedule.
+
+IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
+The development of a new language was necessary for IOA due to their focus on formal verification.
+Ioa++ on the other hand uses the existing C++ language and compilers.
+IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors relying on the programmer to ensure non-blocking operation.
+IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
+Both IOA and ioa++ require the programmer to add code to schedule actions.
+
+%% CORBA thread model
+
+%% %% Lee~\cite{lee2006problem} also identifies a number of strategies for coping with thread-based programming including libraries, patterns, programming languages, and coordination languages.
+%% %% In~\cite{schmitd2000patterns}, Schmidt et al. describe patterns for developing asynchronous and concurrent objects and systems.
+%% %% The Split-C~\cite{culler1993parallel} and Cilk~\cite{blumofe1995cilk} add features to the C programming language for multi-threaded computation.
+%% %% The Erlang~\cite{armstrong1996concurrent}
+
+%% Ada
+
+%% The
+
+%% Events
+
+%% Lynch
+
+%% Garland
+
+%% Ken Goldman
+
+%% proactor/reactor
+%% Observer

### 2011-09-23 papers/paper1/design.tex
Removed listings and revised evaluation for 1000 trials.

@@
 
-An implementation of the component model described in Section~\ref{component_model} must provide a service for executing local actions and facilities for manipulating the set of automata and bindings.
+Ioa++ is an implementation in C++ of the component model described in Section~\ref{component_model} for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.
+Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
 The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
@@
 Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.
-
-\subsection{Implementation}
-
-Ioa++ is implemented in C++ for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.

### 2011-09-23 papers/paper1/evaluation.tex
Removed listings and revised evaluation for 1000 trials.

@@
 The number of actions executed by each automaton was fixed at 1000.
-Each trial was repeated 30 times.
+Each trial was repeated 1000 times.
 All calculations were performed assuming a normal distribution at a confidence level of 95\%.
@@
 The driver for a single trial can be found in examples/random.cpp of the ioa++ package.
-The trials were performed on a MacBook Pro running OS X 10.5.8 with an 2.4 GHz Intel Core 2 Duo processor with 4GB of memory.
+The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
 The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.
@@
 Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
+Confidence intervals are shown with largest being 0.0165.
 When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
@@
 Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
-(Note that the confidence intervals are present but small.)
+Confidence intervals are shown with largest being 0.002155 for Figure~\ref{overhead1} and 0.001781 for Figure~\ref{overhead2}.
 Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
@@
 The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
-From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4211ns $\pm$ 31ns for the single threaded experiments and 4858ns $\pm$ 32ns for the multi-threaded experiments.
-Similary, we calculated the scheduling overhead and found it to be 227ns $\pm$ 2.5ns for the single threaded experiments and 279ns $\pm$ 2.3ns for the multi-threaded experiments.
+From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4316ns $\pm$ 3.5ns for the single threaded experiments and 6067ns $\pm$ 3.6ns for the multi-threaded experiments.
+Similary, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
 Both of these can be improved with better scheduler design.

### 2011-09-23 papers/paper1/programming_model.tex
Removed listings and revised evaluation for 1000 trials.

@@
 The following code declares the asynch\_lcr\_automaton:
+\ifjournal
 \begin{lstlisting}
@@
 \end{lstlisting}
+\fi
 The asynch\_lcr\_automaton is declared as a template where the UID template parameter must model the unique identifier concept.
@@
 The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
+\ifjournal
 \begin{lstlisting}
@@
 \end{lstlisting}
+\fi
 Notice that the member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
@@
 For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue:
+\ifjournal
 \begin{lstlisting}
@@
 \end{lstlisting}
+\fi
 The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function schedule.
@@
 For example, the send action of the asynch\_lcr\_automaton is declared:
+\ifjournal
 \begin{lstlisting}
@@
 \end{lstlisting}
+\fi
 Input actions are declared similarly save the absence of a precondition.
@@
 For the asynch\_lcr\_automaton, the scheduling function is:
+\ifjournal
 \begin{lstlisting}
@@
 \end{lstlisting}
+\fi
 The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this function.
@@
 The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
+\ifjournal
 \begin{lstlisting}
@@
 \end{lstlisting}
+\fi
 Children automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
@@
 This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
+\ifjournal
 \begin{lstlisting}
@@
 \end{lstlisting}
+\fi
 

### 2011-09-24 papers/paper1/component_model.tex
New introduction.

@@
 
-Modularity, well-defined interfaces, and composition are essential properties of reusable software.
-A unit of software that exhibits all of these properties is called a \emph{component}~\cite{szyperski2002component}.
-Modularity implies that a component can be deployed independently of other components.
-Well-defined interfaces and interactions allow components to expose their functionality in a regular way that facilitates reasoning about different interactions among them.
-Composition allows a group of interacting components to be understood as a cohesive unit.
+%% Modularity, well-defined interfaces, and composition are essential properties of reusable software.
+%% A unit of software that exhibits all of these properties is called a \emph{component}~\cite{szyperski2002component}.
+%% Modularity implies that a component can be deployed independently of other components.
+%% Well-defined interfaces and interactions allow components to expose their functionality in a regular way that facilitates reasoning about different interactions among them.
+%% Composition allows a group of interacting components to be understood as a cohesive unit.
 

### 2011-09-24 papers/paper1/design.tex
New introduction.

@@
 
-Ioa++ is an implementation in C++ of the component model described in Section~\ref{component_model} for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.
+ioa++ is an implementation in C++ of the component model described in Section~\ref{component_model} for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.
 Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.

### 2011-09-24 papers/paper1/paper.tex
Related work.
New introduction.

@@
 
-\title{Implementing Distributed Systems Directly with I/O Automata in Ioa++}
+\title{Implementing Distributed Systems Directly with I/O Automata in ioa++}
 \author{
@@
   \IEEEauthorblockA{Department of Computer Science and Engineering\\
-    Campus Box 1045, Washington University\\
+    Washington University in St. Louis\\
+    Campus Box 1045\\
     One Brookings Drive\\
@@
   \IEEEauthorblockN{Guia-Catalin Roman}
-  \IEEEauthorblockA{TODO:  Dr. Roman's address}
+  \IEEEauthorblockA{School of Engineering\\
+    University of New Mexico\\
+    Centennial Engineering Center, Suite 3071\\
+    1 University of New Mexico, , MSC01 1140\\
+    Albuquerque, NM 87131-0001\\
+    gcroman@unm.edu}
 }

### 2011-09-24 papers/paper1/related_work.tex
New introduction.

@@
 
-Ioa++ can be compared directly to the the IOA language~\cite{garland2003ioa} developed by Lynch et al.
+ioa++ can be compared directly to the the IOA language~\cite{garland2003ioa} developed by Lynch et al.
 IOA is programming language for I/O automata that can be used with formal verification tools.
@@
 The development of a new language was necessary for IOA due to their focus on formal verification.
-Ioa++ on the other hand uses the existing C++ language and compilers.
+ioa++ on the other hand uses the existing C++ language and compilers.
 IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors relying on the programmer to ensure non-blocking operation.

### 2011-09-26 papers/paper1/abstract.tex
Incorporating Dr. Roman's revisions.

@@
 We develop a component model for asynchronous and concurrent systems based on the I/O automata formal model and dynamic composition.
-We describe the design and implementation of ioa++ based on that system model and show how ioa++ can be used to develop protocols and other forms of distributed systems.
+We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed systems.
 
@@
 \end{abstract}
-
-\begin{IEEEkeywords}
-TODO; component; formatting; style; styling;
-\end{IEEEkeywords}

### 2011-09-26 papers/paper1/component_model.tex
Incorporating Dr. Roman's revisions.

@@
 
-In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
+\subsection{I/O Automata\label{io_automaton}}
+
+In this subsection we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
 The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
@@
 Often, a local action consists of a precondition and an effect.
-The precondition is a predicate over the state variables that enables the effect when selected.
+The precondition is a predicate over the state variables that enables the effect when selected for execution.
 The effect is a function that computes the next state of the automaton from the current state.
@@
 
-I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
-Each action is only allowed to modify the state of the automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
-The absence of shared state thus ensures that each automaton can be deployed independently.
-The external signature of an I/O automaton constitutes its interface and the result of interacting with an I/O automaton is well-defined since every action is atomic.
-Composition is guaranteed by the model itself.
-
-\subsection{Practical Considerations\label{practical}}
-
 \begin{figure}
@@
 \end{figure}
-
-As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
-However, it lacks certain features needed to develop real components.
-This subsection introduces these features and lays the foundation for adding the dynamics discussed in Section~\ref{dynamics}.
 
@@
 
-Applying recursive encapsulation to I/O automata means that an I/O automaton may contain and use other I/O automata.
-Recursive encapsulation implies that a parent automaton is the result of composing its children.
-This definition excludes a parent from having any state or actions of its own.
-Since it is often convenient to associate state and actions with a parent, we allow a parent to be the composition of its children and an anonymous child that contains the state and actions attributed to the parent.
+Hierarchical decomposition, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
+For convenience, we extend the notion that a parent automaton is the composition of its children to include an anonymous child that contains state and actions attributed to the parent.
 The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.
@@
 Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
-Using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
+However, when using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
 To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
@@
 
+\ifjournal
 These observations create some interesting opportunities for designing and analyzing concurrent software.
@@
 Since the sets are disjoint, the two actions can be executed concurrently.
+\fi
+
+I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
+Each action is only allowed to modify the state of the automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
+The absence of shared state thus ensures that each automaton can be deployed independently.
+The external signature of an I/O automaton constitutes its interface and the result of interacting with an I/O automaton is well-defined since every action is atomic.
+Composition, asynchrony, and concurrency are guaranteed by the model itself.
+
+%% \subsection{Practical Considerations\label{practical}}
+
+%% As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
+%% However, it lacks certain features needed to develop real components.
+%% This subsection introduces these features and lays the foundation for adding the dynamics discussed in Section~\ref{dynamics}.
 
@@
 Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
-To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}\footnote{Use the actual paper.
-I think it has something to do with databases.}.
+To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.
 
@@
 If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
-Consequently, we add the ability to deal with a dynamic set of automata and interactions.
+Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not appropriate for \emph{implementing} a dynamic system of I/O automata.
+%% Consequently, we add the ability to deal with a dynamic set of automata and interactions.
+
+To illustrate, let a system $S=(A,B)$ be a pair consisting of a set of automata $A$ and set of bindings $B$.
+The I/O automata model assumes that $S$ and therefore $A$ and $B$ are both fixed.
+Dynamics are simulated via the manipulation of activation flags.
+A dynamic system implies that the system $S$ be dependent on time $t$ measured by the number of actions that have been executed.
+Thus, $S(t+1)$ need not be equal to $S(t)$.
+A \emph{system action} is an action that \emph{can} change the system $S$.
 
 \paragraph*{System actions}
-The four indivisible operations for managing a dynamic set of automata are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
-Create, bind, unbind, and destroy are collectively called \emph{system actions}.
+We define four indivisible operations for managing a dynamic system of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
 The \emph{create} operation allows an automaton to create a child automaton.
@@
 A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
-A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{practical} would be violated.
+A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{io_automaton} would be violated.
 An unbind fails if the binding does not exist.
@@
 The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the set defined in Section~\ref{practical} \emph{and} and the system automaton.
-However, the opportunities for concurrency explored in Section~\ref{practical} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+Thus, the set of actions implied by any action is the set defined in Section~\ref{io_automata} \emph{and} and the system automaton.
+However, the opportunities for concurrency explored in Section~\ref{io_automata} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
 

### 2011-09-26 papers/paper1/evaluation.tex
New introduction.
Incorporating Dr. Roman's revisions.

@@
 
-Ioa++ permits independent actions to be executed concurrently.
-To test this capability, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
+ioa++ permits independent actions to be executed concurrently.
+The goal of this evaluation is (1) to show the concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
+To test concurrent execution, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
 We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
 We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
-We also vary the complexity of each action to gain insight into how synchronization affects concurrent execution.
+We also vary the complexity of each action to gain insight into how overhead and synchronization affect concurrent execution.
 

### 2011-09-26 papers/paper1/paper.tex
Incorporating Dr. Roman's revisions.

@@
     St. Louis, MO 63130\\
-    {wilsonj,cdgill}@cse.wustl.edu}
+    \{wilsonj,cdgill\}@cse.wustl.edu}
   \and
-  \IEEEauthorblockN{Guia-Catalin Roman}
+  \IEEEauthorblockN{Gruia-Catalin Roman}
   \IEEEauthorblockA{School of Engineering\\
@@
     Centennial Engineering Center, Suite 3071\\
-    1 University of New Mexico, , MSC01 1140\\
+    1 University of New Mexico, MSC01 1140\\
     Albuquerque, NM 87131-0001\\
@@
 \input{abstract}
+
+\begin{IEEEkeywords}
+concurrency; I/O automata; framework; components
+\end{IEEEkeywords}
 

### 2011-09-26 papers/paper1/related_work.tex
Incorporating Dr. Roman's revisions.

@@
 
-ioa++ can be compared directly to the the IOA language~\cite{garland2003ioa} developed by Lynch et al.
-IOA is programming language for I/O automata that can be used with formal verification tools.
-Recently, a compiler for IOA has been developed~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} and used it to implement a number of protocols~\cite{georgiou2009automated}.
+To our knowledge, only the Spectrum Simulation System of Goldman~\cite{goldman1990distributed} and the IOA language developed by Lynch et al.~\cite{garland2003ioa} are the only existing systems that allow one to execute a system expressed as a collection of I/O automata.
+Spectrum focuses on algorithm development through simulation.
+IOA focuses on formal software development including both simulation and verification.
+
+ioa++ can be compared directly to the IOA language due to the addition of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
 \cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''

### 2011-09-27 papers/paper1/abstract.tex
General revision and added conclusion.

@@
 The size and complexity of these systems implies that software re-use in their development, maintenance, and evolution is also essential.
-Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly and easily in reusable software.
+Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly in reusable software.
 % By ``different forms'' we mean single-threaded, multi-threaded, time triggered, event triggered, etc.
@@
 To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
-We develop a component model for asynchronous and concurrent systems based on the I/O automata formal model and dynamic composition.
+We develop a component model for asynchronous and concurrent systems based on I/O automata and dynamic composition.
 We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed systems.
@@
 
-We evaluate the ability to execute actions concurrently in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
+We evaluate concurrency in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
 \end{abstract}

### 2011-09-27 papers/paper1/case_study.tex
General revision and added conclusion.

@@
 The listing in~\cite{lynch1996distributed} consists of approximately 24 lines in a formal language that must be translated to a production language.
-The raw listing for the final version consists of 121 lines (5x).
+The raw listing for the final version consists of 121 lines (\~5x).
 Scheduling accounts for 22 of the 121 lines (20\%) and is the main source of code that cannot be traced back to the formal listing.
@@
 The simulation creates a unidirectional ring of a specified size consisting of alternating asynch\_lcr\_automaton and channel\_automaton.
-(The channel\_automaton was also a direct translation from~\cite{lynch1996distributed}.)
+The channel\_automaton was also a direct translation from~\cite{lynch1996distributed}.
 

### 2011-09-27 papers/paper1/component_model.tex
General revision and added conclusion.

@@
 
-\subsection{I/O Automata\label{io_automaton}}
+In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
+%% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
+We then describe extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time.
 
-In this subsection we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
+\subsection{Background: I/O Automata}
+
 The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
@@
 An input action changes the state of the automaton when it receives a signal or value produced by an output action.
+\ifjournal
 The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.
-
+\fi
 An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
+\ifjournal
 Often, a local action consists of a precondition and an effect.
@@
 An input action presented in this style consists solely of an effect.
+\fi
+An automaton's output and input actions constitute its \emph{external signature}~\cite{lynch1996distributed}.
 
-Together, an automaton's output and input actions constitute its \emph{external signature}~\cite{lynch1996distributed}.
 An automaton communicates with other automata through the actions in its external signature.
@@
 
+\subsection{Component Structure\label{component_structure}}
+
 \paragraph*{Recursive Encapsulation}
@@
 
-The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
+The event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
 The Processor automaton is the root and C and D are its children.
@@
 A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
-A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{io_automaton} would be violated.
+A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{component_structure} would be violated.
 An unbind fails if the binding does not exist.
@@
 The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the set defined in Section~\ref{io_automata} \emph{and} and the system automaton.
-However, the opportunities for concurrency explored in Section~\ref{io_automata} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+Thus, the set of actions implied by any action is the set defined in Section~\ref{component_structure} \emph{and} and the system automaton.
+However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
 

### 2011-09-27 papers/paper1/conclusion.tex
General revision and added conclusion.

@@
 
-\begin{itemize}
-  \item We are going to use it to build the substrate.
-  \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
-  \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
-  \item We can take advantage of multi-core in a very straight-forward way
-  \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
-  \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
-\end{itemize}
+I/O automata are a good foundation for asynchronous and concurrent components due to independent state, well-defined interfaces, and well-defined interactions under composition.
+The ioa++ framework facilitates concurrent execution via dynamic composition and the degree of concurrency is limited by the interactions of the automata and the overhead of the framework.
+The two key challenges when moving from the formal model to an actual implementation were introducing features for managing dynamic constellations of automata and providing a concrete scheduling mechanism.
+We believe that attempts to implement other formal models, e.g., UNITY, will have similar experiences and thus encounter the differences between theory and practice.
+We agree with~\cite{georgiou2009automated} that having a representation of a model that can be compiled is beneficial because it allows one to reason about the program directly from the source code and helps to close the gap between mental model and code listing.
+
+Our primary goal moving forward is to use the ioa++ framework to gain experience building real systems with I/O automata.
+The ioa++ has no support for static composition.
+We hope to resolve this weakness in the future.
+A significant open problem in ioa++ is the design of an efficient dispatcher and scheduler(s).
+
+%% \begin{itemize}
+%%   \item We are going to use it to build the substrate.
+%%   \item Speculate on moving down into operating system (device drivers would be easy, IPC including filesystem replaced by automata)
+%%   \item Speculate on moving down into the hardware level (Local talent, Ivan Sutherland)
+%%   \item We can take advantage of multi-core in a very straight-forward way
+%%   \item New problems in scheduling (Pinning automata to processors to minimize actions that span two processors.  Maximum independent set.)
+%%   \item We are non-blocking all the way.  Combine this with a deterministic implementation of the scheduler and model and there are serious opportunities for real-time.
+%%   \item Static composition
+%%   \item Grandparents
+%%   \item We own the event loop
+%%   \item The procedure
+%% \end{itemize}

### 2011-09-27 papers/paper1/design.tex
General revision and added conclusion.

@@
 
-ioa++ is an implementation in C++ of the component model described in Section~\ref{component_model} for POSIX environments and is available at \url{http://jrwilson.github.com/ioa/}.
+ioa++ is an implementation of the component model described in Section~\ref{component_model} for POSIX environments in C++.
 Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
@@
 Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
-All other actions acquire a read lock to implement single-writer/multiple reader semantics.
+All other actions acquire a read lock to implement single-writer/multiple-reader semantics.
 The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
@@
 This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
-For our framework, we used the Reactor pattern since (1) the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.
+For our framework, we used the Reactor pattern since (1)~the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.
 

### 2011-09-27 papers/paper1/evaluation.tex
General revision and added conclusion.

@@
 ioa++ permits independent actions to be executed concurrently.
-The goal of this evaluation is (1) to show the concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
+The goal of this evaluation is (1) to show that concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
 To test concurrent execution, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
@@
 Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
-The fraction decreases as $\rho$ goes to 1 as the system transitions from acquiring two locks to one lock.
+The fraction decreases as $\rho$ goes to 1 and the system transitions from acquiring two locks to one lock.
 The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.

### 2011-09-27 papers/paper1/programming_model.tex
General revision and added conclusion.

@@
 All automata inherit from the common base class ioa::automaton which implements the system action interface.
+\ifjournal
 The following code declares the asynch\_lcr\_automaton:
-\ifjournal
 \begin{lstlisting}
@@
 The state variables of an automaton are expressed as member variables of the corresponding class.
-The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status:
+The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status.
 \ifjournal
@@
 \fi
-Notice that the member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
+Member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
 State variables are initialized using a constructor.
-For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue:
+For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue.
 \ifjournal
@@
 Actions are member variables that model the action concept.
-Actions contain methods for preconditions, effects, and scheduling and traits that indicate the actions's type (input, output, internal), value status, value type, parameter status, and parameter type.
+Actions contain methods for preconditions, effects, and scheduling and traits that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
 The traits are used to guide the execution of the methods by the model and supply values and parameters as necessary.
@@
 
-To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
+A number of wrappers and macros are available for declaring actions.
+%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
 Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
 The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
+\ifjournal
 For example, the send action of the asynch\_lcr\_automaton is declared:
-\ifjournal
 \begin{lstlisting}
@@
 A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
+\ifjournal
 For the asynch\_lcr\_automaton, the scheduling function is:
-\ifjournal
 \begin{lstlisting}
@@
 Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
+\ifjournal
 The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
-\ifjournal
 \begin{lstlisting}
@@
 The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
+\ifjournal
 The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
@@
 Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.
+\fi
 
@@
 Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received.
+\ifjournal
 This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
-\ifjournal
 \begin{lstlisting}

### 2011-09-27 papers/paper1/related_work.tex
General revision and added conclusion.

@@
 
-This work grew out of the frustration encountered when trying to develop distributed systems with threads.
+This work grew out of the challenges faced when trying to develop distributed systems with threads.
 Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
-These frustrations will increase as concurrent programming is the only foreseeable way to increase performance given the leveling trend of clock frequencies and the advent of multi-core processors\cite{sutter2005software}.
+These same concerns are relevant to trends towards increasingly concurrent programming to increase performance, given the leveling trend of clock frequencies and the advent of multi-core processors\cite{sutter2005software}.
 
-The idea to implement a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
-The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features we believe to be fundamental to developing distributed systems.
-\cite{lynch1996distributed} shows how I/O automata can be applied to a number of problems in distributed computing.
-Event-based proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} can be used to reason about I/O automata.
+The direction implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
+The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features essential to developing distributed systems~\cite{lynch1996distributed}.
+Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.
 
@@
 I/O automata can be viewed as taking the alternate approach by admitting non-determinism while prohibiting shared state.
-In I/O automata, the sequential flow control of threads~\cite{lee2006problem} and CSP~\cite{hoare1978communicating} is replaced by the non-deterministic execution of conditional atomic actions which resembles execution in the UNITY model~\cite{chandy1988parallel}.
+In I/O automata, the sequential flow control seen with threads~\cite{lee2006problem} and CSP~\cite{hoare1978communicating} is replaced by the non-deterministic execution of conditional atomic actions which resembles execution in the UNITY model~\cite{chandy1988parallel}.
 The ability to create new automata can be compared to actor creation in the actor model~\cite{agha1986actors}.
-Actors communicate asynchronously by name using a buffered mail system while I/O automata communicate synchronously and anonymously according to composition.
+Actors communicate by name using a buffered mail system while communications in I/O automata are not buffered and anonymous.
 
-To our knowledge, only the Spectrum Simulation System of Goldman~\cite{goldman1990distributed} and the IOA language developed by Lynch et al.~\cite{garland2003ioa} are the only existing systems that allow one to execute a system expressed as a collection of I/O automata.
+To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing systems that allow one to execute a system expressed as a collection of I/O automata.
 Spectrum focuses on algorithm development through simulation.
@@
 
-ioa++ can be compared directly to the IOA language due to the addition of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
+The ioa++ framework can be compared directly to the IOA language due to the addition of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
 \cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
@@
 IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
-Both IOA and ioa++ require the programmer to add code to schedule actions.
+Both IOA and ioa++ require the programmer to add code to schedule actions and find this to be a common source of errors.
 

### 2011-09-30 papers/paper1/abstract.tex
New case study and related work.

@@
 We develop a component model for asynchronous and concurrent systems based on I/O automata and dynamic composition.
-We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed systems.
+We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed software.
 

### 2011-09-30 papers/paper1/case_study.tex
New case study and related work.

@@
-\section{Case Study:  The TCP LCR Automaton\label{case_study}}
+\section{Example:  The TCP LCR Automaton\label{case_study}}
 
+In this section, we demonstrate that ioa++ can be used to build real distributed systems.
+Real distributed systems are designed using a layered approach where the upper layers contain semantically rich business logic while the bottom layers contain low-level network message-passing.
+The middle layers, i.e., middleware, contain generic high-level coordination mechanisms that abstract away the low-level details of message-passing.
+The lower layers contain abstractions of operating system services such as TCP/UDP sockets (for communicating with other peers) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
+We apply this layered architecture technique to the simple but representative network programming problem of electing a leader in a unidirectional ring of processes connected by TCP sockets.
+
+The ``business logic'' we will implement is quite simple and consists of periodically circulating a token around a ring of nodes.
+If a node has not received a token in a specified amount of time, it initiates a leader election protocol.
+Once a leader has been elected, the leader injects a token and the protocol resumes.
+Nodes can join and leave the ring at any time.
+The protocol should eventually succeed for complete rings.
+These dynamics admit fault situations where a ring might lose its leader or contain multiple leaders.
+A ring containing multiple leaders should stabilize to having a single leader.
+A rare (but possible and valid) situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
+This situation requires no action as a token is circulating in the ring.
+We divide the problem into two sub-problems:  leader election and ring formation.
+
+Nodes in a distributed system may be arranged in a ring so that two consecutive nodes are connected by a directed channel.
+The goal is to designate one of the nodes as the leader of the ring.
+The solution we adopt is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
+The LCR algorithm assumes that every node has a unique identifier (UID).
+When the protocol begins, each node sends its UID to its successor.
+When a node receives a UID from its predecessor that is greater than its own, it forwards the UID to its successor.
+When a node receives its own UID, it elects itself the leader of the ring.
+We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
+Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}.} of the ioa++ package.
+
+A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
+The tcp\_ring\_automaton listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and channel to the successor.
+The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
+The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{constellations} to manage the dynamic constellation of automata and recover from errors.
+The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.
+
+\begin{figure}
+\center
+\includegraphics[width=\columnwidth]{tcp_lcr_automaton}
+\caption{Architecture of the TCP LCR system.}
+\label{tcp_lcr_automaton}
+\end{figure}
+
+The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
+Figure~\ref{tcp_lcr_automaton} shows the layered architecture of the TCP LCR system.
+The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the business logic.
+Thus, ioa++ supports features we expect from a component framework for distributed systems.
+
+%% The key idea is that all of these modules have well-defined interactions when composed, while being inherently concurrent.
+%% One need only focus on the protocol implemented by an automaton when composing it with an existing system, as opposed to worrying about interacting threads and race conditions.
+
+\ifjournal
 In this section, we explore whether ioa++ offers a suitably convenient representation for reifying I/O automata and contains the features necessary for building real distributed systems.
 Our approach is to (1) express an automaton in ioa++ using an existing listing, (2) simulate a distributed system built using the automaton as debugging step, and (3) build a real distributed system by composing the automaton with automata for network services.
-For continuity, we use the asynchronous LCR automaton developed in section~\ref{representation}.
+For continuity, we discuss further the asynchronous LCR automaton described in Section~\ref{programming_model}.
 
 \paragraph*{Translating the asynchronous LCR automaton into ioa++}
-The first version of the code for the asynchronous LCR automaton was translated directly from a listing in~\cite{lynch1996distributed}.
-The final version contains an additional action to reinitialize the automaton and modifications that allow an automaton that elected itself leader to nullify its election.
+The first version of the asynchronous LCR automaton was translated directly from a listing in~\cite{lynch1996distributed}, but was then refined with (1) an additional action to reinitialize the automaton and (2) modifications that allow an automaton that elected itself the leader to nullify its election.
 The motivation for these modifications will be explained when we discuss its use in a real system.
 The listing in~\cite{lynch1996distributed} consists of approximately 24 lines in a formal language that must be translated to a production language.
-The raw listing for the final version consists of 121 lines (\~5x).
+The raw listing for our final version consists of 121 lines (\~5x).
 Scheduling accounts for 22 of the 121 lines (20\%) and is the main source of code that cannot be traced back to the formal listing.
@@
 
-We observe that having a representation of I/O automata in a production language that closely matches a formal representation is of great benefit because it allows one to reason about an automaton directly from the source code.
+We observe that having a representation of I/O automata in a production language (such as ioa++) which also closely matches a formal representation is of great benefit because it thus encourages one to reason about an automaton directly from the source code.
 For example, an invariant in the asynchronous LCR automaton is that all of the UIDs in the send queue of automaton $i$ must be greater than or equal to the UID of automaton $i$.
@@
 The send action removes a UID from the send queue.
-Reasoning about an automaton directly from the source code allows one to use formal methods to develop and debug systems.
+Reasoning about an automaton directly from the source code allows one to use formal methods to develop and debug systems \emph{at the point of implementation directly}.
 
@@
 After the asynchronous LCR automaton was implemented, we developed a simulation to test it.
-The examples/asynch\_lcr.cpp file contains the driver for the simulation.
+The examples/asynch\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/asynch_lcr.cpp}.} file in ioa++ package contains the driver for the simulation.
 The simulation creates a unidirectional ring of a specified size consisting of alternating asynch\_lcr\_automaton and channel\_automaton.
@@
 
-The simulation showed that the asynch\_lcr\_automaton was not correct.
-The first bug we encountered was a silent failure where the ring always failed to elect a leader.
-The problem was that the leader action was never being scheduled so the leader action was added to the scheduling function.
+The simulation showed that the initial asynch\_lcr\_automaton was not correct.
+The first bug we encountered was a silent failure where the ring failed to elect a leader.
+The problem was that the leader action was never being scheduled, and to fix this the leader action was added to the scheduling function.
 The second bug manifested itself by electing the wrong leader, i.e., one whose UID was not the maximum.
 A closer inspection showed that certain UIDs were being sent but never received.
-The key insight into solving the problem was to understand that the automata were executing as the ring was being constructed, i.e., we failed to consider dynamic composition.
+The key insight into solving the problem was to understand that the automata were executing as the ring was being constructed, i.e., we had failed originally to consider dynamic composition.
 The solution was to add binding predicates to the outputs of the asynch\_lcr\_automaton and channel\_automaton.
@@
 The protocol should eventually succeed for complete rings and restart when a ring becomes incomplete.
-These dynamics admit situations where a ring might lose its leader or contain multiple leaders.
+These dynamics admit fault situations where a ring might lose its leader or contain multiple leaders.
 A ring containing multiple leaders should stabilize to having a single leader.
 This is the motivation behind the asynch\_lcr\_automaton being able to withdraw its election.
-A rare but possible and valid situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
+A rare (but possible and valid) situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
+This situation requires no action as a token is circulating in the ring.
 
-The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp (484 lines) of the ioa++ package implements the protocol just described.
+The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} (484 lines) of the ioa++ package implements the protocol just described.
 The tcp\_lcr\_automaton is a composition of a number of automata including an asynch\_lcr\_automaton, an ioa::tcp\_acceptor\_automaton, an ioa::tcp\_connector\_automaton, and two ioa::tcp\_connection\_automaton representing the channel from the predecessor and channel to the successor.
-The TCP automata provided by the ioa++ package are ordinary automata implemented using file descriptors and the techniques described in section~\ref{design}.
-The tcp\_lcr\_automaton uses the techniques for managing dynamic constellations described in section~\ref{representation}.
+The TCP automata provided by the ioa++ package are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
+The tcp\_lcr\_automaton uses the techniques for managing dynamic constellations described in Section~\ref{programming_model}.
 
-The tcp\_lcr\_automaton is validation that I/O automata are a good foundation for building reusable concurrent modules.
-The key idea is that all of these modules have well-defined interactions when composed while being inherently concurrent.
+The tcp\_lcr\_automaton case study confirms that I/O automata can be a good foundation for building reusable concurrent modules.
+The key idea is that all of these modules have well-defined interactions when composed, while being inherently concurrent.
 For example, one could substitute the asynch\_lcr\_automaton with a different automaton to realize a different leader election protocol or a different protocol altogether.
 Alternatively, the tcp\_lcr\_automaton could be refactored to use a different communication mechanism, e.g., UDP.
-One need only focus on the protocol implemented by an automaton when composing it with an existing system as opposed to worrying about interacting threads and race conditions.
-
-%% \begin{figure}
-%% \center
-%% \includegraphics[width=\textwidth]{tcp_lcr_automaton}
-%% \caption{TCP LCR automaton.}
-%% \label{tcp_lcr_automaton}
-%% \end{figure}
-
-%% \paragraph{Dynamics in the TCP LCR automaton.}
-%% A \verb+tcp_lcr_automaton+ continuously attempts to complete the ring by accepting a connection from its predecessor or connecting to its successor.
-%% Accepting can fail if the socket used for listening is currently unavailable.
-%% Connecting can fail if the successor automaton does not exist and is therefore not listening for incoming connections.
-%% The connections themselves will fail when nodes leave the ring.
-%% The \verb+tcp_lcr_automaton+ also contains constraints on the order in which automata are created with respect to each other and with respect to which actions are executed.
-%% For example, the predecessor connection must be created before it is passed to the acceptor via the accept action.
-%% Similarly, the successor connection must be created before it is passed via a constructor to the connector.
+One need only focus on the protocol implemented by an automaton when composing it with an existing system, as opposed to worrying about interacting threads and race conditions.
+\fi
 

### 2011-09-30 papers/paper1/component_model.tex
New case study and related work.

@@
 An action is either an internal action, an output action, or an input action.
-An internal action changes the state of the automaton.
+An internal action only changes the state of the automaton.
 An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
@@
 The scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
-Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
+Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
 Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions,  subject to the precondition using the value produced by the output action.
@@
 
-\paragraph*{Recursive Encapsulation}
-Component frameworks that support recursive encapsulation allow components to contain other components.
+Component frameworks that support \emph{recursive encapsulation} allow components to contain other components.
 Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
@@
 
-Hierarchical decomposition, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
+\emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
 For convenience, we extend the notion that a parent automaton is the composition of its children to include an anonymous child that contains state and actions attributed to the parent.
@@
 Actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
-Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have neither suffix, e.g., \emph{process1}.
-The type associated with the value produced or consumed by an action is in parentheses after the name, e.g., \emph{out!(s)}.
+Output actions are suffixed with \emph{!} (e.g., \emph{detect!}), input actions are suffixed with \emph{?} (e.g., \emph{detected?}), and internal actions have neither suffix (e.g., \emph{process1}).
+The type associated with the value produced or consumed by an action is in parentheses after the name (e.g., \emph{out!(s)}).
 
@@
 A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
-For example, the fully-qualified name of the $in?(s)$ action of the Converter C in Figure~\ref{sys_model} is $[root].C.in?(s)$.
+For example, the fully-qualified name of the $in?(s)$ action of Converter C in Figure~\ref{sys_model} is $[root].C.in?(s)$.
 Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
-The map of bindings for the event detection system is 
+The map of bindings for the event detection system in Figure~\ref{sys_model} is: 
 \begin{displaymath}
@@
    ([root].C.out!(t), [root].D.in?(t)),\\
-   ([root].D.detect!, [root].detected?) \}
+   ([root].D.detect!, [root].detected?) \}.
 \end{split}
@@
 \paragraph*{Binding rules}
-To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
+To support name-based composition under the I/O automata model, the set of bindings must adhere to four essential rules.
 First, the output action and input action of a binding must agree on the type being produced and consumed.
 For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
-The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., a signal).
+The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., use of a signal).
 Second, an input action can be bound to at most one output action.
@@
 The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
-Assuming an countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
+Assuming a countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
 The set of interactions in the I/O automata model is also fixed since automata are statically composed.
-Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
+Dynamic systems are modeled in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
 To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.
 
-While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system results in over-provisioning or inflexibility.
+While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
 To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
@@
 If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
-Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not appropriate for \emph{implementing} a dynamic system of I/O automata.
+Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata directly.
 %% Consequently, we add the ability to deal with a dynamic set of automata and interactions.
 
-To illustrate, let a system $S=(A,B)$ be a pair consisting of a set of automata $A$ and set of bindings $B$.
+To illustrate, let a system $S=(A,B)$ be a pair consisting of a set of automata $A$ and a set of bindings $B$.
 The I/O automata model assumes that $S$ and therefore $A$ and $B$ are both fixed.
 Dynamics are simulated via the manipulation of activation flags.
-A dynamic system implies that the system $S$ be dependent on time $t$ measured by the number of actions that have been executed.
-Thus, $S(t+1)$ need not be equal to $S(t)$.
+A dynamic system implies that the system $S$ must be dependent on time $t$ as a function of the number of actions that have been executed by $t$.
+Thus, $S_{t+1}$ need not be equal to $S_{t}$.
 A \emph{system action} is an action that \emph{can} change the system $S$.
 
-\paragraph*{System actions}
-We define four indivisible operations for managing a dynamic system of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
-The \emph{create} operation allows an automaton to create a child automaton.
-The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
-The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.
+\begin{figure}
+\center
+\includegraphics[width=\columnwidth]{system_action}
+\caption{System actions.
+  A base automaton contains output actions for creating, binding, unbinding, and destroying.
+  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
+  The system automaton returns a result using the appropriate parameterized output action.
+  The base automaton contains input actions to receive the results.}
+\label{system_action}
+\end{figure}
+
+The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
+Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
+The system automaton, in turn, has input actions for receiving requests for system actions, internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
+We define four actions in the system automaton for managing a dynamic system of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
+The \emph{create} action allows an automaton to create a child automaton.
+The \emph{bind} action allows an automaton to bind an input action to an output action while the \emph{unbind} action allows an automaton to separate an input action from an output action.
+The \emph{destroy} action allows an automaton to recursively destroy a child automaton and all bindings associated with it.
+Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.
 
@@
 
-\begin{figure}
-\center
-\includegraphics[width=\columnwidth]{system_action}
-\caption{System actions.
-  A base automaton contains output actions for creating, binding, unbinding, and destroying.
-  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
-  The system automaton returns a result using the appropriate parameterized output action.
-  The base automaton contains input actions to receive the results.}
-\label{system_action}
-\end{figure}
-
-\paragraph*{The system automaton}
-The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
-Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
-The system automaton, in turn, has input actions for receiving requests for system actions, internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
-Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.
-
 Conceptually, the state variables $A$ and $B$ belong to the system automaton.
-However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
-The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the set defined in Section~\ref{component_structure} \emph{and} and the system automaton.
+However, the state encoded by $A$ and $B$ is shared and is used in the execution of every action.
+A \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
+Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
 However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
@@
 To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
-To do this, we take the conjunct of the precondition of an output action with a predicate over the set of bindings $B$ called a \emph{binding predicate} to ensure that the output action is not executed until its set of implied automata is correct.
-Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as previously discussed.
+To do this, we take the conjunction of the precondition of an output action with a predicate over the set of bindings $B$ (called a \emph{binding predicate}) to ensure that the output action is not executed until its set of implied automata is correct.
+Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as was previously discussed.
 Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
-A static system can be implemented by adding binding predicates and logic to create the required constellation of automata.
+A system defined statically can be implemented as a dynamic system adding binding predicates and logic to create the required constellation of automata.

### 2011-09-30 papers/paper1/conclusion.tex
New case study and related work.

@@
-\section{Conclusion and Future Work\label{conclusion}}
+\section{Conclusions\label{conclusion}}
 
-I/O automata are a good foundation for asynchronous and concurrent components due to independent state, well-defined interfaces, and well-defined interactions under composition.
+I/O automata offer a good foundation for asynchronous and concurrent components due to their support for independent state, well-defined interfaces, and well-defined interactions under composition.
 The ioa++ framework facilitates concurrent execution via dynamic composition and the degree of concurrency is limited by the interactions of the automata and the overhead of the framework.
 The two key challenges when moving from the formal model to an actual implementation were introducing features for managing dynamic constellations of automata and providing a concrete scheduling mechanism.
-We believe that attempts to implement other formal models, e.g., UNITY, will have similar experiences and thus encounter the differences between theory and practice.
 We agree with~\cite{georgiou2009automated} that having a representation of a model that can be compiled is beneficial because it allows one to reason about the program directly from the source code and helps to close the gap between mental model and code listing.
+Our primary goal moving forward is to use the ioa++ framework to gain further experience building real systems with I/O automata.
 
-Our primary goal moving forward is to use the ioa++ framework to gain experience building real systems with I/O automata.
-The ioa++ has no support for static composition.
-We hope to resolve this weakness in the future.
-A significant open problem in ioa++ is the design of an efficient dispatcher and scheduler(s).
+%% The ioa++ has no support for static composition.
+%% We hope to resolve this weakness in the future.
+%% A significant open problem in ioa++ is the design of an efficient dispatcher and scheduler(s).
 

### 2011-09-30 papers/paper1/design.tex
New case study and related work.

@@
 
-ioa++ is an implementation of the component model described in Section~\ref{component_model} for POSIX environments in C++.
+ioa++ implements the component model described in Section~\ref{component_model}, for POSIX environments (in C++).
 Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
-The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
+The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
 The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
 All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
-The services provided by the base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
+The services provided by that common base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
 The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
@@
 The controller encodes a \emph{policy} that selects the next local action.
-For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler should be \emph{fair} as defined in Section~\ref{component_model}.
+For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
 The run-time system can be initialized with different controllers to achieve different performance objectives.
+Controllers can use multiple threads of execution to exploit the concurrency available in a system of automata.
+The ioa++ framework is designed so that all native concurrency structures (e.g., threads, mutexes) are encapsulated in the dispatcher and controller.
 
@@
 The controller then has the responsibility of cycling through the actions according to some fair policy.
-Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to trying each action with brute force.
+Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to naively trying each action.
 
-The approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
+A more focused approached used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
 The controller then has the responsibility of processing the set of requested actions according to some fair policy.
@@
 Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
-Consequently, an automaton is given a opportunity to declare actions to the scheduler after one of its actions is executed.
-Since the state and therefore preconditions of all automata are independent, each automaton is responsible for scheduling its own actions.
+Consequently, an automaton is given the opportunity to declare actions to the scheduler after one of its actions is executed.
+Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
 For bootstrapping, each automaton is allowed to schedule actions when it is created.
@@
 Notice that refraining from scheduling disabled actions does not affect the fairness of the scheduler since they cannot cause a state change when selected.
-This result leads to an programming idiom where the precondition of each action is checked before it is added to the schedule.
+This result leads to a guarded programming idiom where the precondition of each action is checked before it is added to the schedule.
 
-In contrast, an automaton should ensure that the scheduler will consider all actions that are currently enabled.
-A simple technique for accomplishing this is to schedule all enabled actions.
+However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
+A simple technique for accomplishing this is to \emph{schedule} all enabled actions.
 More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
-Not declaring an enabled action violates the fairness of the controller since it will never select an action that may cause a state change.
-Not scheduling an (enabled) action is a common mistake.
+However, this must be done carefully.
+Not declaring an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
+Not scheduling an (enabled) action could be another potential mistake.
 
 \paragraph*{Binding predicates and scheduling}
-The preceding discussion on scheduling assumes that an action in one automaton cannot affect preconditions in another automaton.
+%% This assumption is not implicit because we state that the preconditions are independent.
+The preceding discussion on scheduling assumed that an action in one automaton cannot affect preconditions in another automaton.
 This is not entirely correct when one considers the system automaton and binding predicates.
@@
 While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
-To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a successful bind or unbind.
+To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after each successful bind or unbind.
 
@@
 Every action in the set of actions considered by the controller is scheduled by its associated automaton.
-Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
+Since automata can be destroyed, an action in the set owned by the controller might belong to an automaton that no longer exists.
 Our solution is to check if the automaton exists before executing an action.
 This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
-Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)} and that the set of actions in the scheduler turns over comparatively faster than the set of aids.
+Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)}, assuming that automata turn over less frequently overall than the set of actions in the scheduler.
 
 \paragraph*{Delayed actions}
-Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities.
-The framework should permit the development of automata that act as timers and alarms.
-The corresponding feature that must be supported by the framework is scheduling an action at a future time.
-Conceptually, the action is held until the time associated with the action at which point it is added to the set of actions being considered by the scheduler.
+Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities, i.e., the framework should support the development of automata that act as timers and alarms.
+Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
 Actions scheduled in the past or present are released immediately to the scheduler.
@@
 File descriptors are used to communicate with the outside world in a POSIX environment.
-Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
-Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
+Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects may adversely affect the asynchronous and concurrent nature of I/O automata.
+Thus, we require techniques that allow a program (1) to wait until a file descriptor is asynchronously ready for I/O and (2) to perform non-blocking I/O operations.
 This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
-For our framework, we used the Reactor pattern since (1)~the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.
+For our framework, we used the Reactor pattern since (1)~the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less widespread.
 
-Our implementation of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
+Our use of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
 A user automaton creates a file descriptor and configures it for non-blocking I/O.
@@
 \paragraph*{Global FIFO scheduler}
-The scheduler included with the framework is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
+The basic scheduler included with the ioa++ framework distribution is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
 Scheduling is idempotent to prevent unbounded growth of the queue.
@@
 
-\paragraph*{System action protocol}
 Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
 All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
-The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with either success or failure.
+The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with an indication of either success or failure.
 Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
@@
 When the user automaton requests that the child automaton be destroyed, the subject is either (1) added to the DestroySend state if the subject is in CreateRecv or CreateDone or (2) removed from all states if the subject is in CreateSend.
-A subject appearing in both CreateDone and DestroySend transitions to the DestroyRecv state when the automaton makes the request to destroy.
+A subject appearing in both CreateDone and DestroySend transitions to the DestroyRecv state when the automaton makes the request to destroy it.
 The system automaton can process system actions in any order.
 This protocol ensures that an automaton doesn't request that a child automaton be destroyed until it receives confirmation that it was created.
-The subject is forgotten when the system automaton responds with the result of the destroy.
+The subject is forgotten when the system automaton responds with the result of the destroy action.
 
 The protocol for binding is similar but addresses the possibility that a binding can be dissolved at any point due to the asynchronous destruction of automata.
-Define analogous states for the binding state machine: BindSend, BindRecv, BindDone, UnbindSend, and UnbindRecv.
+We define analogous states for the binding state machine: BindSend, BindRecv, BindDone, UnbindSend, and UnbindRecv.
 A binding subject can be in either BindRecv, BindDone, UnbindSend, or UnbindRecv when it receives a result from the system automaton indicating that the binding was dissolved.
@@
 \paragraph*{Automaton and Binding managers}
-The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
-The Automaton Managers and Binding Managers provide an individualized facade to the system action state machines.
+The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating more complicated ones.
+Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
 Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
-Automaton Managers can be observed using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
-Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.
+Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
+Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.
 

### 2011-09-30 papers/paper1/evaluation.tex
New case study and related work.

@@
 
-ioa++ permits independent actions to be executed concurrently.
-The goal of this evaluation is (1) to show that concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
-To test concurrent execution, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
+The ioa++ framework permits independent actions to be executed concurrently.
+The goal of this evaluation is to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising a system and the overhead of the framework.
+To test concurrent execution, we examine a system whose actions can be configured to span the range from having no independent actions to having only independent actions.
 We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
-We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
+We then calculate the speed-up by comparing an execution using one thread to an equivalent execution using two threads.
 We also vary the complexity of each action to gain insight into how overhead and synchronization affect concurrent execution.
@@
 An automaton $R$ contains an input, output, and internal action.
-The output and internal action execute a routine whose complexity is proportional to the parameter $N$.
+The output and internal action effects execute an algorithm whose complexity is proportional to the parameter $N$.
 The automaton executes a fixed number of local actions.
@@
 
-For this experiment, we developed a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action.
+For this experiment, we developed a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{scheduling}.
 When a single thread is used, all actions are executed by the same thread.
@@
 A trial consists of a choice for $\rho$, $N$, and the number of actions to be executed.
-For the experiments, the parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$ and the parameter $N$ was varied from $1$ to $1000000$ by factors of $10$.
+For our experiments, the parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$ and the parameter $N$ was varied from $1$ to $1000000$ by factors of $10$.
 The number of actions executed by each automaton was fixed at 1000.
@@
 
-The driver for a single trial can be found in examples/random.cpp of the ioa++ package.
+The driver for a single trial can be found in examples/random.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/random.cpp}.} of the ioa++ package.
 The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
@@
 Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
-Confidence intervals are shown with largest being 0.0165.
+Confidence intervals are shown but are negligible with the largest being a speed-up factor of $\pm$0.0165.
 When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
@@
 
-Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
-Confidence intervals are shown with largest being 0.002155 for Figure~\ref{overhead1} and 0.001781 for Figure~\ref{overhead2}.
+Figures~\ref{overhead1} and \ref{overhead2} show the average per thread of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions; confidence intervals are again negligible with largest being $\pm$0.002155 for Figure~\ref{overhead1} and $\pm$0.001781 for Figure~\ref{overhead2}.
 Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
@@
 Similary, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
-Both of these can be improved with better scheduler design.
+Both of these potentially can be improved with better scheduler design.
 The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.

### 2011-09-30 papers/paper1/programming_model.tex
New case study and related work.

@@
-\section{I/O Automata Programming Model\label{representation}}
+\section{I/O Automata Programming Model\label{programming_model}}
 
-In this section, we show how I/O automata can be used to construct real programs using ioa++.
-To motivate the presentation, we use the classic problem of electing a leader in a unidirectional ring.
-We first show how I/O automata are represented in the C++ language.
-We then discuss the mechanisms available for dynamic composition.
-We conclude with techniques for managing dynamic constellations of automata.
+In this section, we show how I/O automata can be used to construct programs using ioa++.
+Section~\ref{representation} describes how I/O automata are represented in the C++ language.
+We then discuss the mechanisms available for dynamic composition in Section~\ref{dynamic_composition}.
+Section~\ref{constellations} concludes with techniques for managing dynamic constellations of automata.
 
-\paragraph*{Leader election in a unidirectional ring}
-Nodes in a distributed system are arranged in a ring so that two consecutive nodes are connected by a directed channel.
-The goal is to designate one of the nodes as the leader of the ring.
-The solution we adopt is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
-The LCR algorithm assumes that every node has a unique identifier (UID).
-When the protocol begins, each node sends its UID to its successor.
-When a node receives a UID from its predecessor that is greater than its own, it forwards the UID to its successor.
-When a node receives its own UID, it elects itself the leader of the ring.
-Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp of the ioa++ package.
+\subsection{I/O Automata Representation\label{representation}}
 
-\subsection{I/O Automata Representation}
-
-\paragraph*{Declaring an automaton}
 All automata inherit from the common base class ioa::automaton which implements the system action interface.
@@
 \end{lstlisting}
+The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept.
+The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
 \fi
-The asynch\_lcr\_automaton is declared as a template where the UID template parameter must model the unique identifier concept.
-The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
 
 \paragraph*{Declaring and initializing state variables}
-The state variables of an automaton are expressed as member variables of the corresponding class.
+The state variables of an automaton are expressed as member variables of its corresponding class or class template.
+\ifjournal
 The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status.
-\ifjournal
 \begin{lstlisting}
@@
 \fi
-Member variables are declared private which is in keeping with the notion that the state of each automaton is independent.
-State variables are initialized using a constructor.
-For the asynch\_lcr\_automaton, the flags are both false and the UID is initialized and added to the send queue.
+Member variables are declared private, which is in keeping with the notion that the state of each automaton is independent, and are initialized using a constructor.
+The constructor can also bootstrap the scheduler using the ioa::schedule function discussed below.
 \ifjournal
+For the asynch\_lcr\_automaton, the flags are both set initially false and the UID is initialized and added to the send queue.
 \begin{lstlisting}
@@
 \end{lstlisting}
+The constructor also bootstraps the scheduler by scheduling all enabled local actions using the ioa::schedule function discussed below.
 \fi
-The constructor also bootstraps the scheduler by scheduling all enabled local actions using the member function schedule.
-The schedule member function is discussed in the paragraph on scheduling.
 
 \paragraph*{Declaring/defining actions}
-Actions are member variables that model the action concept.
-Actions contain methods for preconditions, effects, and scheduling and traits that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
-The traits are used to guide the execution of the methods by the model and supply values and parameters as necessary.
+Actions are member variables that model the \emph{action} concept.
+Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
+These traits are used to guide the execution of the methods by the model and to supply values and parameters as necessary.
 The traits are also used to check that input actions are bound to output actions with the same value status and type.
@@
 
-A number of wrappers and macros are available for declaring actions.
+A number of wrappers and macros are provided by ioa++ for declaring actions.
 %%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
-Local actions are declared using three functions corresponding to a precondition, effect, and scheduling call.
+Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
 The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
@@
 \fi
-Input actions are declared similarly save the absence of a precondition.
+Input actions are declared similarly except for the absence of a precondition.
 The precondition and scheduling functions are not allowed to change the state of the automaton and are consequently declared const.
-The macros for actions that are subject to binding by other automata belong in a public section.
+The macros for actions that are subject to binding by other automata are declared public.
 
@@
 A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
+Another common pattern is to have all scheduling functions dispatch to a single scheduling function to avoid replicating scheduling logic.
 \ifjournal
@@
 \end{lstlisting}
+The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this common member function.
 \fi
-The various scheduling functions in the asynch\_lcr\_automaton and the constructor all dispatch to this function.
 
-\subsection{Dynamics}
+\subsection{Dynamics\label{dynamic_composition}}
 
 \paragraph*{Dynamic composition}
-The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
+%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
 Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
@@
 \fi
-Children automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
-An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating the child automaton.
+Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
+An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating child automata.
 Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
@@
 Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
-The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of actions bound to the specified action.
+The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of other actions bound to the specified action.
 The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
-Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received.
+Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received if produced.
 \ifjournal
@@
 
-\subsection{Managing Dynamic Constellations}
+\subsection{Managing Dynamic Constellations\label{constellations}}
 
 In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the ``nuts and bolts'' presented in the previous section.
-This list is neither exhaustive nor authoritative but rather a distillation of our experiences working with I/O automata.
+This list is not exhaustive but rather serves as a distillation of our experiences so far working with I/O automata in ioa++.
 
 \paragraph*{System action events}
-Determining the outcome of a system action event, e.g., that a child automaton was created, is necessary when there are constraints on the order in which system actions must occur.
-For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata supply the output action and input action are created before attempting to bind.
-Similarly, an automaton might compose with an automaton that it does not manage and therefore expects the automaton to exist before attempting to bind.
+Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
+For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
+Similarly, an automaton might compose with an automaton that it does not manage and therefore may expect the automaton to exist before attempting to bind.
 To enable such behavior, user automata can observe ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.
@@
 Using this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
-The two strategies for handling an error are to either reset or replace the portion of the constellation that encountered the error.
+The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.
 
@@
 An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
-Errors need not be viewed as exceptional circumstances but a necessary part of the protocol implemented by the automaton.
 Automaton designers should strive to write automata that are capable of being reset as this is the simplest error handling strategy.
 
-\paragraph*{Handling errors by replacing}
+\paragraph*{Handling errors by replacement}
 The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
-To replace a child automaton, the parent automaton must destroy the child, create a new child, and bind to the new child.
-For reasons already discussed, the bind must occur after the create.
-This constraint is enforced by the ioa::binding\_manager class.
-Additionally, the destroy must occur before the first bind to ensure that all bindings associated with the first child have been dissolved.
-Attempts to bind before the destroy will fail because either the output action or the input action will be unavailable, i.e., still bound to the first child.
+To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
+Recall that system actions are asynchronous and can be executed in any order.
+Thus, the user might might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
+If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
+Note that executing the bind action before the create will also fail but this case is prevent by the ioa::binding\_manager class.
+Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.
 A general solution is to observe the automaton and wait for it to be destroyed.
@@
 If all of the bindings of a child automaton are with its parent, then the child can be replaced using parameters.
-The parent automaton associates a parameter, typically the child automaton, with all bindings to the child automaton.
-When the child needs to be replaced, a new child is created and the current parameter is updated to reflect this change.
-The old bindings can be disabled by checking their parameter against the new parameter.
-When using replacement with parameters, there is no need to wait for the child automaton to be destroyed since the parameter causes all bindings to be different.
+Instead of binding the new child to the exact same actions of the parent that are being used by the old child, which is the root problem in the preceding scenario, the new child is bound to actions parameterized by the new child.
+The parameters serve to differentiate the actions and obviate the need to wait for the old child to be destroyed.
+To use this technique, the parent remembers the parameter associated with the current child and becomes unresponsive to actions not bearing that parameter.

### 2011-09-30 papers/paper1/related_work.tex
New case study and related work.

@@
 
-This work grew out of the challenges faced when trying to develop distributed systems with threads.
+Our development of ioa++ grew out of the challenges faced when trying to develop distributed systems with threads.
 Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
@@
 
-The direction implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
+The direct implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
+Structured programming simplifies development by allowing programmers to reason about their programs directly from the source code; regardless of whether of not the code it formally verified.
 The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features essential to developing distributed systems~\cite{lynch1996distributed}.
-Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.
+%% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.
 
-Comparing I/O automata to every model for asynchronous, concurrent, and distributed computing is beyond the scope of this paper so we focus on exemplars.
-The I/O automata model is based on state transitions which matches the imperative style that dominates modern concurrent programming.
-This is similar to the motivation for UNITY~\cite{chandy1988parallel} and is in contrast to functional models such as the $\pi$-calculus~\cite{milner1992calculus} and actors~\cite{agha1986actors}.
-The I/O automata model assumes independent state with explicit communication like communicating sequential processes (CSP)~\cite{hoare1978communicating} but unlike threads~\cite{lee2006problem} and the UNITY model~\cite{chandy1988parallel}.
-All models for asynchronous, concurrent, and distributed computing admit non-determinism.
-Lee points out that the difficulties of thread-based programming come from a need to ``prune'' non-determinism when paired with shared state and adopts the view that software components should be deterministic with respect to concurrency save a few that introduce non-determinism in a controlled way~\cite{lee2006problem}.
-Programs in UNITY are composed by concatenating program texts and resolving shared variables~\cite{chandy1988parallel}.
-For this reason, they share many of the same problems as threads under composition.
-I/O automata can be viewed as taking the alternate approach by admitting non-determinism while prohibiting shared state.
-In I/O automata, the sequential flow control seen with threads~\cite{lee2006problem} and CSP~\cite{hoare1978communicating} is replaced by the non-deterministic execution of conditional atomic actions which resembles execution in the UNITY model~\cite{chandy1988parallel}.
-The ability to create new automata can be compared to actor creation in the actor model~\cite{agha1986actors}.
-Actors communicate by name using a buffered mail system while communications in I/O automata are not buffered and anonymous.
+Incorporating a formal model of concurrency into a language to make concurrent programming easier is an established technique.
+We consider three exemplars:  Communicating Sequential Processes, Actors, and threads.
+The Communicating Sequential Processes (CSP) model~\cite{hoare1978communicating} is an extremely popular technique for modeling concurrent systems.
+The fundamental concurrency control primitive in CSP is the \emph{rendezvous} which allows processes to synchronize and possibly exchange values.
+CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence programming language design, e.g., Go~\cite{go}.
+An actor~\cite{agha1986actors} is a behavior that can perform a number of actions when receiving a message including (1) replacing itself with a new behavior, (2) sending messages to other actors, and (3) creating new actors.
+The Actor model influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.
+The collective formalization of threads, e.g., semaphores~\cite{dijkstra1968cooperating}, monitors~\cite{hoare1974monitors}, all appear in popular libraries, e.g., POSIX threads~\cite{butenhof1997programming}, and languages, e.g., Java~\cite{christopher2000high}.
 
-To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing systems that allow one to execute a system expressed as a collection of I/O automata.
+%% Frameworks for reusable asynchronous and concurrent modules
+%%   Collective formalization of threads
+%%     CORBA (asynchronous method invocation)/RMI/RPC (function call stack model, ~threads)
+%%   CSP
+%%   Actors
+
+%% Comparing I/O automata to every model for asynchronous, concurrent, and distributed computing is beyond the scope of this paper, so we focus instead on exemplars.
+%% The I/O automata model is based on state transitions which matches the imperative style that dominates modern concurrent programming.
+%% This is similar to the motivation for UNITY~\cite{chandy1988parallel} and is in contrast to functional models such as the $\pi$-calculus~\cite{milner1992calculus} and actors~\cite{agha1986actors}.
+%% The I/O automata model assumes independent state with explicit communication, like communicating sequential processes (CSP)~\cite{hoare1978communicating} but unlike threads~\cite{lee2006problem} or the UNITY model~\cite{chandy1988parallel}.
+
+%% All models for asynchronous, concurrent, and distributed computing admit non-determinism.
+%% Lee points out that the difficulties of thread-based programming come from a need to ``prune'' non-determinism when paired with shared state and adopts the view that software components should be deterministic with respect to concurrency save a few that introduce non-determinism in a controlled way~\cite{lee2006problem}.
+%% Programs in UNITY are composed by concatenating program texts and resolving shared variables~\cite{chandy1988parallel}.
+%% For this reason, they share many of the same concerns as threads under composition.
+%% I/O automata can be viewed as taking an alternate approach by admitting non-determinism while prohibiting shared state.
+%% In I/O automata, the sequential flow control seen with threads~\cite{lee2006problem} and CSP~\cite{hoare1978communicating} is replaced by the non-deterministic execution of conditional atomic actions which resembles execution in the UNITY model~\cite{chandy1988parallel}.
+%% The ability to create new automata can be compared to actor creation in the actor model~\cite{agha1986actors}.
+%% Actors communicate by name using a buffered mail system while communications in I/O automata are not buffered and anonymous.
+
+To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing approaches that allow one to execute a system expressed as a collection of I/O automata.
 Spectrum focuses on algorithm development through simulation.
@@
 
-The ioa++ framework can be compared directly to the IOA language due to the addition of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
+The ioa++ framework can be compared directly to the IOA language due to the availability of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
 \cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
-Programming written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
+Programs written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
 IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
-IOA lacks support for procedure calls which complicates modeling and interfacing with external services; especially when a procedure call blocks.
+IOA lacks support for procedure calls which complicates modeling and interfacing with external services: especially when a procedure call blocks.
 The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
@@
 The development of a new language was necessary for IOA due to their focus on formal verification.
-ioa++ on the other hand uses the existing C++ language and compilers.
-IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors relying on the programmer to ensure non-blocking operation.
+On the other hand ioa++ uses the existing C++ language and compilers.
+IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors, relying on the programmer to ensure non-blocking operation.
 IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
-Both IOA and ioa++ require the programmer to add code to schedule actions and find this to be a common source of errors.
+Both IOA and ioa++ require the programmer to add code to schedule actions, which may be a common source of errors.
+Scheduling in IOA is deterministic and the programmer provides the policy.
+Scheduling in ioa++ is non-deterministic according to the global policy encoded by the controller.
 

### 2011-09-30 examples/tcp_lcr.cpp

@@ -25,6 +25,370 @@
 
 // In seconds.
 #define CONNECTION_TIMEOUT 10
+
+class tcp_ring_automaton :
+  public ioa::automaton,
+  private ioa::observer
+{
+private:
+  ioa::handle_manager<tcp_ring_automaton> m_self;
+  const ioa::inet_address m_recv_address;
+  const ioa::inet_address m_send_address;
+  ioa::automaton_manager<ioa::tcp_acceptor_automaton>* m_acceptor;
+  ioa::automaton_manager<ioa::tcp_connection_automaton>* m_predecessor;
+  bool m_accept_flag;
+  ioa::automaton_manager<ioa::tcp_connection_automaton>* m_successor;
+  ioa::automaton_manager<ioa::tcp_connector_automaton>* m_connector;
+  bool m_successor_connected;
+  bool m_clear_to_send;
+  std::queue<std::string> m_send;
+  std::queue<std::string> m_recv;
+
+public:
+  tcp_ring_automaton (const ioa::inet_address& recv_address,
+		     const ioa::inet_address& send_address) :
+    m_self (ioa::get_aid ()),
+    m_recv_address (recv_address),
+    m_send_address (send_address)
+  {
+    schedule ();
+  }
+
+private:
+  void observe (ioa::observable*) {
+    // Something related to our connections changed.
+    schedule ();
+  }
+
+  void schedule () const {
+    if (create_acceptor_precondition ()) {
+      ioa::schedule (&tcp_ring_automaton::create_acceptor);
+    }
+    if (create_predecessor_precondition ()) {
+      ioa::schedule (&tcp_ring_automaton::create_predecessor);
+    }
+    if (create_successor_precondition ()) {
+      ioa::schedule (&tcp_ring_automaton::create_successor);
+    }
+    if (accept_precondition (m_acceptor)) {
+      ioa::schedule (&tcp_ring_automaton::accept, m_acceptor);
+    }
+    if (connect_precondition ()) {
+      ioa::schedule (&tcp_ring_automaton::connect);
+    }
+    if (send_successor_precondition (m_successor)) {
+      ioa::schedule (&tcp_ring_automaton::send_successor, m_successor);
+    }
+    if (receive_precondition ()) {
+      ioa::schedule (&tcp_ring_automaton::receive);
+    }
+  }
+
+  bool create_acceptor_precondition () const {
+    return m_acceptor == 0;
+  }
+
+  void create_acceptor_effect () {
[…]

### 2011-09-30 papers/paper1/component_model.tex
Marking phrases we should reconsider.

@@
 %% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
-We then describe extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time.
+We then describe \textbf{extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time}.
 
@@
 Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
-Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions,  subject to the precondition using the value produced by the output action.
+Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
 The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
 The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.
-
-\begin{figure}
-\center
-\includegraphics[width=\columnwidth]{system_model}
-\caption{Example event detection system.}
-\label{sys_model}
-\end{figure}
 
@@
 
+\begin{figure}
+\center
+\includegraphics[width=\columnwidth]{system_model}
+\caption{Example event detection system.}
+\label{sys_model}
+\end{figure}
+
 Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
@@
 
-The event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
+The event detector system illustrated in Figure~\ref{sys_model} contains three component automata:  an anonymous Processor, a Converter named C, and a Detector named D.
 The Processor automaton is the root and C and D are its children.
@@
 
+%% This is only true for static systems.  Dynamic system can do crazy things.
 Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
@@
 However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
-Graphically, an arrow originating in one component must terminate in another component, even if the latter is a child of the former as Figure~\ref{sys_model} illustrates.
+Graphically, an arrow originating in one component must terminate in another component, \textbf{even if the latter is a child of the former} as Figure~\ref{sys_model} illustrates.
 %% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
@@
 \to \{D, root\}$.
-Thus, $process1$ and $out!(s)$ can be executed concurrently while $process1$ and $detect!$ can not.
+Thus, $out!(s)$ and $process1$ can be executed concurrently while $process1$ and $detect!$ can not.
 
@@
 If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
-Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata directly.
+Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata \textbf{directly}.
 %% Consequently, we add the ability to deal with a dynamic set of automata and interactions.
@@
 Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
-However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
 
@@
 Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
-A system defined statically can be implemented as a dynamic system adding binding predicates and logic to create the required constellation of automata.
+A system defined statically can be implemented as a dynamic system by adding binding predicates and logic to create the required constellation of automata.

### 2011-09-30 papers/paper1/design.tex
Marking phrases we should reconsider.

@@
 The controller encodes a \emph{policy} that selects the next local action.
-For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
+For the framework to be a valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
 The run-time system can be initialized with different controllers to achieve different performance objectives.
 Controllers can use multiple threads of execution to exploit the concurrency available in a system of automata.
-The ioa++ framework is designed so that all native concurrency structures (e.g., threads, mutexes) are encapsulated in the dispatcher and controller.
+The ioa++ framework is designed so that all native concurrency mechanisms (e.g., threads, mutexes) are encapsulated in the dispatcher and controller.
 
@@
 
-A more focused approached used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
+A more focused approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
 The controller then has the responsibility of processing the set of requested actions according to some fair policy.
@@
 Not declaring an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
-Not scheduling an (enabled) action could be another potential mistake.
+Not scheduling an (enabled) action \textbf{could be another potential mistake}.
 

### 2011-09-30 papers/paper1/abstract.tex
refactored abstract and intro, updated my config

--- 2011-09-27 papers/paper1/abstract.tex
@@
 We develop a component model for asynchronous and concurrent systems based on I/O automata and dynamic composition.
-We describe the design and implementation of the component model and show how ioa++ can be used to develop protocols and other forms of distributed systems.
+We describe the design and implementation of the component model in ioa++ and show how it can be used to develop protocols and other forms of distributed software.
 
-Our experiences demonstrate that ioa++ can help simplify developing distributed systems, by supporting reasoning about asynchronous and concurrent programs directly from the source code using the I/O automata formal model.
+Our experiences demonstrate that ioa++ can help simplify developing distributed systems, by supporting reasoning about asynchronous and concurrent programs directly from source code using the I/O automata formal model.
 Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one concurrency model to another.
-Whereas reasoning about the correctness of modules in a threaded program requires considering interactions among the various threads, reasoning about the correctness of modules in ioa++ is limited to the modules themselves and a well-defined set of composition rules.
+% I've just commented this out instead of trying to weave it in elsewhere, especially since I think it is already somewhat implied by the threads discussion in the intro - Chris
+%Whereas reasoning about the correctness of modules in a threaded program requires considering interactions among the various threads, reasoning about the correctness of modules in ioa++ is limited to the modules themselves and a well-defined set of composition rules.
 
-We evaluate concurrency in ioa++ through a series of micro-benchmarks which confirm that concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
+We evaluate concurrency performance in ioa++ through a series of micro-benchmarks which confirm that the framework's ability to exploit potentially concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
 \end{abstract}

### 2011-09-30 papers/paper1/case_study.tex
Another pass.

@@
 A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
-The tcp\_ring\_automaton listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and channel to the successor.
+The tcp\_ring\_automaton listed in examples/tcp\_ring\_automaton.hpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}.} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
 The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.

### 2011-09-30 papers/paper1/design.tex
Another pass.

@@
 This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
-Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)}, assuming that automata turn over less frequently overall than the set of actions in the scheduler.
+Our solution assigns each automaton a unique identifier assuming that the set of automata turn over less frequently overall than the set of actions in the scheduler.
 
@@
 \paragraph*{Automaton and Binding managers}
-The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating more complicated ones.
+The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple constellations but becomes unwieldy when creating more complicated ones.
 Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
@@
 
+\ifjournal
 \paragraph*{Generators}
@@
 Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.
+\fi

### 2011-09-30 papers/paper1/evaluation.tex
Another pass.

@@
 From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4316ns $\pm$ 3.5ns for the single threaded experiments and 6067ns $\pm$ 3.6ns for the multi-threaded experiments.
-Similary, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
+Similarly, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
 Both of these potentially can be improved with better scheduler design.
-The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.
+The synchronization time depends on the interactions among the automata comprising the system, the scheduler, and the locking scheme used by the framework.
 We plan to address scheduling and synchronization in future work.

### 2011-09-30 papers/paper1/programming_model.tex
Another pass.

@@
 The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept.
-The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
+The metaprogramming facilities of the C++ language aid in the development of general purpose reusable components.
 \fi
@@
 Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
-The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
+%%The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
 \ifjournal
@@
 \fi
-Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
-An automaton manager object requires an ioa::automaton object for its system action implementation and a generator for allocating child automata.
-Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
+%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
+An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
+%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
 A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
@@
 The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
-Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manange is destroyed or unbound respectively.
+Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
 \fi
@@
 If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
-Note that executing the bind action before the create will also fail but this case is prevent by the ioa::binding\_manager class.
+Note that executing the bind action before the create will also fail but this case is prevented by the ioa::binding\_manager class.
 Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.

### 2011-09-30 papers/paper1/related_work.tex
refactored abstract and intro, updated my config
Another pass.

@@
 Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
-These same concerns are relevant to trends towards increasingly concurrent programming to increase performance, given the leveling trend of clock frequencies and the advent of multi-core processors\cite{sutter2005software}.
+Given the leveling of processor clock frequencies and the advent of multi-core, these same concerns are relevant to trends that look to deliver performance increases through increasingly concurrent programs~\cite{sutter2005software}.
 
 The direct implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
-Structured programming simplifies development by allowing programmers to reason about their programs directly from the source code; regardless of whether of not the code it formally verified.
+Structured programming simplifies development by allowing programmers to reason about their programs directly from the source code; regardless of whether of not the code is formally verified.
 The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features essential to developing distributed systems~\cite{lynch1996distributed}.
@@
 The fundamental concurrency control primitive in CSP is the \emph{rendezvous} which allows processes to synchronize and possibly exchange values.
-CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence programming language design, e.g., Go~\cite{go}.
+CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence the design of modern programming languages, e.g., Go~\cite{go}.
 An actor~\cite{agha1986actors} is a behavior that can perform a number of actions when receiving a message including (1) replacing itself with a new behavior, (2) sending messages to other actors, and (3) creating new actors.
 The Actor model influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.
-The collective formalization of threads, e.g., semaphores~\cite{dijkstra1968cooperating}, monitors~\cite{hoare1974monitors}, all appear in popular libraries, e.g., POSIX threads~\cite{butenhof1997programming}, and languages, e.g., Java~\cite{christopher2000high}.
+The collective formalization of threads, i.e., semaphores~\cite{dijkstra1968cooperating}, monitors~\cite{hoare1974monitors}, etc., appears in popular libraries, e.g., POSIX threads~\cite{butenhof1997programming}, and languages, e.g., Java~\cite{christopher2000high}.
 
@@
 The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
-The next action to be executed is determined by a schedule function that resolves the non-determinism in the schedule.
+The next action to be executed is selected by a function that resolves the non-determinism in the schedule.
 
@@
 Both IOA and ioa++ require the programmer to add code to schedule actions, which may be a common source of errors.
-Scheduling in IOA is deterministic and the programmer provides the policy.
-Scheduling in ioa++ is non-deterministic according to the global policy encoded by the controller.
+In IOA, the programmer provides a deterministic scheduling policy.
+Scheduling in ioa++, however, is non-deterministic according to the global policy encoded by the controller.
 

### 2011-09-30 examples/tcp_lcr.cpp

@@ -19,374 +19,9 @@
 #include <ioa/tcp_connector_automaton.hpp>
 #include <ioa/global_fifo_scheduler.hpp>
 #include "asynch_lcr_automaton.hpp"
+#include "tcp_ring_automaton.hpp"
 
 #include <uuid/uuid.h>
-#include <iostream>
-
-// In seconds.
-#define CONNECTION_TIMEOUT 10
-
-class tcp_ring_automaton :
-  public ioa::automaton,
-  private ioa::observer
-{
-private:
-  ioa::handle_manager<tcp_ring_automaton> m_self;
-  const ioa::inet_address m_recv_address;
-  const ioa::inet_address m_send_address;
-  ioa::automaton_manager<ioa::tcp_acceptor_automaton>* m_acceptor;
-  ioa::automaton_manager<ioa::tcp_connection_automaton>* m_predecessor;
-  bool m_accept_flag;
-  ioa::automaton_manager<ioa::tcp_connection_automaton>* m_successor;
-  ioa::automaton_manager<ioa::tcp_connector_automaton>* m_connector;
-  bool m_successor_connected;
-  bool m_clear_to_send;
-  std::queue<std::string> m_send;
-  std::queue<std::string> m_recv;
-
-public:
-  tcp_ring_automaton (const ioa::inet_address& recv_address,
-		     const ioa::inet_address& send_address) :
-    m_self (ioa::get_aid ()),
-    m_recv_address (recv_address),
-    m_send_address (send_address)
-  {
-    schedule ();
-  }
-
-private:
-  void observe (ioa::observable*) {
-    // Something related to our connections changed.
-    schedule ();
-  }
-
-  void schedule () const {
-    if (create_acceptor_precondition ()) {
-      ioa::schedule (&tcp_ring_automaton::create_acceptor);
-    }
-    if (create_predecessor_precondition ()) {
-      ioa::schedule (&tcp_ring_automaton::create_predecessor);
-    }
-    if (create_successor_precondition ()) {
-      ioa::schedule (&tcp_ring_automaton::create_successor);
-    }
-    if (accept_precondition (m_acceptor)) {
-      ioa::schedule (&tcp_ring_automaton::accept, m_acceptor);
-    }
-    if (connect_precondition ()) {
-      ioa::schedule (&tcp_ring_automaton::connect);
-    }
-    if (send_successor_precondition (m_successor)) {
-      ioa::schedule (&tcp_ring_automaton::send_successor, m_successor);
-    }
[…]

### 2011-09-30 examples/tcp_ring_automaton.hpp

@@ -1,0 +1,388 @@
+/*
+   Copyright 2011 Justin R. Wilson
+
+   Licensed under the Apache License, Version 2.0 (the "License");
+   you may not use this file except in compliance with the License.
+   You may obtain a copy of the License at
+
+       http://www.apache.org/licenses/LICENSE-2.0
+
+   Unless required by applicable law or agreed to in writing, software
+   distributed under the License is distributed on an "AS IS" BASIS,
+   WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
+   See the License for the specific language governing permissions and
+   limitations under the License.
+*/
+
+#ifndef __tcp_ring_automaton_hpp__
+#define __tpc_ring_automaton_hpp__
+
+#include <ioa/ioa.hpp>
+#include <iostream>
+
+// In seconds.
+#define CONNECTION_TIMEOUT 10
+
+class tcp_ring_automaton :
+  public ioa::automaton,
+  private ioa::observer
+{
+private:
+  ioa::handle_manager<tcp_ring_automaton> m_self;
+  const ioa::inet_address m_recv_address;
+  const ioa::inet_address m_send_address;
+  ioa::automaton_manager<ioa::tcp_acceptor_automaton>* m_acceptor;
+  ioa::automaton_manager<ioa::tcp_connection_automaton>* m_predecessor;
+  bool m_accept_flag;
+  ioa::automaton_manager<ioa::tcp_connection_automaton>* m_successor;
+  ioa::automaton_manager<ioa::tcp_connector_automaton>* m_connector;
+  bool m_successor_connected;
+  bool m_clear_to_send;
+  std::queue<std::string> m_send;
+  std::queue<std::string> m_recv;
+
+public:
+  tcp_ring_automaton (const ioa::inet_address& recv_address,
+		     const ioa::inet_address& send_address) :
+    m_self (ioa::get_aid ()),
+    m_recv_address (recv_address),
+    m_send_address (send_address)
+  {
+    schedule ();
+  }
+
+private:
+  void observe (ioa::observable*) {
+    // Something related to our connections changed.
+    schedule ();
+  }
+
+  void schedule () const {
+    if (create_acceptor_precondition ()) {
+      ioa::schedule (&tcp_ring_automaton::create_acceptor);
+    }
+    if (create_predecessor_precondition ()) {
[…]

### 2011-09-30 papers/paper1/case_study.tex
Minor tweaks.

--- 2011-09-30 papers/paper1/case_study.tex
@@
 We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
-Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}.} of the ioa++ package.
+Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}} of the ioa++ package.
 
 A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
-The tcp\_ring\_automaton listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and channel to the successor.
+The tcp\_ring\_automaton listed in examples/tcp\_ring\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
 The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
@@
 
-The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}.} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
+The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
 Figure~\ref{tcp_lcr_automaton} shows the layered architecture of the TCP LCR system.

### 2011-09-30 papers/paper1/component_model.tex
refactored abstract and intro, updated my config
Minor tweaks.

@@
 %% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
-We then describe \textbf{extensions to support dynamic semantics for systems with participants that may arrive and depart at run-time}.
+We then describe extensions to support semantics for dynamic systems, i.e., systems whose set of constituent automata changes over time.
 
@@
 
+%The I/O automata formalism was developed to model reactive systems~\cite{lynch1987hierarchical} and has been used in the design and verification of concurrent and distributed systems.
+I/O automata are a good foundation for such components because they have independent state, straightforward interfaces, and precise semantics for event generation, distribution, and consumption resulting in well-defined interactions under composition.
 \emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
@@
 However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
-Graphically, an arrow originating in one component must terminate in another component, \textbf{even if the latter is a child of the former} as Figure~\ref{sys_model} illustrates.
+Graphically, an arrow originating in one component must terminate in another component as Figure~\ref{sys_model} illustrates.
 %% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
@@
 If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
-Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata \textbf{directly}.
+Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata.
 %% Consequently, we add the ability to deal with a dynamic set of automata and interactions.

### 2011-09-30 papers/paper1/design.tex
Minor tweaks.

@@
 A more focused approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
+The act of declaring an action to the scheduler is called \emph{scheduling}.
 The controller then has the responsibility of processing the set of requested actions according to some fair policy.
@@
 Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
-Consequently, an automaton is given the opportunity to declare actions to the scheduler after one of its actions is executed.
+Consequently, an automaton is given the opportunity to schedule actions to the scheduler after one of its actions is executed.
 Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
@@
 However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
-A simple technique for accomplishing this is to \emph{schedule} all enabled actions.
+A simple technique for accomplishing this is to schedule all enabled actions.
 More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
 However, this must be done carefully.
-Not declaring an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
-Not scheduling an (enabled) action \textbf{could be another potential mistake}.
+Not scheduling an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
+Not scheduling an (enabled) action is a common mistake.
 

### 2011-09-30 papers/paper1/evaluation.tex
Minor tweaks.

@@
 
-The driver for a single trial can be found in examples/random.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/random.cpp}.} of the ioa++ package.
+The driver for a single trial can be found in examples/random.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/random.cpp}} of the ioa++ package.
 The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.

### 2011-10-01 papers/paper1/component_model.tex
refactored section 2, moved part of it to 3

@@
 %% We then describe extensions to support dynamic semantics as participants in the system may arrive and depart at run-time.
-We then describe extensions to support semantics for dynamic systems, i.e., systems whose set of constituent automata changes over time.
+We then describe extensions to support semantics for \emph{dynamic systems,} i.e., systems whose set of constituent automata changes over time.
 
@@
 Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
-Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions using the value produced by the output action subject to the precondition.
+Executing a composed output action consists of atomically evaluating the precondition, and if it is true applying the effect of the output action and all composed input actions using the value produced by the output action.
 The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
@@
 
+\paragraph*{Motivation for I/O automata based components}
 %The I/O automata formalism was developed to model reactive systems~\cite{lynch1987hierarchical} and has been used in the design and verification of concurrent and distributed systems.
@@
 
-Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
-The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
+To illustrate how components can be expressed naturally using I/O automata, Figure~\ref{sys_model} shows a simple event detector system comprising three component automata: an anonymous Processor, a Converter named C, and a Detector named D.
+The Processor automaton is the root and C and D are its children.
+Automata are depicted as solid rectangles. The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
 The name of a named automaton appears after its type, e.g., Converter C.
@@
 
-The event detector system illustrated in Figure~\ref{sys_model} contains three component automata:  an anonymous Processor, a Converter named C, and a Detector named D.
-The Processor automaton is the root and C and D are its children.
-
-\paragraph*{Explicit Composition}
+\paragraph*{Explicit composition}
 Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
-However, when using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
+However, when using recursive encapsulation, developers must have the ability to compose component automata regardless of the naming convention used by the original developers.
 To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
@@
 %% This is only true for static systems.  Dynamic system can do crazy things.
-Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
+Each component automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
 The automaton that prescribes a binding is said to \emph{own} that binding.
@@
 Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
-The map of bindings for the event detection system in Figure~\ref{sys_model} is: 
+The map of bindings for the event detection system shown in Figure~\ref{sys_model} is: 
 \begin{displaymath}
@@
 For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
-This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
+This stronger rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
 Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
@@
 However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
-Graphically, an arrow originating in one component must terminate in another component as Figure~\ref{sys_model} illustrates.
+Graphically, an arrow originating in one component automaton must terminate in another component automaton as Figure~\ref{sys_model} illustrates.
 %% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.
@@
 
-I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for a reusable, asynchronous, and concurrent software system.
-Each action is only allowed to modify the state of the automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
-The absence of shared state thus ensures that each automaton can be deployed independently.
-The external signature of an I/O automaton constitutes its interface and the result of interacting with an I/O automaton is well-defined since every action is atomic.
-Composition, asynchrony, and concurrency are guaranteed by the model itself.
+I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for developing reusable components for asynchronous and concurrent software systems.
+Each action is only allowed to modify the state of the component automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
+The absence of shared state thus ensures that each component automaton can be deployed independently.
+The external signature of a component automaton constitutes its interface and the result of interacting with it is well-defined since every action is atomic.
+Composition, atomicity, asynchrony, and concurrency are thus well supported by the model itself.
 
@@
 The set of interactions in the I/O automata model is also fixed since automata are statically composed.
-Dynamic systems are modeled in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
+Dynamic systems can be modeled in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
 To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.
 
+\paragraph*{Limitations of the static model}
 While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
@@
 To illustrate, let a system $S=(A,B)$ be a pair consisting of a set of automata $A$ and a set of bindings $B$.
-The I/O automata model assumes that $S$ and therefore $A$ and $B$ are both fixed.
-Dynamics are simulated via the manipulation of activation flags.
-A dynamic system implies that the system $S$ must be dependent on time $t$ as a function of the number of actions that have been executed by $t$.
-Thus, $S_{t+1}$ need not be equal to $S_{t}$.
-A \emph{system action} is an action that \emph{can} change the system $S$.
+The I/O automata model assumes that $S$ and therefore $A$ and $B$ are both fixed, so that dynamics must be \emph{simulated} via the manipulation of activation flags.
 
+\paragraph*{Support for dynamic systems}
+To address the limitations of the static representation of dynamic systems, ioa++ provides
+architectural support for dynamic systems by separating the concerns of \emph{enacting} versus \emph{initiating} 
+actions for managing dynamics into distinct \emph{system} and \emph{base} automata respectively.
+We define four \emph{system actions} for managing dynamic systems of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
+The \emph{create} action allows an automaton to create a child automaton.
+The \emph{bind} action allows an automaton to associate an input action with an output action while the \emph{unbind} action allows an automaton to dissociate an input action from an output action.
+The \emph{destroy} action allows an automaton to destroy a child automaton and all bindings associated with it (recursively).
 \begin{figure}
@@
 \includegraphics[width=\columnwidth]{system_action}
-\caption{System actions.
-  A base automaton contains output actions for creating, binding, unbinding, and destroying.
-  The system automaton contains parameterized input actions for receiving the requests to create, binding, unbind, and destroy.
-  The system automaton returns a result using the appropriate parameterized output action.
-  The base automaton contains input actions to receive the results.}
+\caption{System automaton and actions.}
 \label{system_action}
@@
 
-The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
-Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
-The system automaton, in turn, has input actions for receiving requests for system actions, internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
-We define four actions in the system automaton for managing a dynamic system of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
-The \emph{create} action allows an automaton to create a child automaton.
-The \emph{bind} action allows an automaton to bind an input action to an output action while the \emph{unbind} action allows an automaton to separate an input action from an output action.
-The \emph{destroy} action allows an automaton to recursively destroy a child automaton and all bindings associated with it.
-Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.
+% A dynamic system implies that the system $S$ must be dependent on time $t$ as a function of the number of actions that have been executed by $t$.
+% Thus, $S_{t+1}$ need not be equal to $S_{t}$.
+% A \emph{system action} is an action that \emph{can} change the system $S$.
+% The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
+As Figure~\ref{system_action} illustrates, the system automaton provides: parameterized input actions for receiving requests to create, bind, unbind, and destroy automata; internal actions for evaluating and enacting system action requests; and corresponding parameterized output actions through which to return results of the requests.
+The base automaton provides output actions that trigger requests to create, bind, unbind, and destroy automata, along with
+input actions to receive the results.
+Component automata \emph{must} inherit from the base automaton, which as we describe in Section~\ref{framework_architecture} is implemented in ioa++ as a C++ class.
+Each component automaton is thus composed (not bound) with the system automaton and inherits output actions for requesting system actions and input actions for receiving the results of system actions.
 
-System actions can fail for a variety of reasons.
-A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
-A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{component_structure} would be violated.
+We note that system action requests may fail for a variety of reasons.
+A create fails if the automaton that should be added to the set of automata, i.e., the automaton being created, already exists.
+A bind fails if the automaton associated with the output action or the automaton associated with the input action doesn't exist, or if any of the binding rules outlined in Section~\ref{component_structure} would be violated.
 An unbind fails if the binding does not exist.
-Note that the owner of binding is taken into account when determining if the binding exists so only the owner of a binding can dissolve it.
+Note that the owner of the binding is taken into account when determining whether the binding exists, such that only the owner of a binding can dissolve it.
 A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.
@@
 
-Conceptually, the state variables $A$ and $B$ belong to the system automaton.
-However, the state encoded by $A$ and $B$ is shared and is used in the execution of every action.
-A \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
-However, the opportunities for concurrency explored in Section~\ref{component_structure} are not lost when one considers that only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
-
 \paragraph*{Binding predicates}
@@
 
-To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
+To allow a dynamic system to provide semantics consistent with a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
 To do this, we take the conjunction of the precondition of an output action with a predicate over the set of bindings $B$ (called a \emph{binding predicate}) to ensure that the output action is not executed until its set of implied automata is correct.

### 2011-10-01 papers/paper1/design.tex
refactored section 2, moved part of it to 3

@@
 
-ioa++ implements the component model described in Section~\ref{component_model}, for POSIX environments (in C++).
+The ioa++ framework implements the component model described in Section~\ref{component_model}, for POSIX environments (in C++).
 Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
 The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
-The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
+
+Conceptually, state variables belong to the system automaton.
+However, the state they encode is shared and is used in the execution of every action.
+For a system $S=(A,B)$ consisting of a set of automata $A$ and a set of bindings $B$,
+a \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
+Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
+However, the opportunities for concurrency discussed in Section~\ref{component_structure} are not lost when one considers that only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+The scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
 All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.

### 2011-10-01 papers/paper1/component_model.tex
changes through section 3

@@
 input actions to receive the results.
-Component automata \emph{must} inherit from the base automaton, which as we describe in Section~\ref{framework_architecture} is implemented in ioa++ as a C++ class.
-Each component automaton is thus composed (not bound) with the system automaton and inherits output actions for requesting system actions and input actions for receiving the results of system actions.
+User-defined component automata (i.e., \emph{user automata}) inherit from the base automaton, which as we describe in Section~\ref{framework_architecture} is implemented in ioa++ as a C++ class. 
+Each user automaton is thus composed (not bound) with the system automaton and inherits output actions for requesting system actions and input actions for receiving the results of system actions.
 

### 2011-10-01 papers/paper1/design.tex
changes through section 3

@@
 
-The ioa++ framework implements the component model described in Section~\ref{component_model}, for POSIX environments (in C++).
-Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
+The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} (including services for executing local actions and facilities for manipulating the set of automata and bindings), for POSIX environments .
 The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
@@
 Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
-However, the opportunities for concurrency discussed in Section~\ref{component_structure} are not lost when one considers that only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.
+The opportunities for concurrent execution discussed in Section~\ref{component_structure} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
 The scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
-All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
-The services provided by that common base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
-The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
-In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
-An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
+
+All user automata inherit from a common Base Automaton class that provides facilities for asynchronously creating and destroying child automata and bindings.
+% The services provided by that common base class are indicated by the Base Automaton shown in Figure~\ref{framework_architecture}.
+The Base Automaton class contains statically composed System Actions that are used to communicate with the system automaton.
+In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a user automaton.
+Instead, an Automaton Manager provides a common higher-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
 Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
@@
 
-The scheduler is responsible both for selecting and for executing local actions.
-These two activities can be decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
+The scheduler is responsible for both selecting and executing local actions.
+These two activities are decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
 
 The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
-Input actions are executed with output actions according to the current set of bindings.
-The dispatcher allows executing actions concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
-Since concurrent access to the dispatcher is allowed, the module implementing the dispatcher must be thread-safe.
+Input actions are executed with corresponding output actions, according to the current set of bindings.
+The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
 
+Since concurrent access to the dispatcher is allowed, the dispatcher implementation the dispatcher must be thread-safe.
 Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
@@
 The run-time system can be initialized with different controllers to achieve different performance objectives.
-Controllers can use multiple threads of execution to exploit the concurrency available in a system of automata.
-The ioa++ framework is designed so that all native concurrency mechanisms (e.g., threads, mutexes) are encapsulated in the dispatcher and controller.
+To leverage the power of different operating system mechanisms without exposing users to their complexities directly, the ioa++ framework is designed so that all native concurrency and synchronization mechanisms (e.g., threads and mutexes) are encapsulated within the dispatcher and controller.
+For example, a controller can be written to use multiple threads of execution to exploit fully the concurrency inherently available in a system of automata,
+while still providing concurrency semantics in a manner consistent with the I/O automata component model.
 
@@
 The controller then has the responsibility of cycling through the actions according to some fair policy.
-Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to naively trying each action.
+Since the scheduler has no knowledge about which actions are \emph{enabled} or \emph{disabled} (have a precondition that is true or false respectively), this approach reduces to naively trying each action.
 
-A more focused approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
-The act of declaring an action to the scheduler is called \emph{scheduling}.
+A more focused approach, which is the default approach used in our framework implementation, is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
+%The act of declaring an action to the scheduler is called \emph{scheduling}.
 The controller then has the responsibility of processing the set of requested actions according to some fair policy.
-Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
 Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
-Consequently, an automaton is given the opportunity to schedule actions to the scheduler after one of its actions is executed.
+Consequently, an automaton is given the opportunity to declare actions to the scheduler (i.e., to \emph{schedule} them) after one of its actions is executed.
 Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
@@
 In this approach, scheduling disabled actions is unnecessary.
-Let $a$ be a scheduled disabled action.
-In order for $a$ to become enabled, the automaton associated with the action must change state by executing an enabled action $b$ sometime after $a$ is scheduled.
-Consequently, the scheduling of the disabled action $a$ can be deferred until the execution of action $b$.
+In order for a scheduled disabled action $a$ to become enabled, the automaton associated with the action must change state by executing another enabled action $b$ sometime after $a$ is scheduled.
+Consequently, the scheduling of the disabled action $a$ can be deferred until the execution of some other action.
 Notice that refraining from scheduling disabled actions does not affect the fairness of the scheduler since they cannot cause a state change when selected.
@@
 
-However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
+However, an automaton must still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
 A simple technique for accomplishing this is to schedule all enabled actions.
 More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
-However, this must be done carefully.
-Not scheduling an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
-Not scheduling an (enabled) action is a common mistake.
+However, this must be done carefully, since not considering an enabled action would violate the fairness of the controller.
+% since it will never select an action that may cause a state change.
+% Not scheduling an (enabled) action is a common mistake.
 
@@
 Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
-The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
+The bind action changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
 While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
@@
 This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
-Our solution assigns each automaton a unique identifier assuming that the set of automata turn over less frequently overall than the set of actions in the scheduler.
+Our solution assigns each automaton a unique identifier assuming that the set of automata in the system turn over less frequently overall than the set of actions that have been scheduled.
 
 \paragraph*{Delayed actions}
-Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities, i.e., the framework should support the development of automata that act as timers and alarms.
+Scheduling activities for a future time is a useful feature for many systems because it allows the system to sleep between activities.
+For example, the framework should support the development of automata that act as timers and alarms.
 Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
@@
 This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
-For our framework, we used the Reactor pattern since (1)~the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less widespread.
+The ioa++ framework uses the Reactor pattern since (1) it is simpler and (2) many operating systems have non-blocking I/O and some kind of synchronous I/O (de)multiplexing, whereas asynchronous I/O is less widespread.
 
@@
 
-\paragraph*{Global FIFO scheduler}
-The basic scheduler included with the ioa++ framework distribution is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
-Scheduling is idempotent to prevent unbounded growth of the queue.
-The global FIFO scheduler uses the Reactor pattern to implement delayed actions and file descriptor events.
+\paragraph*{Basic and multi-threaded schedulers}
+The ioa++ framework distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
+Scheduling is idempotent to prevent unbounded growth of the queue, and both scheduler implementations use the Reactor pattern to implement delayed actions and file descriptor events.
 
@@
 Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
-Subjects allow an automaton to have multiple outstanding requests.
+Subjects allow an automaton to have multiple requests outstanding at once.
 

### 2011-10-01 papers/paper1/design.tex
updates through section 4

@@
 
-All user automata inherit from a common Base Automaton class that provides facilities for asynchronously creating and destroying child automata and bindings.
-% The services provided by that common base class are indicated by the Base Automaton shown in Figure~\ref{framework_architecture}.
-The Base Automaton class contains statically composed System Actions that are used to communicate with the system automaton.
-In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a user automaton.
-Instead, an Automaton Manager provides a common higher-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
+All user automata inherit from a common base class (indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}) that provides facilities for asynchronously creating and destroying child automata and bindings.
+The Base Automaton layer provides statically composed system actions that are used to communicate with the system automaton.
+In most situations, the services provided by the Base Automaton layer are too low-level to be used directly by a user automaton.
+Instead, an Automaton Manager provides a common higher-level interface for asynchronously managing a single child automaton using the services of the Base Automaton layer.
 Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

### 2011-10-01 papers/paper1/programming_model.tex
updates through section 4

@@
-\section{I/O Automata Programming Model\label{programming_model}}
+\section{Programming Model\label{programming_model}}
 
-In this section, we show how I/O automata can be used to construct programs using ioa++.
+In this section, we show how I/O automata can be used to construct component based systems using ioa++.
 Section~\ref{representation} describes how I/O automata are represented in the C++ language.
@@
 
-All automata inherit from the common base class ioa::automaton which implements the system action interface.
+As was mentioned in Section~\ref{design}, all user automata inherit from the common base class (ioa::automaton) which implements the system action interface.
 \ifjournal
@@
 \end{lstlisting}
-The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept.
-The metaprogramming facilities of the C++ language aid in the development of general purpose reusable components.
-\fi
+The asynch\_lcr\_automaton is declared as a C++ template where the UID template parameter must model the \emph{unique identifier} concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
+The generic programming facilities of the C++ language aid in the development of general purpose reusable components.
 
 \paragraph*{Declaring and initializing state variables}
+\fi
 The state variables of an automaton are expressed as member variables of its corresponding class or class template.
@@
 \paragraph*{Declaring/defining actions}
-Actions are member variables that model the \emph{action} concept.
+Actions are member variables that model the \emph{action} 
+\ifjournal
+concept.
+\else
+concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
+\fi
 Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
@@
 However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
-A common pattern is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
-Another common pattern is to have all scheduling functions dispatch to a single scheduling function to avoid replicating scheduling logic.
+A common idiom is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
+Another common idiom is to have all scheduling functions dispatch to a single scheduling function to avoid replicating scheduling logic.
 \ifjournal
@@
 A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
-The ioa::automaton\_manager class and ioa::binding\_manager class contain methods for destroying and unbinding respectively.
+The ioa::automaton\_manager class and ioa::binding\_manager class also contain methods for destroying and unbinding respectively.
 \ifjournal
@@
 
-In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the ``nuts and bolts'' presented in the previous section.
-This list is not exhaustive but rather serves as a distillation of our experiences so far working with I/O automata in ioa++.
+In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the basic versions presented in the previous discussion.
+This list is not exhaustive but rather serves as a distillation of our experiences so far working with ioa++.
 
@@
 For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
-Similarly, an automaton might compose with an automaton that it does not manage and therefore may expect the automaton to exist before attempting to bind.
-To enable such behavior, user automata can observe ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.
+Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
+To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.
 
@@
 In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
-Using this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
+Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
 The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.
@@
 If all of the bindings of a child automaton are with its parent, then the child can be replaced using parameters.
-Instead of binding the new child to the exact same actions of the parent that are being used by the old child, which is the root problem in the preceding scenario, the new child is bound to actions parameterized by the new child.
+Instead of binding the new child to the exact same actions of the parent that are being used by the old child, which is the key problem in the preceding scenario, the new child is bound to actions parameterized by the new child.
 The parameters serve to differentiate the actions and obviate the need to wait for the old child to be destroyed.

### 2011-10-01 papers/paper1/case_study.tex
revised motivation at the start of the case study

@@
 
-In this section, we demonstrate that ioa++ can be used to build real distributed systems.
-Real distributed systems are designed using a layered approach where the upper layers contain semantically rich business logic while the bottom layers contain low-level network message-passing.
-The middle layers, i.e., middleware, contain generic high-level coordination mechanisms that abstract away the low-level details of message-passing.
-The lower layers contain abstractions of operating system services such as TCP/UDP sockets (for communicating with other peers) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
+In this section, we illustrate how ioa++ can be used to build real distributed systems.
+In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
+The lower layers provide services such as TCP/UDP sockets (for communicating between endsystems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
+The intermediate layers, i.e., frameworks like ioa++, contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.
+
 We apply this layered architecture technique to the simple but representative network programming problem of electing a leader in a unidirectional ring of processes connected by TCP sockets.  
+In order to focus on how ioa++ supports distributed applications in general, rather than on the details of any
+particular application, the application logic in this example consists simply of periodically circulating a token around a ring of nodes.
+If a node has not received a token in a specified amount of time it initiates a leader election protocol.
+Once a leader has been elected, the leader injects a token and the protocol resumes.
 
-The ``business logic'' we will implement is quite simple and consists of periodically circulating a token around a ring of nodes.
-If a node has not received a token in a specified amount of time, it initiates a leader election protocol.
-Once a leader has been elected, the leader injects a token and the protocol resumes.
-Nodes can join and leave the ring at any time.
-The protocol should eventually succeed for complete rings.
-These dynamics admit fault situations where a ring might lose its leader or contain multiple leaders.
-A ring containing multiple leaders should stabilize to having a single leader.
-A rare (but possible and valid) situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
-This situation requires no action as a token is circulating in the ring.
-We divide the problem into two sub-problems:  leader election and ring formation.
+Despite the simplicity of the application logic, the resulting \emph{dynamic} distributed system serves to motivate why and illustrate how a number of crucial features are provided by ioa++.
+For example, nodes can join and leave the ring at any time, and yet the protocol should eventually succeed for complete rings when they stabilize.
+These dynamics admit fault situations where a ring might lose its leader or temporarily contain multiple leaders, and yet each ring should stabilize to having a single leader.
+Another possible and valid (if potentially rare) situation is a ring circulating a token without a leader, i.e., the leader sends a token, leaves the ring, and is replaced before the token makes a full cycle.
+% This situation requires no action as a token is circulating in the ring.
+We divide these problems into two sub-problems:  leader election and ring formation.
 
@@
 The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
-Figure~\ref{tcp_lcr_automaton} shows the layered architecture of the TCP LCR system.
-The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the business logic.
-Thus, ioa++ supports features we expect from a component framework for distributed systems.
+
+Figure~\ref{tcp_lcr_automaton} shows the resulting layered architecture of the TCP LCR system.
+The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the application logic.
+As this simple but representative example illustrates, the ioa++ framework readily supports the development of architecturally layered component-based distributed systems involving non-trivial dynamics.
 

### 2011-10-01 papers/paper1/evaluation.tex
minor edits to evaluation

@@
 The ioa++ framework permits independent actions to be executed concurrently.
-The goal of this evaluation is to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising a system and the overhead of the framework.
-To test concurrent execution, we examine a system whose actions can be configured to span the range from having no independent actions to having only independent actions.
+The goal of this evaluation is to show that the degree of exploitable concurrency depends only on (1) the interactions of the I/O automata comprising a system, and (2) the overhead of the framework which this evaluation serves to quantify for our current implementation of ioa++.
+
+To evaluate concurrent execution, we examine a system whose actions can be configured to span the range from having no independent actions to having only independent actions.
 We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
@@
 
-An automaton $R$ contains an input, output, and internal action.
+An automaton of type $R$ contains an input, output, and internal action.
 The output and internal action effects execute an algorithm whose complexity is proportional to the parameter $N$.
@@
 
-For this experiment, we developed a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{scheduling}.
-When a single thread is used, all actions are executed by the same thread.
+For this experiment, we used a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{scheduling}.
+When the scheduler is configured to use a single thread, all actions are executed by that same thread.
 When two threads are used, one thread executes the actions of one $R$ automaton while the other thread executes the actions of the other $R$ automaton.
@@
 
-Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
-Confidence intervals are shown but are negligible with the largest being a speed-up factor of $\pm$0.0165.
+Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with two threads versus one thread.
+Confidence intervals are shown but are negligible with the largest being a speed-up range of $\pm$0.0165.
 When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
@@
 The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
-For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.
+For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends very little time on synchronization when $\rho = 1$.
 
-This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
+This experiment indicates that although concurrent execution is possible with ioa++, a significant speed-up will only be achieved if (1) enough independent actions are enabled and (2) the durations of the independent actions are large relative to the overhead of ioa++.
 The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
@@
 The synchronization time depends on the interactions among the automata comprising the system, the scheduler, and the locking scheme used by the framework.
-We plan to address scheduling and synchronization in future work.
+We plan to optimize scheduling and synchronization performance in ioa++ as future work.

### 2011-10-01 papers/paper1/component_model.tex
Fixes through page 6.

@@
 % The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
-As Figure~\ref{system_action} illustrates, the system automaton provides: parameterized input actions for receiving requests to create, bind, unbind, and destroy automata; internal actions for evaluating and enacting system action requests; and corresponding parameterized output actions through which to return results of the requests.
+As Figure~\ref{system_action} illustrates, the system automaton provides: parameterized input actions for receiving requests to create, bind, unbind, and destroy automata; internal actions for evaluating and enacting system action requests; and corresponding parameterized output actions for returning responses to system action requests.
+(Parameterized actions are indicated with [].)
 The base automaton provides output actions that trigger requests to create, bind, unbind, and destroy automata, along with
@@
 To do this, we take the conjunction of the precondition of an output action with a predicate over the set of bindings $B$ (called a \emph{binding predicate}) to ensure that the output action is not executed until its set of implied automata is correct.
-Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as was previously discussed.
 Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.

### 2011-10-01 papers/paper1/conclusion.tex
related work and conclusions tweake

@@
 I/O automata offer a good foundation for asynchronous and concurrent components due to their support for independent state, well-defined interfaces, and well-defined interactions under composition.
-The ioa++ framework facilitates concurrent execution via dynamic composition and the degree of concurrency is limited by the interactions of the automata and the overhead of the framework.
+The ioa++ framework facilitates concurrent execution via dynamic composition and the degree of concurrency is limited only by the interactions of the automata and the overhead of the framework.
 The two key challenges when moving from the formal model to an actual implementation were introducing features for managing dynamic constellations of automata and providing a concrete scheduling mechanism.
-We agree with~\cite{georgiou2009automated} that having a representation of a model that can be compiled is beneficial because it allows one to reason about the program directly from the source code and helps to close the gap between mental model and code listing.
-Our primary goal moving forward is to use the ioa++ framework to gain further experience building real systems with I/O automata.
+We agree with~\cite{georgiou2009automated} that having a representation of a model that can be compiled is beneficial because it allows one to reason about the program directly from the source code and helps to close the gap between the developer's mental model and the code listing.
+Our primary goal moving forward is to use the ioa++ framework to gain further experience building real systems with I/O automata, and to support optimization of its performance with respect to an increasingly diverse range of use cases.
 

### 2011-10-01 papers/paper1/related_work.tex
related work and conclusions tweake

@@
 Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
-Given the leveling of processor clock frequencies and the advent of multi-core, these same concerns are relevant to trends that look to deliver performance increases through increasingly concurrent programs~\cite{sutter2005software}.
+Given the shift from higher clock frequencies to larger numbers of cores as the driver of hardware performance, these same concerns are relevant to trends that look to deliver performance improvements through increasingly concurrent programs~\cite{sutter2005software}.
 
-The direct implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
-Structured programming simplifies development by allowing programmers to reason about their programs directly from the source code; regardless of whether of not the code is formally verified.
-The I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} was selected because it supports features essential to developing distributed systems~\cite{lynch1996distributed}.
+Leveraging a formal model of concurrency to make concurrent programming easier is a well established technique.
+We consider three exemplars that are most relevant to the discussion in this paper: Communicating Sequential Processes, Actors, and I/O automata.
+Of these, we selected the I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} because of its support for features essential to developing distributed systems~\cite{lynch1996distributed}.
 %% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.
 
-Incorporating a formal model of concurrency into a language to make concurrent programming easier is an established technique.
-We consider three exemplars:  Communicating Sequential Processes, Actors, and threads.
-The Communicating Sequential Processes (CSP) model~\cite{hoare1978communicating} is an extremely popular technique for modeling concurrent systems.
+The Communicating Sequential Processes (CSP) model~\cite{hoare1978communicating} is a popular technique for modeling concurrent systems.
 The fundamental concurrency control primitive in CSP is the \emph{rendezvous} which allows processes to synchronize and possibly exchange values.
 CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence the design of modern programming languages, e.g., Go~\cite{go}.
+
 An actor~\cite{agha1986actors} is a behavior that can perform a number of actions when receiving a message including (1) replacing itself with a new behavior, (2) sending messages to other actors, and (3) creating new actors.
-The Actor model influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.
-The collective formalization of threads, i.e., semaphores~\cite{dijkstra1968cooperating}, monitors~\cite{hoare1974monitors}, etc., appears in popular libraries, e.g., POSIX threads~\cite{butenhof1997programming}, and languages, e.g., Java~\cite{christopher2000high}.
+The Actor model has influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.
+
+% The direct implementation of a formal model to simplify concurrent programming was inspired by the application of structured programming~\cite{dijkstra1968letters} to programming language design.
+% Structured programming simplifies development by allowing programmers to reason about their programs directly from the source code; regardless of whether of not the code is formally verified.
+% The collective formalization of threads, i.e., semaphores~\cite{dijkstra1968cooperating}, monitors~\cite{hoare1974monitors}, etc., appears in popular libraries, e.g., POSIX threads~\cite{butenhof1997programming}, and languages, e.g., Java~\cite{christopher2000high}.
 
@@
 
+% Other efforts to implement I/O automata have focused on simulation and verification~\cite{goldman1990distributed},~\cite{georgiou2009automated}.
+% As observed in~\cite{georgiou2009automated}, a benefit of implementing distributed systems directly with I/O automata is the ability to reason about the behavior of a system or component directly from the source code using techniques from the I/O automata formalism.
+% However, where~\cite{georgiou2009automated} introduces a new language and uses the Message Passing Interface (MPI)~\cite{gropp1999using} library for communication, ioa++ uses the existing C++ language and exposes native operating system services.
+
 To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing approaches that allow one to execute a system expressed as a collection of I/O automata.
@@
 \cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
-Programs written in IOA are compiled to Java and executed on hosts communicating using the MPI library.
-IOA requires programs to presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
-IOA lacks support for procedure calls which complicates modeling and interfacing with external services: especially when a procedure call blocks.
+Programs written in IOA are compiled into Java and executed on hosts communicating using the MPI library.
+IOA requires programs to be presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
+IOA lacks support for procedure calls, which complicates modeling and interfacing with external services, especially when a procedure call may block.
 The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
@@
 IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
-The development of a new language was necessary for IOA due to their focus on formal verification.
-On the other hand ioa++ uses the existing C++ language and compilers.
+The development of a new language was necessary for IOA due to its focus on formal verification.
+On the other hand ioa++ uses the existing C++ language and compilers, towards easing portability of ioa++ systems to new platforms.
 IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors, relying on the programmer to ensure non-blocking operation.

### 2011-10-01 papers/paper1/design.tex

@@
 
-The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} (including services for executing local actions and facilities for manipulating the set of automata and bindings), for POSIX environments .
+The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} for POSIX environments .
 The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
 
-Conceptually, state variables belong to the system automaton.
+Conceptually, the set of automata $A$ and set of bindings $B$ belong to the system automaton.
 However, the state they encode is shared and is used in the execution of every action.
-For a system $S=(A,B)$ consisting of a set of automata $A$ and a set of bindings $B$,
-a \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
+A \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
 Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
 The opportunities for concurrent execution discussed in Section~\ref{component_structure} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
+Binding predicates do not modify $B$ and therefore do not affect concurrent execution.
 The scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
@@
 
-A more focused approach, which is the default approach used in our framework implementation, is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
+A more focused approach, which is the approach used in our framework implementation, is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
 %The act of declaring an action to the scheduler is called \emph{scheduling}.

### 2011-10-01 papers/paper1/case_study.tex
Updates through page 10.

@@
 In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
-The lower layers provide services such as TCP/UDP sockets (for communicating between endsystems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
+The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
 The intermediate layers, i.e., frameworks like ioa++, contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.

### 2011-10-01 papers/paper1/programming_model.tex
Updates through page 10.

@@
 A number of wrappers and macros are provided by ioa++ for declaring actions.
-%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in section~\ref{practical} are available.
+%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in Section~\ref{practical} are available.
 Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
@@
 However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
-A common idiom is to schedule all local actions using one member function where each action is guarded by its precondition as described in section~\ref{scheduling}.
-Another common idiom is to have all scheduling functions dispatch to a single scheduling function to avoid replicating scheduling logic.
+A common idiom is for each scheduling function in an automaton to call a common member function that schedules all enabled local actions to avoid replicating scheduling logic.
+The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{scheduling}.
+
 \ifjournal
@@
 \ifjournal
-The tcp\_lcr\_automaton discussed in section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
+The tcp\_lcr\_automaton discussed in Section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
 \begin{lstlisting}
@@
 The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
-Most often, ioa::binding\_count is used in the precondition of outputs to ensure that the value generated by the output will be received if produced.
+Most often, ioa::binding\_count is used in the precondition of output actions to ensure that the value generated by the output action will be received if produced.
 \ifjournal
@@
 Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
-For example, the Binding Manager mentioned in section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
+For example, the Binding Manager mentioned in Section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
 Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.

### 2011-10-01 papers/paper1/component_model.tex
Updates through page 10.
presubmission tweaks

@@
 Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
-The scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
+The scheduler is assumed to be \emph{fair,} meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
 Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
@@
 
+\begin{figure*}
+\center
+\includegraphics[width=\textwidth]{system_model}
+\caption{Example event detection system.}
+\label{sys_model}
+\end{figure*}
 Component frameworks that support \emph{recursive encapsulation} allow components to contain other components.
@@
 The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.
-
-\begin{figure}
-\center
-\includegraphics[width=\columnwidth]{system_model}
-\caption{Example event detection system.}
-\label{sys_model}
-\end{figure}
 
@@
 \paragraph*{Concurrent execution}
-Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
+Recall that composing I/O automata consists of concatenating state variables and folding input actions into output actions and that the result of composition is a single equivalent automaton.
 Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
@@
 The set of interactions in the I/O automata model is also fixed since automata are statically composed.
-Dynamic systems can be modeled in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
+Dynamic systems can be \emph{modeled} in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
 To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.
@@
 The \emph{destroy} action allows an automaton to destroy a child automaton and all bindings associated with it (recursively).
-\begin{figure}
+
+\begin{figure*}
 \center
-\includegraphics[width=\columnwidth]{system_action}
+\includegraphics[width=\textwidth]{system_action}
 \caption{System automaton and actions.}
 \label{system_action}
-\end{figure}
-
+\end{figure*}
 % A dynamic system implies that the system $S$ must be dependent on time $t$ as a function of the number of actions that have been executed by $t$.
@@
 %% The bind operation can be defined $\{exists (w) \land exists (o_a) \land exists (i_a) \land free (i_a, i_n, i_p) \land \lnot involved(o_a, o_n, o_p, i_a) \land o_a \neq i_a \} \quad bind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\}$ which says that in order for the binding to succeed the owner $w$ exists, automaton $o_a$ exists, automaton $i_a$ exists, the input action $(i_a, i_n, i_p)$ must not be bound, the output action $(o_n, o_p)$ of automaton $o_a$ cannot be bound to an input action in automaton $i_a$, and the automata $o_a$ and $i_a$ cannot be the same.
-%% The precondition for $bind$ is same as the binding rules derived in section~\ref{system_model} augmented with existence tests for the owner, output automaton, and input automaton.
+%% The precondition for $bind$ is same as the binding rules derived in Section~\ref{system_model} augmented with existence tests for the owner, output automaton, and input automaton.
 %% The unbind operation is similar to destroy: $\{(w, o_a, o_n, o_p, i_a, i_n, i_p) \in B\} \quad unbind (w, o_a, o_n, o_p, i_a, i_n, i_p) \quad \{(w, o_a, o_n, o_p, i_a, i_n, i_p) \notin B\}$.

### 2011-10-01 papers/paper1/design.tex
presubmission tweaks

@@
 
-The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} for POSIX environments .
+The ioa++ framework is a C++ implementation of the component model described in Section~\ref{component_model} for POSIX environments.
 The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
@@
 
-Since concurrent access to the dispatcher is allowed, the dispatcher implementation the dispatcher must be thread-safe.
+Since concurrent access to the dispatcher is allowed, the dispatcher implementation must be thread-safe.
 Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.

### 2011-12-14 papers/paper1/component_model.tex
Fixed typos.

@@
 A create fails if the automaton that should be added to the set of automata, i.e., the automaton being created, already exists.
-A bind fails if the automaton associated with the output action or the automaton associated with the input action doesn't exist, or if any of the binding rules outlined in Section~\ref{component_structure} would be violated.
+A bind fails if the automaton associated with the output action or the automaton associated with the input action does not exist, or if any of the binding rules outlined in Section~\ref{component_structure} would be violated.
 An unbind fails if the binding does not exist.
@@
 % Are partial successes allowed?
-% Furthermore, such  transactions can be arbitrarily complex which doesn't bode well for efficiency.
+% Furthermore, such  transactions can be arbitrarily complex which does not bode well for efficiency.
 

### 2011-12-14 papers/paper1/evaluation.tex
presubmission tweaks
Fixed typos.

@@
 The driver for a single trial can be found in examples/random.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/random.cpp}} of the ioa++ package.
-The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
+The trials were performed on a Mac Pro running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
 The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.
@@
 Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
-The fraction decreases as $\rho$ goes to 1 and the system transitions from acquiring two locks to one lock.
+The fraction decreases as $\rho$ goes from 0 to 1 because the system transitions from always acquiring two locks to always acquiring one lock.
 The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.

### 2011-12-14 papers/paper1/programming_model.tex
presubmission tweaks
Fixed typos.

@@
 Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
-These traits are used to guide the execution of the methods by the model and to supply values and parameters as necessary.
+These traits are used to guide the execution of the methods by the dispatcher and to supply values and parameters as necessary.
 The traits are also used to check that input actions are bound to output actions with the same value status and type.
@@
 Recall that system actions are asynchronous and can be executed in any order.
-Thus, the user might might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
+Thus, the user might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
 If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.

### 2012-02-23 papers/paper2/abstract.tex
Removing material.

\begin{abstract}
The ability to realize sophisticated distributed reactive systems such as those found in cyber-physical, pervasive, or enterprise computing environments is predicated on a suitable model and implementation of asynchrony and concurrency.
The size and complexity of these systems implies that software re-use in their development, maintenance, and evolution is also essential.
Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly in reusable software.

To that end, this paper presents the ioa++ framework for developing distributed reactive systems, which is the primary contribution of this research.
We develop a component model in which the asynchrony and concurrency of reactive distributed systems can be represented naturally based on I/O automata and dynamic composition.
We describe the design and implementation of that component model in ioa++ and illustrate its use in developing a prototype distributed coordination protocol.
We also evaluate concurrency performance in ioa++ through a series of micro-benchmarks which confirm that the framework’s ability to exploit potentially concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
\end{abstract}

%% \begin{abstract}
%% The ability to realize sophisticated distributed systems such as those found in cyber-physical, pervasive, or enterprise computing environments is predicated on a suitable model and implementation of asynchrony and concurrency.
%% The size and complexity of these systems implies that software re-use in their development, maintenance, and evolution is also essential.
%% Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly in reusable software.
%% % By ``different forms'' we mean single-threaded, multi-threaded, time triggered, event triggered, etc.

%% To that end, this paper presents the \emph{ioa++ framework} for developing distributed systems, which is the primary contribution of this research.
%% We develop a component model for asynchronous and concurrent systems based on I/O automata and dynamic composition.
%% We describe the design and implementation of the component model in ioa++ and show how it can be used to develop protocols and other forms of distributed software.

%% Our experiences demonstrate that ioa++ can help simplify developing distributed systems, by supporting reasoning about asynchronous and concurrent programs directly from source code using the I/O automata formal model.
%% Common concurrency and interaction semantics allow simple modules to be aggregated into complex systems without building adapters that convert from one concurrency model to another.
%% % I've just commented this out instead of trying to weave it in elsewhere, especially since I think it is already somewhat implied by the threads discussion in the intro - Chris
%% %Whereas reasoning about the correctness of modules in a threaded program requires considering interactions among the various threads, reasoning about the correctness of modules in ioa++ is limited to the modules themselves and a well-defined set of composition rules.

%% We evaluate concurrency performance in ioa++ through a series of micro-benchmarks which confirm that the framework's ability to exploit potentially concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
%% \end{abstract}

### 2012-02-23 papers/paper2/case_study.tex
Removing material.

--- 2011-10-01 papers/paper1/case_study.tex
@@
 The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
-The intermediate layers, i.e., frameworks like ioa++, contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.
+The intermediate layers contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.
 

### 2012-02-23 papers/paper2/component_model.tex
Removing material.

--- 2011-12-14 papers/paper1/component_model.tex
@@
 
-The Input/Output (I/O) automata model was developed by Lynch and Tuttle to model reactive concurrent and asynchronous systems~\cite{lynch1987hierarchical},\cite{lynch1996distributed}.
+The Input/Output (I/O) automata model was developed to model reactive concurrent and asynchronous systems~\cite{lynch1996distributed}.
 An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
@@
 \fi
-An automaton's output and internal actions constitute its \emph{local signature}~\cite{lynch1996distributed}.
 \ifjournal
@@
 \fi
-An automaton's output and input actions constitute its \emph{external signature}~\cite{lynch1996distributed}.
-
-An automaton communicates with other automata through the actions in its external signature.
 I/O automata are \emph{composed} by (1)~concatenating the state variables of the constituent automata, (2)~consolidating the effects of all input actions with the same name, and (3)~incorporating the effects of input actions into output actions with the same name.
-Composing automata that have identically named output actions would be ambiguous and thus is not allowed.
-The result of composition is a new automaton, so that the principles for reasoning about the behavior of a composed system are the same as the principles for reasoning about a single automaton.
-Often, the properties of the resulting aggregate automaton can be reasoned about using only the properties of the constituent automata.
-%% A useful operation in composite automata is \emph{hiding}~\cite{lynch1996distributed} which converts an output action to an internal action.
-
-Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
+Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action (an output or internal action~\cite{lynch1996distributed}) and then if the precondition is true applying the effect.
 The scheduler is assumed to be \emph{fair,} meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
-Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if and only if the precondition is true.
-Executing a composed output action consists of atomically evaluating the precondition, and if it is true applying the effect of the output action and all composed input actions using the value produced by the output action.
-The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
-The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.
 
@@
 \end{figure*}
-Component frameworks that support \emph{recursive encapsulation} allow components to contain other components.
-Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
-The contained component is the \emph{child} while the containing component is the \emph{parent}.
-The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
-A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.
 
-\paragraph*{Motivation for I/O automata based components}
+\paragraph*{Motivation for I/O automata based components.}
 %The I/O automata formalism was developed to model reactive systems~\cite{lynch1987hierarchical} and has been used in the design and verification of concurrent and distributed systems.
 I/O automata are a good foundation for such components because they have independent state, straightforward interfaces, and precise semantics for event generation, distribution, and consumption resulting in well-defined interactions under composition.
-\emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1987hierarchical},~\cite{lynch1994atomic}.
-For convenience, we extend the notion that a parent automaton is the composition of its children to include an anonymous child that contains state and actions attributed to the parent.
+\emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1994atomic} which allows a \emph{parent} to contain zero or more \emph{children}.
 The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.
@@
 
-\paragraph*{Explicit composition}
-Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
-However, when using recursive encapsulation, developers must have the ability to compose component automata regardless of the naming convention used by the original developers.
-To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
-Such an association is called a \emph{binding} and consists of a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.
-
-%% This is only true for static systems.  Dynamic system can do crazy things.
-Each component automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
-The automaton that prescribes a binding is said to \emph{own} that binding.
-The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
-A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
-For example, the fully-qualified name of the $in?(s)$ action of Converter C in Figure~\ref{sys_model} is $[root].C.in?(s)$.
-Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
-The map of bindings for the event detection system shown in Figure~\ref{sys_model} is: 
-\begin{displaymath}
-\begin{split}
-\{ ([root].out!(s), [root].C.in?(s)),\\
-   ([root].C.out!(t), [root].D.in?(t)),\\
-   ([root].D.detect!, [root].detected?) \}.
-\end{split}
-\end{displaymath}
-
-\paragraph*{Binding rules}
-To support name-based composition under the I/O automata model, the set of bindings must adhere to four essential rules.
+\paragraph*{Explicit composition.}
+Composition in the I/O automata model is defined at the automaton level and is based on actions having the same name.
+To prepare for dynamic composition, we redefine composition at the action level.
+An explicit association between an output action and input action is called a \emph{binding}.
+%% The automaton that prescribes a binding is said to \emph{own} the binding.
+%% A binding is denoted $(output, input, owner)$.
+The set of bindings in a system must adhere to four essential rules.
 First, the output action and input action of a binding must agree on the type being produced and consumed.
-For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
-The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., use of a signal).
 Second, an input action can be bound to at most one output action.
-Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
-A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
-For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
-This stronger rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
-Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
-%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
-%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
+Third, an output action cannot be bound to more than one input action in the same automaton.
 Fourth, an output action cannot be bound to an input action in the same automaton.
-Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
-The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
-To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
-However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
+The second, third, and fourth rules exist because they admit ambiguity into the relationship between the states of the associated automata.
 Graphically, an arrow originating in one component automaton must terminate in another component automaton as Figure~\ref{sys_model} illustrates.
@@
 
-\paragraph*{Concurrent execution}
+\paragraph*{Concurrent execution.}
 Recall that composing I/O automata consists of concatenating state variables and folding input actions into output actions and that the result of composition is a single equivalent automaton.
@@
 
-\paragraph*{Parameters and action types}
+\paragraph*{Parameters and action types.}
 Actions in the I/O automata model can be defined using parameters.
@@
 
-\paragraph*{Example}
+\paragraph*{Example.}
 Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
@@
 
-I/O automata thus provide a natural component abstraction and are therefore a suitable foundation for developing reusable components for asynchronous and concurrent software systems.
-Each action is only allowed to modify the state of the component automaton with which it is associated, which serves to encapsulate behavior and avoid unexpected side effects.
-The absence of shared state thus ensures that each component automaton can be deployed independently.
-The external signature of a component automaton constitutes its interface and the result of interacting with it is well-defined since every action is atomic.
-Composition, atomicity, asynchrony, and concurrency are thus well supported by the model itself.
-
 %% \subsection{Practical Considerations\label{practical}}
@@
 
-The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
-Assuming a countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
-The set of interactions in the I/O automata model is also fixed since automata are statically composed.
-Dynamic systems can be \emph{modeled} in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
-To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.
+%% The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
+%% Assuming a countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
+%% The set of interactions in the I/O automata model is also fixed since automata are statically composed.
+%% Dynamic systems can be \emph{modeled} in that approach by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
+%% To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for activation and deactivation are defined~\cite{lynch1994atomic}.
 
-\paragraph*{Limitations of the static model}
-While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
-To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
-System resources are wasted if the actual number of participants is much smaller than $N$.
-If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
-Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata.
-%% Consequently, we add the ability to deal with a dynamic set of automata and interactions.
+%% \paragraph*{Limitations of the static model.}
+%% While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system may result in either over-provisioning or inflexibility.
+%% To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
+%% System resources are wasted if the actual number of participants is much smaller than $N$.
+%% If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
+%% Thus, the techniques for \emph{modeling} a dynamic system of I/O automata are not necessarily appropriate for \emph{implementing} a dynamic system of I/O automata.
+%% %% Consequently, we add the ability to deal with a dynamic set of automata and interactions.
 
-To illustrate, let a system $S=(A,B)$ be a pair consisting of a set of automata $A$ and a set of bindings $B$.
-The I/O automata model assumes that $S$ and therefore $A$ and $B$ are both fixed, so that dynamics must be \emph{simulated} via the manipulation of activation flags.
+%% To illustrate, let a system $S=(A,B)$ be a pair consisting of a set of automata $A$ and a set of bindings $B$.
+%% The I/O automata model assumes that $S$ and therefore $A$ and $B$ are both fixed, so that dynamics must be \emph{simulated} via the manipulation of activation flags.
 
-\paragraph*{Support for dynamic systems}
-To address the limitations of the static representation of dynamic systems, ioa++ provides
-architectural support for dynamic systems by separating the concerns of \emph{enacting} versus \emph{initiating} 
-actions for managing dynamics into distinct \emph{system} and \emph{base} automata respectively.
+\paragraph*{Support for dynamic systems.}
+The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed} and models dynamic systems by activating and deactivating automata as appropriate~\cite{lynch1994atomic}.
+While appropriate for modeling, this assumption is not conducive to building flexible and efficient systems because it requires \emph{a priori} bounds on all resources.
+To address the limitations of the static representation of dynamic systems, ioa++ provides architectural support for dynamic systems by separating the concerns of \emph{enacting} versus \emph{initiating} actions for managing dynamics into distinct \emph{system} and \emph{base} automata respectively.
 We define four \emph{system actions} for managing dynamic systems of automata: \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
@@
 
-We note that system action requests may fail for a variety of reasons.
-A create fails if the automaton that should be added to the set of automata, i.e., the automaton being created, already exists.
-A bind fails if the automaton associated with the output action or the automaton associated with the input action does not exist, or if any of the binding rules outlined in Section~\ref{component_structure} would be violated.
-An unbind fails if the binding does not exist.
-Note that the owner of the binding is taken into account when determining whether the binding exists, such that only the owner of a binding can dissolve it.
-A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.
-
 %% Let $A$ be the set of automata that exist where each element in $A$ is a pair $(p, a)$ where $p$ is the parent of $a$.
@@
 
-\paragraph*{Binding predicates}
-The static configuration of a system as prescribed by the I/O automata model guarantees that the set of implied automata for an action is always the same.
-This property is necessary and desirable for reasoning about the behavior of the system.
-Admitting dynamics removes this guarantee since the set of implied automata will change according to the various system actions being executed.
-To be specific, the set of implied automata for internal actions does not change but the set of implied automata for output actions can be changed with a successful bind, unbind, or destroy.
-
-To allow a dynamic system to provide semantics consistent with a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
-To do this, we take the conjunction of the precondition of an output action with a predicate over the set of bindings $B$ (called a \emph{binding predicate}) to ensure that the output action is not executed until its set of implied automata is correct.
-Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
-A system defined statically can be implemented as a dynamic system by adding binding predicates and logic to create the required constellation of automata.
+\paragraph*{Binding predicates.}
+Consider a simple system consisting of two automata where automaton A reliably transfers messages to automaton B.
+In a dynamic system, one cannot guarantee that the message will be received because there is no way to determine if the output action producing the message is bound to the input action receiving the message.
+Since we can guarantee that the set of binding does not change during the execution of an action, we allow automata to query the current set of bindings using a \emph{binding predicate}.
+Binding predicates are typically used the precondition of output actions to guarantee delivery.

### 2012-02-23 papers/paper2/conclusion.tex
Removing material.

--- 2011-10-01 papers/paper1/conclusion.tex
@@
 
-I/O automata offer a good foundation for asynchronous and concurrent components due to their support for independent state, well-defined interfaces, and well-defined interactions under composition.
+I/O automata are a good basis for asynchronous and concurrent components due to their support for independent state, well-defined interfaces, and well-defined interactions under composition.
 The ioa++ framework facilitates concurrent execution via dynamic composition and the degree of concurrency is limited only by the interactions of the automata and the overhead of the framework.
-The two key challenges when moving from the formal model to an actual implementation were introducing features for managing dynamic constellations of automata and providing a concrete scheduling mechanism.
-We agree with~\cite{georgiou2009automated} that having a representation of a model that can be compiled is beneficial because it allows one to reason about the program directly from the source code and helps to close the gap between the developer's mental model and the code listing.
+Two key challenges when moving from the formal model to an actual implementation were introducing features for managing dynamic constellations of automata and providing a concrete scheduling mechanism.
+We agree with~\cite{georgiou2009automated} that having a model that can be compiled is beneficial for reasoning about the program directly from the source code and helps to close the gap between the developer's mental model and the code listing.
 Our primary goal moving forward is to use the ioa++ framework to gain further experience building real systems with I/O automata, and to support optimization of its performance with respect to an increasingly diverse range of use cases.

### 2012-02-23 papers/paper2/design.tex
Removing material.

--- 2011-10-01 papers/paper1/design.tex
@@
 \center
-\includegraphics[width=\columnwidth]{architecture}
+\includegraphics[width=.6\columnwidth]{architecture}
 \caption{Framework architecture.}
@@
 These two activities are decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
-
 The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
@@
 The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
-
-Since concurrent access to the dispatcher is allowed, the dispatcher implementation must be thread-safe.
-Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
-Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
-All other actions acquire a read lock to implement single-writer/multiple-reader semantics.
-The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
-The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
-Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.
-
 The controller encodes a \emph{policy} that selects the next local action.
@@
 The run-time system can be initialized with different controllers to achieve different performance objectives.
-To leverage the power of different operating system mechanisms without exposing users to their complexities directly, the ioa++ framework is designed so that all native concurrency and synchronization mechanisms (e.g., threads and mutexes) are encapsulated within the dispatcher and controller.
-For example, a controller can be written to use multiple threads of execution to exploit fully the concurrency inherently available in a system of automata,
-while still providing concurrency semantics in a manner consistent with the I/O automata component model.
+The ioa++ framework distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
 
-\paragraph*{Explicit dynamic scheduling}
-The I/O automata model assumes that the scheduler knows about every local action in the system.
-There are two approaches for achieving this in practice.
-In the first approach, automata declare \emph{statically} all of their actions to the scheduler.
-The controller then has the responsibility of cycling through the actions according to some fair policy.
-Since the scheduler has no knowledge about which actions are \emph{enabled} or \emph{disabled} (have a precondition that is true or false respectively), this approach reduces to naively trying each action.
-
-A more focused approach, which is the approach used in our framework implementation, is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
+\paragraph*{Explicit dynamic scheduling.}
+In our framework implementation, automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
 %The act of declaring an action to the scheduler is called \emph{scheduling}.
-The controller then has the responsibility of processing the set of requested actions according to some fair policy.
 Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
@@
 For bootstrapping, each automaton is allowed to schedule actions when it is created.
+Scheduling is idempotent to prevent unbounded growth of scheduler data structures.
 
-In this approach, scheduling disabled actions is unnecessary.
-In order for a scheduled disabled action $a$ to become enabled, the automaton associated with the action must change state by executing another enabled action $b$ sometime after $a$ is scheduled.
-Consequently, the scheduling of the disabled action $a$ can be deferred until the execution of some other action.
-Notice that refraining from scheduling disabled actions does not affect the fairness of the scheduler since they cannot cause a state change when selected.
-This result leads to a guarded programming idiom where the precondition of each action is checked before it is added to the schedule.
-
-However, an automaton must still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
-A simple technique for accomplishing this is to schedule all enabled actions.
-More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
-However, this must be done carefully, since not considering an enabled action would violate the fairness of the controller.
-% since it will never select an action that may cause a state change.
-% Not scheduling an (enabled) action is a common mistake.
-
-\paragraph*{Binding predicates and scheduling}
+\paragraph*{Binding predicates and scheduling.}
 %% This assumption is not implicit because we state that the preconditions are independent.
@@
 
-\paragraph*{Actions of no-longer-existing automata}
-Every action in the set of actions considered by the controller is scheduled by its associated automaton.
-Since automata can be destroyed, an action in the set owned by the controller might belong to an automaton that no longer exists.
-Our solution is to check if the automaton exists before executing an action.
-This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
-Our solution assigns each automaton a unique identifier assuming that the set of automata in the system turn over less frequently overall than the set of actions that have been scheduled.
-
-\paragraph*{Delayed actions}
+\paragraph*{Delayed actions.}
 Scheduling activities for a future time is a useful feature for many systems because it allows the system to sleep between activities.
@@
 
-\paragraph*{File descriptor events}
+\paragraph*{File descriptor events.}
 File descriptors are used to communicate with the outside world in a POSIX environment.
@@
 
-\paragraph*{Basic and multi-threaded schedulers}
-The ioa++ framework distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
-Scheduling is idempotent to prevent unbounded growth of the queue, and both scheduler implementations use the Reactor pattern to implement delayed actions and file descriptor events.
-
 \subsection{System Actions\label{system_action_section}}
@@
 The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with an indication of either success or failure.
-Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
-Subjects allow an automaton to have multiple requests outstanding at once.
-
-Automata use a simple protocol for creating and destroying child automata.
-Each child automaton is represented by a unique subject which is in one or two of five states: CreateSend, CreateRecv, CreateDone, DestroySend, and DestroyRecv.
-Subjects start in the CreateSend state which indicates that the request for this subject is waiting to be sent to the system automaton.
-Once the request has been sent, the subject transitions to the CreateRecv state and waits for a response from the system automaton.
-Once the system automaton sends the response, the subject transitions to the CreateDone state.
-When the user automaton requests that the child automaton be destroyed, the subject is either (1) added to the DestroySend state if the subject is in CreateRecv or CreateDone or (2) removed from all states if the subject is in CreateSend.
-A subject appearing in both CreateDone and DestroySend transitions to the DestroyRecv state when the automaton makes the request to destroy it.
-The system automaton can process system actions in any order.
-This protocol ensures that an automaton doesn't request that a child automaton be destroyed until it receives confirmation that it was created.
-The subject is forgotten when the system automaton responds with the result of the destroy action.
-
-The protocol for binding is similar but addresses the possibility that a binding can be dissolved at any point due to the asynchronous destruction of automata.
-We define analogous states for the binding state machine: BindSend, BindRecv, BindDone, UnbindSend, and UnbindRecv.
-A binding subject can be in either BindRecv, BindDone, UnbindSend, or UnbindRecv when it receives a result from the system automaton indicating that the binding was dissolved.
-
-\paragraph*{Automaton and Binding managers}
 The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple constellations but becomes unwieldy when creating more complicated ones.
@@
 \ifjournal
-\paragraph*{Generators}
+\paragraph*{Generators.}
 A \emph{generator} is a single-use factory for creating an instance of an automaton.

### 2012-02-23 papers/paper2/paper.tex
Removing material.

--- 2011-09-22 papers/paper1/paper.tex
@@
-\documentclass[10pt, conference, compsocconf]{IEEEtran}
+\documentclass{llncs}
 
@@
 
-\title{Implementing Distributed Systems Directly with I/O Automata in Ioa++}
-\author{
-  \IEEEauthorblockN{Justin R. Wilson and Christopher D. Gill}
-  \IEEEauthorblockA{Department of Computer Science and Engineering\\
-    Campus Box 1045, Washington University\\
-    One Brookings Drive\\
-    St. Louis, MO 63130\\
-    {wilsonj,cdgill}@cse.wustl.edu}
+\title{Implementing Reactive Systems Directly\\with I/O Automata in ioa++}
+\author{Justin R. Wilson\inst{1} \and Christopher D. Gill\inst{1} \and Gruia-Catalin Roman\inst{2}}
+
+\institute{
+  Dept. of Computer Science and Engineering, 
+  Washington University, St. Louis, MO
 \and
-  \IEEEauthorblockN{Guia-Catalin Roman}
-  \IEEEauthorblockA{TODO:  Dr. Roman's address}
+  School of Engineering,
+  University of New Mexico, Albuquerque, NM
+  \{wilsonj,cdgill\}@cse.wustl.edu gcroman@unm.edu
 }
@@
 \input{abstract}
+
+\keywords{concurrency, I/O automata, components}
 
@@
 
-\bibliographystyle{IEEEtran}
+\bibliographystyle{splncs03}
 \bibliography{bibliography}{}

### 2012-02-23 papers/paper2/programming_model.tex
Removing material.

--- 2011-12-14 papers/paper1/programming_model.tex
@@
 
-\paragraph*{Declaring and initializing state variables}
+\paragraph*{Declaring and initializing state variables.}
 \fi
@@
 
-\paragraph*{Declaring/defining actions}
+\paragraph*{Declaring/defining actions.}
 Actions are member variables that model the \emph{action} 
@@
 
-\paragraph*{Scheduling}
+\paragraph*{Scheduling.}
 Automata submit actions to the scheduler using the ioa::schedule function.
@@
 
-\paragraph*{Dynamic composition}
+\paragraph*{Dynamic composition.}
 %% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
@@
 
-\paragraph*{Binding predicates}
+\paragraph*{Binding predicates.}
 Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
@@
 
-\paragraph*{System action events}
+\paragraph*{System action events.}
 Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
@@
 
-\paragraph*{Error reporting}
+\paragraph*{Error reporting.}
 The event semantics of I/O automata provide a natural mechanism for reporting errors.
@@
 
-\paragraph*{Handling errors by resetting}
+\paragraph*{Handling errors by resetting.}
 An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
@@
 
-\paragraph*{Handling errors by replacement}
+\paragraph*{Handling errors by replacement.}
 The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.

### 2012-02-23 papers/paper2/related_work.tex
Removing material.

--- 2011-10-01 papers/paper1/related_work.tex
@@
 Leveraging a formal model of concurrency to make concurrent programming easier is a well established technique.
-We consider three exemplars that are most relevant to the discussion in this paper: Communicating Sequential Processes, Actors, and I/O automata.
-Of these, we selected the I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} because of its support for features essential to developing distributed systems~\cite{lynch1996distributed}.
+We consider three exemplars that are most relevant to the discussion in this paper: Communicating Sequential Processes~\cite{hoare1978communicating}, Actors~\cite{agha1986actors}, and I/O automata~\cite{lynch1996distributed}.
+Of these, we selected the I/O automata model because of its support for features essential to developing distributed systems.
 %% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.
-
-The Communicating Sequential Processes (CSP) model~\cite{hoare1978communicating} is a popular technique for modeling concurrent systems.
-The fundamental concurrency control primitive in CSP is the \emph{rendezvous} which allows processes to synchronize and possibly exchange values.
-CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence the design of modern programming languages, e.g., Go~\cite{go}.
-
-An actor~\cite{agha1986actors} is a behavior that can perform a number of actions when receiving a message including (1) replacing itself with a new behavior, (2) sending messages to other actors, and (3) creating new actors.
-The Actor model has influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.
 
@@
 The development of a new language was necessary for IOA due to its focus on formal verification.
-On the other hand ioa++ uses the existing C++ language and compilers, towards easing portability of ioa++ systems to new platforms.
+% On the other hand ioa++ uses the existing C++ language and compilers.
 IOA encapsulates external services using built in ``mediators'' while ioa++ exposes operating system services via file descriptors, relying on the programmer to ensure non-blocking operation.

### 2012-02-24 papers/paper2/case_study.tex
Down to 17 pages.

@@
 The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
-The intermediate layers contain generic coordination mechanisms that serve to abstract away the low-level details from the application logic.
+Intermediate layers offer generic coordination mechanisms that serve to abstract away the low-level details from the application logic.
 
@@
 
-Nodes in a distributed system may be arranged in a ring so that two consecutive nodes are connected by a directed channel.
-The goal is to designate one of the nodes as the leader of the ring.
-The solution we adopt is the asynchronous LCR algorithm presented by Lynch in Chapter 15 of~\cite{lynch1996distributed} which is in turn based on the work of LeLann~\cite{le1977distributed} and Chang and Roberts~\cite{chang1979improved}.
+Nodes in a distributed system may be arranged in a ring with one of the nodes as the leader of the ring, e.g., elected per the asynchronous LCR algorithm~\cite{lynch1996distributed}.
 The LCR algorithm assumes that every node has a unique identifier (UID).
@@
 We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
-Our implementation of the asynchronous LCR automaton is located in examples/asynch\_lcr\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}} of the ioa++ package.
+Our implementation of the asynchronous LCR automaton is also freely available as open-source\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}}.
 
@@
 The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
-The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{constellations} to manage the dynamic constellation of automata and recover from errors.
+The tcp\_ring\_automaton uses timeouts and the techniques listed in Section~\ref{programming_model} to manage the dynamic constellation of automata and recover from errors.
 The tcp\_ring\_automaton contains two external actions:  one for receiving messages from a predecessor and one for sending messages to a successor.
 
-\begin{figure}
-\center
-\includegraphics[width=\columnwidth]{tcp_lcr_automaton}
-\caption{Architecture of the TCP LCR system.}
-\label{tcp_lcr_automaton}
-\end{figure}
+%% \begin{figure}
+%% \center
+%% \includegraphics[width=\columnwidth]{tcp_lcr_automaton}
+%% \caption{Architecture of the TCP LCR system.}
+%% \label{tcp_lcr_automaton}
+%% \end{figure}
 
 The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
-
-Figure~\ref{tcp_lcr_automaton} shows the resulting layered architecture of the TCP LCR system.
-The bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the application logic.
+In the resulting layered architecture of the TCP LCR system, the bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the application logic.
 As this simple but representative example illustrates, the ioa++ framework readily supports the development of architecturally layered component-based distributed systems involving non-trivial dynamics.

### 2012-02-24 papers/paper2/component_model.tex
Down to 17 pages.

@@
 
-\subsection{Background: I/O Automata}
-
+\paragraph*{I/O automata.}
 The Input/Output (I/O) automata model was developed to model reactive concurrent and asynchronous systems~\cite{lynch1996distributed}.
@@
 
-\subsection{Component Structure\label{component_structure}}
-
 \begin{figure*}
@@
 I/O automata are a good foundation for such components because they have independent state, straightforward interfaces, and precise semantics for event generation, distribution, and consumption resulting in well-defined interactions under composition.
-\emph{Hierarchical decomposition}, the modeling dual of recursive encapsulation, is an established technique for I/O automata~\cite{lynch1994atomic} which allows a \emph{parent} to contain zero or more \emph{children}.
+\emph{Hierarchical decomposition} is an established technique for I/O automata~\cite{lynch1994atomic} which allows a \emph{parent} to contain zero or more \emph{children}.
 The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.
@@
 \paragraph*{Explicit composition.}
-Composition in the I/O automata model is defined at the automaton level and is based on actions having the same name.
+Composition of I/O automata is based on actions having the same name.
 To prepare for dynamic composition, we redefine composition at the action level.
@@
 \paragraph*{Concurrent execution.}
-Recall that composing I/O automata consists of concatenating state variables and folding input actions into output actions and that the result of composition is a single equivalent automaton.
 Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
@@
 
-\subsection{Dynamics\label{dynamics}}
-
 %% The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.

### 2012-02-24 papers/paper2/design.tex
Down to 17 pages.

@@
 The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
-
 Conceptually, the set of automata $A$ and set of bindings $B$ belong to the system automaton.
 However, the state they encode is shared and is used in the execution of every action.
-A \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_structure} and the system automaton.
-The opportunities for concurrent execution discussed in Section~\ref{component_structure} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
+The \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
+Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_model} and the system automaton.
+The opportunities for concurrent execution discussed in Section~\ref{component_model} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
 Binding predicates do not modify $B$ and therefore do not affect concurrent execution.
@@
 
-\subsection{Scheduler\label{scheduling}}
-
 The scheduler is responsible for both selecting and executing local actions.
@@
 The controller encodes a \emph{policy} that selects the next local action.
-For the framework to be a valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
-The run-time system can be initialized with different controllers to achieve different performance objectives.
-The ioa++ framework distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
+For fidelity to the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
+The ioa++ distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
 
 \paragraph*{Explicit dynamic scheduling.}
-In our framework implementation, automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
-%The act of declaring an action to the scheduler is called \emph{scheduling}.
-Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
-Consequently, an automaton is given the opportunity to declare actions to the scheduler (i.e., to \emph{schedule} them) after one of its actions is executed.
+Automata declare \emph{dynamically} the actions they would like the scheduler to consider, and executing an action may change the state of the automaton which in turn may cause the set of enabled actions to change.
+Consequently, an automaton is given the opportunity to declare new actions to the scheduler after one of its actions is executed.
 Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
@@
 %% This assumption is not implicit because we state that the preconditions are independent.
-The preceding discussion on scheduling assumed that an action in one automaton cannot affect preconditions in another automaton.
-This is not entirely correct when one considers the system automaton and binding predicates.
+In a static system, executing an action in one automaton cannot change a precondition in another automaton.
+In a dynamic system, executing an action in the system automaton can change preconditions in other automata through binding predicates.
 For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
@@
 The bind action changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
-While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
+This example is representative of situations where an automaton is waiting for one or more of its outputs to be bound.
 To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after each successful bind or unbind.
@@
 \paragraph*{Delayed actions.}
-Scheduling activities for a future time is a useful feature for many systems because it allows the system to sleep between activities.
-For example, the framework should support the development of automata that act as timers and alarms.
+Scheduling activities for a future time may be useful to allow the system to sleep between activities.
+For example, the framework should support automata that act as timers and alarms.
 Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
@@
 Thus, we require techniques that allow a program (1) to wait until a file descriptor is asynchronously ready for I/O and (2) to perform non-blocking I/O operations.
-This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
-The ioa++ framework uses the Reactor pattern since (1) it is simpler and (2) many operating systems have non-blocking I/O and some kind of synchronous I/O (de)multiplexing, whereas asynchronous I/O is less widespread.
+This causes an action to be scheduled when a file descriptor is ready for reading or writing.
+A user automaton creates a file descriptor and configures it for non-blocking I/O.
+The automaton then tells the scheduler to release a local action whenever the automaton becomes ready for reading or writing.
+Subsequent requests made using the same file descriptor and event are ignored.
+Once the file descriptor is ready, the action is released for consideration by the controller.
 
-Our use of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
-A user automaton creates a file descriptor and configures it for non-blocking I/O.
-The automaton then tells the scheduler to release an action (typically an internal action) whenever the automaton becomes ready for reading (or writing).
-Subsequent requests made using the same file descriptor and event (ready for reading, ready for writing) are ignored.
-Once the file descriptor is ready, the action is released for consideration by the controller.
-The scheduler also provides a special function for closing file descriptors that purges them from the reactor.
-
-\subsection{System Actions\label{system_action_section}}
-
-Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
+\paragraph*{System actions.}
+Our approach to system actions follows the model introduced in Section~\ref{component_model}.
 All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.

### 2012-02-24 papers/paper2/evaluation.tex
Removing material.
Down to 17 pages.

--- 2011-12-14 papers/paper1/evaluation.tex
@@
 
-For this experiment, we used a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{scheduling}.
+For this experiment, we used a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{design}.
 When the scheduler is configured to use a single thread, all actions are executed by that same thread.
@@
 
-\begin{figure}
-\center
-\includegraphics[width=\columnwidth]{speed_up}
-\caption{Speed-up for system $S$.\label{speed_up}}
-\end{figure}
-
 Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with two threads versus one thread.
@@
 When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.
-
 The speed-up shows a strong dependence on the duration of each local action which is proportional to $N$.
@@
 \center
-\includegraphics[width=\columnwidth]{overhead1}
+\includegraphics[width=.72\columnwidth]{speed_up}
+\caption{Speed-up for system $S$.\label{speed_up}}
+\end{figure}
+
+\begin{figure}
+\center
+\includegraphics[width=.72\columnwidth]{overhead1}
 \caption{Overhead for system $S$ when using one thread.\label{overhead1}}
@@
 \center
-\includegraphics[width=\columnwidth]{overhead2}
+\includegraphics[width=.72\columnwidth]{overhead2}
 \caption{Overhead for system $S$ when using two threads.\label{overhead2}}
@@
 The fraction decreases as $\rho$ goes from 0 to 1 because the system transitions from always acquiring two locks to always acquiring one lock.
-The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
+The multi-threaded execution shows that the system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
 For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends very little time on synchronization when $\rho = 1$.

### 2012-02-24 papers/paper2/programming_model.tex
Down to 17 pages.

@@
 In this section, we show how I/O automata can be used to construct component based systems using ioa++.
-Section~\ref{representation} describes how I/O automata are represented in the C++ language.
-We then discuss the mechanisms available for dynamic composition in Section~\ref{dynamic_composition}.
-Section~\ref{constellations} concludes with techniques for managing dynamic constellations of automata.
-
-\subsection{I/O Automata Representation\label{representation}}
-
+We then discuss the mechanisms available for dynamic composition and techniques for managing dynamic constellations of automata.
 As was mentioned in Section~\ref{design}, all user automata inherit from the common base class (ioa::automaton) which implements the system action interface.
@@
 \paragraph*{Declaring/defining actions.}
-Actions are member variables that model the \emph{action} 
-\ifjournal
-concept.
-\else
-concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
-\fi
-Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
+Actions are member variables that contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
 These traits are used to guide the execution of the methods by the dispatcher and to supply values and parameters as necessary.
 The traits are also used to check that input actions are bound to output actions with the same value status and type.
-Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.
-
-A number of wrappers and macros are provided by ioa++ for declaring actions.
-%%To facilitate declaring actions, a number of wrappers and macros corresponding to the action types listed in Section~\ref{practical} are available.
+Bindings are checked for type compatibility at compile time.
 Local actions are declared using three functions corresponding respectively to precondition, effect, and scheduling.
-%%The macros assume the functions are named action\_name\_precondition, action\_name\_effect, and action\_name\_schedule.
 \ifjournal
@@
 A common idiom is for each scheduling function in an automaton to call a common member function that schedules all enabled local actions to avoid replicating scheduling logic.
-The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{scheduling}.
+The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{design}.
 
@@
 \fi
-
-\subsection{Dynamics\label{dynamic_composition}}
 
@@
 \paragraph*{Binding predicates.}
-Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
-The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of other actions bound to the specified action.
+Recall that automata require the ability to inspect the current set of bindings.
+The binding predicate supplied by ioa++, ioa::binding\_count, returns the number of other actions bound to the specified action.
 The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
@@
 
-\subsection{Managing Dynamic Constellations\label{constellations}}
-
-In this section, we introduce techniques and principles for managing dynamic constellations of automata and bindings that go beyond the basic versions presented in the previous discussion.
-This list is not exhaustive but rather serves as a distillation of our experiences so far working with ioa++.
-
-\paragraph*{System action events.}
-Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
-For example, the Binding Manager mentioned in Section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
+\paragraph*{Managing Dynamic Constellations}
+Determining the outcome of a system action event is necessary when there are constraints on the order in which system actions must occur.
+For example, the Binding Manager mentioned in Section~\ref{design} waits until the automata that supply the output action and input action are created, before attempting to bind.
 Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
@@
 
-\paragraph*{Error reporting.}
 The event semantics of I/O automata provide a natural mechanism for reporting errors.
@@
 Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
-The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.
-
-\paragraph*{Handling errors by resetting.}
+The two techniques for handling an error are either to reset or to replace the portion of the composed automata that encountered the error.
 An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
-Automaton designers should strive to write automata that are capable of being reset as this is the simplest error handling strategy.
-
-\paragraph*{Handling errors by replacement.}
-The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
 To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.

### 2012-02-24 papers/paper2/related_work.tex
Down to 17 pages.

@@
 Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
-Given the shift from higher clock frequencies to larger numbers of cores as the driver of hardware performance, these same concerns are relevant to trends that look to deliver performance improvements through increasingly concurrent programs~\cite{sutter2005software}.
+Given the shift from higher clock frequencies to larger numbers of cores as the main driver of hardware performance, these same concerns are relevant to trends that look to deliver performance improvements through increasingly concurrent programs~\cite{sutter2005software}.
 
-Leveraging a formal model of concurrency to make concurrent programming easier is a well established technique.
-We consider three exemplars that are most relevant to the discussion in this paper: Communicating Sequential Processes~\cite{hoare1978communicating}, Actors~\cite{agha1986actors}, and I/O automata~\cite{lynch1996distributed}.
-Of these, we selected the I/O automata model because of its support for features essential to developing distributed systems.
+Formal models of concurrency are also relevant to the discussion in this paper, e.g., Communicating Sequential Processes~\cite{hoare1978communicating}, Actors~\cite{agha1986actors}, and I/O automata~\cite{lynch1996distributed}.
+Of these, we selected the I/O automata model because it offers a straightforward representation of reactive systems.
 %% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.
@@
 To our knowledge, the Spectrum Simulation System~\cite{goldman1990distributed} and the IOA language~\cite{garland2003ioa} are the only existing approaches that allow one to execute a system expressed as a collection of I/O automata.
-Spectrum focuses on algorithm development through simulation.
-IOA focuses on formal software development including both simulation and verification.
-
-The ioa++ framework can be compared directly to the IOA language due to the availability of a compiler~\cite{tsai2002code},~\cite{tauber2004verifiable},~\cite{tauber2004compiling} that has been used to implement a number of protocols~\cite{georgiou2009automated}.
-\cite{georgiou2009automated} lists six challenges when developing the compiler: ``Program structuring,'' ``IOA Programs and external services,'' ``Modeling procedure calls,'' ``Composing automata,'' ``Nondeterminism'', and ``Implementing datatypes.''
-Programs written in IOA are compiled into Java and executed on hosts communicating using the MPI library.
-IOA requires programs to be presented in a ``node-channel'' form that allows them to be composed with built in ``mediators'' that implement external services such as MPI.
-IOA lacks support for procedure calls, which complicates modeling and interfacing with external services, especially when a procedure call may block.
-The IOA compiler composes automata statically using a ``composer'' resulting in a ``node automaton.''
-The next action to be executed is selected by a function that resolves the non-determinism in the schedule.
-
+Spectrum focuses on algorithm development through simulation while the focus of ioa++ is to support the direct implementation of real systems.
 IOA and ioa++ represent two fundamentally different approaches to programming directly with I/O automata.
@@
 IOA focuses on static composition to produce a single node program while ioa++ focuses on dynamic composition and allows concurrent execution.
-Both IOA and ioa++ require the programmer to add code to schedule actions, which may be a common source of errors.
+Both IOA and ioa++ require the programmer to add code to schedule actions.
 In IOA, the programmer provides a deterministic scheduling policy.

### 2012-02-24 papers/paper2/case_study.tex
Down to 16 pages.

@@
 In practice, distributed systems are designed using a layered approach where the highest layers contain (often semantically rich) application logic.
-The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks), e.g., \cite{schmidt2001c++},~\cite{schmidt2002c++}.
+The lower layers provide services such as TCP/UDP sockets (for communicating between end systems) and timers (for detecting time-outs and performing periodic tasks).
 Intermediate layers offer generic coordination mechanisms that serve to abstract away the low-level details from the application logic.
@@
 We add actions for starting an election and withdrawing an election which are necessary to cope with the dynamics discussed previously.
-Our implementation of the asynchronous LCR automaton is also freely available as open-source\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}}.
+Our asynchronous LCR automaton is also freely available as open-source\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/asynch_lcr_automaton.hpp}}.
 
 A node in the ring passively waits for a connection from its successor and actively tries to connect to its successor.
-The tcp\_ring\_automaton listed in examples/tcp\_ring\_automaton.hpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}} of the ioa++ package implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
+The tcp\_ring\_automaton\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_ring_automaton.hpp}} implements this logic and is a composition of an automaton for accepting TCP connections (acceptor), an automaton for making TCP connections (connector), and two automata for TCP connections representing the channel from the predecessor and the channel to the successor.
 The TCP automata are provided by the ioa++ package and are ordinary automata implemented using file descriptors and the techniques described in Section~\ref{design}.
@@
 
-The tcp\_lcr\_automaton which is listed in examples/tcp\_lcr.cpp\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} of the ioa++ package implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
+The tcp\_lcr\_automaton\footnote{\url{github.com/jrwilson/ioa/blob/master/examples/tcp_lcr.cpp}} implements the complete protocol by composing an asynch\_lcr\_automaton and a tcp\_ring\_automaton.
 In the resulting layered architecture of the TCP LCR system, the bottom layer consists of component automata for TCP sockets, the middle layer consists of a component automaton that supports coordination via a unidirectional ring, and the top layer implements the application logic.

### 2012-02-24 papers/paper2/component_model.tex
Down to 16 pages.

@@
 \emph{Hierarchical decomposition} is an established technique for I/O automata~\cite{lynch1994atomic} which allows a \emph{parent} to contain zero or more \emph{children}.
-The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.
-
-To illustrate how components can be expressed naturally using I/O automata, Figure~\ref{sys_model} shows a simple event detector system comprising three component automata: an anonymous Processor, a Converter named C, and a Detector named D.
-The Processor automaton is the root and C and D are its children.
+The \emph{root automaton} is at the top of the hierarchy represents the entire system as composition is applied recursively upward.
+To illustrate how components can be expressed naturally using I/O automata, Figure~\ref{sys_model} shows a simple event detector comprising three component automata: an anonymous Processor, a Converter named C, and a Detector named D.
+The Processor is the root and C and D are its children.
 Automata are depicted as solid rectangles. The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
-The name of a named automaton appears after its type, e.g., Converter C.
 Actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
@@
 Fourth, an output action cannot be bound to an input action in the same automaton.
-The second, third, and fourth rules exist because they admit ambiguity into the relationship between the states of the associated automata.
+The second, third, and fourth rules prevent ambiguity in the relationship between the states of the associated automata.
 Graphically, an arrow originating in one component automaton must terminate in another component automaton as Figure~\ref{sys_model} illustrates.
@@
 \paragraph*{Concurrent execution.}
-Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
+Since each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
 Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
@@
 input actions to receive the results.
-User-defined component automata (i.e., \emph{user automata}) inherit from the base automaton, which as we describe in Section~\ref{framework_architecture} is implemented in ioa++ as a C++ class. 
+User-defined component automata (i.e., \emph{user automata}) inherit from the base automaton.
 Each user automaton is thus composed (not bound) with the system automaton and inherits output actions for requesting system actions and input actions for receiving the results of system actions.

### 2012-02-24 papers/paper2/conclusion.tex
Down to 16 pages.

@@
 We agree with~\cite{georgiou2009automated} that having a model that can be compiled is beneficial for reasoning about the program directly from the source code and helps to close the gap between the developer's mental model and the code listing.
-Our primary goal moving forward is to use the ioa++ framework to gain further experience building real systems with I/O automata, and to support optimization of its performance with respect to an increasingly diverse range of use cases.
+Our primary goal moving forward is to use the ioa++ framework to gain further experience building real systems with I/O automata, and to support optimization of its performance.
 

### 2012-02-24 papers/paper2/design.tex
Down to 16 pages.

@@
 The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
-Conceptually, the set of automata $A$ and set of bindings $B$ belong to the system automaton.
-However, the state they encode is shared and is used in the execution of every action.
-The \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
-Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_model} and the system automaton.
-The opportunities for concurrent execution discussed in Section~\ref{component_model} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
-Binding predicates do not modify $B$ and therefore do not affect concurrent execution.
+
 The scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
+A \emph{dispatcher} executes actions and a \emph{controller} selects the next action.
+Input actions are executed with corresponding output actions, according to the current set of bindings.
+The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
+The controller encodes a \emph{policy} that selects the next local action.
+For fidelity to the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
+The ioa++ distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
+
+%% Conceptually, the set of automata $A$ and set of bindings $B$ belong to the system automaton.
+%% However, the state they encode is shared and is used in the execution of every action.
+%% The \emph{scheduler} uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
+%% Thus, the set of actions implied by any action is the union of the set defined in Section~\ref{component_model} and the system automaton.
+%% The opportunities for concurrent execution discussed in Section~\ref{component_model} hinge on the fact that \emph{only} actions that modify $A$ and $B$ must be serialized.
+%% Binding predicates do not modify $B$ and therefore do not affect concurrent execution.
 
@@
 Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.
-
-The scheduler is responsible for both selecting and executing local actions.
-These two activities are decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
-The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
-Input actions are executed with corresponding output actions, according to the current set of bindings.
-The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
-The controller encodes a \emph{policy} that selects the next local action.
-For fidelity to the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
-The ioa++ distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
 
@@
 Subsequent requests made using the same file descriptor and event are ignored.
-Once the file descriptor is ready, the action is released for consideration by the controller.
+Once the file descriptor is ready, the action is released to the controller.
 

### 2012-02-24 papers/paper2/evaluation.tex
Down to 16 pages.

@@
 
-Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with two threads versus one thread.
-Confidence intervals are shown but are negligible with the largest being a speed-up range of $\pm$0.0165.
-When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
-Consequently, every action must be serialized yielding a maximum speed-up of 1.
-When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.
-The speed-up shows a strong dependence on the duration of each local action which is proportional to $N$.
-A small value for $N$ means that very little time is spent executing automaton code.
-Consequently, a greater fraction of time is spent executing framework code which includes various synchronization calls to the pthreads library.
-For a small enough $N$, this overhead dominates the execution time.
-This situation is exacerbated by multiple threads since they will actively interfere with one another which can be seen in the slow-down for small values of $N$.
-Conversely, when $N$ is large, relatively little time is spent executing framework and synchronization code.
-Consequently, contention is reduced allowing for greater speed-ups as indicated by Figure~\ref{speed_up}.
-
 \begin{figure}
@@
 
+Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with two threads versus one thread.
+Confidence intervals are shown but are negligible with the largest being a speed-up range of $\pm$0.0165.
+When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
+Consequently, every action must be serialized yielding a maximum speed-up of 1.
+When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.
+The speed-up shows a strong dependence on the duration of each local action which is proportional to $N$.
+A small value for $N$ means that very little time is spent executing automaton code.
+Consequently, a greater fraction of time is spent executing framework code which includes various synchronization calls to the pthreads library.
+For a small enough $N$, this overhead dominates the execution time.
+This situation is exacerbated by multiple threads since they will actively interfere with one another which can be seen in the slow-down for small values of $N$.
+Conversely, when $N$ is large, relatively little time is spent executing framework and synchronization code.
+Consequently, contention is reduced allowing for greater speed-ups as indicated by Figure~\ref{speed_up}.
+
 Figures~\ref{overhead1} and \ref{overhead2} show the average per thread of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions; confidence intervals are again negligible with largest being $\pm$0.002155 for Figure~\ref{overhead1} and $\pm$0.001781 for Figure~\ref{overhead2}.
@@
 Both of these potentially can be improved with better scheduler design.
-The synchronization time depends on the interactions among the automata comprising the system, the scheduler, and the locking scheme used by the framework.
-We plan to optimize scheduling and synchronization performance in ioa++ as future work.
+Synchronization time depends on the interactions among the automata comprising the system, the scheduler, and the locking scheme used by the framework.
+We plan to optimize scheduling and synchronization in ioa++ as future work.

### 2012-02-24 papers/paper2/programming_model.tex
Down to 16 pages.

@@
 In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
-Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
-The two techniques for handling an error are either to reset or to replace the portion of the composed automata that encountered the error.
+Thus, errors can be propagated up the hierarchy and resolved at an appropriate level.
+Two techniques for handling an error are either to reset or to replace the portion of the composed automata that encountered the error.
 An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
-To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
+To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with it), create a new child, and bind to the new child.
 Recall that system actions are asynchronous and can be executed in any order.

### 2012-02-25 papers/paper2/abstract.tex
Two lines shy.

@@
 \begin{abstract}
-The ability to realize sophisticated distributed reactive systems such as those found in cyber-physical, pervasive, or enterprise computing environments is predicated on a suitable model and implementation of asynchrony and concurrency.
+The ability to realize sophisticated distributed reactive systems such as those found in cyber-physical or pervasive computing environments is predicated on a suitable model and implementation of asynchrony and concurrency.
 The size and complexity of these systems implies that software re-use in their development, maintenance, and evolution is also essential.
 Taken together, these requirements demand appropriate programming abstractions through which different forms of asynchrony and concurrency can be realized directly in reusable software.
-
 To that end, this paper presents the ioa++ framework for developing distributed reactive systems, which is the primary contribution of this research.
@@
 We also evaluate concurrency performance in ioa++ through a series of micro-benchmarks which confirm that the framework’s ability to exploit potentially concurrent execution is limited only by the interactions of the automata that comprise the system and the overhead of synchronization.
+
+\keywords{concurrency, I/O automata, components}
 \end{abstract}

### 2012-02-25 papers/paper2/component_model.tex
Two lines shy.

@@
 % The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
-As Figure~\ref{system_action} illustrates, the system automaton provides: parameterized input actions for receiving requests to create, bind, unbind, and destroy automata; internal actions for evaluating and enacting system action requests; and corresponding parameterized output actions for returning responses to system action requests.
-(Parameterized actions are indicated with [].)
+As Figure~\ref{system_action} illustrates, the system automaton provides: parameterized input actions (marked []) for receiving requests to create, bind, unbind, and destroy automata; internal actions for evaluating and enacting system action requests; and corresponding parameterized output actions for returning responses to system action requests.
 The base automaton provides output actions that trigger requests to create, bind, unbind, and destroy automata, along with
 input actions to receive the results.
-User-defined component automata (i.e., \emph{user automata}) inherit from the base automaton.
+User-defined component automata inherit from the base automaton.
 Each user automaton is thus composed (not bound) with the system automaton and inherits output actions for requesting system actions and input actions for receiving the results of system actions.

### 2012-02-25 papers/paper2/design.tex
Two lines shy.

@@
 The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
-
-The scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
+The \emph{scheduler} is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
 A \emph{dispatcher} executes actions and a \emph{controller} selects the next action.

### 2012-02-25 papers/paper2/programming_model.tex
Two lines shy.

@@
 \paragraph*{Dynamic composition.}
-%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
-Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
-\ifjournal
-The tcp\_lcr\_automaton discussed in Section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
-\begin{lstlisting}
-ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
-lcr = ioa::make_automaton_manager (this,
-  ioa::make_generator<asynch_lcr_automaton<uuid> > (
-    m_u));
-ioa::make_binding_manager (this,
-  lcr, &asynch_lcr_automaton<uuid>::send,
-  &m_self, &tcp_lcr_automaton::send_lcr);
-ioa::make_binding_manager (this,
-  &m_self, &tcp_lcr_automaton::receive_lcr,
-  lcr, &asynch_lcr_automaton<uuid>::receive);
-ioa::make_binding_manager (this,
-  lcr, &asynch_lcr_automaton<uuid>::leader,
-  &m_self, &tcp_lcr_automaton::leader_lcr);
-ioa::make_binding_manager (this,
-  &m_self, &tcp_lcr_automaton::init_lcr,
-  lcr, &asynch_lcr_automaton<uuid>::init);
-\end{lstlisting}
-\fi
-%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
-An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
-%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
-A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
-The ioa::automaton\_manager class and ioa::binding\_manager class also contain methods for destroying and unbinding respectively.
-\ifjournal
-The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
-From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
-The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
-Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
-\fi
+Dynamic composition is accomplished by creating and removing child automata and bindings.
+An ioa::automaton\_manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
+An ioa::binding\_manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
 

### 2012-02-25 papers/paper2/related_work.tex
Down to 16 pages.
Two lines shy.

@@
 Lee elaborates on the problems faced when using threads for concurrency in~\cite{lee2006problem}.
-Given the shift from higher clock frequencies to larger numbers of cores as the main driver of hardware performance, these same concerns are relevant to trends that look to deliver performance improvements through increasingly concurrent programs~\cite{sutter2005software}.
-
+Given the shift from higher clock frequencies to larger numbers of cores as the main driver of hardware performance, these same concerns are relevant to trends towards increasingly concurrent programs~\cite{sutter2005software}.
 Formal models of concurrency are also relevant to the discussion in this paper, e.g., Communicating Sequential Processes~\cite{hoare1978communicating}, Actors~\cite{agha1986actors}, and I/O automata~\cite{lynch1996distributed}.
